### 2016-04-27 agda.tex
initial migration from TFP16 repo

% !TEX root = hazelnut-popl17.tex
In the previous section, we gave an overview of the most important rules in
the semantics of Hazelnut, and stated the important theorems. In a few
cases, we sketched out why these theorems will hold.

In order to formally verify that our design meets the stated objectives, we
are preparing a formalization of the grammar, the judgements given above
and the metatheory in the proof assistant Agda \cite{norell:thesis}. We
refer readers unfamiliar with Agda to the Agda Wiki, hosted
at \url{http://wiki.portal.chalmers.se/agda/}.

Our formalization of Hazelnut is under development at
\url{http://github.com/hazelgrove/agda-tfp16}. At the time of submission, we 
have completed much of the initial groundwork, but the proofs of 
Theorem \ref{thrm:actsafe} and Theorem \ref{thrm:actdet} (and a few  
lemmas that we elided here) are under construction.

The documentation 
includes a more detailed discussion of the technical representation
decisions made. The core idea of our formalization is to encode each judgement as a
dependent type. The rules of the judgements become the constructors of the
type, and derivations of theorems values of the type. This is a rich
setting that allows proofs to take advantage of pattern matching on the
shape of derivations, closely matching on-paper proofs of similar
properties.

The formalization differs from the calculus defined here in a few small 
ways. Most interestingly, instead of giving separate movement rules for each 
form of Z-expression, we abstract
over different corresponding forms that happen to have the same arity, e.g. 
additions and applications. Formalizing this intuition
reduces the number of cases we need to consider 
somewhat, but more importantly allows us to write a slightly more
general calculus -- it will be easier to extend Hazelnut with more 
interesting language features with this generic infrastructure in place.

Because the metatheory in this paper is largely concerned with the statics
of Hazelnut rather than its dynamics, we adopt Barendregt's convention for
bound variables and avoid substitution entirely \cite{urban}. Future
iterations will need a more mature technique for reasoning about
binding---likely de Bruijn indices or abstract binding
trees \cite{lh09unibind,Pouillard11}.

### 2016-04-27 example.tex
initial migration from TFP16 repo

% !TEX root = hazelnut-popl17.tex
\begin{figure*}[t!]
\[
\begin{array}{|c||c|c||l|l|}
\hline
\# & \textbf{H-Expression} & \textbf{Z-Expression} & \textbf{Next Action} & \textbf{Semantics}
\\
\hline
1 &
\hhole{} &
\zwsel{\hhole{}}
&
\aConstruct{\flam{x}} & \refrule{\ref{r:conelamhole}}
\\ 2 &
\hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}}
&
\aConstruct{\fnum{}} & \refrule{\ref{r:contnum}}
\\ 3 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}}
&
\aMove{\dNext{}} & \refrule{\ref{r:movenextsib}}
\\ 4 &
&
\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
\aConstruct{\fnum{}} & \refrule{\ref{r:contnum}}
\\ 5 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}}
&
\aMove{\dParent{}} & \refrule{\ref{r:moveparent}}
\\ 6 &
&
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
\\ 7 &
&
\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}}
&
\aMove{\dChild{}} & \refrule{\ref{r:movefirstchild-lam}}
\\ 8 &
&
\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}}
&
\aConstruct{\fvar{x}} & \refrule{\ref{r:conevar}}
\\ 9 &
\hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
&
\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
&
\quad\textrm{---{}---}
&
\quad\textrm{---{}---}
\\
\hline
\multicolumn{5}{c}{\text{... now assume a context where $id : \tarr{\tnum}{\tnum}$ ...}}\\
\hline
10 &
\hhole{} &
\zwsel{\hhole{}}
&
\aConstruct{\fasc} & \refrule{\ref{r:constructasc}}
\\
11 &
\hhole{} : \hhole{} &
\hhole{} : \zwsel{ \hhole{}}
&
\aConstruct{\fnum{}} & \refrule{\ref{r:contnum}}
\\
12 &
\hhole{} :\tnum{} &
\hhole{} : \zwsel{\tnum{}}
&
\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
\\
%13 &
%\hhole{} :\tnum{} &
%\zwsel{\hhole{}} : \tnum{}
%&
%\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
%\\
13 &
%\hhole{} :\tnum{}
 &
\zwsel{\hhole{}} : \tnum{}
&
\aConstruct{ \fvar{id}} & \refrule{\ref{r:conevar2}}
\\
14 &
\hhole{\textrm{$id$}} : \tnum{} &
\hhole{\zwsel{{\textrm{$id$}}}} : \tnum{}
&
\aConstruct{\fap{}} & \refrule{\ref{r:coneapfn}}
\\
15 &
\hhole{\hap{{{\textrm{$id$}}}}{{\hhole{}}}} : \tnum{}
&
\hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hhole{}}}} : \tnum{}
&
\aConstruct{\fnumlit{3}} &  \refrule{\ref{r:conenumnum}}
\\
16 &
\hhole{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
&
\hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hnum{3}}}} : \tnum{}
&
\aMove{\dParent{}} &  \refrule{\ref{r:moveparent-ap2}}
\\
17 &
%\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
&
\hhole{\zwsel{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}} : \tnum{}
&
\aMove{\dParent{}} &  \refrule{\ref{r:moveparent-hole}}
\\
18 &
%\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
&
\zwsel{\hhole{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{}
&
\aFinish &  \refrule{\ref{r:finishana}}
\\
19 &
{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
&
\zwsel{{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{}
&
\quad\textrm{---{}---} & \quad\textrm{---{}---}
\\
\hline

%% 11 &
%% {\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
%% &
%% \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
%% &
%% \aMove{\dParent{}} & ?
%% \\ 12 &
%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
%% &
%% \zwsel{\hlam{x}{{{x}}}} : \tarr{\tnum{}}{\tnum{}}
%% &
%% \aMove{\dParent{}} & \refrule{15b} bad
%% \\ 13 &
%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
%% &
%% \zwsel{\hlam{x}{{{x}}} : \tarr{\tnum{}}{\tnum{}}}
%% &
%% \aConstruct{\fap{}} & \refrule{20h} bad
%% \\ 14 &
%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{}}
%% &
%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{}}}
%% &
%% \aConstruct{\fnumlit{3}} & \refrule{20l} bad
%% \\ 15 &
%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{3}}
%% %&
%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{3}}}
%% %&
%% %\aFinish{} & ?
%% %\\
%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{{\hnum{3}}}
%% &
%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{{\hnum{3}}}}
%% &
%% \quad\textrm{---{}---}
%% &
%% \quad\textrm{---{}---}
%% \\
%% \hline
\end{array}
\]
\caption{Constructing an identity function in Hazelnut (Lines~1--9), then  applying this function (assumed bound to $id$, not shown) to an argument~(Lines 10--19). The formal syntax and referenced rules in the final column are described in Section \ref{sec:hazel}.}
\label{fig:first-example}
\end{figure*}
%
Figure~\ref{fig:first-example} gives an example of the Hazelnut user
performing two simple programming tasks.
The syntactic forms in this figure will be formally defined in Sec. \ref{sec:hazel}. For now, we will develop only the necessary intuitions. In the first task (Lines 1-9), the user constructs the identity function over numbers. In the second task (Lines 10-19), the user applies this function (assumed to be bound to a variable, $id$), to the number expression $\hnum{3}$.
Each of these tasks is carried out interactively, through the sequence of \emph{actions} shown in the  column labeled \textbf{Next Action}. For reference, we cite the relevant rules from Sec. \ref{sec:hazel} in the final column.

The second and third columns of the
table show the program as it is being constructed in two forms. The second column shows it as an \textbf{H-expression}, which is an expression that can contain \emph{holes}, delimited by $\llparenthesis$ and $\rrparenthesis$. The third column shows a corresponding \textbf{Z-expression}. Z-expressions are H-expressions with a single focus on some sub-term, delimited by $\triangleright$ and $\triangleleft$. The focus need not be on a hole.
% on working on filling just one of the holes.
Each action produces a new Z-expression, but this may or may not correspond to a new H-expression (in particular, some actions only move the focus, without changing the structure of the term.)
% to the hole in
%focus in the Z-Expression to produce the next line, which may or may not
%produce a substantively different H-Expression.

Line~1 begins with the simplest initial expression: an H-expression
consisting of a single hole. The corresponding Z-Expression has that hole in focus,
indicated by the syntax~$\zwsel{\hehole}$. Focus determines the locus of action. The first action the user performs is $\aConstruct{\flam{x}}$, which replaces the hole with a lambda abstraction binding the variable $x$. This results in the program on line
2, consisting of a lambda abstraction ascribed an arrow type with holes in all positions. The argument type hole is in focus. The
user proceeds to fill these holes using construction and movement actions, resulting in the final expression on Line 9. With no holes remaining, this expression is \emph{complete}.% (though it could, of course, undergo further actions nevertheless.)

So far, editing has proceeded in an essentially type-directed, outside-in fashion -- the user first specified the type of the function, then produced a body of that type by the action on Line 8. Lines 10-12 similarly begin in a type-directed manner with the user giving an explicit type ascription, indicating that the expression that they are constructing will have type $\tnum$.

However, on Line 13, the user performs the $\aConstruct{\fvar{id}}$ action. Notice that $id$ has type $\tarr{\tnum}{\tnum}$, which is not consistent with the type $\tnum$ given in the ascription. Na\"ively, this would produce a type error, leaving the program in a well-formed but semantically undefined state. One way to avoid this state is to simply not make this action available in the program configuration on Line 12. This is inflexible, forcing an outside-in approach to program construction (i.e. the user would need to construct the function application form before constructing the variable $id$.) Instead, Hazelnut permits this action, but places the variable $id$ inside a hole. This defers the consistency check that would normally occur: a hole can be checked against any type, as long as its contents have some type. The cursor is placed inside the hole. The user then proceeds to apply $id$ to the number expression $\hnum{3}$. At this point, the expression inside the hole has a type consistent with the ascription, so the user can \emph{finish} the hole. In our simple formalism, this requires moving the cursor to the hole (in practice, the system might find the nearest parent of hole form.) The result is the complete, well-typed program shown on Line 19 (notice that \emph{complete} is distinct from \emph{closed} -- the variable $id$ is free on Line 19, so this is not a closed program.)

%% The third column~(\textbf{Next Action}) lists the first user action:
%% Constructing a lambda abstraction using variable~$x$.
%% %
%% The final column~(\textbf{Semantics}) indicates the semantic rule for this
%% action, Rule (\ref{r:conelamhole}), which gives general semantics for
%% introducing lambda terms into holes.
%% %
%% In Section~\ref{sec:hazel}, we list this rule, and the other rules used in
%% this final column. In total, these rules give a formal semantics to the
%% user actions, which relate each line's Z-Expression to the Z-Expression on
%% the subsequent line.

%% In addition to introducing the lambda term, and its variable, the
%% first user action~$\aConstruct{\flam{x}}$ also introduces a type
%% ascription for this function, as an arrow type, with holes for the
%% type of its domain and codomain.
%% %
%% The actions for Lines~2--5 consist of the user filling these holes
%% with the basetype $\tnum{}$.
%% %
%% To do so, the user constructs the type constructor twice (Lines 2 and
%% 4), and navigates between the holes with a move action (Line~3).
%% %
%% Generally, the move action~$\dNext$ moves the focus from one
%% sub-structure to the next sibling sub-structure of the (common) parent
%% structure; in this case, it moves from the domain type of the arrow
%% type to the codomain of the arrow type.
%% %

### 2016-04-27 implementation.tex
initial migration from TFP16 repo

\subsection{Implementation Concepts}

\begin{figure}
\centering
\includegraphics[width=4in]{Implementation_Diagram}
\caption{Implementation Concepts. Each Action--Zipper Structure--View combination is considered to appear ``instantaneously'' on the timeline.}
\label{fig:FRP}
\end{figure}

The key question that must be answered for any implementation strategy is: how do we model a stream of actions from a user? Let us assume that these actions are chosen (using some input device, preferably, a keyboard) from some ``palette'' that never presents the user with actions that are not semantically well-defined, according to the action semantics defined earlier.
%In a traditional editor, the input from the user is a stream of characters, and there are no guarantees that at any point that the program is syntactically well-formed, so the designer leaves editing as an .
%In contrast, in a structure editor, the input from the user is a stream of operations.
As such, each new action will ``atomically'' generate a new Z-expression. 
This insight leads us to conclude that a natural way to implement this editor would be using event-based Functional Reactive Programming~\cite{Wan:2000:FRP:349299.349331} (FRP).
Figure~\ref{fig:FRP} illustrates the concept of an FRP-based implementation of a  structure editor organized like Hazelnut.
The input from the user is a stream of actions.  Each action results in a change to the underlying abstract model (i.e., a new Z-expression is created after each action.)
Each model change results in an updated \emph{view} which is then presented to the user.  The user can then consider this new view when they choose a new action as input.

\subsection{HZ}
We explore the concepts presented in the paper in HZ, our implementation of Hazelnut.
In order to reach a wide audience, we decided to implement HZ in the web browser.
In order to take advantage of all the benefits of FRP, we chose to implement HZ using OCaml\footnote{https://ocaml.org/}, the \texttt{js\_of\_ocaml} compiler\footnote{http://ocsigen.org/js\_of\_ocaml/} and the OCaml React library\footnote{http://erratique.ch/software/react}.

At the time of the writing of this paper, our implementation of HZ includes encodings of Z-expression as presented in this paper.
We consider this ZExp to be our model. 
HZ renders the model as a string embedded in HTML.
Currently we support only the delete action.  Other actions are currently under development. We anticipate having a substantially more functional implementation by the time this work is presented (our focus thusfar has been on the metatheory.) 
The work-in-progress code as well as directions for how to compile and run it can be found here: \url{https://github.com/hazelgrove/impl-tfp16}.

A substantially simpler system that we developed while exploring the ideas that led to Hazelnut can be found at the following URL:
\url{http://www.cs.cmu.edu/~comar/nestedpairs/}.

### 2016-04-27 intro.tex
initial migration from TFP16 repo

% !TEX root = hazelnut-popl17.tex

%% Programs (and, by the Curry-Howard correspondence, proofs) are rich
%% inductive structures. This fact is well understood amongst researchers
%% and experienced programmers, but still somewhat obscure amongst
%% programmers at large because programmers normally construct and interact
%% with programs only indirectly, e.g. using a text editor composed with a
%% parser.

%% There are some benefits to this approach, to be sure, but the structural
%% mismatch between programs and their textual representations also imposes
%% various burdens.  For example, the primitive edit actions available in a
%% text editor (e.g. inserting or deleting a character or word) do not
%% always correspond to sensible structural transformations.

% spj: describe the problem; state our contributions; STOP. one page max.

% When constructing a program or proof in a language with rich type
% structure, skilled programmers generally follow a \emph{type discipline}
% where they first determine the type of the expression that they are
% constructing in order to constrain the mental search space that they are
% operating within.

% For example, if the programmer knows that an expression of type
% $\tarr{\tnum}{\tnum}$ is needed, then it is often the case (though, of course, not
% necessarily the case) that the expression will take the form $$\hlam{\mathit{x}}{e}$$
% for some variable $x$ and function body $e$. If the programmer chooses this
% form, then after picking a suitable variable name, her focus will be on
% constructing a suitable body, $e$. Following the type discipline, $e$ must
% be of type $\tnum$, and so this process can begin anew.

% The problem is that when using a text
% editor to construct a program, it is easy, and indeed necessary, to deviate from this disciplined process.  Rather, text editors operate on sequences of
% characters (i.e. \emph{text}.)

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use are often presented with text that does not correspond to a well-defined program. This may be because the programmer is in the midst of a sequence of
text edit actions that leave the program ``temporarily'' malformed or ill-defined, or because the programmer has made a mistake. This complicates the programming process, both because the language definition provides no reasoning principles relevant to such text, and, relatedly, because the editor can no longer provide useful services (e.g. syntax highlighting, or semantics-aware code completion.) %Tool designers often develop \emph{ad hoc} heuristic methods that allow one to 
Source code editors sometimes develop \emph{ad hoc} workarounds for this problem, e.g. they might use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.)

% stuck editing a representation of the program instead of the structures
% themselves. The editor does not restrict what the programmer may do: you
% can delete characters that belong, insert ones that don't, forget things
% that were needed, and so. There's nothing stopping us from accidentally
% writing $$\lambda \mathit{x:num}.\mathit{(x,x)}$$ even though it's obvious
% that building a pair can't hope to form a natural number.

% The type structure of the language makes this sort of error obvious: it's
% not that you're adding characters that make your program incorrect, or even
% malformed; you're adding characters that can't possibly create a structure
% you want because of the type. Simply put, the primitive operations
% available in text editors do not always correspond to sensible
% transformations on the structure of the program.

\emph{Structure editors} have made some progress toward addressing this
problem by allowing programmers to edit the tree structure of a program
directly. Each edit action leaves the program being constructed in a
structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also generally the case that programs can be written more quickly using a structure editor. However, we will not talk about ``edit costs'' here, because we would like to abstract away from details like whether a keyboard or a mouse (or some other input device) is being used.} We give examples of notable structure editors in Sec. \ref{sec:rw}.

The problem is that it is still possible to leave the program in a semantically meaningless (i.e. undefined) state, because language definitions generally only give meaning to complete, well-typed terms. This makes it difficult for humans and tools to reason about types and binding during the development process.  %More sophisticated semantic reasoning principles thus remain .

%% TODO: some sort of transition that talks about how you still don't have
%% semantic reasoning principles because you can be left in a semantically
%% ill-defined state

In this paper, we present our ongoing work on a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e. Hazelnut is a \emph{structure editor calculus}.) In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics. Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state. In fact, edit actions maintain an even stronger invariant -- when acting on an expression whose type is determined by its surroundings (e.g. an expression in function argument position), only edit actions consistent with that type are permitted (we will formally state this invariant in Sec. \ref{sec:actions}.) This does not imply that programs need to be constructed in a strictly outside-in manner, however, because an expression that has a type that is not yet consistent with its surroundings will automatically be put into a {hole} that defers consistency analysis until the hole is \emph{finished}.

The remainder of the paper is organized as follows:
\begin{itemize}
  \item We begin with an example to develop the reader's intuitions in Section
    \ref{sec:example}.

  \item We then formally define Hazelnut in Section \ref{sec:hazel} and state the important metatheoretic properties.

  \item In Section \ref{sec:mech}, we describe our ongoing effort to formalize the semantics and metatheory of Hazelnut in Agda.

  \item In Section \ref{sec:impl}, we describe our ongoing effort to implement Hazelnut in a web browser, using a functional reactive model for user interaction.
  \item In Section \ref{sec:rw}, we give an overview of related work.
  \item We conclude in Section \ref{sec:future} by describing our vision for this work going 
    forward.
\end{itemize}

### 2016-04-27 syntax-popl17.tex
initial migration from TFP16 repo

%\documentclass[12pt]{article}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
\makeatletter

\renewcommand\topfraction{0.85}
\renewcommand\bottomfraction{0.85}
\renewcommand\textfraction{0.1}
\renewcommand\floatpagefraction{0.85}

\AtBeginDocument{%
 \abovedisplayskip=3pt
 \abovedisplayshortskip=0pt
 \belowdisplayskip=2pt
 \belowdisplayshortskip=0pt
}

\setlength{\floatsep}{10pt}
\setlength{\textfloatsep}{12pt}

\usepackage[compact]{titlesec}  
\titlespacing*{\section}{0pt}{3pt}{3pt}
\titlespacing*{\subsection}{0pt}{4pt}{3pt}
\titlespacing*{\subsubsection}{0pt}{4pt}{3pt}
\titlespacing*{\paragraph}{0pt}{3pt}{5pt}
\setlength{\skip\footins}{3pt}

\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink

\usepackage{enumitem}

\input{macros}
\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

%\titlebanner{}        % These are ignored unless
\preprintfooter{Draft}   % 'preprint' option specified.

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus}
%\subtitle{Modular Type Constructors}

\authorinfo{\vspace{-2px}}{}{}
%\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
%         {Carnegie Mellon University}
%         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Programs are rich inductive structures, but programmers typically construct and manipulate them indirectly, through flat textual representations. This indirection  comes at a cost: programmers must understand the  subtleties of parsing, and it can require several text editor actions to make a single syntactically and semantically well-defined change. 
During these sequences of edit actions, or when the programmer makes a mistake, programmers and programming tools must contend with malformed text or semantically ill-defined programs.

\emph{Structure editors} promise to alleviate these burdens by exposing only edit actions that  produce sensible changes to the program structure.
Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically ill-defined as they are being constructed.
In this paper, we report on our ongoing efforts to develop Hazelnut, a minimal  structure editor defined in a principled type-theoretic style where all edit actions leave the program in both a syntactically and semantically well-defined state. Uniquely, Hazelnut does not force the programmer to construct the program in a strictly ``outside-in'' fashion.
Formally, Hazelnut is a bidirectionally typed lambda calculus extended with
1) \emph{holes} (which mark subterms that are being  constructed from the inside out);
2) a \emph{focus model}; and 3) a bidirectional \emph{action model} equipped with a useful \emph{action sensibility} theorem.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation

\section{Introduction}\label{sec:introduction}
\input{intro}

\section{Programming in Hazelnut}\label{sec:example}
\input{example}

\section{Hazelnut, Formally}
\label{sec:hazel}
Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

Whenever an action is performed on a well-typed expression, it produces another well-typed expression in a \emph{sensible} manner. More specifically, the action semantics satisfies a crucial \emph{sensibility theorem}, stated in Sec. \ref{sec:actions}.
\end{itemize}

In our overview of the semantics below, we will reproduce only the most interesting rules, and in some cases we will do so ``out of order.'' The appendix (and our Agda formalization, see Sec. \ref{sec:mech}) defines the complete collection of rules in their dependency order.
\subsection{Holes}\label{sec:holes}
\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \tau,\htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & e,\hexp & ::= &
  \hexp : \htau ~\vert~
  x ~\vert~
  \hlam{x}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}

The syntax of H-types and H-expressions is given in Figure \ref{fig:hexp-syntax}. Most of the forms correspond directly to those of the simply-typed lambda calculus extended with type $\tnum$. The number expression corresponding to the number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$. 
In addition to these standard forms, \emph{empty holes} are drawn $\hehole$ and \emph{non-empty H-expression holes} are drawn $\hhole{\hexp}$. In our simple calculus, all well-formed type expressions are valid types, so we do not need non-empty H-type holes.%Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.

We refer to terms that do not contain subterms of hole form as \emph{complete}. Informally, we will use metavariables $\tau$ and $e$ rather than $\htau$ and $\hexp$ for complete H-types and H-expressions, respectively. Formally, we can derive $\hcomplete{\tau}$ when $\tau$ is a complete H-type, and $\hcomplete{e}$ when $e$ is a complete H-expression. We omit the straightforward definitions of these judgements for concision. The dynamics of Hazelnut, which we need not detail here, is defined only  over complete H-expressions (i.e. we can only ``run'' a complete program, though see Sec. \ref{sec:future}.)

The statics of Hazelnut is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100}, i.e. around the following mutually defined typing judgements:
\[\arraycolsep=15pt\begin{array}{ll}
%\textbf{Judgement Form} & \textbf{Description}\\
\hana{\hGamma}{\hexp}{\htau} & \text{$\hexp$ analyzes against $\htau$}\\
\hsyn{\hGamma}{\hexp}{\htau} & \text{$\hexp$ synthesizes $\htau$}
\end{array}\]
where typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$.
Derivations of the type analysis judgement establish that $\hexp$ can appear where an expression of type $\htau$ is expected. Derivations of the type synthesis judgement determine a type that can be assigned to $\hexp$ even in positions where an expected type is not known (e.g. at the top level.) Algorithmically, the type is an ``input'' of the type analysis judgement, but an ``output'' of the type synthesis judgement. %The rules describe a \emph{local type inference} scheme, i.e. type ascriptions are unnecessary when an expression is being analyzed against a known type. 
Making a judgemental distinction between these two notions will be essential for giving a sensible action semantics to our system (Sec. \ref{sec:actions}.)

 %We use the metavariable $\Gamma$ for \emph{complete typing contexts}, i.e. typing contexts where each hypothesis mentions only complete types.

\begin{subequations}\label{rules:syn-ana}
Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any \emph{compatible} type. This is expressed by the \emph{subsumption rule}:
\begin{equation}\label{rule:ana-subsume}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau'}\\
  \tcompat{\htau}{\htau'}
}{
  \hana{\hGamma}{\hexp}{\htau}
}
\end{equation}
The \emph{H-type compatibility judgement}, $\tcompat{\htau}{\htau'}$, reduces to syntactic equality for complete H-types. For incomplete H-types, the rules are given after we discuss the semantics of holes below.

First, let us briefly review the standard constructs. 
Type ascription allows the user to state  a type for the ascribed expression to be analyzed against:
\begin{equation}\label{rule:syn-asc}
\inferrule{
  \hana{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hexp : \htau}{\htau}
}
\end{equation}

A variable synthesizes the type that the context assigns to it:
\begin{equation}\label{rule:syn-var}
\inferrule{ }{
  \hsyn{\hGamma, x : \htau}{x}{\htau}
}
\end{equation}

Functions are not themselves annotated with types, so they can only appear in analytic position:
\begin{equation}\label{rule:syn-lam}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation}
(It would be straightforward to also add a ``half-annotated'' lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out of our calculus \cite{DBLP:conf/tldi/ChlipalaPH05}.)

For function application, if the expression in function position synthesizes an arrow type, the argument is analyzed against the synthesized argument type:
\begin{equation}\label{rule:syn-ap}
\inferrule{
  \hsyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}\\
  \hana{\hGamma}{\hexp_2}{\htau_2}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
}
\end{equation}

Numbers synthesize type $\tnum$:
\begin{equation}\label{rule:syn-num}
\inferrule{ }{
  \hsyn{\hGamma}{\hnum{n}}{\tnum}
}
\end{equation}

Addition operates like a function over numbers:
\begin{equation}\label{rule:syn-plus}
\inferrule{
  \hana{\hGamma}{\hexp_1}{\tnum}\\
  \hana{\hGamma}{\hexp_2}{\tnum}
}{
  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
}
\end{equation}

The rules given so far are sufficient to type complete H-expressions. The remaining rules give H-expressions with holes a well-defined static semantics.

The empty hole synthesizes the hole type:
\begin{equation}\label{rule:syn-ehole}
\inferrule{ }{
  \hsyn{\hGamma}{\hehole}{\tehole}
}
\end{equation}

A non-empty hole contains an H-expression that is ``under construction''. The inner expression must synthesize some type, but the non-empty hole synthesizes only the hole type:
\begin{equation}\label{rule:syn-hole}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
}
\end{equation}
The type compatibility judgement $\tcompat{\htau}{\htau'}$, which appeared as a premise in the subsumption rule, makes the hole type compatible with any other type:
\begin{subequations}\label{rules:tcompat}
\begin{equation}\label{rule:tcompat-hole}
\inferrule{ }{
  \tcompat{\htau}{\tehole}
}
\end{equation}
The remaining rules, given in the appendix, establish that type compatibility is symmetric and reflexive (but not transitive.)
% \begin{equation}\label{rule:tcompat-comm}
% \inferrule{
%   \tcompat{\htau}{\htau'}
% }{
%   \tcompat{\htau'}{\htau}
% }
% \end{equation}
% \begin{equation}\label{rule:tcompat-num}
% \inferrule{ }{
%   \tcompat{\tnum}{\tnum}
% }
% \end{equation}
% \begin{equation}\label{rule:tcompat-arr}
% \inferrule{
%   \tcompat{\htau_1}{\htau_1'}\\
%   \tcompat{\htau_2}{\htau_2'}
% }{
%   \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
% }
% \end{equation}
\end{subequations}
Consequently, by subsumption, we can derive that $\hana{id : \tarr{\tnum}{\tnum}}{\hhole{id}}{\tnum}$, as is necessary to synthesize a type for the H-expression on Line 14 of Fig. \ref{fig:first-example}. %In other words, this mechanism is essential if  users are to able to construct a program in anything but an ``outside in'' fashion.

The final rule handles function applications where the expression in function position synthesizes a hole type, rather than an arrow type. We treat it as if it had instead synthesized $\tarr{\tehole}{\tehole}$:
\begin{equation}\label{rule:syn-ap-2}
\inferrule{
  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
  \hana{\hGamma}{\hexp_2}{\tehole}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}
}
\end{equation}

The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}. Their system (which was not bidirectionally typed nor an editor model) also needed to define two rules for function application. In general, when a premise requires that a synthesized type be of a particular form, we need a special case where the synthesized hole type is treated instead as if it were the ``holey-est'' type of that form.\footnote{Alternatively, we might add a rule that allows expressions that synthesize hole type to then non-deterministically synthesize any other type, but maintaining determinism is useful in practice, so we avoid this approach.}

\end{subequations}
\subsection{Focus Model}\label{sec:cursors}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} ~\vert~
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} ~\vert~
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} ~\vert~
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau ~\vert~
  \hexp : \ztau ~\vert~
  \hlam{x}{\zexp} ~\vert~
  \hap{\zexp}{\hexp} ~\vert~
  \hap{\hexp}{\zexp} ~\vert~
  \hadd{\zexp}{\hexp} ~\vert~
  \hadd{\hexp}{\zexp} ~\vert~
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single cursor.}
\label{fig:zexp-syntax}
\end{figure}

In order to identify a single subtree of an H-type or H-expression as the current focus of action, we apply Huet's \emph{zipper pattern} \cite{JFP::Huet1997}. The syntax of Z-types, $\ztau$, and Z-expressions, $\zexp$, is given in Figure \ref{fig:zexp-syntax}. The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that is the current focus. All other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the focus will be found. All other sub-terms are H-types or H-expressions. Taken together, every syntactically well-formed Z-type and Z-expression contains exactly one focused H-type or H-expression.

We write $\removeSel{\ztau}$ for the H-type constructed by removing the focus marker from the Z-type $\ztau$. This straightforward metafunction is defined as follows:
\begin{align*}
%\removeSel{(\zlsel{\htau})} & = \htau\\
\removeSel{(\zwsel{\htau})} & = \htau\\
%\removeSel{(\zrsel{\htau})} & = \htau\\
\removeSel{(\tarr{\ztau}{\htau})} & = \tarr{\removeSel{\ztau}}{\htau}\\
\removeSel{(\tarr{\htau}{\ztau})} & = \tarr{\htau}{\removeSel{\ztau}}
\end{align*}

Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by removing the focus marker from the Z-expression $\zexp$. The definition of this metafunction is analagous, so we leave it in the appendix for concision.
% \begin{align*}
% %\removeSel{(\zlsel{\hexp})} & = \hexp\\
% \removeSel{(\zwsel{\hexp})} & = \hexp\\
% %\removeSel{(\zrsel{\hexp})} & = \hexp\\
% \removeSel{(\zexp : \htau)} & = \removeSel{\zexp} : \htau\\
% \removeSel{(\hexp : \ztau)} & = \hexp : \removeSel{\ztau}\\
% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\
% \removeSel{(\hap{\zexp}{\hexp})} & = \hap{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hap{\hexp}{\zexp})} & = \hap{\hexp}{\removeSel{\zexp}}\\
% \removeSel{(\hadd{\zexp}{\hexp})} & = \hadd{\removeSel{\zexp}}{\hexp}\\
% \removeSel{(\hadd{\hexp}{\zexp})} & = \hadd{\hexp}{\removeSel{\zexp}}\\
% \removeSel{\hhole{\zexp}} &= \hhole{\removeSel{\zexp}}
% \end{align*}

\subsection{Action Semantics}\label{sec:actions}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{llcllll}
\mathsf{Action} & \alpha & ::= &
  \aMove{\delta} ~\vert~
  %\aSelect{\delta} ~\vert~
  \aDel ~\vert~
  %\aReplace{\htau} ~\vert~
  %\aReplace{\hexp} ~\vert~
  \aConstruct{\varphi} ~\vert~
  \aFinish\\
\mathsf{Direction} & \delta & ::= &
  \dChild ~\vert~
  \dParent ~\vert~
  \dNext ~\vert~
  \dPrev\\
\mathsf{Shape} & \varphi & ::= &
  \farr ~\vert~
  \fnum \\
& & \vert &
  \fasc ~\vert~
  \fvar{x} ~\vert~
  \flam{x} ~\vert~
  \fap ~\vert~
  \farg ~\vert~
  \fnumlit{n} ~\vert~
  \fplus
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of actions.}
\label{fig:action-syntax}
\vspace{-8px}
\end{figure}

The syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, or \emph{shapes}, $\varphi$, is given in Figure \ref{fig:action-syntax}. Actions are performed on Z-types and Z-expressions according to the \emph{action semantics} of Hazelnut, which is organized around three judgements:
\[\arraycolsep=10pt\begin{array}{ll}
%\textbf{Judgement Form} & \textbf{Description}\\
\performTyp{\ztau}{\alpha}{\ztau'} & \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'} & \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
& \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'} & \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
& \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
& \text{against $\htau$}
\end{array}\]

As suggested by the descriptions above, the action semantics maintains the following \emph{action sensibility} theorem:
\begin{theorem}[Action Sensibility] Both of the following hold:
\label{thrm:actsafe}
\begin{enumerate}
\item If $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ then
  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
\item If $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
  $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ then
  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{enumerate}
\end{theorem}
In words, every action leaves the program in a semantically well-defined state. More specifically, the first clause of Theorem \ref{thrm:actsafe} establishes that actions performed on expressions that synthesize a type can only produce expressions that also synthesize some (possibly different) type. The second clause establishes that actions performed on expressions in analytic position (e.g. those under type ascriptions or in argument position, see above) can only produce expressions that can also be analyzed against the expected type.% Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.

It is also useful to maintain a \emph{deterministic} action semantics, i.e. every well-defined action should produce a unique Z-type or Z-expression. Formally, this is stated as follows:
\begin{theorem}[Action Determinism] All of the following hold:
\label{thrm:actdet}
\begin{enumerate}
\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
  $\zexp' = \zexp''$ and $\htau' = \htau''$.
% \item If all of

%   \begin{quote}
%     \begin{enumerate}
%     \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
%     \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
%     \item $\tcompat{\htau}{\htau'}$, and
%     \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
%       $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
%     \end{enumerate}
%   \end{quote}
%   hold, then $\zexp' = \zexp''$.
\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
  \zexp''$.
\end{enumerate}
\end{theorem}

In order to maintain determinism, we will need to supplement the definition of type compatibility above with a definition for \emph{type incompatibility}, $\tincompat{\htau}{\htau'}$. The key rule establishes that arrow types are incompatible with the $\tnum$ type:
\begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
\end{subequations}
The remaining rules, given in the appendix, establish that type incompatibility is symmetric and covariant.
\subsubsection{Subsumption}

The action semantics includes a subsumption rule much like the one from the underlying semantics of H-expressions:
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the expression synthesizes a type, then we defer to the synthetic action performance judgement, as long as it produces an expression that synthesizes a type compatible with the type provided for analysis. It is easy to see that this satisfies Theorem 1 by applying the IH and subsumption.

\subsubsection{Relative Movement} Movement actions change the focus but do not change the underlying H-type or H-expression (so action sensibility is easy to show for these rules as well.)

The rules for relative movement within Z-types are given below and should be self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }{
      \aMove{\dChild}
    }{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{\htau_2}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\tarr{\htau_1}{\htau_2}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{\zwsel{\htau_1}}{{\htau_2}}
    }{
      \aMove{\dNext}
    }{
      {\tarr{\htau_1}{\zwsel{\htau_2}}}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \tarr{{\htau_1}}{\zwsel{\htau_2}}
    }{
      \aMove{\dPrev}
    }{
      {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
    }
  }
\end{equation}
% \begin{equation}
% \inferrule{
%   \performTyp{
%     \ztau
%   }{
%     \aMove{\delta}
%   }{
%     \ztau'
%   }
% }{
%   \performTyp{
%     \tarr{\ztau}{\htau}
%   }{
%     \aMove{\delta}
%   }{
%     \tarr{\ztau'}{\htau}
%   }
% }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{
%       \ztau
%     }{
%       \aMove{\delta}
%     }{
%       \ztau'
%     }
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aMove{\delta}
%     }{
%       \tarr{\htau}{\ztau}
%     }
%   }
% \end{equation}
\end{subequations}
% The final two rules above recurse into the zipper structure.

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
\end{subequations}
\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

Again, the rule for Z-types is self-explanatory:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \zwsel{\htau}
    }{
      \aDel
    }{
      \zwsel{\tehole}
    }
  }
\end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\ztau}{\htau}}{\aDel}{\tarr{\ztau'}{\htau}}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performTyp{\tarr{\htau}{\ztau}}{\aDel}{\tarr{\htau}{\ztau'}}
%   }
% \end{equation}
\end{subequations}

Deletion within a Z-expression is similarly straightforward:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
  }
\end{equation}
%\end{subequations}
% The base case turns into a hole:
%\begin{subequations}
% \begin{equation}
% \inferrule{ }{
%   \performDel{\zwsel{\hexp}}{\hehole}
% }
% \end{equation}
% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
% \begin{equation}
%   \inferrule{
%     \performDel{\zexp}{\zexp'}
%   }{
%     \performDel{\zexp : \htau}{\zexp' : \htau}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
%   }
% \end{equation}

\end{subequations}
\subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.

Again, let us begin with type actions. The $\aConstruct{\farr}$ action constructs an arrow type. The focused H-type becomes the argument type, and the focus is placed on an empty return type hole:
\begin{subequations}
  \begin{equation}
    \label{r:contarr}
  \inferrule{ }{
    \performTyp{
      \zwsel{\htau}
    }{
      \aConstruct{\farr}
    }{
      \tarr{\htau}{\zwsel{\tehole}}
    }
  }
\end{equation}

The $\aConstruct{\fnum}$ action replaces an empty Z-type hole with the $\tnum$ type:
  \begin{equation}
    \label{r:contnum}
  \inferrule{ }{
    \performTyp{
      \zwsel{\tehole}
    }{
      \aConstruct{\fnum}
    }{
      \zwsel{\tnum}
    }
  }
\end{equation}

% Construction proceeds recursively down the zipper:
%   \begin{equation}
%     \label{r:contarrL}
%   \inferrule{
%     \performTyp{\ztau}{\aConstruct{\varphi}}{\ztau'}
%   }{
%     \performTyp{
%       \tarr{\ztau}{\htau}
%     }{
%       \aConstruct{\varphi}
%     }{
%       \tarr{\ztau'}{\htau}
%     }
%   }
% \end{equation}
%   \begin{equation}
%     \label{r:contarrR}
%   \inferrule{
%     \performTyp{\ztau}{\aConstruct{\varphi}}{\ztau'}
%   }{
%     \performTyp{
%       \tarr{\htau}{\ztau}
%     }{
%       \aConstruct{\varphi}
%     }{
%       \tarr{\htau}{\ztau'}
%     }
%   }
% \end{equation}
\end{subequations}

\begin{subequations}

Moving on to expression actions, we start to see more interesting rules. The $\aConstruct{\fasc}$ action operates differently depending on whether the focused expression synthesizes a type or is being analyzed against a type. In the first case, the ascribed type is the synthesized type:
\begin{equation}
  \label{r:constructasc}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}{\htau}
  }
\end{equation}
In the second case, the ascribed type is the type provided for analysis:
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}
  }
\end{equation}

The $\aConstruct{\fvar{x}}$ action places the variable $x$ into the focused empty hole. If that hole is being asked to synthesize a type, then the result of the action synthesizes the type assigned to $x$ in the context:
\begin{equation}
  \label{r:conevar}
  \inferrule{ }{
    \performSyn{\hGamma, x : \htau}{\zwsel{\hehole}}{\tehole}{\aConstruct{\fvar{x}}}{\zwsel{x}}{\htau}
  }
\end{equation}
If the focused empty hole is being analyzed against a type that is inconsistent with the type assigned to $x$ by the context, $x$ is placed inside a hole:
\begin{equation}
 \label{r:conevar2}
  \inferrule{
    \tincompat{\htau}{\htau'}
  }{
    \performAna{\hGamma, x : \htau'}{\zwsel{\hehole}}{\htau}{\aConstruct{\fvar{x}}}{\hhole{\zwsel{x}}}
  }
\end{equation}
The rule above featured in the example in Section \ref{sec:example}.

Notice that no rule was necessary for the case where the hole was being analyzed against a type compatible with the variable's type, because this case is handled by the action subsumption rule.

The $\aConstruct{\flam{x}}$ action places a lambda term binding $x$ into an empty hole. If the focused empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the focus in the argument type position:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the focused empty hole is being analyzed against an arrow type, then no ascription is necessary:
\begin{equation}
  \inferrule{ }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\tarr{\htau_1}{\htau_2}}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}

If the focused empty hole is being analyzed against a type that is incompatible with an arrow type, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$ is inserted inside a hole, to maintain Theorem \ref{thrm:actsafe}:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hhole{
        \hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}
      }}
  }
\end{equation}

The $\aConstruct{\fap}$ action applies the expression in focus to a hole. If the focused expression synthesizes a function type, then the rule is straightforward:
\begin{equation}
  \label{r:coneapfn}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\tarr{\htau_1}{\htau_2}}
      {\aConstruct{\fap}}
      {\hap{\hexp}{\zwsel{\hehole}}}
      {\htau_2}
  }
\end{equation}

If the focused expression synthesizes a hole type, then we can treat it as if it synthesized the $\tarr{\tehole}{\tehole}$ type, exactly as described in Sec. \ref{sec:holes}:
\begin{equation}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\tehole}
      {\aConstruct{\fap}}
      {\hap{\hexp}{\zwsel{\hehole}}}
      {\tehole}
  }
\end{equation}

Finally, if the focused expression synthesizes a type that is incompatible with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fap}}
      {\hap{\hhole{\hexp}}{\zwsel{\hehole}}}
      {\tehole}
  }
\end{equation}

The $\aConstruct{\farg}$ action places the focused expression instead in the argument position of an application. Because the function position is always an empty hole in this situation, we only need a single rule:
\begin{equation}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\farg}}
      {\hap{\zwsel{\hehole}}{\hexp}}
      {\tehole}
  }
\end{equation}

The $\aConstruct{\fnumlit{n}}$ action places the number expression $\hnum{n}$ into an empty hole. If the focused hole is being asked to synthesize a type, then the rule is straightforward:
\begin{equation}
  \label{r:conenumnum}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\fnumlit{n}}}
      {\zwsel{\hnum{n}}}
      {\tnum}
  }
\end{equation}
If the focused hole is being analyzed against a type that is incompatible with $\tnum$, then we must place the number expression inside a hole:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tnum}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\fnumlit{n}}}
      {\hhole{\zwsel{\hnum{n}}}}
  }
\end{equation}

Finally, the $\aConstruct{\fplus}$ action constructs a plus expression with the focused expression as its first argument. If the focused expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
\begin{equation}
  \inferrule{
    \tcompat{\htau}{\tnum}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fplus}}
      {\hadd{\hexp}{\zwsel{\hehole}}}
      {\tnum}
  }
\end{equation}

Otherwise, we must place the focused expression inside a hole:
\begin{equation}
  \inferrule{
    \tincompat{\htau}{\tnum}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hexp}}
      {\htau}
      {\aConstruct{\fplus}}
      {\hadd{\hhole{\hexp}}{\zwsel{\hehole}}}
      {\tnum}
  }
\end{equation}
\end{subequations}
Notice that we do not have an action that explicitly wraps an expression in a non-empty hole. These arise implicitly when an action that would not na\"ively satisfy Theorem \ref{thrm:actsafe} is performed (see Figure \ref{fig:first-example}.)

\subsubsection{Finishing}
The final action we will consider in Hazelnut is $\aFinish$, which finishes the focused non-empty hole.

If the focused non-empty hole appears in synthetic position, then it can always be finished:
\begin{subequations}
  \begin{equation}
    \label{r:finishana}
  \inferrule{
    \hsyn{\hGamma}{\hexp}{\htau'}
  }{
    \performSyn
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\tehole}
      {\aFinish}
      {\zwsel{\hexp}}
      {\htau'}
  }
\end{equation}

If the focused non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the wrapped expression is consistent with the type the hole is being analyzed against. This amounts to analyzing those contents against the provided type (by subsumption):
\begin{equation}
  \inferrule{
    \hana{\hGamma}{\hexp}{\htau}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hhole{\hexp}}}
      {\htau}
      {\aFinish}
      {\zwsel{\hexp}}
  }
\end{equation}
\end{subequations}

\subsubsection{Zipper Cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression. We also need to define the recursive cases, which propagate the action into the subtree where the focus appears. These rules follow the structure of the corresponding rules in the statics of H-expressions.

\begin{subequations}
For example, when the focus is in the expression position of an ascription, we use the analytic action performance judgement:
\begin{equation}
\inferrule{
  \performAna
    {\hGamma}
    {\zexp}
    {\htau}
    {\alpha}
    {\zexp'}
}{
  \performSyn
    {\hGamma}
    {\zexp : \htau}
    {\htau}
    {\alpha}
    {\zexp' : \htau}
    {\htau}
}
\end{equation}

When the focus is in the type position of an ascription, we must re-check the ascribed expression because the type might have changed (in practice, one would optimize this check to only occur if the type actually was changed):
\begin{equation}
\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
}{
  \performSyn
    {\hGamma}
    {\hexp : \ztau}
    {\removeSel{\ztau}}
    {\alpha}
    {\hexp : \ztau'}
    {\removeSel{\ztau'}}
}
\end{equation}

If the focus is in the body of a lambda expression, then we must use the analytic action performance rule:
\begin{equation}
\inferrule{
  \performAna
    {\hGamma, x : \htau_1}
    {\zexp}
    {\htau_2}
    {\alpha}
    {\zexp'}
}{
  \performAna
    {\hGamma}
    {\hlam{x}{\zexp}}
    {\tarr{\htau_1}{\htau_2}}
    {\alpha}
    {\hlam{x}{\zexp'}}
}
\end{equation}

There are two rules that handle the case where the focus is in the function position of an application, corresponding to the two application rules in the statics. Each involves rechecking the argument against the new function type:
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
    \performSyn
      {\hGamma}
      {\zexp}
      {\htau_2}
      {\alpha}
      {\zexp'}
      {\tarr{\htau_3}{\htau_4}}\\
    \hana{\hGamma}{\hexp}{\htau_3}
  }{
    \performSyn
      {\hGamma}
      {\hap{\zexp}{\hexp}}
      {\htau_1}
      {\alpha}
      {\hap{\zexp'}{\hexp}}
      {\htau_4}
  }
\end{equation}
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
    \performSyn
      {\hGamma}
      {\zexp}
      {\htau_2}
      {\alpha}
      {\zexp'}
      {\tehole}\\
    \hana{\hGamma}{\hexp}{\tehole}
  }{
    \performSyn
      {\hGamma}
      {\hap{\zexp}{\hexp}}
      {\htau_1}
      {\alpha}
      {\hap{\zexp'}{\hexp}}
      {\tehole}
  }
\end{equation}

Similarly, there are two rules that handle the case where the focus is in the argument position:
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\hexp}{\tarr{\htau_2}{\htau}}\\
    \performAna
      {\hGamma}
      {\zexp}
      {\htau_2}
      {\alpha}
      {\zexp'}
  }{
    \performSyn
      {\hGamma}
      {\hap{\hexp}{\zexp}}
      {\htau}
      {\alpha}
      {\hap{\hexp}{\zexp'}}
      {\htau}
  }
\end{equation}
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\hexp}{\tehole}\\
    \performAna
      {\hGamma}
      {\zexp}
      {\tehole}
      {\alpha}
      {\zexp'}
  }{
    \performSyn
      {\hGamma}
      {\hap{\hexp}{\zexp}}
      {\tehole}
      {\alpha}
      {\hap{\hexp}{\zexp'}}
      {\tehole}
  }
\end{equation}

The rules for the addition operator follow from the statics directly:
\begin{equation}
  \inferrule{
    \performAna
      {\hGamma}
      {\zexp}
      {\tnum}
      {\alpha}
      {\zexp'}
  }{
    \performSyn
      {\hGamma}
      {\hadd{\zexp}{\hexp}}
      {\tnum}
      {\alpha}
      {\hadd{\zexp'}{\hexp}}
      {\tnum}
  }
\end{equation}
\begin{equation}
  \inferrule{
    \performAna
      {\hGamma}
      {\zexp}
      {\tnum}
      {\alpha}
      {\zexp'}
  }{
    \performSyn
      {\hGamma}
      {\hadd{\hexp}{\zexp}}
      {\tnum}
      {\alpha}
      {\hadd{\hexp}{\zexp'}}
      {\tnum}
  }
\end{equation}

Finally, if the focus is inside a non-empty hole, we special case the situation where the action results in a doubly-nested empty hole, $\hhole{\hehole}$, to eliminate the nesting (given our current action semantics, only the delete action can cause this form to arise and the form $\hhole{\hhole{\zexp}}$ cannot arise):
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
    \performSyn
      {\hGamma}
      {\zexp}
      {\htau}
      {\alpha}
      {\zexp'}
      {\htau'}\\
    \zexp' \neq \zwsel{\hehole}
  }{
    \performSyn
      {\hGamma}
      {\hhole{\zexp}}
      {\tehole}
      {\alpha}
      {\hhole{\zexp'}}
      {\tehole}
  }
\end{equation}
\begin{equation}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
    \performSyn
      {\hGamma}
      {\zexp}
      {\htau}
      {\alpha}
      {\zwsel{\hehole}}
      {\tehole}\\
  }{
    \performSyn
      {\hGamma}
      {\hhole{\zexp}}
      {\tehole}
      {\alpha}
      {\zwsel{\hehole}}
      {\tehole}
  }
\end{equation}

\end{subequations}

\section{Mechanization}
\label{sec:mech}\label{sec:mt}
\input{agda}

\section{Implementation}
\label{sec:impl}
\input{implementation}

\section{Related Work}\label{sec:rw}
%\subsection{Structure Editors}

Structured editing has been recognized as a way to avoid the possibility of syntax errors for decades.  An early example is the
The Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981}, first published in 1981.
The synthesizer generator~\cite{Reps:1984:SG:390010.808247} allows the user to create an attribute-grammar specification that then can be used to generate a structured editor.
CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language. Barista \cite{ko_barista:_2006} is a modern take on the same basic concept.%These early systems were developed  of the systems are rooted in the type-theoretic tradition.

Novice programmers have been a common target for structure editors. For example,
GNOME\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates.
Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.
Touchdevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students.
Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students. These are largely drag-and-drop user interfaces with a limited action model and an unclear semantics.

Not all structure editors are for educational purposes. For example,
mbeddr \cite{voelter_mbeddr:_2012} is an extensible C-based Programming Language and IDE (nominally, for programming embedded systems.)
mbeddr is build on top of the commercial JetBrains MPS framework for constructing structure editors.
Another popular approach is to bring elements of structured editing into a traditional editor.
Codelets \cite{oney_codelets:_2012} uses structured editing to add interactive documentation and examples in an editor. 
Our previous work on Graphite~\cite{Omar:2012:ACC:2337223.2337324} allows developers to associate structured editing interfaces called  \emph{palettes} with types. Graphite is integrated into a text-based program editor (Eclipse.)

Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g. Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes. These  systems are also, to our knowledge, not formally well-defined but rather exist only as part of system implementations.

Perhaps the systems most similar in spirit to Hazelnut are Lamdu~\cite{lamdu} and Unison~\cite{unison}. Like Hazelnut, these are both statically typed functional language editors. In both cases, the language is similar to Haskell. In Lamdu, the editor uses structure editing to enable Live Programming, where the code is always being executed as it is being written.

Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified. For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model. Unison is also a rather large body of Haskell code, though its action model appears superficially more similar to ours. We maintain what we believe to be a stronger action sensibility invariant than Unison (i.e. in Unison, one must construct expressions from the outside-in.) These systems are rich sources of interesting ideas, however -- there is room enough for many different approaches in this (re-)emerging space.

%Drag-and-drop / for novices: lots of examples, e.g. Alice and others
%
%Contemporary: Lamdu, MPS/Mbeddr, TouchDevelop
%
%Hybrid: Cyrus' active code completion paper

%\subsection{Refactoring Models}
%(Michael, can you fill this section out?)

%\subsection{Formal Editor Models}
%Need to do a search to see what else has been done...

\section{Discussion \& Conclusion}
\label{sec:future}
This paper presented Hazelnut, a type theoretic structure editor calculus. Our aim is to take a principled approach to its design by formally specifying its semantics, providing strong metatheoretic guarantees, mechanizing its semantics and metatheory in Agda and implementing it using the concepts of  functional reactive programming. As of this submission, we have achieved reasonable confidence in the formal system presented above, and have transitioned our focus toward the mechanization and implementation efforts. By the time of presentation, we anticipate having complete or nearly complete versions of these.

\subsection{Future Work}
Hazelnut is, obviously, a very limited language at its core. So the most obvious avenue for future work is to increase the expressive power of this language. Our plan is to simultaneously maintain a mechanization and implementation (following, for example, Standard ML) as we proceed, ultimately producing the first large-scale, formally verified bidirectionally typed language codesigned with a type-aware editor. It may be that certain language features are unnecessary given a sufficiently advanced type-aware structure editor (e.g. SML's \texttt{open}?), while other features may only be practical with editor support. We intend to use Hazelnut and derivative systems thereof as a platform for rigorously exploring such questions.

There are various aspects of the editor model that we have not yet formalized. For example, our action model does not consider how actions are actually entered using, for example, key combinations or chords. It also did not provide any specific model of how available actions will be determined for presentation to the user. In practice, we would want also to rank available actions in some reasonable manner (perhaps based on usage data gathered from other users or code repositories.)

Another research direction is in exploring how types can be used to control the presentation of expressions in the editor. For example, following our approach in a textual setting on \emph{type-specific languages} (TSLs), it should be possible to have the type that an expression is being analyzed against define alternative display forms and interaction modes \cite{TSLs}.

Finally, we did not consider any aspects of \emph{collaborative programming}, such as a packaging system, a differencing algorithm for use in a source control system, support for multiple simultaneous focii for different users, and so on. These are all interesting avenues for future work.

On the theoretical side, the notion of having one of many possible holes in a term in focus has a very strong intuitive connection
  to the proof theoretic notion of focusing \cite{Simmons11tr}. Beyond just
  the name, both seem to involve, in some sense, a search through the space of possible
  ways to finish a derivation. We intend to explore this connection to see
  if it's coincidental or more meaningful and welcome insights in this regard.

We already discussed a connection to gradual typing \cite{Siek06a}. We hope to explore this connection more thoroughly. In particular, it may be possible to better support exploratory and live programming by allowing even programs with holes in them to execute as long as those holes are only in the type portions, by deferring to the semantics given in work on gradual typing.

It may also be possible to give a dynamics to incomplete expressions. Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and 
$\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
explored an equational theory for terms with holes, but intend to once our
formalization effort is more mature.

\begin{quote}
In any case, these are but steps toward more graphical program-description
systems, for we will not forever stay confined to mere strings of symbols.

--- Marvin Minsky, Turing Award lecture
\end{quote}% We recommend abbrvnat bibliography style.
\clearpage
\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\bibliography{../research}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...

\end{document}

### 2016-04-27 make.sh
super-simple makefile for pdflatex+bibtex

@@ -1,0 +1,6 @@
+#!/bin/sh
+NAME=hazelnut-popl17
+pdflatex $NAME
+bibtex   $NAME
+pdflatex $NAME
+pdflatex $NAME

### 2016-04-28 hazelnut-popl17.tex
new abstract

--- 2016-04-27 syntax-popl17.tex
@@
 
-\authorinfo{\vspace{-2px}}{}{}
-%\authorinfo{~}{~}{~}
+%\authorinfo{\vspace{-2px}}{}{}
+\authorinfo{~}{~}{~}
 %\authorinfo{Cyrus Omar \and Jonathan Aldrich}
@@
 \begin{abstract}
-Programs are rich inductive structures, but programmers typically construct and manipulate them indirectly, through flat textual representations.
-This indirection  comes at a cost: programmers must understand the  subtleties of parsing, and it can require several text editor actions to make a single syntactically and semantically well-defined change.
-During these sequences of edit actions, or when the programmer makes a mistake, programmers and programming tools must contend with malformed text or semantically ill-defined programs.
-
-\emph{Structure editors} promise to alleviate these burdens by exposing only edit actions that  produce sensible changes to the program structure.
-Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}.
-They also focus primarily on syntactic well-formedness, so programs can still be left semantically ill-defined as they are being constructed.
-In this paper, we report on our ongoing efforts to develop Hazelnut, a minimal  structure editor defined in a principled type-theoretic style where all edit actions leave the program in both a syntactically and semantically well-defined state.
-Uniquely, Hazelnut does not force the programmer to construct the program in a strictly ``outside-in'' fashion.
-Formally, Hazelnut is a bidirectionally typed lambda calculus extended with
-1) \emph{holes} (which mark subterms that are being  constructed from the inside out);
-2) a \emph{focus model}; and 3) a bidirectional \emph{action model} equipped with a useful \emph{action sensibility} theorem.
+Programmers typically construct and manipulate well-typed programs indirectly, as text that must pass through a parser and typechecker.
+Not all text survives this journey.
+In particular, text that arises transiently during the editing process, or when the programmer makes a mistake, is commonly found to be malformed or ill-typed.
+Contending with (i.e.
+reasoning about) malformed text or well-formed but meaningless syntax trees is a difficult task, for programmers and  tools alike.
+
+%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
+% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}.
+They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
+
+We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the program in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a foundation for \emph{live programming environments}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms.
+Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}.
+Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems.
+%Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
 \end{abstract}
@@
 %\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
-\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single cursor.}
+\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single focus.}
 \label{fig:zexp-syntax}

### 2016-04-28 intro.tex
new abstract

@@
 syntax highlighting, or semantics-aware code completion.) %Tool designers often develop \emph{ad hoc} heuristic methods that allow one to
+
 Source code editors sometimes develop \emph{ad hoc} workarounds for this problem, e.g.

### 2016-04-28 hazelnut-popl17.tex
updated abstract

@@
 \begin{abstract}
-Programmers typically construct and manipulate well-typed programs indirectly, as text that must pass through a parser and typechecker.
+Well-typed expressions are meaningful inductive structures, but programmers typically construct and manipulate them indirectly, as text that must pass through a parser and typechecker.
 Not all text survives this journey.
-In particular, text that arises transiently during the editing process, or when the programmer makes a mistake, is commonly found to be malformed or ill-typed.
+In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
 Contending with (i.e.
-reasoning about) malformed text or well-formed but meaningless syntax trees is a difficult task, for programmers and  tools alike.
+reasoning about) malformed text or well-formed but meaningless expressions is a difficult task, for programmers and their tools alike.
 
@@
 
-We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the program in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a foundation for \emph{live programming environments}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms.
+We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms.
 Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}.

### 2016-05-04 hazelnut-popl17.tex
added TODO to cite polymorphic blocks paper

@@
 \begin{abstract}
-Well-typed expressions are meaningful inductive structures, but programmers typically construct and manipulate them indirectly, as text that must pass through a parser and typechecker.
+Well-typed expressions are rich inductive structures, but programmers typically construct and manipulate them only indirectly, as text that must pass through a parser and typechecker.
 Not all text survives this journey.
@@
 
+TODO: cite http://cseweb.ucsd.edu/~lerner/pb.html
+
 %Drag-and-drop / for novices: lots of examples, e.g. Alice and others

### 2016-05-04 hazelnut-popl17.tex
added TODO to cite andy ko's paper on design requirements

@@
 
-TODO: cite http://cseweb.ucsd.edu/~lerner/pb.html
+TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}
+
+TODO: cite \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}
 

### 2016-05-15 hazelnut-popl17.tex
clarification of why the premise in one of the rules for application actions makes sense

--- 2016-05-04 hazelnut-popl17.tex
@@
 
-If the focused empty hole is being analyzed against a type that is incompatible with an arrow type, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$ is inserted inside a hole, to maintain Theorem \ref{thrm:actsafe}:
+If the focused empty hole is being analyzed against a type that is
+incompatible with any arrow type, expressed in the premise as an arrow with
+two holes, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
+is inserted inside a hole, to maintain Theorem \ref{thrm:actsafe}:
 \begin{equation}

### 2016-05-25 hazelnut-popl17.tex
wrote figures for dynamics
progress on intro

--- 2016-05-04 hazelnut-popl17.tex
@@
 
+\usepackage{todonotes}
+
 \input{macros}
@@
 \begin{abstract}
-Well-typed expressions are rich inductive structures, but programmers typically construct and manipulate them only indirectly, as text that must pass through a parser and typechecker.
+Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker.
 Not all text survives this journey.
 In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
-Contending with (i.e.
-reasoning about) malformed text or well-formed but meaningless expressions is a difficult task, for programmers and their tools alike.
+% Not all text that arises during the programming process survives the journey through a parser and typechecker.
+In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
+Contending with (i.e. reasoning about) malformed text, or well-formed but meaningless expressions, is a difficult task, for programmers and tools alike.
 
@@
 
-We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms.
+This paper introduces Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms.
 Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}.

### 2016-05-25 intro.tex
progress on intro

@@
 %Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
-Programmers and the tools that they use are often presented with text that does not correspond to a well-defined program. This may be because the programmer is in the midst of a sequence of
-text edit actions that leave the program ``temporarily'' malformed or ill-defined, or because the programmer has made a mistake.
-This complicates the programming process, both because the language definition provides no reasoning principles relevant to such text, and, relatedly, because the editor can no longer provide useful services (e.g.
-syntax highlighting, or semantics-aware code completion.) %Tool designers often develop \emph{ad hoc} heuristic methods that allow one to
+Programmers and the tools that they use often encounter text that does not correspond to a meaningful program. This may be because the programmer is in the midst of a sequence of 
+edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake.
+It is difficult for tools to help programmers execute a corrective course of action in such situations (e.g.
+by providing syntax highlighting and semantics-aware code completion.) %because the language definition is necessarily silent about the structure or meaning of such text.
+%For example, a code editor cannot assist the programmer in these situations.
 
-Source code editors sometimes develop \emph{ad hoc} workarounds for this problem, e.g.
-they might use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.)
+Some editors have developed \emph{ad hoc} workarounds for this problem, e.g.
+they might use whitespace to guess where a construct is likely to end, attempt to insert closing delimiters automatically, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are quite complex and sometimes misleading.
 

### 2016-05-25 hazelnut-popl17.tex

@@
+\def\OPTIONConf{0}%
 %\documentclass[12pt]{article}
@@
 \renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}
+
+\usepackage{joshuadunfield}
+\usepackage{llproof}
+\usepackage{rulelinks}
 
@@
 \end{subequations}
+
+\clearpage
+
+\section{Dynamics of Hazelnut}
+\label{sec:dynamics}
+\input{dynamics}
+
+\clearpage
 

### 2016-06-22 hazelnut-popl17.tex
updated abstract

@@
 % authoryear    To obtain author/year citation style instead of numeric.
+\usepackage[T1]{fontenc}
 \usepackage{amsmath}
@@
 \usepackage{todonotes}
+\usepackage{listings}
+\usepackage{microtype}
 \sloppy
@@
 \begin{abstract}
-Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker.
+% Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker.
 Not all text survives this journey.
-In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
+In particular, text that arises transiently, or when the programmer makes a mistake, is often malformed or ill-typed.
 % Not all text that arises during the programming process survives the journey through a parser and typechecker.
 In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
-Contending with (i.e. reasoning about) malformed text, or well-formed but meaningless expressions, is a difficult task, for programmers and tools alike.
+% Contending with malformed program text or well-formed but meaningless syntax trees is difficult for programmers and their tools alike.
 
@@
 They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
-
-This paper introduces Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms.
-Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}.
-Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems.
+% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.
+
+\emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors.
+This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text.
+%Structure editors for professional programmers are also under active development (e.g.
+\texttt{mbeddr}.)
+
+This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful.
+Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner.
+To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}.
+Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.
+
+We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself.
+To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut.
+We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application.
+Our reference implementation is written using \lstinline{js_of_ocaml}.
+
 %Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
@@
 \end{equation}
+and this for the example
+\begin{equation}\label{r:movefirstchild-lam}
+\inferrule{ }{
+  \performMove{
+    \zwsel{\hlam{x}{\hexp}}
+  }{
+    \aMove{\dChild}
+  }{
+    \hlam{x}{\zwsel{\hexp}}
+  }
+}
+\end{equation}
+
+\begin{equation}\label{r:moveparent-ap2}
+  \inferrule{ }{
+    \performMove{
+      \hap{{\hexp_1}}{\zwsel{\hexp_2}}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hap{\hexp_1}{\hexp_2}}
+    }
+  }
+\end{equation}
+
+\begin{equation}\label{r:moveparent-hole}
+  \inferrule{ }{
+    \performMove{
+      \hhole{\zwsel{\hexp}}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hhole{\hexp}}
+    }
+  }
+\end{equation}
+
 \end{subequations}
@@
 %\subsection{Structure Editors}
+
+TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}
 

### 2016-06-22 intro.tex
updated abstract

@@
 Programmers and the tools that they use often encounter text that does not correspond to a meaningful program. This may be because the programmer is in the midst of a sequence of 
-edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake.
-It is difficult for tools to help programmers execute a corrective course of action in such situations (e.g.
-by providing syntax highlighting and semantics-aware code completion.) %because the language definition is necessarily silent about the structure or meaning of such text.
-%For example, a code editor cannot assist the programmer in these situations.
+edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake\todo{mention YoungSeok's data here?}.
+It is difficult for tools to help programmers determine and execute a corrective course of action in such situations (e.g.
+by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.) %because the language definition is silent %For example, a code editor cannot assist the programmer in these situations.
 
 Some editors have developed \emph{ad hoc} workarounds for this problem, e.g.
-they might use whitespace to guess where a construct is likely to end, attempt to insert closing delimiters automatically, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are quite complex and sometimes misleading.
+they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading.
 
@@
 
-\emph{Structure editors} have made some progress toward addressing this
-problem by allowing programmers to edit the tree structure of a program
+\emph{Structure editors} have long promised to address this
+problem by allowing programmers to edit the tree structure of a program, or some rich projection thereof, 
 directly. Each edit action leaves the program being constructed in a
-structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also generally the case that programs can be written more quickly using a structure editor.
-However, we will not talk about ``edit costs'' here, because we would like to abstract away from details like whether a keyboard or a mouse (or some other input device) is being used.} We give examples of notable structure editors in Sec.
+structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor.
+However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
+whether a keyboard or a mouse is in use.} We give examples of notable structure editors in Sec.
 \ref{sec:rw}.

### 2016-06-22 hazelnut-popl17.tex
added TODOs

@@
 \usepackage{listings}
+\lstset{tabsize=2, 
+basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
 \usepackage{microtype}
@@
 \label{sec:hazel}
-Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
+\todo{revise text}\todo{add let and ifz}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$.
+Its major constituents, introduced by example in the previous section, are:
 \begin{itemize}
@@
 we can only ``run'' a complete program, though see Sec.
-\ref{sec:future}.)
+\ref{sec:future}.)\todo{get rid of e/edot distinction}\todo{change this to refer to dynamics section}
 
@@
 
-The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:
+The rules for relative movement within Z-expressions are similar.
+Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:\todo{remove prevSib}
 \begin{subequations}
@@
 \end{equation}
-and this for the example
+and this for the example\todo{revise}
 \begin{equation}\label{r:movefirstchild-lam}
@@
 \end{equation}
+
+Reachability\todo{state reachability theorem}
 
@@
 
+Constructibility\todo{state constructibility theorem}
 \end{subequations}
@@
 
-\section{Dynamics of Hazelnut}
-\label{sec:dynamics}
+\section{Dynamics of Hazelnut}\label{sec:dynamics}
+
+\subsection{Complete Expressions}
+... canonical forms and type safety with e complete in it ...
+
+\subsection{Holes as TODOs}
+... add an e err and define new type safety ...
+
+\subsection{Indeterminate Expressions}
 \input{dynamics}
@@
 \clearpage
+
+\section{Extending Hazelnut}\label{sec:extending}
+\todo{write this section} Add sum types.
+
+H-expression
+
+Z-expression + Erasure rules
+
+Statics
+
+Movement actions
+
+Construct actions
 
@@
 \label{sec:future}
-This paper presented Hazelnut, a type theoretic structure editor calculus.
+\todo{revise}This paper presented Hazelnut, a type theoretic structure editor calculus.
 Our aim is to take a principled approach to its design by formally specifying its semantics, providing strong metatheoretic guarantees, mechanizing its semantics and metatheory in Agda and implementing it using the concepts of  functional reactive programming.
@@
 explored an equational theory for terms with holes, but intend to once our
-formalization effort is more mature.
+formalization effort is more mature.\todo{fix bib}
 
@@
 
-\bibliography{../research}
+\bibliography{../research}}
 %\softraggedright

### 2016-06-22 implementation.tex
added TODOs

@@
 
+\todo{make figure take up only one column}\todo{change terminology in figure}\todo{revise}
 The key question that must be answered for any implementation strategy is: how do we model a stream of actions from a user?

### 2016-06-22 intro.tex
added TODOs

@@
 
-%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
-Programmers and the tools that they use often encounter text that does not correspond to a meaningful program. This may be because the programmer is in the midst of a sequence of 
-edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake\todo{mention YoungSeok's data here?}.
-It is difficult for tools to help programmers determine and execute a corrective course of action in such situations (e.g.
-by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.) %because the language definition is silent %For example, a code editor cannot assist the programmer in these situations.
+%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
+Most programming languages define a textual syntax.
+This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs.
+While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.
+
+Cognitively, it requires that programmers understand the subtleties of text parsing (e.g.
+the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.\todo{citation?}
+
+It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}.
+This may be because the programmer is in the midst of a sequence of
+edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake.
+The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g.
+by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
 
@@
 
-\emph{Structure editors} have long promised to address this
-problem by allowing programmers to edit the tree structure of a program, or some rich projection thereof, 
-directly. Each edit action leaves the program being constructed in a
+\emph{Structure editors} are program editors where every edit action leaves the program being manipulated in a 
 structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor.
 However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
-whether a keyboard or a mouse is in use.} We give examples of notable structure editors in Sec.
-\ref{sec:rw}.
+whether a keyboard or a mouse is in use.} Scratch etc.
+for novices.
+mbeddr for pros.
+Lamdu and Unison ongoing efforts with a functional foundation.
+We give other examples of notable structure editors in Sec.
+\ref{sec:rw}.\todo{revise}
 
@@
 undefined) state, because language definitions generally only give meaning to complete, well-typed terms.
-This makes it difficult for humans and tools to reason about types and binding during the development process.
-%More sophisticated semantic reasoning principles thus remain .
+This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
+Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g.
+Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .
 
@@
 
-In this paper, we present our ongoing work on a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e.
+\todo{revise remainder of intro}In this paper, we present our ongoing work on a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e.
 Hazelnut is a \emph{structure editor calculus}.) In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics.
@@
   \item We then formally define Hazelnut in Section \ref{sec:hazel} and state the important metatheoretic properties.
+  \item Dynamics \ref{sec:dynamics}
 

### 2016-06-22 hazelnut-popl17.tex
added .bib

@@
 \label{sec:hazel}
-\todo{revise text}\todo{add let and ifz}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$.
+\todo{revise text}\todo{add let and ifz}\todo{make eqn numbering work better}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$.
 Its major constituents, introduced by example in the previous section, are:
@@
 The syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, or \emph{shapes}, $\varphi$, is given in Figure \ref{fig:action-syntax}.
-Actions are performed on Z-types and Z-expressions according to the \emph{action semantics} of Hazelnut, which is organized around three judgements:
+Actions are performed on Z-types and Z-expressions according to the \emph{action semantics} of Hazelnut, which is organized around three judgements:\todo{fix spacing}
 \[\arraycolsep=10pt\begin{array}{ll}
@@
 
-\bibliography{../research}}
+\bibliography{bibliography}
 %\softraggedright

### 2016-06-25 intro.tex
added citations

@@
 %Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program.
-Most programming languages define a textual syntax.
+
+Every day, programmers use text-based tools to construct and manipulate programs.  
+These programs are written in languages which are defined by a textual syntax. 
+While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.
+
+%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
+%Most programming languages define a textual syntax.
 This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs.
@@
 Cognitively, it requires that programmers understand the subtleties of text parsing (e.g.
-the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.\todo{citation?}
+the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.
+One study~\cite{Altadmri:2015:MCI:2676723.2677258} of novice programmer error found that the most common of all errors was an syntactic error, specifically unbalanced parenthesis.
+% Too much detail?
 
-It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}.
+It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program
+By analyzing the same data set used in~\cite{6883030}, which contains 1,460 hours of fine-grained code editing logs collected from 21 programmers, we found that 44.2\% of the code edits resulted in syntactically malformed state of the source code.
+%\todo{mention YoungSeok's data here}. 
 This may be because the programmer is in the midst of a sequence of 

### 2016-06-25 intro.tex
added more citations

@@
 
-It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program
+It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program.
 By analyzing the same data set used in~\cite{6883030}, which contains 1,460 hours of fine-grained code editing logs collected from 21 programmers, we found that 44.2\% of the code edits resulted in syntactically malformed state of the source code.
 %\todo{mention YoungSeok's data here}. 
-This may be because the programmer is in the midst of a sequence of 
-edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake.
+These malformed states could be errors, but they also may be because the programmer is in the midst of a sequence of edit actions that leaves the text temporarily malformed or ill-typed.
+%, or because the programmer has made a mistake. 
 The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g.
-by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
+by providing syntax highlighting~\cite{sarkar2015impact} and semantics-aware code completion services~\cite{Mooty:2010:CCC:1915084.1916348}.)
 

### 2016-06-25 intro.tex
did one pass over the intro, added a paragraph about Cognitive Dimensions.

@@
 % transformations on the structure of the program.
-
-\emph{Structure editors} are program editors where every edit action leaves the program being manipulated in a 
+One attempt to solve this problem has been \emph{Structure editors},
+where every edit action leaves the program being manipulated in a 
 structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor.
 However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
-whether a keyboard or a mouse is in use.} Scratch etc.
-for novices.
-mbeddr for pros.
-Lamdu and Unison ongoing efforts with a functional foundation.
-We give other examples of notable structure editors in Sec.
-\ref{sec:rw}.\todo{revise}
+whether a keyboard or a mouse is in use.}
+%Scratch etc. for novices. mbeddr for pros. Lamdu and Unison ongoing efforts with a functional foundation. We give other examples of notable structure editors in Sec. \ref{sec:rw}.\todo{revise}
 
-The problem is that it is still possible to leave the program in a semantically meaningless (i.e.
-undefined) state, because language definitions generally only give meaning to complete, well-typed terms.
+However, these syntactically correct states can be semantically meaningless (i.e.
+undefined), because the language definitions generally only give meaning to complete, well-typed terms.
+(e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
 This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
-Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g.
+Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
+%Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g.
 Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .
@@
 
-\todo{revise remainder of intro}In this paper, we present our ongoing work on a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e.
-Hazelnut is a \emph{structure editor calculus}.) In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics.
+%\todo{revise remainder of intro}
+
+In this paper, we present a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e. Hazelnut is a \emph{structure editor calculus}). 
+By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state.
+%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics.
 Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.
@@
 \ref{sec:actions}.) This does not imply that programs need to be constructed in a strictly outside-in manner, however, because an expression that has a type that is not yet consistent with its surroundings will automatically be put into a {hole} that defers consistency analysis until the hole is \emph{finished}.
+
+By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
+By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?
+In progress, unfinished, incomple?} and yet still in a well defined type state.
+This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.
+
+\todo{Talk about dynamics.}
 

### 2016-06-27 hazelnut-popl17.tex
tweaks

@@
 
-\emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors.
+\emph{Structure editors} allow programmers to edit the tree structure of a program directly.
 This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text.
@@
 
-This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful.
-Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner.
-To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}.
-Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.
-
-We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself.
-To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut.
-We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application.
+This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not only syntactically well-formed but also statically and dynamically meaningful.
+Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct terms in an unnatural ``outside-in'' manner.
+To avoid this problem, Hazelnut's {action semantics} automatically places terms assigned a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}.
+Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems.
+
+We intend Hazelnut to serve as a principled foundational calculus, rather than as an end-user tool itself.
+To that end, we demonstrate how Hazelnut's rich metatheory, which we have mechanized in Agda, guides an extension to Hazelnut.
+We also  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program.
 Our reference implementation is written using \lstinline{js_of_ocaml}.

### 2016-06-27 intro.tex
tweaks

--- 2016-06-25 intro.tex
@@
 %Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
-
-Every day, programmers use text-based tools to construct and manipulate programs.  
-These programs are written in languages which are defined by a textual syntax. 
-While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.
-
-%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
-%Most programming languages define a textual syntax.
-This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs.
-While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.
+Most contemporary programming languages define a textual syntax so that programmers can use text editors and other text-based tools to construct and manipulate programs.
+While this design is of substantial practical utility, it also introduces some fundamental complexity into the programming process.
 
 Cognitively, it requires that programmers understand the subtleties of text parsing (e.g.
-the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.
-One study~\cite{Altadmri:2015:MCI:2676723.2677258} of novice programmer error found that the most common of all errors was an syntactic error, specifically unbalanced parenthesis.
-% Too much detail?
+the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.\todo{citation?}
 
-It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program
-By analyzing the same data set used in~\cite{6883030}, which contains 1,460 hours of fine-grained code editing logs collected from 21 programmers, we found that 44.2\% of the code edits resulted in syntactically malformed state of the source code.
-%\todo{mention YoungSeok's data here}. 
+It also complicates matters for tool designers because tools are often confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}.
 This may be because the programmer is in the midst of a sequence of

### 2016-06-28 hazelnut-popl17.tex
adding plausible statements of construct and reach

--- 2016-06-22 hazelnut-popl17.tex
@@
 
-Reachability\todo{state reachability theorem}
+\paragraph{Reachability}
+
+We give the following theorem to demonstrate that the rules for movement do
+indeed capture the process of moving the focus to any editable position
+within a term. Intuitively, given two zippered expressions that are the
+same up to the erasure of focus, its proof gives a list of composable
+actions that, when applied to the first term produces the second.
+
+As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
+every action in such a list, we couldn't hope to construct an ill-typed
+term, so the statement here matches the mutually recursive structure of the
+bidirectional typing judgements.
+
+\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
+\label{thrm:reach}
+Both of the following hold:
+\begin{enumerate}
+  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
+   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
+   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
+   = \zexp'$.
+
+  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
+   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
+   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
+   = \zexp'$.
+\end{enumerate}
+\end{theorem}
 
@@
 
-Constructibility\todo{state constructibility theorem}
+\paragraph{Constructibility}
+
+We give the following theorem to demonstrate that the rules for the action
+$\aConstruct{\varphi}$ do indeed capture the process of building any term
+of the form $\varphi$. Intuitively, its proof gives a list of composable
+actions that, when applied one sequentially to an empty hole in focus,
+produce any well-typed expression.
+
+As with Theorem \ref{thrm:reach}, the statement of this theorem must match
+the mutually recursive structure of the bidirectional typing judgements.
+
+\begin{theorem}[Constructability]\todo{why $\htau$ and not $\ztau$?}
+\label{thrm:construct}
+Both of the following hold:
+\begin{enumerate}
+  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
+      L \in \mathit{list} ~\mathtt{action}$ such that
+      $\mathit{iterate}~ L~ \zwsel{\hhole{}} = \zexp'$ and
+      $\removeSel{\zexp'} = \hexp$
+
+  \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
+      L \in \mathit{list} ~\mathtt{action}$ such that $\mathit{iterate}~
+      L~ \zwsel{\hhole{}} = \zexp'$ and $\removeSel{\zexp'} = \hexp$
+\end{enumerate}
+\end{theorem}
+
+Because actions are defined on the zippered form of expressions, we argue
+that we can construct a term that is the same as the target up to the
+erasure of focus rather than on the nose. Since any form inside a zippered
+term may be in focus, there are linearly many such terms for a given
+$\hexp$---Theorem \ref{thrm:construct} can be composed with
+Theorem \ref{thrm:reach} to reach any one that might be desired in
+particular.
+
 \end{subequations}

### 2016-06-28 hazelnut-popl17.tex
Adding some todos, commentary in other areas of the action semantics. im starting to think of these theorems as semantic checksums. is that a good name?

@@
 \item \textbf{Z-types} and \textbf{Z-expressions} (Sec.
-\ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
+\ref{sec:cursors}), which superimpose a single \emph{focus}\todo{cursor?} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
 \item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.
@@
 \end{equation}
+\end{subequations}
 
@@
 every action in such a list, we couldn't hope to construct an ill-typed
-term, so the statement here matches the mutually recursive structure of the
-bidirectional typing judgements.
+term. Therefore the statement here matches the mutually recursive structure
+of the bidirectional typing judgements.
 
@@
 
-\end{subequations}
 \subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.
@@
 % \end{equation}
-
 \end{subequations}
+
+\paragraph{Deletability}\todo{do we care?}
+The meaning of the deletion action is to remove the entire term in focus,
+and its definition does not depend on the structure of the term being
+deleted. Therefore, there is no theorem to state that acts as a checksum on
+the rules analogous to Theorem \ref{thrm:reach} or \ref{thrm:construct},
+since we could not have forgotten any rules corresponding to the structure
+of the terms.
+
 \subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.
@@
 
+\paragraph{Finishability}\todo{do we care?}
+
+As with deletion, since the meaning of the finish action does not depend on
+the structure of the term being finished other than requiring that it must
+be one of the two hole forms, there is no checksum theorem to state for
+this action.
+
 \subsubsection{Zipper Cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression.
@@
 
-\paragraph{Constructibility}
+\paragraph{Constructibility}\todo{why is this with the zipper rules not the
+    construct rules?}
 
@@
 actions that, when applied one sequentially to an empty hole in focus,
-produce any well-typed expression.
+build any well-typed expression.
 

### 2016-06-28 hazelnut-popl17.tex
adding some todos

@@
 \end{equation}
-(It would be straightforward to also add a ``half-annotated'' lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out of our calculus \cite{DBLP:conf/tldi/ChlipalaPH05}.)
+It would be straightforward to also add a ``half-annotated'' lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out of our calculus \cite{DBLP:conf/tldi/ChlipalaPH05}.
 
@@
 
+\todo{redundant text w/ citation}
 The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}.
@@
 As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
-every action in such a list, we couldn't hope to construct an ill-typed
+every action in such a list, it is possible consider moving to an ill-typed
 term. Therefore the statement here matches the mutually recursive structure
@@
 \paragraph{Finishability}\todo{do we care?}
-
 As with deletion, since the meaning of the finish action does not depend on

### 2016-06-28 hazelnut-popl17.tex
adding some todos

@@
 
-\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus}
+\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
+ Calculus}
 %\subtitle{Modular Type Constructors}
@@
 % Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.
-
+\todo{why not title just ``hazelnut: a structure editor calculus''?}
 \emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors.
@@
 
-TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}
+\todo{TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}}
 
@@
 
-TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}
-
-TODO: cite \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}
+\todo{TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}}
+
+\todo{TODO: cite} \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}
 
@@
 
+\todo{new action form that makes actions extensible given proofs of
+admissiblity of a derived form, like prevSib}
+
 \begin{quote}

### 2016-06-28 hazelnut-popl17.tex

--- 2016-06-28 hazelnut-popl17.tex
@@
 
-This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful.
-Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner.
-To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}.
-Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.
-
-We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself.
-To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut.
-We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application.
+This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes two more steps in this direction: Hazelnut's edit actions operate over terms that are not only syntactically well-formed but also statically and dynamically meaningful.
+Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct terms in an unnatural ``outside-in'' manner.
+To avoid this problem, Hazelnut's {action semantics} automatically places terms assigned a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}.
+Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems.
+
+We intend Hazelnut to serve as a principled foundational calculus, rather than as an end-user tool itself.
+To that end, we demonstrate how Hazelnut's rich metatheory, which we have mechanized in Agda, guides an extension to Hazelnut.
+We also  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program.
 Our reference implementation is written using \lstinline{js_of_ocaml}.
@@
 \item \textbf{Z-types} and \textbf{Z-expressions} (Sec.
-\ref{sec:cursors}), which superimpose a single \emph{focus}\todo{cursor?} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
+\ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
 \item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.
@@
 \end{equation}
+
+Reachability\todo{state reachability theorem}
+
 \end{subequations}
-
-\paragraph{Reachability}
-
-We give the following theorem to demonstrate that the rules for movement do
-indeed capture the process of moving the focus to any editable position
-within a term. Intuitively, given two zippered expressions that are the
-same up to the erasure of focus, its proof gives a list of composable
-actions that, when applied to the first term produces the second.
-
-As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
-every action in such a list, we couldn't hope to construct an ill-typed
-term. Therefore the statement here matches the mutually recursive structure
-of the bidirectional typing judgements.
-
-\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
-\label{thrm:reach}
-Both of the following hold:
-\begin{enumerate}
-  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-   = \zexp'$.
-
-  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-   = \zexp'$.
-\end{enumerate}
-\end{theorem}
-
 \subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.
@@
 % \end{equation}
+
 \end{subequations}
-
-\paragraph{Deletability}\todo{do we care?}
-The meaning of the deletion action is to remove the entire term in focus,
-and its definition does not depend on the structure of the term being
-deleted. Therefore, there is no theorem to state that acts as a checksum on
-the rules analogous to Theorem \ref{thrm:reach} or \ref{thrm:construct},
-since we could not have forgotten any rules corresponding to the structure
-of the terms.
-
 \subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.
@@
 
-\paragraph{Finishability}\todo{do we care?}
-
-As with deletion, since the meaning of the finish action does not depend on
-the structure of the term being finished other than requiring that it must
-be one of the two hole forms, there is no checksum theorem to state for
-this action.
-
 \subsubsection{Zipper Cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression.
@@
 
-\paragraph{Constructibility}\todo{why is this with the zipper rules not the
-    construct rules?}
-
-We give the following theorem to demonstrate that the rules for the action
-$\aConstruct{\varphi}$ do indeed capture the process of building any term
-of the form $\varphi$. Intuitively, its proof gives a list of composable
-actions that, when applied one sequentially to an empty hole in focus,
-build any well-typed expression.
-
-As with Theorem \ref{thrm:reach}, the statement of this theorem must match
-the mutually recursive structure of the bidirectional typing judgements.
-
-\begin{theorem}[Constructability]\todo{why $\htau$ and not $\ztau$?}
-\label{thrm:construct}
-Both of the following hold:
-\begin{enumerate}
-  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
-      L \in \mathit{list} ~\mathtt{action}$ such that
-      $\mathit{iterate}~ L~ \zwsel{\hhole{}} = \zexp'$ and
-      $\removeSel{\zexp'} = \hexp$
-
-  \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
-      L \in \mathit{list} ~\mathtt{action}$ such that $\mathit{iterate}~
-      L~ \zwsel{\hhole{}} = \zexp'$ and $\removeSel{\zexp'} = \hexp$
-\end{enumerate}
-\end{theorem}
-
-Because actions are defined on the zippered form of expressions, we argue
-that we can construct a term that is the same as the target up to the
-erasure of focus rather than on the nose. Since any form inside a zippered
-term may be in focus, there are linearly many such terms for a given
-$\hexp$---Theorem \ref{thrm:construct} can be composed with
-Theorem \ref{thrm:reach} to reach any one that might be desired in
-particular.
-
+Constructibility\todo{state constructibility theorem}
 \end{subequations}

### 2016-06-28 intro.tex

--- 2016-06-25 intro.tex
@@
 %Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
+% Programming languages, and therefore programs themselves, are rich mathematical structures. 
+Well-typed programs are rich semantic structures, but human programmers typically construct and manipulate them only indirectly: by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics.
+This indirection is of substantial practical utility, to be sure -- text editors and other text-based tools benefit from decades of development effort.
+However, it also introduces some fundamental complexity into the programming process.
 
-Every day, programmers use text-based tools to construct and manipulate programs.  
-These programs are written in languages which are defined by a textual syntax. 
-While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.
+% Every day, programmers use text-based tools to construct and manipulate programs.  
+% These programs are written in languages which defined by a textual syntax. 
+% While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.
 
+% It also complicates matters for tool designers because tools are often confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}.
+This may be because the programmer is in the midst of a sequence of
+% edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake.
+The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g.
+by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
 %Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
@@
 
-Cognitively, it requires that programmers understand the subtleties of text parsing (e.g.
-the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.
-One study~\cite{Altadmri:2015:MCI:2676723.2677258} of novice programmer error found that the most common of all errors was an syntactic error, specifically unbalanced parenthesis.
+First, it requires that programmers understand the various subtleties of textual syntax (e.g.
+the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers \cite{Altadmri:2015:MCI:2676723.2677258}.
+%For example, one study of novice programmers found syntax errors (e.g.
+unbalanced parentheses) to be the most common class of errors.
 % Too much detail?
 
-It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program.
-By analyzing the same data set used in~\cite{6883030}, which contains 1,460 hours of fine-grained code editing logs collected from 21 programmers, we found that 44.2\% of the code edits resulted in syntactically malformed state of the source code.
-%\todo{mention YoungSeok's data here}. 
-These malformed states could be errors, but they also may be because the programmer is in the midst of a sequence of edit actions that leaves the text temporarily malformed or ill-typed.
+Moreover, the fact that not all sequences of characters correspond to a well-formed, meaningful program complicates matters for tool designers.
+Tools are confronted with such text quite frequently -- by analyzing a dataset gathered by Yoon and Myers \cite{6883030} (which contains 1,460 hours of fine-grained edit logs collected from 21 programmers) we found that 44.2\% of code edits resulted in malformed edit states.
+Some additional percentage of edit states were well-formed but ill-typed (the dataset did not include enough information to determine the exact percentage.)
+Some of these edit states were due to programmer error, while others arose simply because the programmer was in the midst of a sequence of edit actions that left the text temporarily malformed or ill-typed.
+The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g.
+syntax highlighting~\cite{sarkar2015impact} and semantics-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348}) and report meaningful error messages\todo{citations}.
 %, or because the programmer has made a mistake. 
-The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g.
-by providing syntax highlighting~\cite{sarkar2015impact} and semantics-aware code completion services~\cite{Mooty:2010:CCC:1915084.1916348}.)
+% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action
 
 Some editors have developed \emph{ad hoc} workarounds for this problem, e.g.
-they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading.
+they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, use regular expressions to highlight malformed program text, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading.
+% This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.
 

### 2016-07-03 hazelnut-popl17.tex
added colored highlighting to zipper focus; adding sum types

--- 2016-06-28 hazelnut-popl17.tex
@@
  Calculus}
-%\subtitle{Modular Type Constructors}
 
@@
 % Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.
-\todo{why not title just ``hazelnut: a structure editor calculus''?}
-\emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors.
-This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text.
-%Structure editors for professional programmers are also under active development (e.g.
-\texttt{mbeddr}.)
-
-This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful.
-Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner.
-To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}.
-Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.
-
-We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself.
-To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut.
-We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application.
+
+\emph{Structure editors} allow programmers to edit the tree structure of a program directly.
+This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they do not need to contend with malformed program text.
+
+This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{focus} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: it's {edit actions} operate over statically meaningful terms.
+Na\"ively, this prohibition on ill-typed edit states would force the programmer to construct terms in a rigid ``outside-in'' manner.
+To avoid this problem, the {action semantics} automatically places terms assigned a type that is inconsistent with the expected type {inside} a {hole}.
+This safely defers the type consistency check until the term inside the hole is \emph{finished}.
+
+Hazelnut is a foundational type-theoretic account of typed structure editing, rather than an end-user tool itself.
+To that end, we describe how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to the calculus.
+We also discuss various plausible evaluation strategies for terms with holes, and in so doing reveal connections with gradual typing and contextual modal type theory (the Curry-Howard interpretation of contextual modal logic.) Finally, we  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program.
 Our reference implementation is written using \lstinline{js_of_ocaml}.
@@
 
+In addition to eliminating malformed edit states, structure editors also promise to increase the speed of development.
+However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
+whether a keyboard or a mouse is in use.\todo{integrate this}
+
+By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
+By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?
+In progress, unfinished, incomple?} and yet still in a well defined type state.
+This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}
+
 \subsection{Future Work}

### 2016-07-03 intro.tex

@@
 Well-typed programs are rich semantic structures, but human programmers typically construct and manipulate them only indirectly: by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics.
-This indirection is of substantial practical utility, to be sure -- text editors and other text-based tools benefit from decades of development effort.
-However, it also introduces some fundamental complexity into the programming process.
+This indirection is of substantial practical utility, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.
 
@@
 
-Moreover, the fact that not all sequences of characters correspond to a well-formed, meaningful program complicates matters for tool designers.
-Tools are confronted with such text quite frequently -- by analyzing a dataset gathered by Yoon and Myers \cite{6883030} (which contains 1,460 hours of fine-grained edit logs collected from 21 programmers) we found that 44.2\% of code edits resulted in malformed edit states.
-Some additional percentage of edit states were well-formed but ill-typed (the dataset did not include enough information to determine the exact percentage.)
-Some of these edit states were due to programmer error, while others arose simply because the programmer was in the midst of a sequence of edit actions that left the text temporarily malformed or ill-typed.
+The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers.
+In particular, program editors confront meaningless text quite frequently: in a dataset collected by Yoon and Myers consisting of 1460 hours of fine-grained edit logs from 21 programmers, 44.2\% of edits resulted in a malformed edit state \cite{6883030}.
+Some additional percentage of edit states were well-formed but ill-typed (the dataset did not record the imported libraries, so an exact percentage could not be determined.)
+Some of these edit states could be attributed to programmer error, while others arose simply because the programmer was in the midst of a sequence of edit actions that left the text temporarily malformed or ill-typed.
 The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g.
-syntax highlighting~\cite{sarkar2015impact} and semantics-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348}) and report meaningful error messages\todo{citations}.
+syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring \todo{citation}) and to report localized, meaningful error messages\todo{citations}.
 %, or because the programmer has made a mistake. 
@@
 Some editors have developed \emph{ad hoc} workarounds for this problem, e.g.
-they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, use regular expressions to highlight malformed program text, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading.
+they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer.
 % This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.
@@
 % transformations on the structure of the program.
-One attempt to solve this problem has been \emph{Structure editors},
-where every edit action leaves the program being manipulated in a 
-structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor.
-However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
-whether a keyboard or a mouse is in use.}
-%Scratch etc. for novices. mbeddr for pros. Lamdu and Unison ongoing efforts with a functional foundation. We give other examples of notable structure editors in Sec. \ref{sec:rw}.\todo{revise}
+These complications have motivated a long line of research into \emph{structure editors}, i.e. program editors where edit actions maintain non-trivial structural invariants.
 
-However, these syntactically correct states can be semantically meaningless (i.e.
+Most structure editors are \emph{syntactic structure editors}, i.e.
+the edit states are trees with \emph{holes} representing branches that have yet to be constructed, and the edit actions are context-free tree transformations.
+Syntactic structure editors like Scratch \cite{Resnick:2009:SP:1592761.1592779} are popular in childhood educational settings, in part because they eliminate the possibility of syntax errors\todo{papers with case studies?}.
+
+The Scratch language has a trivial static semantics (i.e.
+it requires only that every assignable has been declared) and an \emph{ad hoc} dynamic semantics (i.e.
+if evaluation encounters a hole or a dynamic error, it goes on to the next command.) Researchers have also developed syntactic structure editors for  languages with a more conventional semantics.
+For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}.
+These editors all have interesting user interfaces, but there are several foundational problems that complicate their design.
+% \begin{itemize}
+% \item
+
+First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e.
+``the implementation is the specification''.
+This makes it difficult to reason methodically about types and binding, and therefore, to design sound and useful editor services.
+
+Developing a formal semantics and metatheory in the type theoretic style for the underlying language (e.g.
+following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}) would go only part of the way towards addressing the challenges faced by structure editor designers, because a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e.
+terms with holes.
+Moreover, a syntactic structure editor does not guarantee that every edit state is statically meaningful, so incorporating holes into the static semantics would in any case be only a minor salve.
+Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.
+
+%This brings us to the contributions of this paper. 
+This paper defines Hazelnut,  a \emph{typed structure editor calculus} that addresses these foundational problems. 
+In particular, Hazelnut is organized around a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}.
+Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e.
+well-typed) expression with a single \emph{focus} of action (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Actions act at the expression or type in focus (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
+a function argument), only actions consistent with that type are permitted.
+
+Na\"ively, enforcing this injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner.
+For example, the programmer would often need to construct the outer ``function application'' form before identifying the intended function.
+To avoid this problem, the action semantics automatically places expressions assigned a type that is inconsistent with the expected type inside a {hole}.
+This safely defers the type consistency check until the expression inside the hole is \emph{finished}.
+In other words, holes appear both at the leaves of the tree and at the internal nodes that remain under construction.
+%In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}.
+%Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)
+
+% By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state.
+%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics.
+Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.
+
+%However, these syntactically correct states can be semantically meaningless (i.e.
 undefined), because the language definitions generally only give meaning to complete, well-typed terms.
 (e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
-This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
-Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
+% This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
+% Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
 %Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g.
@@
 
-%% TODO: some sort of transition that talks about how you still don't have
-%% semantic reasoning principles because you can be left in a semantically
-%% ill-defined state
-
-%\todo{revise remainder of intro}
-
-In this paper, we present a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e. Hazelnut is a \emph{structure editor calculus}). 
-By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state.
-%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics.
-Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.
-In fact, edit actions maintain an even stronger invariant -- when acting on an expression whose type is determined by its surroundings (e.g.
-an expression in function argument position), only edit actions consistent with that type are permitted (we will formally state this invariant in Sec.
-\ref{sec:actions}.) This does not imply that programs need to be constructed in a strictly outside-in manner, however, because an expression that has a type that is not yet consistent with its surroundings will automatically be put into a {hole} that defers consistency analysis until the hole is \emph{finished}.
-
-By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
-By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?
-In progress, unfinished, incomple?} and yet still in a well defined type state.
-This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.
-
-\todo{Talk about dynamics.}
-
-The remainder of the paper is organized as follows:
-\begin{itemize}
-  \item We begin with an example to develop the reader's intuitions in Section
+The remainder of the paper is organized as follows. We begin with two example edit sequences to develop the reader's intuitions in Section
     \ref{sec:example}.
-
-  \item We then formally define Hazelnut in Section \ref{sec:hazel} and state the important metatheoretic properties.
-  \item Dynamics \ref{sec:dynamics}
-
-  \item In Section \ref{sec:mech}, we describe our ongoing effort to formalize the semantics and metatheory of Hazelnut in Agda.
-
-  \item In Section \ref{sec:impl}, we describe our ongoing effort to implement Hazelnut in a web browser, using a functional reactive model for user interaction.
-  \item In Section \ref{sec:rw}, we give an overview of related work.
-  \item We conclude in Section \ref{sec:future} by describing our vision for this work going 
-    forward.
+We then give a detailed overview of Hazelnut's semantics and metatheory, which has been mechanized in Agda, in Section \ref{sec:hazel}.
+Hazelnut is designed as a {foundational} calculus, i.e.
+as a calculus that language and editor designers extend with higher level constructs.
+We show how Hazelnut's rich mechanized metatheory guides and constrains the development of such an extension in Section \ref{sec:extending}.
+We consider several possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
+In so doing, we discover interesting connections with gradual typing and contextual modal type theory.
+The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes.
+It also suggests a principled design for an ``edit and resume'' feature.
+In Section \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
+Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
+In Section \ref{sec:rw}, we summarize related work.
+We conclude in Section \ref{sec:future} by describing our vision of a principled science of structure editing.
 \end{itemize}

### 2016-07-03 example.tex
updated abstract
added TODOs
more tweaks to intro

@@
 % !TEX root = hazelnut-popl17.tex
-\begin{figure*}[t!]
-\[
-\begin{array}{|c||c|c||l|l|}
-\hline
-\# & \textbf{H-Expression} & \textbf{Z-Expression} & \textbf{Next Action} & \textbf{Semantics}
-\\
-\hline
-1 &
-\hhole{} &
-\zwsel{\hhole{}}
-&
-\aConstruct{\flam{x}} & \refrule{\ref{r:conelamhole}}
-\\ 2 &
-\hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
-\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}}
-&
-\aConstruct{\fnum{}} & \refrule{\ref{r:contnum}}
-\\ 3 &
-\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
-\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}}
-&
-\aMove{\dNext{}} & \refrule{\ref{r:movenextsib}}
-\\ 4 &
-&
-\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
-&
-\aConstruct{\fnum{}} & \refrule{\ref{r:contnum}}
-\\ 5 &
-\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
-\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}}
-&
-\aMove{\dParent{}} & \refrule{\ref{r:moveparent}}
-\\ 6 &
-&
-\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
-&
-\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
-\\ 7 &
-&
-\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}}
-&
-\aMove{\dChild{}} & \refrule{\ref{r:movefirstchild-lam}}
-\\ 8 &
-&
-\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}}
-&
-\aConstruct{\fvar{x}} & \refrule{\ref{r:conevar}}
-\\ 9 &
-\hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
-&
-\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
-&
-\quad\textrm{---{}---}
-&
-\quad\textrm{---{}---}
-\\
-\hline
-\multicolumn{5}{c}{\text{... now assume a context where $id : \tarr{\tnum}{\tnum}$ ...}}\\
-\hline
-10 &
-\hhole{} &
-\zwsel{\hhole{}}
-&
-\aConstruct{\fasc} & \refrule{\ref{r:constructasc}}
-\\
-11 &
-\hhole{} : \hhole{} &
-\hhole{} : \zwsel{ \hhole{}}
-&
-\aConstruct{\fnum{}} & \refrule{\ref{r:contnum}}
-\\
-12 &
-\hhole{} :\tnum{} &
-\hhole{} : \zwsel{\tnum{}}
-&
-\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
-\\
-%13 &
-%\hhole{} :\tnum{} &
-%\zwsel{\hhole{}} : \tnum{}
-%&
-%\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
-%\\
-13 &
-%\hhole{} :\tnum{}
- &
-\zwsel{\hhole{}} : \tnum{}
-&
-\aConstruct{ \fvar{id}} & \refrule{\ref{r:conevar2}}
-\\
-14 &
-\hhole{\textrm{$id$}} : \tnum{} &
-\hhole{\zwsel{{\textrm{$id$}}}} : \tnum{}
-&
-\aConstruct{\fap{}} & \refrule{\ref{r:coneapfn}}
-\\
-15 &
-\hhole{\hap{{{\textrm{$id$}}}}{{\hhole{}}}} : \tnum{}
-&
-\hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hhole{}}}} : \tnum{}
-&
-\aConstruct{\fnumlit{3}} &  \refrule{\ref{r:conenumnum}}
-\\
-16 &
-\hhole{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
-&
-\hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hnum{3}}}} : \tnum{}
-&
-\aMove{\dParent{}} &  \refrule{\ref{r:moveparent-ap2}}
-\\
-17 &
-%\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
-&
-\hhole{\zwsel{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}} : \tnum{}
-&
-\aMove{\dParent{}} &  \refrule{\ref{r:moveparent-hole}}
-\\
-18 &
-%\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
-&
-\zwsel{\hhole{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{}
-&
-\aFinish &  \refrule{\ref{r:finishana}}
-\\
-19 &
-{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
-&
-\zwsel{{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{}
-&
-\quad\textrm{---{}---} & \quad\textrm{---{}---}
-\\
-\hline
 
-%% 11 &
-%% {\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \aMove{\dParent{}} & ?
-%% \\ 12 &
-%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \zwsel{\hlam{x}{{{x}}}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \aMove{\dParent{}} & \refrule{15b} bad
-%% \\ 13 &
-%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \zwsel{\hlam{x}{{{x}}} : \tarr{\tnum{}}{\tnum{}}}
-%% &
-%% \aConstruct{\fap{}} & \refrule{20h} bad
-%% \\ 14 &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{}}
-%% &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{}}}
-%% &
-%% \aConstruct{\fnumlit{3}} & \refrule{20l} bad
-%% \\ 15 &
-%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{3}}
-%% %&
-%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{3}}}
-%% %&
-%% %\aFinish{} & ?
-%% %\\
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{{\hnum{3}}}
-%% &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{{\hnum{3}}}}
-%% &
-%% \quad\textrm{---{}---}
-%% &
-%% \quad\textrm{---{}---}
-%% \\
-%% \hline
-\end{array}
-\]
-\caption{Constructing an identity function in Hazelnut (Lines~1--9), then  applying this function (assumed bound to $id$, not shown) to an argument~(Lines 10--19).
-The formal syntax and referenced rules in the final column are described in Section \ref{sec:hazel}.}
-\label{fig:first-example}
-\end{figure*}
 %
-Figure~\ref{fig:first-example} gives an example of the Hazelnut user
+\todo{add let and ifz}\todo{add let to example}\todo{finalize table}Figure~\ref{fig:first-example} gives an example of the Hazelnut user
 performing two simple programming tasks.

### 2016-07-03 intro.tex
more tweaks to intro

@@
 % !TEX root = hazelnut-popl17.tex
-
-%% Programs (and, by the Curry-Howard correspondence, proofs) are rich
-%% inductive structures. This fact is well understood amongst researchers
-%% and experienced programmers, but still somewhat obscure amongst
-%% programmers at large because programmers normally construct and interact
-%% with programs only indirectly, e.g. using a text editor composed with a
-%% parser.
 
@@
 % Programming languages, and therefore programs themselves, are rich mathematical structures.
-Well-typed programs are rich semantic structures, but human programmers typically construct and manipulate them only indirectly: by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics.
-This indirection is of substantial practical utility, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.
+Programmers typically construct and manipulate well-typed expressions only indirectly, by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics.
+This indirection has practical benefits, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.
 
@@
 
-First, it requires that programmers understand the various subtleties of textual syntax (e.g.
+First, it requires that programmers understand the various subtleties of the textual syntax (e.g.
 the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers \cite{Altadmri:2015:MCI:2676723.2677258}.
@@
 The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers.
-In particular, program editors confront meaningless text quite frequently: in a dataset collected by Yoon and Myers consisting of 1460 hours of fine-grained edit logs from 21 programmers, 44.2\% of edits resulted in a malformed edit state \cite{6883030}.
-Some additional percentage of edit states were well-formed but ill-typed (the dataset did not record the imported libraries, so an exact percentage could not be determined.)
-Some of these edit states could be attributed to programmer error, while others arose simply because the programmer was in the midst of a sequence of edit actions that left the text temporarily malformed or ill-typed.
+In particular, program editors encounter meaningless text on a regular basis: in a dataset collected by Yoon and Myers consisting of 1460 hours of fine-grained edit logs from 21 programmers, 44.2\% of edits resulted in a malformed edit state \cite{6883030}.
+Some additional percentage of edit states were well-formed but ill-typed (the dataset did not gather the imported libraries, so an exact percentage could not be determined.)
+Some of these edit states arose because the programmer was in the midst of a sequence of edits that left the edit state temporarily malformed or ill-typed, while others could be attributed to programmer error.
 The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g.
-syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring \todo{citation}) and to report localized, meaningful error messages\todo{citations}.
+syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring \todo{citation}), and to report localized, meaningful error messages\todo{citations}.
 %, or because the programmer has made a mistake. 
@@
 
-Some editors have developed \emph{ad hoc} workarounds for this problem, e.g.
-they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer.
+Many editors have developed \emph{ad hoc} workarounds for this problem, e.g.
+they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or continue past a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer.
 % This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.
@@
 % transformations on the structure of the program.
-These complications have motivated a long line of research into \emph{structure editors}, i.e. program editors where edit actions maintain non-trivial structural invariants.
+These complications have motivated a long line of research into \emph{structure editors}, i.e.
+program editors where the edit states are program structures, rather than text.
+This eliminates the possibility of syntax errors.
 
 Most structure editors are \emph{syntactic structure editors}, i.e.
-the edit states are trees with \emph{holes} representing branches that have yet to be constructed, and the edit actions are context-free tree transformations.
-Syntactic structure editors like Scratch \cite{Resnick:2009:SP:1592761.1592779} are popular in childhood educational settings, in part because they eliminate the possibility of syntax errors\todo{papers with case studies?}.
-
+the edit states are syntax trees with \emph{holes} that stand for branches of the tree that have yet to be constructed, and the edit actions are context-free tree transformations.
+For example, Scratch is a syntactic structure editor that has achieved success as a tool for teaching children how to program \cite{Resnick:2009:SP:1592761.1592779}.
 The Scratch language has a trivial static semantics (i.e.
 it requires only that every assignable has been declared) and an \emph{ad hoc} dynamic semantics (i.e.
-if evaluation encounters a hole or a dynamic error, it goes on to the next command.) Researchers have also developed syntactic structure editors for  languages with a more conventional semantics.
+if evaluation encounters a hole or a dynamic error, it goes on to the next command.)
+
+Researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics.
 For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}.
-These editors all have interesting user interfaces, but there are several foundational problems that complicate their design.
+Each of these editors has an innovative user interface, but the non-trivial type and binding structure of the underlying language poses several challenges that complicates its design.
 % \begin{itemize}
@@
 
-First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e.
+First, there is an all too familiar challenge: these languages do not have a clear formal semantics and metatheory, i.e.
 ``the implementation is the specification''.
-This makes it difficult to reason methodically about types and binding, and therefore, to design sound and useful editor services.
+This makes it difficult to reason methodically about types and binding, so, in turn, it is difficult to design sound and useful editor services.
 
-Developing a formal semantics and metatheory in the type theoretic style for the underlying language (e.g.
-following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}) would go only part of the way towards addressing the challenges faced by structure editor designers, because a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e.
+Even if a formal semantics and metatheory for the underlying language were to be developed (e.g.
+following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}), it would go only part of the way towards addressing the challenges faced by structure editor designers.
+First, a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e.
 terms with holes.
-Moreover, a syntactic structure editor does not guarantee that every edit state is statically meaningful, so incorporating holes into the static semantics would in any case be only a minor salve.
+Moreover, a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed -- so incorporating holes into the static semantics would in any case be only a minor salve.
 Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.
@@
 We conclude in Section \ref{sec:future} by describing our vision of a principled science of structure editing.
-\end{itemize}
+
+\begin{figure}[t!]
+\[
+\begin{array}{|c||c|c||l|l|}
+\hline
+\# & \textbf{Z-Expression} & 
+%\textbf{H-Expression} & 
+% \textbf{Type} & 
+\textbf{Next Action} & \textbf{Rule}
+\\
+\hline
+1 &
+\zwsel{\hhole{}} & 
+% \hhole{} &
+% \tehole 
+% &
+\aConstruct{\flam{x}} & 
+\text{(\ref{r:conelamhole})}
+\\ 2 &
+\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} & 
+% \hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
+% \tarr{\tehole}{\tehole} &
+\aConstruct{\fnum{}} &
+\text{(\ref{r:contnum})}
+\\ 3 &
+\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
+% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
+% \tarr{\tnum}{\tehole} & 
+\aMove{\dNext{}} & 
+\text{({\ref{r:movenextsib}})}
+\\ 4 &
+\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
+&
+% \text{\textquotedbl}&
+% \tarr{\tnum}{\tehole} & 
+\aConstruct{\fnum{}} & 
+\text{(\ref{r:contnum})}
+\\ 5 &
+\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} & 
+% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
+% \tarr{\tnum}{\tnum} &
+\aMove{\dParent{}} & 
+\text{(\ref{r:moveparent})}
+\\ 6 &
+\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
+&
+% \text{\textquotedbl}&
+% \tarr{\tnum}{\tnum} & 
+\aMove{\dPrev{}} & 
+\text{(\ref{r:moveprevsib})}
+\\ 7 &
+% &
+\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} & 
+% \tarr{\tnum}{\tnum} &
+% &
+\aMove{\dChild{}} & 
+\text{(\ref{r:movefirstchild-lam})}
+\\ 8 &
+% &
+\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
+% \tarr{\tnum}{\tnum} & 
+\aConstruct{\fvar{x}} & 
+\text{(\ref{r:conevar})}
+\\ 9 &
+% \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
+% &
+\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
+% \tarr{\tnum}{\tnum} & 
+\quad\textrm{---}
+&
+\quad\textrm{---}
+\\
+\hline
+% \multicolumn{5}{c}{\text{... now assume a context where $id : \tarr{\tnum}{\tnum}$ ...}}\\
+% \hline
+% 10 &
+% % \hhole{} &
+% \zwsel{\hhole{}} & 
+% \tehole &
+% \aConstruct{\fasc} & 
+% \text{(\ref{r:constructasc})}
+% \\
+% 11 &
+% % \hhole{} : \hhole{} &
+% \hhole{} : \zwsel{ \hhole{}} &
+% \tehole & 
+% \aConstruct{\fnum{}} & 
+% \text{(\ref{r:contnum})}
+% \\
+% 12 &
+% % \hhole{} :\tnum{} &
+% \hhole{} : \zwsel{\tnum{}} &
+% \tnum &
+% \aMove{\dPrev{}} & 
+% \text{(\ref{r:moveprevsib})}
+% \\
+% % %13 &
+% % %\hhole{} :\tnum{} &
+% % %\zwsel{\hhole{}} : \tnum{}
+% % %&
+% % %\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
+% % %\\
+% 13 &
+% % \hhole{} :\tnum{}
+%  % &
+% \zwsel{\hhole{}} : \tnum{} &
+% \tnum & 
+% \aConstruct{ \fvar{id}} & 
+% \text{(\ref{r:conevar2})}
+% \\
+% 14 &
+% % \hhole{\textrm{$id$}} : \tnum{} &
+% \hhole{\zwsel{{\textrm{$id$}}}} : \tnum{} &
+% \tnum & 
+% \aConstruct{\fap{}} & 
+% \text{(\ref{r:coneapfn})}
+% \\
+% 15 &
+% % \hhole{\hap{{{\textrm{$id$}}}}{{\hhole{}}}} : \tnum{}
+% % &
+% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hhole{}}}} : \tnum{} &
+% \tnum & 
+% \aConstruct{\fnumlit{3}} &  
+% \text{(\ref{r:conenumnum})}
+% \\
+% 16 &
+% % \hhole{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hnum{3}}}} : \tnum{} &
+% \tnum & 
+% \aMove{\dParent{}} &  
+% \text{(\ref{r:moveparent-ap2})}
+% \\
+% 17 &
+% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \hhole{\zwsel{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}} : \tnum{} &
+% \tnum & 
+% \aMove{\dParent{}} &  
+% \text{(\ref{r:moveparent-hole})}
+% \\
+% 18 &
+% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \zwsel{\hhole{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
+% \tnum & 
+% \aFinish &  
+% \text{(\ref{r:finishana})}
+% \\
+% 19 &
+% % {\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \zwsel{{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
+% \tnum & 
+% \quad\textrm{---} & 
+% \quad\textrm{---}
+% \\
+% \hline
+
+%% 11 &
+%% {\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \aMove{\dParent{}} & ?
+%% \\ 12 &
+%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \zwsel{\hlam{x}{{{x}}}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \aMove{\dParent{}} & \refrule{15b} bad
+%% \\ 13 &
+%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \zwsel{\hlam{x}{{{x}}} : \tarr{\tnum{}}{\tnum{}}}
+%% &
+%% \aConstruct{\fap{}} & \refrule{20h} bad
+%% \\ 14 &
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{}}
+%% &
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{}}}
+%% &
+%% \aConstruct{\fnumlit{3}} & \refrule{20l} bad
+%% \\ 15 &
+%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{3}}
+%% %&
+%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{3}}}
+%% %&
+%% %\aFinish{} & ?
+%% %\\
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{{\hnum{3}}}
+%% &
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{{\hnum{3}}}}
+%% &
+%% \quad\textrm{---{}---}
+%% &
+%% \quad\textrm{---{}---}
+%% \\
+%% \hline
+\end{array}
+\]
+\caption{Constructing an identity function in Hazelnut (Lines~1--9), then  applying this function (assumed bound to $id$, not shown) to an argument~(Lines 10--19).
+The formal syntax and referenced rules in the final column are described in Section \ref{sec:hazel}.}
+\label{fig:first-example}
+\end{figure}

### 2016-07-03 intro.tex
tweaks

@@
 % Programming languages, and therefore programs themselves, are rich mathematical structures.
-Programmers typically construct and manipulate well-typed expressions only indirectly, by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics.
+Programmers typically construct and manipulate well-typed expressions only indirectly, by editing text that is first parsed according to a textual syntax and then typechecked according to a static semantics.
 This indirection has practical benefits, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.
@@
 First, it requires that programmers understand the various subtleties of the textual syntax (e.g.
-the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers \cite{Altadmri:2015:MCI:2676723.2677258}.
+the relative precedence of the various forms.) This can be particularly challenging for novice programmers \cite{Altadmri:2015:MCI:2676723.2677258}.
 %For example, one study of novice programmers found syntax errors (e.g.
@@
 The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers.
-In particular, program editors encounter meaningless text on a regular basis: in a dataset collected by Yoon and Myers consisting of 1460 hours of fine-grained edit logs from 21 programmers, 44.2\% of edits resulted in a malformed edit state \cite{6883030}.
+In particular, program editors must contend with meaningless text on a regular basis: in a dataset collected by Yoon and Myers consisting of 1460 hours of fine-grained edit logs from 21 programmers, 44.2\% of edits resulted in a malformed edit state \cite{6883030}.
 Some additional percentage of edit states were well-formed but ill-typed (the dataset did not gather the imported libraries, so an exact percentage could not be determined.)
@@
 The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g.
-syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring \todo{citation}), and to report localized, meaningful error messages\todo{citations}.
+syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \todo{citation}), and to report localized, meaningful error messages\todo{citations}.\todo{heuristics}
 %, or because the programmer has made a mistake. 
@@
 
-Many editors have developed \emph{ad hoc} workarounds for this problem, e.g.
+% Many editors have developed \emph{ad hoc} workarounds for this problem, e.g.
 they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or continue past a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer.
@@
 These complications have motivated a long line of research into \emph{structure editors}, i.e.
-program editors where the edit states are program structures, rather than text.
-This eliminates the possibility of syntax errors.
+program editors where the edit state is a  program structure, rather than text.
+% Eliminating text eliminates the possibility of syntax errors.
 
 Most structure editors are \emph{syntactic structure editors}, i.e.
-the edit states are syntax trees with \emph{holes} that stand for branches of the tree that have yet to be constructed, and the edit actions are context-free tree transformations.
+the edit state is a syntax tree with \emph{holes} that stand for branches of the tree that have yet to be constructed, and the edit actions are context-free tree transformations.
 For example, Scratch is a syntactic structure editor that has achieved success as a tool for teaching children how to program \cite{Resnick:2009:SP:1592761.1592779}.
-The Scratch language has a trivial static semantics (i.e.
-it requires only that every assignable has been declared) and an \emph{ad hoc} dynamic semantics (i.e.
-if evaluation encounters a hole or a dynamic error, it goes on to the next command.)
 
-Researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics.
+The Scratch language has a trivial static semantics, but researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics.
 For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}.
-Each of these editors has an innovative user interface, but the non-trivial type and binding structure of the underlying language poses several challenges that complicates its design.
+Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language poses several challenges that complicates its design.
 % \begin{itemize}
@@
 ``the implementation is the specification''.
-This makes it difficult to reason methodically about types and binding, so, in turn, it is difficult to design sound and useful editor services.
+This makes it difficult to reason methodically about types and binding, so, in turn, it is difficult to design reliable editor services.
 
-Even if a formal semantics and metatheory for the underlying language were to be developed (e.g.
+Even if a formal semantics and metatheory for the underlying language were forthcoming (e.g.
 following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}), it would go only part of the way towards addressing the challenges faced by structure editor designers.
@@
 
-%This brings us to the contributions of this paper. 
-This paper defines Hazelnut,  a \emph{typed structure editor calculus} that addresses these foundational problems. 
-In particular, Hazelnut is organized around a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}.
+This paper seeks to address these foundational challenges.
+In particular, we introduce Hazelnut,  a \emph{typed structure editor}  organized around a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}.
 Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e.
-well-typed) expression with a single \emph{focus} of action (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Actions act at the expression or type in focus (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
+well-typed) expression with a single superimposed \emph{focus} (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Actions act relative to the expression or type in focus (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
 a function argument), only actions consistent with that type are permitted.
@@
 Na\"ively, enforcing this injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner.
-For example, the programmer would often need to construct the outer ``function application'' form before identifying the intended function.
-To avoid this problem, the action semantics automatically places expressions assigned a type that is inconsistent with the expected type inside a {hole}.
+For example, the programmer would often need to construct the outer function application form before identifying the intended function.
+To avoid this problem, Hazelnut automatically places newly constructed expressions inside a hole if the expression's type is inconsistent with the expected type.
 This safely defers the type consistency check until the expression inside the hole is \emph{finished}.
-In other words, holes appear both at the leaves of the tree and at the internal nodes that remain under construction.
+In other words, holes appear both at the leaves and at the internal nodes of the syntax tree that remain under construction.
 %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}.
@@
 
-The remainder of the paper is organized as follows. We begin with two example edit sequences to develop the reader's intuitions in Section
-    \ref{sec:example}.
-We then give a detailed overview of Hazelnut's semantics and metatheory, which has been mechanized in Agda, in Section \ref{sec:hazel}.
-Hazelnut is designed as a {foundational} calculus, i.e.
+The remainder of the paper is organized as follows:
+\begin{itemize}[noitemsep,nolistsep]
+\item We begin in Section
+    \ref{sec:example} with two example edit sequences to develop the reader's intuitions.  
+\item We then give a detailed overview of Hazelnut's semantics and metatheory, which has been mechanized in Agda, in Section \ref{sec:hazel}. 
+\item Hazelnut is designed as a {foundational} calculus, i.e.
 as a calculus that language and editor designers extend with higher level constructs.
 We show how Hazelnut's rich mechanized metatheory guides and constrains the development of such an extension in Section \ref{sec:extending}.
-We consider several possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
+\item In Section \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
+Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
+
+\item We consider several possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
 In so doing, we discover interesting connections with gradual typing and contextual modal type theory.
@@
 It also suggests a principled design for an ``edit and resume'' feature.
-In Section \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
-Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
-In Section \ref{sec:rw}, we summarize related work.
-We conclude in Section \ref{sec:future} by describing our vision of a principled science of structure editing.
+
+\item In Section \ref{sec:rw}, we summarize the related work.
+\item We conclude in Section \ref{sec:future} by summarizing our vision of a principled science of structure editing rooted in type theory, and suggest a number of future directions.
+\end{itemize}
 
@@
 \]
-\caption{Constructing an identity function in Hazelnut (Lines~1--9), then  applying this function (assumed bound to $id$, not shown) to an argument~(Lines 10--19).
-The formal syntax and referenced rules in the final column are described in Section \ref{sec:hazel}.}
+\caption{Constructing an identity function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
 \label{fig:first-example}

### 2016-07-03 intro.tex
intro shortened

@@
 For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}.
-Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language poses several challenges that complicates its design.
+Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design.
+The reason for this is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed.
+Indeed, it is not entirely clear what it should mean for an edit state that contains holes to be statically meaningful.
+Consequently, heuristics analagous to those that pervade textual program editors also pervade these editors.
+
 % \begin{itemize}
@@
 
-First, there is an all too familiar challenge: these languages do not have a clear formal semantics and metatheory, i.e.
+% First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e.
 ``the implementation is the specification''.
@@
 
-Even if a formal semantics and metatheory for the underlying language were forthcoming (e.g.
+% Even if a formal semantics and metatheory for the underlying language were forthcoming (e.g.
 following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}), it would go only part of the way towards addressing the challenges faced by structure editor designers.
@@
 terms with holes.
-Moreover, a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed -- so incorporating holes into the static semantics would in any case be only a minor salve.
+Moreover, a syntactic structure editor  -- so incorporating holes into the static semantics would in any case be only a minor salve.
 Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.
@@
 \end{itemize}
-
-\begin{figure}[t!]
-\[
-\begin{array}{|c||c|c||l|l|}
-\hline
-\# & \textbf{Z-Expression} & 
-%\textbf{H-Expression} & 
-% \textbf{Type} & 
-\textbf{Next Action} & \textbf{Rule}
-\\
-\hline
-1 &
-\zwsel{\hhole{}} & 
-% \hhole{} &
-% \tehole 
-% &
-\aConstruct{\flam{x}} & 
-\text{(\ref{r:conelamhole})}
-\\ 2 &
-\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} & 
-% \hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
-% \tarr{\tehole}{\tehole} &
-\aConstruct{\fnum{}} &
-\text{(\ref{r:contnum})}
-\\ 3 &
-\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
-% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
-% \tarr{\tnum}{\tehole} & 
-\aMove{\dNext{}} & 
-\text{({\ref{r:movenextsib}})}
-\\ 4 &
-\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
-&
-% \text{\textquotedbl}&
-% \tarr{\tnum}{\tehole} & 
-\aConstruct{\fnum{}} & 
-\text{(\ref{r:contnum})}
-\\ 5 &
-\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} & 
-% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
-% \tarr{\tnum}{\tnum} &
-\aMove{\dParent{}} & 
-\text{(\ref{r:moveparent})}
-\\ 6 &
-\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
-&
-% \text{\textquotedbl}&
-% \tarr{\tnum}{\tnum} & 
-\aMove{\dPrev{}} & 
-\text{(\ref{r:moveprevsib})}
-\\ 7 &
-% &
-\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} & 
-% \tarr{\tnum}{\tnum} &
-% &
-\aMove{\dChild{}} & 
-\text{(\ref{r:movefirstchild-lam})}
-\\ 8 &
-% &
-\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
-% \tarr{\tnum}{\tnum} & 
-\aConstruct{\fvar{x}} & 
-\text{(\ref{r:conevar})}
-\\ 9 &
-% \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
-% &
-\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
-% \tarr{\tnum}{\tnum} & 
-\quad\textrm{---}
-&
-\quad\textrm{---}
-\\
-\hline
-% \multicolumn{5}{c}{\text{... now assume a context where $id : \tarr{\tnum}{\tnum}$ ...}}\\
-% \hline
-% 10 &
-% % \hhole{} &
-% \zwsel{\hhole{}} & 
-% \tehole &
-% \aConstruct{\fasc} & 
-% \text{(\ref{r:constructasc})}
-% \\
-% 11 &
-% % \hhole{} : \hhole{} &
-% \hhole{} : \zwsel{ \hhole{}} &
-% \tehole & 
-% \aConstruct{\fnum{}} & 
-% \text{(\ref{r:contnum})}
-% \\
-% 12 &
-% % \hhole{} :\tnum{} &
-% \hhole{} : \zwsel{\tnum{}} &
-% \tnum &
-% \aMove{\dPrev{}} & 
-% \text{(\ref{r:moveprevsib})}
-% \\
-% % %13 &
-% % %\hhole{} :\tnum{} &
-% % %\zwsel{\hhole{}} : \tnum{}
-% % %&
-% % %\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
-% % %\\
-% 13 &
-% % \hhole{} :\tnum{}
-%  % &
-% \zwsel{\hhole{}} : \tnum{} &
-% \tnum & 
-% \aConstruct{ \fvar{id}} & 
-% \text{(\ref{r:conevar2})}
-% \\
-% 14 &
-% % \hhole{\textrm{$id$}} : \tnum{} &
-% \hhole{\zwsel{{\textrm{$id$}}}} : \tnum{} &
-% \tnum & 
-% \aConstruct{\fap{}} & 
-% \text{(\ref{r:coneapfn})}
-% \\
-% 15 &
-% % \hhole{\hap{{{\textrm{$id$}}}}{{\hhole{}}}} : \tnum{}
-% % &
-% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hhole{}}}} : \tnum{} &
-% \tnum & 
-% \aConstruct{\fnumlit{3}} &  
-% \text{(\ref{r:conenumnum})}
-% \\
-% 16 &
-% % \hhole{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hnum{3}}}} : \tnum{} &
-% \tnum & 
-% \aMove{\dParent{}} &  
-% \text{(\ref{r:moveparent-ap2})}
-% \\
-% 17 &
-% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \hhole{\zwsel{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}} : \tnum{} &
-% \tnum & 
-% \aMove{\dParent{}} &  
-% \text{(\ref{r:moveparent-hole})}
-% \\
-% 18 &
-% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \zwsel{\hhole{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
-% \tnum & 
-% \aFinish &  
-% \text{(\ref{r:finishana})}
-% \\
-% 19 &
-% % {\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \zwsel{{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
-% \tnum & 
-% \quad\textrm{---} & 
-% \quad\textrm{---}
-% \\
-% \hline
-
-%% 11 &
-%% {\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \aMove{\dParent{}} & ?
-%% \\ 12 &
-%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \zwsel{\hlam{x}{{{x}}}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \aMove{\dParent{}} & \refrule{15b} bad
-%% \\ 13 &
-%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \zwsel{\hlam{x}{{{x}}} : \tarr{\tnum{}}{\tnum{}}}
-%% &
-%% \aConstruct{\fap{}} & \refrule{20h} bad
-%% \\ 14 &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{}}
-%% &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{}}}
-%% &
-%% \aConstruct{\fnumlit{3}} & \refrule{20l} bad
-%% \\ 15 &
-%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{3}}
-%% %&
-%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{3}}}
-%% %&
-%% %\aFinish{} & ?
-%% %\\
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{{\hnum{3}}}
-%% &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{{\hnum{3}}}}
-%% &
-%% \quad\textrm{---{}---}
-%% &
-%% \quad\textrm{---{}---}
-%% \\
-%% \hline
-\end{array}
-\]
-\caption{Constructing an identity function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
-\label{fig:first-example}
-\end{figure}

### 2016-07-03 intro.tex
final tweaks to intro, started working on example

@@
 
-First, it requires that programmers understand the various subtleties of the textual syntax (e.g.
-the relative precedence of the various forms.) This can be particularly challenging for novice programmers \cite{Altadmri:2015:MCI:2676723.2677258}.
+First, it requires that programmers learn the subtleties of the textual syntax (e.g.
+the precedence of the various forms.) This can be particularly challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}.
 %For example, one study of novice programmers found syntax errors (e.g.
@@
 The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers.
-In particular, program editors must contend with meaningless text on a regular basis: in a dataset collected by Yoon and Myers consisting of 1460 hours of fine-grained edit logs from 21 programmers, 44.2\% of edits resulted in a malformed edit state \cite{6883030}.
-Some additional percentage of edit states were well-formed but ill-typed (the dataset did not gather the imported libraries, so an exact percentage could not be determined.)
-Some of these edit states arose because the programmer was in the midst of a sequence of edits that left the edit state temporarily malformed or ill-typed, while others could be attributed to programmer error.
-The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g.
-syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \todo{citation}), and to report localized, meaningful error messages\todo{citations}.\todo{heuristics}
+In particular, program editors must contend with meaningless text on a regular basis.
+In a dataset analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of edit states were syntactically malformed \cite{6883030}.
+Some additional percentage of edit states were well-formed but ill-typed (the dataset did not contain enough information to determine the exact percentage.)
+Some of these meaningless edit states arose because the programmer was in the midst of a sequence of edits that left the edit state transiently meaningless, while others could be attributed to the programmer making a mistake.
+Because the language definitions is silent about these edit states, it is difficult to design useful editor services, e.g.
+syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \todo{citation}.
+Editors must either disable these services when they encounter these edit states or develop \emph{ad hoc} heuristics, which can confuse or mislead the programmer.
 %, or because the programmer has made a mistake. 
@@
 Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design.
-The reason for this is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed.
-Indeed, it is not entirely clear what it should mean for an edit state that contains holes to be statically meaningful.
-Consequently, heuristics analagous to those that pervade textual program editors also pervade these editors.
+The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is well-formed.
+Indeed, it is not entirely clear what it should mean for term with holes to be statically meaningful.
+Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.
 
@@
 
-This paper seeks to address these foundational challenges.
-In particular, we introduce Hazelnut,  a \emph{typed structure editor}  organized around a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}.
+This paper seeks to address these foundational challenges from first principles.
+We introduce Hazelnut,  a \emph{typed structure editor}  based on a small bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}.
 Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e.
@@
 Na\"ively, enforcing this injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner.
-For example, the programmer would often need to construct the outer function application form before identifying the intended function.
-To avoid this problem, Hazelnut automatically places newly constructed expressions inside a hole if the expression's type is inconsistent with the expected type.
-This safely defers the type consistency check until the expression inside the hole is \emph{finished}.
-In other words, holes appear both at the leaves and at the internal nodes of the syntax tree that remain under construction.
+For example, the programmer might need to construct the outer function application form before identifying the intended function.
+To avoid this problem, Hazelnut leaves newly constructed expressions inside a hole if the expression's type is inconsistent with the expected type.
+This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}.
+In other words, holes appear both at the leaves and at the internal nodes of the typed syntax tree that remain under construction.
 %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}.
@@
 
-The remainder of the paper is organized as follows:
+The remainder of this paper is organized as follows:
 \begin{itemize}[noitemsep,nolistsep]
@@
     \ref{sec:example} with two example edit sequences to develop the reader's intuitions.  
-\item We then give a detailed overview of Hazelnut's semantics and metatheory, which has been mechanized in Agda, in Section \ref{sec:hazel}. 
+\item We then give a detailed overview of Hazelnut's semantics and metatheory, which have been mechanized using the Agda proof assistant, in Section \ref{sec:hazel}.
 \item Hazelnut is designed as a {foundational} calculus, i.e.
@@
 
-\item We consider several possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
+\item This paper is primarily concerned with the static semantics of incomplete expressions.
+However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
 In so doing, we discover interesting connections with gradual typing and contextual modal type theory.
 The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes.
-It also suggests a principled design for an ``edit and resume'' feature.
+It also suggests a principled design for an ``edit and resume'' feature.\todo{revise this one}
 
 \item In Section \ref{sec:rw}, we summarize the related work.
-\item We conclude in Section \ref{sec:future} by summarizing our vision of a principled science of structure editing rooted in type theory, and suggest a number of future directions.
+\item We conclude in Section \ref{sec:future} by summarizing this vision of a principled science of structure editing rooted in type theory, and suggest a number of future directions.
 \end{itemize} 
+\todo{summarize contents of the supplement here?}

### 2016-07-04 example.tex
intro shortened
progress on sec 2

@@
 %
-\todo{add let and ifz}\todo{add let to example}\todo{finalize table}Figure~\ref{fig:first-example} gives an example of the Hazelnut user
-performing two simple programming tasks.
-The syntactic forms in this figure will be formally defined in Sec.
+Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from an empty hole via a sequence of {actions}.
+We will discuss the syntactic forms and the referenced rules in Sec.
 \ref{sec:hazel}.
-For now, we will develop only the necessary intuitions.
-In the first task (Lines 1-9), the user constructs the identity function over numbers.
+For now, let us develop only the high-level intuitions.
+
+%In the first task (Lines 1-9), the user constructs the identity function over numbers.
 In the second task (Lines 10-19), the user applies this function (assumed to be bound to a variable, $id$), to the number expression $\hnum{3}$.
-Each of these tasks is carried out interactively, through the sequence of \emph{actions} shown in the  column labeled \textbf{Next Action}.
+% Each of these tasks is carried out interactively, through the sequence of \emph{actions} shown in the  column labeled \textbf{Next Action}.
 For reference, we cite the relevant rules from Sec.
@@
 
-The second and third columns of the
-table show the program as it is being constructed in two forms.
+% The second and third columns of the
+% table show the program as it is being constructed in two forms.
 The second column shows it as an \textbf{H-expression}, which is an expression that can contain \emph{holes}, delimited by $\llparenthesis$ and $\rrparenthesis$.
@@
 % on working on filling just one of the holes.
-Each action produces a new Z-expression, but this may or may not correspond to a new H-expression (in particular, some actions only move the focus, without changing the structure of the term.)
+% Each action produces a new Z-expression, but this may or may not correspond to a new H-expression (in particular, some actions only move the focus, without changing the structure of the term.)
 % to the hole in
@@
 
-Line~1 begins with the simplest initial expression: an H-expression
-consisting of a single hole. The corresponding Z-Expression has that hole in focus,
-indicated by the syntax~$\zwsel{\hehole}$.
-Focus determines the locus of action.
-The first action the user performs is $\aConstruct{\flam{x}}$, which replaces the hole with a lambda abstraction binding the variable $x$.
+The edit state in Hazelnut is a {Z-expression}, $\zexp$.
+Every Z-expression has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the {cursor}.
+The selected H-expression or H-type is typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively.
+For example, on Line 1 of Figure \ref{fig:first-example}, the empty expression hole, $\hhole{}$, is under the cursor.
+
+Actions act relative to the cursor.
+The first action the user performs is $\aConstruct{\flam{x}}$, which instantiates the hole with a lambda abstraction binding the variable $x$.
 This results in the program on line
-2, consisting of a lambda abstraction ascribed an arrow type with holes in all positions. The argument type hole is in focus. The
+2, consisting of a lambda abstraction ascribed an arrow type with holes in all positions.
+The cursor is placed on the argument type hole.
+Type holes, like empty expression holes, are also typeset $\tehole$ for visual consistency.
+
+The
 user proceeds to fill these holes using construction and movement actions, resulting in the final expression on Line 9.
@@
 Lines 10-12 similarly begin in a type-directed manner with the user giving an explicit type ascription, indicating that the expression that they are constructing will have type $\tnum$.
+
+\begin{figure}[t!]
+\[
+\begin{array}{|c||c|l||l|l|}
+\hline
+\# & \textbf{Z-Expression} & 
+%\textbf{H-Expression} & 
+% \textbf{Type} & 
+\textbf{Next Action} & \textbf{Rule}
+\\
+\hline
+1 &
+\zwsel{\hhole{}} & 
+% \hhole{} &
+% \tehole 
+% &
+\aConstruct{\flam{x}} & 
+\text{(\ref{r:conelamhole})}
+\\ 2 &
+\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} & 
+% \hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
+% \tarr{\tehole}{\tehole} &
+\aConstruct{\fnum{}} &
+\text{(\ref{r:contnum})}
+\\ 3 &
+\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
+% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
+% \tarr{\tnum}{\tehole} & 
+\aMove{\dNext{}} & 
+\text{({\ref{r:movenextsib}})}
+\\ 4 &
+\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
+&
+% \text{\textquotedbl}&
+% \tarr{\tnum}{\tehole} & 
+\aConstruct{\fnum{}} & 
+\text{(\ref{r:contnum})}
+\\ 5 &
+\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} & 
+% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
+% \tarr{\tnum}{\tnum} &
+\aMove{\dParent{}} & 
+\text{(\ref{r:moveparent})}
+\\ 6 &
+\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
+&
+% \text{\textquotedbl}&
+% \tarr{\tnum}{\tnum} & 
+\aMove{\dPrev{}} & 
+\text{(\ref{r:moveprevsib})}
+\\ 7 &
+% &
+\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} & 
+% \tarr{\tnum}{\tnum} &
+% &
+\aMove{\dChild{}} & 
+\text{(\ref{r:movefirstchild-lam})}
+\\ 8 &
+% &
+\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
+% \tarr{\tnum}{\tnum} & 
+\aConstruct{\fvar{x}} & 
+\text{(\ref{r:conevar})}
+\\ 9 &
+% \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
+% &
+\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
+% \tarr{\tnum}{\tnum} & 
+\quad\textrm{---}
+&
+\quad\textrm{---}
+\\
+\hline
+% \multicolumn{5}{c}{\text{... now assume a context where $id : \tarr{\tnum}{\tnum}$ ...}}\\
+% \hline
+% 10 &
+% % \hhole{} &
+% \zwsel{\hhole{}} & 
+% \tehole &
+% \aConstruct{\fasc} & 
+% \text{(\ref{r:constructasc})}
+% \\
+% 11 &
+% % \hhole{} : \hhole{} &
+% \hhole{} : \zwsel{ \hhole{}} &
+% \tehole & 
+% \aConstruct{\fnum{}} & 
+% \text{(\ref{r:contnum})}
+% \\
+% 12 &
+% % \hhole{} :\tnum{} &
+% \hhole{} : \zwsel{\tnum{}} &
+% \tnum &
+% \aMove{\dPrev{}} & 
+% \text{(\ref{r:moveprevsib})}
+% \\
+% % %13 &
+% % %\hhole{} :\tnum{} &
+% % %\zwsel{\hhole{}} : \tnum{}
+% % %&
+% % %\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
+% % %\\
+% 13 &
+% % \hhole{} :\tnum{}
+%  % &
+% \zwsel{\hhole{}} : \tnum{} &
+% \tnum & 
+% \aConstruct{ \fvar{id}} & 
+% \text{(\ref{r:conevar2})}
+% \\
+% 14 &
+% % \hhole{\textrm{$id$}} : \tnum{} &
+% \hhole{\zwsel{{\textrm{$id$}}}} : \tnum{} &
+% \tnum & 
+% \aConstruct{\fap{}} & 
+% \text{(\ref{r:coneapfn})}
+% \\
+% 15 &
+% % \hhole{\hap{{{\textrm{$id$}}}}{{\hhole{}}}} : \tnum{}
+% % &
+% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hhole{}}}} : \tnum{} &
+% \tnum & 
+% \aConstruct{\fnumlit{3}} &  
+% \text{(\ref{r:conenumnum})}
+% \\
+% 16 &
+% % \hhole{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hnum{3}}}} : \tnum{} &
+% \tnum & 
+% \aMove{\dParent{}} &  
+% \text{(\ref{r:moveparent-ap2})}
+% \\
+% 17 &
+% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \hhole{\zwsel{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}} : \tnum{} &
+% \tnum & 
+% \aMove{\dParent{}} &  
+% \text{(\ref{r:moveparent-hole})}
+% \\
+% 18 &
+% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \zwsel{\hhole{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
+% \tnum & 
+% \aFinish &  
+% \text{(\ref{r:finishana})}
+% \\
+% 19 &
+% % {\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
+% % &
+% \zwsel{{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
+% \tnum & 
+% \quad\textrm{---} & 
+% \quad\textrm{---}
+% \\
+% \hline
+
+%% 11 &
+%% {\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \aMove{\dParent{}} & ?
+%% \\ 12 &
+%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \zwsel{\hlam{x}{{{x}}}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \aMove{\dParent{}} & \refrule{15b} bad
+%% \\ 13 &
+%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
+%% &
+%% \zwsel{\hlam{x}{{{x}}} : \tarr{\tnum{}}{\tnum{}}}
+%% &
+%% \aConstruct{\fap{}} & \refrule{20h} bad
+%% \\ 14 &
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{}}
+%% &
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{}}}
+%% &
+%% \aConstruct{\fnumlit{3}} & \refrule{20l} bad
+%% \\ 15 &
+%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{3}}
+%% %&
+%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{3}}}
+%% %&
+%% %\aFinish{} & ?
+%% %\\
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{{\hnum{3}}}
+%% &
+%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{{\hnum{3}}}}
+%% &
+%% \quad\textrm{---{}---}
+%% &
+%% \quad\textrm{---{}---}
+%% \\
+%% \hline
+\end{array}
+\]
+\caption{Constructing an identity function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
+\label{fig:first-example}
+\end{figure}
 

### 2016-07-04 intro.tex
progress on sec 2

@@
 Some additional percentage of edit states were well-formed but ill-typed (the dataset did not contain enough information to determine the exact percentage.)
-Some of these meaningless edit states arose because the programmer was in the midst of a sequence of edits that left the edit state transiently meaningless, while others could be attributed to the programmer making a mistake.
+Some of these meaningless edit states arose because the programmer was in the midst of a sequence of edits that left the edit state transiently meaningless, while others arose because the programmer made a logical mistake.
 Because the language definitions is silent about these edit states, it is difficult to design useful editor services, e.g.
-syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \todo{citation}.
-Editors must either disable these services when they encounter these edit states or develop \emph{ad hoc} heuristics, which can confuse or mislead the programmer.
+syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \cite{mens2004survey}.
+Editors must either disable these editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics, which can be misleading.
 %, or because the programmer has made a mistake.
@@
 Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design.
-The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is well-formed.
-Indeed, it is not entirely clear what it should mean for term with holes to be statically meaningful.
-Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.
+The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed.
+As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics.
+%Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful. 
+% Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.
 
@@
 
-This paper seeks to address these foundational challenges from first principles.
-We introduce Hazelnut,  a \emph{typed structure editor}  based on a small bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}.
+This paper develops a principled solution to this problem.
+We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}, respectively.
 Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e.
-well-typed) expression with a single superimposed \emph{focus} (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Actions act relative to the expression or type in focus (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
+well-typed) H-expression with a single superimposed \emph{cursor}.
+We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because we follow Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
 a function argument), only actions consistent with that type are permitted.
@@
 For example, the programmer might need to construct the outer function application form before identifying the intended function.
-To avoid this problem, Hazelnut leaves newly constructed expressions inside a hole if the expression's type is inconsistent with the expected type.
+To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type.
 This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}.
-In other words, holes appear both at the leaves and at the internal nodes of the typed syntax tree that remain under construction.
+In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remains under construction.
 %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}.
@@
 
-\item This paper is primarily concerned with the static semantics of incomplete expressions.
+\item This paper is primarily concerned with develop a statics and an action model.
 However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.

### 2016-07-04 example.tex
more writing

@@
 % !TEX root = hazelnut-popl17.tex
+The reader is encouraged to follow along with the examples that we will discuss in this section using the implementation.
 
 %
+\subsection{Example 1: Constructing the Increment Function}
+
 Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from an empty hole via a sequence of {actions}.
@@
 The first action the user performs is $\aConstruct{\flam{x}}$, which instantiates the hole with a lambda abstraction binding the variable $x$.
-This results in the program on line
-2, consisting of a lambda abstraction ascribed an arrow type with holes in all positions.
+This results in the program on Line 2, consisting of a lambda abstraction ascribed an arrow type with holes in all positions.
 The cursor is placed on the argument type hole.
-Type holes, like empty expression holes, are also typeset $\tehole$ for visual consistency.
+Type holes are also typeset $\tehole$ for visual consistency.
 
-The
-user proceeds to fill these holes using construction and movement actions, resulting in the final expression on Line 9.
-With no holes remaining, this expression is \emph{complete}.% (though it could, of course, undergo further actions nevertheless.)
-
+\subsection{Example 2: Applying the Increment Function}
 So far, editing has proceeded in an essentially type-directed, outside-in fashion -- the user first specified the type of the function, then produced a body of that type by the action on Line 8.
@@
 \]
-\caption{Constructing an identity function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
+\caption{Constructing an increment function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
 \label{fig:first-example}

### 2016-07-04 intro.tex
more writing

@@
 The remainder of this paper is organized as follows:
-\begin{itemize}[noitemsep,nolistsep]
+\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
 \item We begin in Section
@@
 \item We then give a detailed overview of Hazelnut's semantics and metatheory, which have been mechanized using the Agda proof assistant, in Section \ref{sec:hazel}.
-\item Hazelnut is designed as a {foundational} calculus, i.e.
-as a calculus that language and editor designers extend with higher level constructs.
-We show how Hazelnut's rich mechanized metatheory guides and constrains the development of such an extension in Section \ref{sec:extending}.
+\item Hazelnut is a {foundational} calculus, i.e.
+a calculus that language and editor designers are expected to extend with higher level constructs.
+We show how Hazelnut's rich mechanized metatheory guides the development of an extension of Hazelnut with simple sum types in Section \ref{sec:extending}.
 \item In Section \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
@@
 
-\item This paper is primarily concerned with develop a statics and an action model.
-However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
-In so doing, we discover interesting connections with gradual typing and contextual modal type theory.
+% \item However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}. 
+% so doing, we discover interesting connections with gradual typing and contextual modal type theory.
 The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes.
-It also suggests a principled design for an ``edit and resume'' feature.\todo{revise this one}
+\todo{revise this one}
 
-\item In Section \ref{sec:rw}, we summarize the related work.
-\item We conclude in Section \ref{sec:future} by summarizing this vision of a principled science of structure editing rooted in type theory, and suggest a number of future directions.
+\item In Section \ref{sec:rw}, we summarize related work.
+In particular, much of the technical machinery needed to handle type holes coincides with that developed in the study of gradual type systems.
+Similarly, one interpretation of expression holes is as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work.
+%This suggests a principled design for an ``edit and resume'' feature.
+
+\item We conclude in Section \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
 \end{itemize} 
-\todo{summarize contents of the supplement here?}
+The supplemental material includes 1) a typeset listing of Hazelnut's rules in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our reference implementation of Hazelnut, both in source form and pre-compiled to JavaScript for use in a web browser.

### 2016-07-05 example.tex
finish sec 2 except for some rule references

@@
 Actions act relative to the cursor.
-The first action the user performs is $\aConstruct{\flam{x}}$, which instantiates the hole with a lambda abstraction binding the variable $x$.
-This results in the program on Line 2, consisting of a lambda abstraction ascribed an arrow type with holes in all positions.
+The first action that we perform is $\aConstruct{\flam{x}}$, which instantiates the hole with a lambda abstraction binding the variable $x$.
+This results in the Z-expression on Line 2, consisting of a lambda abstraction with an arrow type ascription.
 The cursor is placed on the argument type hole.
-Type holes are also typeset $\tehole$ for visual consistency.
+Type holes are typeset like empty expression holes for visual consistency.
 
-\subsection{Example 2: Applying the Increment Function}
-So far, editing has proceeded in an essentially type-directed, outside-in fashion -- the user first specified the type of the function, then produced a body of that type by the action on Line 8.
-Lines 10-12 similarly begin in a type-directed manner with the user giving an explicit type ascription, indicating that the expression that they are constructing will have type $\tnum$.
+The actions on Lines 2-4 complete the type ascription.
+In particular, the $\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and the $\aMove{\dNext}$ action moves the cursor to the next hole.
 
 \begin{figure}[t!]
-\[
-\begin{array}{|c||c|l||l|l|}
+$\arraycolsep=4px
+\begin{array}{|r||l|l||l|l|}
 \hline
@@
 % \tarr{\tnum}{\tnum} & 
-\aMove{\dPrev{}} & 
+\aMove{\dParent{}} & 
 \text{(\ref{r:moveprevsib})}
 \\ 7 & 
+\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} & 
+\aMove{\dChild} & 
+\text{(...)}
+\\ 8 &
 % &
@@
 \text{(\ref{r:movefirstchild-lam})}
-\\ 8 &
+\\ 9 &
 % &
@@
 \text{(\ref{r:conevar})}
-\\ 9 &
+\\ 10 &
 % \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
@@
 % \tarr{\tnum}{\tnum} & 
-\quad\textrm{---}
+{\aConstruct{\fplus}}
 &
-\quad\textrm{---}
-\\
-\hline
+\text{(...)}
+\\ 11 & 
+\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} & 
+\aConstruct{\fnumlit{1}} & 
+\text{(...)}
+\\ 12 & 
+\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} & 
+\textrm{---} & 
+{\textrm{---}}
+\\ \hline
 % \multicolumn{5}{c}{\text{... now assume a context where $id : \tarr{\tnum}{\tnum}$ ...}}\\
@@
 \end{array}
-\]
+$
 \caption{Constructing an increment function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
@@
 
-However, on Line 13, the user performs the $\aConstruct{\fvar{id}}$ action.
+\todo{rule references}The actions on Lines 5-8 move the cursor to the function body.
+For simplicity, Hazelnut defines only a small collection of primitive movement actions.
+In practice, an editor would also support compound movement actions, e.g.
+an action that moves the cursor directly to the next remaining hole, defined in terms of these primitive actions.
+
+Lines 9-11 complete the function body by first constructing the variable $x$, then  constructing the plus form, and finally constructing the number $\hnum{1}$.
+Notice that we did not need to construct the function body in an ``outside-in'' manner, i.e.
+we constructed $x$ before constructing the outer plus form.
+The transient function body, $x$, was coincidentally of type $\tnum$, so Hazelnut needed to make no special allowances in this case.
+
+\subsection{Example 2: Applying the Increment Function}
+
+\begin{figure}[t!]
+\colorbox{light-gray}{~~~~~~... now assume a context where $incr : \tarr{\tnum}{\tnum}$ ...~~~~~~}
+% \vspace{-6px}
+
+$\arraycolsep=4px
+\begin{array}{|r||l|l||l|l|}
+% \\
+\hline
+\# & \textbf{Z-Expression} & 
+%\textbf{H-Expression} & 
+% \textbf{Type} & 
+\textbf{Next Action} & \textbf{Rule}
+\\
+\hline
+13 &
+\zwsel{\hhole{}} & 
+% \hhole{} &
+% \tehole 
+% &
+\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}& 
+\text{(...)}
+\\ 14 & 
+\zwsel{incr} & 
+\aConstruct{\fap} & 
+\text{(...)}
+\\ 15 & 
+incr(\zwsel{\hhole{}}) & 
+\aConstruct{\fvar{incr}} & 
+\text{(...)}
+\\ 16 & 
+incr(\hhole{\zwsel{incr}}) & 
+\aConstruct{\fap} & 
+\text{(...)}
+\\ 17 & 
+incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} & 
+\aConstruct{\fnumlit{3}} & 
+\text{(...)}
+\\ 18 & 
+incr(\hhole{incr(\zwsel{\hnum{3}})}) & 
+\aMove{\dParent}& 
+\text{(...)}
+\\ 19 & 
+incr(\hhole{\zwsel{incr(\hnum{3})}}) & 
+\aMove{\dParent} & 
+\text{(...)}
+\\ 20 &
+incr(\zwsel{\hhole{incr(\hnum{3})}})& 
+\aFinish & 
+\text{(...)}
+\\ 21 & 
+incr(\zwsel{incr(\hnum{3})}) & 
+\textrm{---} & 
+{\textrm{---}}
+\\ \hline
+\end{array}
+$
+\caption{Appyling the increment function.}
+\label{fig:second-example}
+\end{figure}
+
+Figure \ref{fig:second-example} shows an edit sequence that constructs the expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the increment function from Figure \ref{fig:first-example} (we omit these steps for concision.)
+
+Line 13 constructs a variable identifying the function we wish to apply, i.e.
+$incr$.
+Line 14 then constructs the application form with $incr$ in function position, leaving the cursor on a hole in the argument position.
+
+We need to apply $incr$ again, so we perform the same action on Line 15 as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. Na\"ively, however, this would result in the following edit state: 
+\[
+incr(\zwsel{incr})
+\]
+The problem is that this is ill-typed, i.e.
+the argument of $incr$ must be of type $\tnum$ but is here of type $\tarr{\tnum}{\tnum}$.
+To maintain our edit invariant, this result would not be permitted.
+Consequently, the programmer would instead need to perform the $\aConstruct{\fap}$ action on Line 15.
+This would result in the following edit state (which is well-typed according to the semantics we will define in the next section):
+\[
+incr(\hhole{}(\zwsel{\hhole{}}))
+\]
+The programmer is not able to mention the intended function before constructing the outer application form.
+
+Hazelnut supports a more flexible strategy: rather than disallowing the $\aConstruct{\fvar{incr}}$ action on Line 15, it puts $incr$ inside a hole:
+\[
+incr(\hhole{\zwsel{incr}})
+\]
+As we will detail, this defers the necessary type compatibility check, just as the empty hole does. The expression inside a non-empty hole must itself be well-typed.
+
+Lines 16-17 proceed to apply the inner mention of $incr$ to a number literal, $\hnum{3}$.
+Finally, Lines 18-19 move the cursor to the non-empty hole and Line 20 performs the $\aFinish$ action.
+This action checks that the expression inside the hole is compatible with the type the hole is being checked against, as indeed it is, and removes the hole, resulting in the final edit state on Line 20.
+In practice, an editor might automatically perform the $\aFinish$ action as soon as it is possible to do so, but for simplicity, in Hazelnut it must be performed explicitly.
+
+% So far, editing has proceeded in an essentially type-directed, outside-in fashion -- the user first specified the type of the function, then produced a body of that type by the action on Line 8.
+Lines 10-12 similarly begin in a type-directed manner with the user giving an explicit type ascription, indicating that the expression that they are constructing will have type $\tnum$.
+
+% However, on Line 13, the user performs the $\aConstruct{\fvar{id}}$ action.
 Notice that $id$ has type $\tarr{\tnum}{\tnum}$, which is not consistent with the type $\tnum$ given in the ascription.

### 2016-07-05 example.tex
finished sec 2

@@
 \begin{figure}[t!]
+\begin{center}
 $\arraycolsep=4px
@@
 \aMove{\dNext{}} & 
-\text{({\ref{r:movenextsib}})}
+\text{({\ref{rule:move-nextSib-arr}})}
 \\ 4 &
@@
 \aMove{\dParent{}} & 
-\text{(\ref{r:moveparent})}
+\text{(\ref{rule:move-parent-arr-right})}
 \\ 6 &
@@
 \aMove{\dParent{}} & 
-\text{(\ref{r:moveprevsib})}
+\text{(\ref{rule:move-parent-asc-right})}
 \\ 7 & 
@@
 \aMove{\dChild} & 
-\text{(...)}
+\text{(\ref{r:movefirstchild})}
 \\ 8 &
@@
 &
-\text{(...)}
+\text{(\ref{rule:construct-plus-compat})}
 \\ 11 & 
@@
 \aConstruct{\fnumlit{1}} & 
-\text{(...)}
+\text{(\ref{r:conenumnum})}
 \\ 12 & 
@@
 \\ \hline
-% \multicolumn{5}{c}{\text{... now assume a context where $id : \tarr{\tnum}{\tnum}$ ...}}\\
-% \hline
-% 10 &
-% % \hhole{} &
-% \zwsel{\hhole{}} & 
-% \tehole &
-% \aConstruct{\fasc} & 
-% \text{(\ref{r:constructasc})}
-% \\
-% 11 &
-% % \hhole{} : \hhole{} &
-% \hhole{} : \zwsel{ \hhole{}} &
-% \tehole & 
-% \aConstruct{\fnum{}} & 
-% \text{(\ref{r:contnum})}
-% \\
-% 12 &
-% % \hhole{} :\tnum{} &
-% \hhole{} : \zwsel{\tnum{}} &
-% \tnum &
-% \aMove{\dPrev{}} & 
-% \text{(\ref{r:moveprevsib})}
-% \\
-% % %13 &
-% % %\hhole{} :\tnum{} &
-% % %\zwsel{\hhole{}} : \tnum{}
-% % %&
-% % %\aMove{\dPrev{}} & \refrule{\ref{r:moveprevsib}}
-% % %\\
-% 13 &
-% % \hhole{} :\tnum{}
-%  % &
-% \zwsel{\hhole{}} : \tnum{} &
-% \tnum & 
-% \aConstruct{ \fvar{id}} & 
-% \text{(\ref{r:conevar2})}
-% \\
-% 14 &
-% % \hhole{\textrm{$id$}} : \tnum{} &
-% \hhole{\zwsel{{\textrm{$id$}}}} : \tnum{} &
-% \tnum & 
-% \aConstruct{\fap{}} & 
-% \text{(\ref{r:coneapfn})}
-% \\
-% 15 &
-% % \hhole{\hap{{{\textrm{$id$}}}}{{\hhole{}}}} : \tnum{}
-% % &
-% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hhole{}}}} : \tnum{} &
-% \tnum & 
-% \aConstruct{\fnumlit{3}} &  
-% \text{(\ref{r:conenumnum})}
-% \\
-% 16 &
-% % \hhole{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \hhole{\hap{{{\textrm{$id$}}}}{\zwsel{\hnum{3}}}} : \tnum{} &
-% \tnum & 
-% \aMove{\dParent{}} &  
-% \text{(\ref{r:moveparent-ap2})}
-% \\
-% 17 &
-% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \hhole{\zwsel{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}} : \tnum{} &
-% \tnum & 
-% \aMove{\dParent{}} &  
-% \text{(\ref{r:moveparent-hole})}
-% \\
-% 18 &
-% % %\hhole{\hap{{{\textrm{id}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \zwsel{\hhole{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
-% \tnum & 
-% \aFinish &  
-% \text{(\ref{r:finishana})}
-% \\
-% 19 &
-% % {\hap{{{\textrm{$id$}}}}{{\hnum{3}}}} : \tnum{}
-% % &
-% \zwsel{{{\hap{{{\textrm{$id$}}}}{{\hnum{3}}}}}} : \tnum{} &
-% \tnum & 
-% \quad\textrm{---} & 
-% \quad\textrm{---}
-% \\
-% \hline
-
-%% 11 &
-%% {\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \aMove{\dParent{}} & ?
-%% \\ 12 &
-%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \zwsel{\hlam{x}{{{x}}}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \aMove{\dParent{}} & \refrule{15b} bad
-%% \\ 13 &
-%% %{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}
-%% &
-%% \zwsel{\hlam{x}{{{x}}} : \tarr{\tnum{}}{\tnum{}}}
-%% &
-%% \aConstruct{\fap{}} & \refrule{20h} bad
-%% \\ 14 &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{}}
-%% &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{}}}
-%% &
-%% \aConstruct{\fnumlit{3}} & \refrule{20l} bad
-%% \\ 15 &
-%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\hhole{3}}
-%% %&
-%% %\hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{\hhole{3}}}
-%% %&
-%% %\aFinish{} & ?
-%% %\\
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{{\hnum{3}}}
-%% &
-%% \hapP{{\textrm{id}} : \tarr{\tnum{}}{\tnum{}}}{\zwsel{{\hnum{3}}}}
-%% &
-%% \quad\textrm{---{}---}
-%% &
-%% \quad\textrm{---{}---}
-%% \\
-%% \hline
 \end{array}
-$
+$\end{center}\vspace{-6px}
 \caption{Constructing an increment function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
@@
 
-\todo{rule references}The actions on Lines 5-8 move the cursor to the function body.
+The actions on Lines 5-8 move the cursor to the function body.
 For simplicity, Hazelnut defines only a small collection of primitive movement actions.
 In practice, an editor would also support compound movement actions, e.g.
-an action that moves the cursor directly to the next remaining hole, defined in terms of these primitive actions.
+an action that moves the cursor directly to the next remaining hole, defined in terms of these primitive movement actions.
 
@@
 \begin{figure}[t!]
-\colorbox{light-gray}{~~~~~~... now assume a context where $incr : \tarr{\tnum}{\tnum}$ ...~~~~~~}
+\begin{center}
+\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{53px}}
 % \vspace{-6px}
-
 $\arraycolsep=4px
@@
 \aConstruct{\fvar{incr}} \hphantom{~\,\,~~}& 
-\text{(...)}
+\text{(\ref{r:conevar})}
 \\ 14 & 
@@
 \aConstruct{\fap} & 
-\text{(...)}
+\text{(\ref{r:coneapfn})}
 \\ 15 & 
@@
 \aConstruct{\fvar{incr}} & 
-\text{(...)}
+\text{(\ref{r:conevar2})}
 \\ 16 & 
@@
 \aConstruct{\fap} & 
-\text{(...)}
+\text{(\ref{r:coneapfn})}
 \\ 17 & 
@@
 \aConstruct{\fnumlit{3}} & 
-\text{(...)}
+\text{(\ref{r:conenumnum})}
 \\ 18 & 
@@
 \aMove{\dParent}& 
-\text{(...)}
+\text{(\ref{r:moveparent-ap2})}
 \\ 19 & 
@@
 \aMove{\dParent} & 
-\text{(...)}
+\text{(\ref{r:moveparent-hole})}
 \\ 20 &
@@
 \aFinish & 
-\text{(...)}
+\text{(\ref{r:finishana})}
 \\ 21 & 
@@
 \end{array}
-$
+$\end{center}\vspace{-6px}
 \caption{Appyling the increment function.}
@@
 \]
-The problem is that this is ill-typed, i.e.
-the argument of $incr$ must be of type $\tnum$ but is here of type $\tarr{\tnum}{\tnum}$.
-To maintain our edit invariant, this result would not be permitted.
-Consequently, the programmer would instead need to perform the $\aConstruct{\fap}$ action on Line 15.
-This would result in the following edit state (which is well-typed according to the semantics we will define in the next section):
+This edit state is ill-typed, i.e. the argument of $incr$ must be of type $\tnum$ but is here of type $\tarr{\tnum}{\tnum}$. To maintain our edit invariant, this edit state cannot be allowed to arise.
+
+The programmer could alternatively have performed the $\aConstruct{\fap}$ action on Line 15.
+This would result in the following edit state, which is well-typed according to the static semantics we will define in the next section:
 \[
@@
 \]
-The programmer is not able to mention the intended function before constructing the outer application form.
+The problem is that the programmer is not able to identify the intended function before constructing the outer application form. This is qualitatively unnatural.
 
@@
 \]
-As we will detail, this defers the necessary type compatibility check, just as the empty hole does. The expression inside a non-empty hole must itself be well-typed.
+As we will detail, this defers the necessary type compatibility check, exactly as the empty hole does.
+One way to think about non-empty holes is as a formalization of the heuristic concept of the ``red underline'' that a text or syntactic structure editor displays where a type incompatibility is detected.
+The difference is that discovering a type incompatibility does not leave the entire program formally meaningless.
 
+The expression inside a non-empty hole must itself be well-typed, so we can continue to edit it.
 Lines 16-17 proceed to apply the inner mention of $incr$ to a number literal, $\hnum{3}$.
 Finally, Lines 18-19 move the cursor to the non-empty hole and Line 20 performs the $\aFinish$ action.
-This action checks that the expression inside the hole is compatible with the type the hole is being checked against, as indeed it is, and removes the hole, resulting in the final edit state on Line 20.
-In practice, an editor might automatically perform the $\aFinish$ action as soon as it is possible to do so, but for simplicity, in Hazelnut it must be performed explicitly.
+This action checks that the expression inside the hole is compatible with the type the hole is being checked against, as indeed it now is, and removes the hole, resulting in the final edit state on Line 20.
+In practice, an editor might automatically perform the $\aFinish$ action as soon as it is possible to do so, but for simplicity, Hazelnut formally requires that it be performed explicitly.
 

### 2016-07-05 example.tex
tweaks to sec 2 and 3

@@
 
-Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from an empty hole via a sequence of {actions}.
-We will discuss the syntactic forms and the referenced rules in Sec.
+Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from the empty hole via the indicated sequence of {actions}.
+We will introduce Hazelnut's formal syntax and define the referenced rules in Sec.
 \ref{sec:hazel}.
-For now, let us develop only the high-level intuitions.
+First, let us build some high-level intuitions.
 
@@
 The edit state in Hazelnut is a {Z-expression}, $\zexp$.
-Every Z-expression has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the {cursor}.
-The selected H-expression or H-type is typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively.
-For example, on Line 1 of Figure \ref{fig:first-example}, the empty expression hole, $\hhole{}$, is under the cursor.
+Every Z-expression has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the {cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively.
+For example, on Line 1, the empty expression hole, $\hhole{}$, is under the cursor.
 
@@
 $\end{center}\vspace{-6px}
-\caption{Constructing an increment function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
+\caption{Constructing an increment function in Hazelnut.}
 \label{fig:first-example}
@@
 For simplicity, Hazelnut defines only a small collection of primitive movement actions.
-In practice, an editor would also support compound movement actions, e.g.
-an action that moves the cursor directly to the next remaining hole, defined in terms of these primitive movement actions.
+In practice, an editor would also define compound movement actions, e.g.
+an action that moves the cursor directly to the next remaining hole, in terms of these primitive movement actions.
 
@@
 we constructed $x$ before constructing the outer plus form.
-The transient function body, $x$, was coincidentally of type $\tnum$, so Hazelnut needed to make no special allowances in this case.
+Luckily, the transient function bodies, $x$ and $x + \hhole{}$, could be checked against type $\tnum$ (as we will detail in Sec.
+\ref{sec:holes}.)
 
@@
 \begin{center}
-\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{53px}}
+\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
 % \vspace{-6px}
@@
 
-Figure \ref{fig:second-example} shows an edit sequence that constructs the expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the increment function from Figure \ref{fig:first-example} (we omit these steps for concision.)
+Figure \ref{fig:second-example} shows an edit sequence that constructs the expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the increment function from Figure \ref{fig:first-example}.
 
-Line 13 constructs a variable identifying the function we wish to apply, i.e.
-$incr$.
+We begin on Line 13 by constructing the variable $incr$.
 Line 14 then constructs the application form with $incr$ in function position, leaving the cursor on a hole in the argument position.
+Notice that we did not need to construct the outer application form before identifying the function being applied.
+Again, the transient edit state happens to be well-typed, so we needed to make no special allowances for this order of actions.
 
-We need to apply $incr$ again, so we perform the same action on Line 15 as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. Na\"ively, however, this would result in the following edit state: 
+We now need to apply $incr$ again, so we perform the same action on Line 15 as we did on Line 13, i.e.
+$\aConstruct{\fvar{incr}}$.
+In a syntactic structure editor, performing such an action would result in the following edit state:
 \[
@@
 \]
-This edit state is ill-typed, i.e. the argument of $incr$ must be of type $\tnum$ but is here of type $\tarr{\tnum}{\tnum}$. To maintain our edit invariant, this edit state cannot be allowed to arise.
+This edit state is ill-typed (after \emph{cursor erasure}): the argument of $incr$ must be of type $\tnum$ but here it is of type $\tarr{\tnum}{\tnum}$.
+Hazelnut cannot allow such an edit state to arise.
 
 The programmer could alternatively have performed the $\aConstruct{\fap}$ action on Line 15.
-This would result in the following edit state, which is well-typed according to the static semantics we will define in the next section:
+This would result in the following edit state, which is well-typed according to the static semantics that we will define in the next section:
 \[
@@
 \]
-The problem is that the programmer is not able to identify the intended function before constructing the outer application form. This is qualitatively unnatural.
+The problem is that the programmer is not able to identify the intended function before constructing the outer function application form.
+This stands in contrast to Lines 13-14.
+% This is qualitatively unnatural.
 
-Hazelnut supports a more flexible strategy: rather than disallowing the $\aConstruct{\fvar{incr}}$ action on Line 15, it puts $incr$ inside a hole:
+Hazelnut's action semantics addresses this problem: rather than disallowing the $\aConstruct{\fvar{incr}}$ action on Line 15, it leaves $incr$ inside a hole:
 \[
@@
 \]
-As we will detail, this defers the necessary type compatibility check, exactly as the empty hole does.
-One way to think about non-empty holes is as a formalization of the heuristic concept of the ``red underline'' that a text or syntactic structure editor displays where a type incompatibility is detected.
-The difference is that discovering a type incompatibility does not leave the entire program formally meaningless.
+This defers the type consistency check, exactly as an empty hole in the same position does.
+One way to think about non-empty holes is as an internalization of the ``red underline'' that text or syntactic structure editors display to indicate a type inconsistency.
+By internalizing this concept, the presence of a type inconsistency does not leave the entire program formally meaningless.
 
-The expression inside a non-empty hole must itself be well-typed, so we can continue to edit it.
+The expression inside a non-empty hole must itself be well-typed, so the programmer can continue to edit it.
 Lines 16-17 proceed to apply the inner mention of $incr$ to a number literal, $\hnum{3}$.
 Finally, Lines 18-19 move the cursor to the non-empty hole and Line 20 performs the $\aFinish$ action.
-This action checks that the expression inside the hole is compatible with the type the hole is being checked against, as indeed it now is, and removes the hole, resulting in the final edit state on Line 20.
-In practice, an editor might automatically perform the $\aFinish$ action as soon as it is possible to do so, but for simplicity, Hazelnut formally requires that it be performed explicitly.
+The $\aFinish$ action removes the hole if the type of the expression inside the hole is consistent with the expected type, as it now is.
+This results in the final edit state on Line 20, as desired.
+In practice, the editor might automatically perform the $\aFinish$ action as soon as it becomes possible, but for simplicity, Hazelnut formally requires that it be performed explicitly.
 

### 2016-07-05 hazelnut-popl17.tex
progress on sec 2
more writing
finish sec 2 except for some rule references
finished sec 2
tweaks to sec 2 and 3

@@
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\definecolor{light-gray}{gray}{0.9}
 \usepackage{mathpazo}
@@
 \AtBeginDocument{%
- \abovedisplayskip=3pt
+ \abovedisplayskip=2pt
  \abovedisplayshortskip=0pt
@@
 \usepackage[compact]{titlesec}
-\titlespacing*{\section}{0pt}{3pt}{3pt}
-\titlespacing*{\subsection}{0pt}{4pt}{3pt}
-\titlespacing*{\subsubsection}{0pt}{4pt}{3pt}
-\titlespacing*{\paragraph}{0pt}{3pt}{5pt}
+\titlespacing*{\section}{0pt}{4pt}{2pt}
+\titlespacing*{\subsection}{0pt}{4pt}{2pt}
+\titlespacing*{\subsubsection}{0pt}{4pt}{2pt}
+\titlespacing*{\paragraph}{0pt}{4pt}{2pt}
 \setlength{\skip\footins}{3pt}
@@
 \usepackage{todonotes}
+\usepackage{xcolor}
+%\usepackage{adjustbox}
 
@@
 %\authorinfo{\vspace{-2px}}{}{}
-\authorinfo{~}{~}{~}
+\authorinfo{~}{~}{\vspace{-10px}}
 %\authorinfo{Cyrus Omar \and Jonathan Aldrich}
@@
 \label{sec:hazel}
-\todo{revise text}\todo{add let and ifz}\todo{make eqn numbering work better}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$.
-Its major constituents, introduced by example in the previous section, are:
-\begin{itemize}
-\item \textbf{H-types} and \textbf{H-expressions} (Sec.
-\ref{sec:holes}), which are terms with \emph{holes}.
-Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
-\item \textbf{Z-types} and \textbf{Z-expressions} (Sec.
-\ref{sec:cursors}), which superimpose a single \emph{focus}\todo{cursor?} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
-\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.
+\input{formally}
 
-Whenever an action is performed on a well-typed expression, it produces another well-typed expression in a \emph{sensible} manner.
-More specifically, the action semantics satisfies a crucial \emph{sensibility theorem}, stated in Sec.
-\ref{sec:actions}.
-\end{itemize}
-
-In our overview of the semantics below, we will reproduce only the most interesting rules, and in some cases we will do so ``out of order.'' The appendix (and our Agda formalization, see Sec.
-\ref{sec:mech}) defines the complete collection of rules in their dependency order.
-\subsection{Holes}\label{sec:holes}
-\begin{figure}[t]
-$\arraycolsep=4pt\begin{array}{lllllll}
-\mathsf{HTyp} & \tau,\htau & ::= &
-  \tarr{\htau}{\htau} ~\vert~
-  \tnum ~\vert~
-  \tehole\\
-\mathsf{HExp} & e,\hexp & ::= &
-  \hexp : \htau ~\vert~
-  x ~\vert~
-  \hlam{x}{\hexp} ~\vert~
-  \hap{\hexp}{\hexp} ~\vert~
-  \hnum{n} ~\vert~
-  \hadd{\hexp}{\hexp} ~\vert~
-  \hehole ~\vert~
-  \hhole{\hexp}
-\end{array}$
-%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
-\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
-\label{fig:hexp-syntax}
-\end{figure}
-
-The syntax of H-types and H-expressions is given in Figure \ref{fig:hexp-syntax}.
-Most of the forms correspond directly to those of the simply-typed lambda calculus extended with type $\tnum$.
-The number expression corresponding to the number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$.
-In addition to these standard forms, \emph{empty holes} are drawn $\hehole$ and \emph{non-empty H-expression holes} are drawn $\hhole{\hexp}$.
-In our simple calculus, all well-formed type expressions are valid types, so we do not need non-empty H-type holes.%Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.
-
-We refer to terms that do not contain subterms of hole form as \emph{complete}.
-Informally, we will use metavariables $\tau$ and $e$ rather than $\htau$ and $\hexp$ for complete H-types and H-expressions, respectively.
-Formally, we can derive $\hcomplete{\tau}$ when $\tau$ is a complete H-type, and $\hcomplete{e}$ when $e$ is a complete H-expression.
-We omit the straightforward definitions of these judgements for concision.
-The dynamics of Hazelnut, which we need not detail here, is defined only  over complete H-expressions (i.e.
-we can only ``run'' a complete program, though see Sec.
-\ref{sec:future}.)\todo{get rid of e/edot distinction}\todo{change this to refer to dynamics section}
-
-The statics of Hazelnut is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100}, i.e. around the following mutually defined typing judgements:
-\[\arraycolsep=15pt\begin{array}{ll}
-%\textbf{Judgement Form} & \textbf{Description}\\
-\hana{\hGamma}{\hexp}{\htau} & \text{$\hexp$ analyzes against $\htau$}\\
-\hsyn{\hGamma}{\hexp}{\htau} & \text{$\hexp$ synthesizes $\htau$}
-\end{array}\]
-where typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$.
-Derivations of the type analysis judgement establish that $\hexp$ can appear where an expression of type $\htau$ is expected.
-Derivations of the type synthesis judgement determine a type that can be assigned to $\hexp$ even in positions where an expected type is not known (e.g.
-at the top level.) Algorithmically, the type is an ``input'' of the type analysis judgement, but an ``output'' of the type synthesis judgement.
-%The rules describe a \emph{local type inference} scheme, i.e.
-type ascriptions are unnecessary when an expression is being analyzed against a known type.
-Making a judgemental distinction between these two notions will be essential for giving a sensible action semantics to our system (Sec. \ref{sec:actions}.)
-
- %We use the metavariable $\Gamma$ for \emph{complete typing contexts}, i.e. typing contexts where each hypothesis mentions only complete types.
-
-\begin{subequations}\label{rules:syn-ana}
-Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any \emph{compatible} type.
-This is expressed by the \emph{subsumption rule}:
-\begin{equation}\label{rule:ana-subsume}
-\inferrule{
-  \hsyn{\hGamma}{\hexp}{\htau'}\\
-  \tcompat{\htau}{\htau'}
-}{
-  \hana{\hGamma}{\hexp}{\htau}
-}
-\end{equation}
-The \emph{H-type compatibility judgement}, $\tcompat{\htau}{\htau'}$, reduces to syntactic equality for complete H-types.
-For incomplete H-types, the rules are given after we discuss the semantics of holes below.
-
-First, let us briefly review the standard constructs.
-Type ascription allows the user to state  a type for the ascribed expression to be analyzed against:
-\begin{equation}\label{rule:syn-asc}
-\inferrule{
-  \hana{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hexp : \htau}{\htau}
-}
-\end{equation}
-
-A variable synthesizes the type that the context assigns to it:
-\begin{equation}\label{rule:syn-var}
-\inferrule{ }{
-  \hsyn{\hGamma, x : \htau}{x}{\htau}
-}
-\end{equation}
-
-Functions are not themselves annotated with types, so they can only appear in analytic position:
-\begin{equation}\label{rule:syn-lam}
-\inferrule{
-  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
-}{
-  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
-}
-\end{equation}
-It would be straightforward to also add a ``half-annotated'' lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out of our calculus \cite{DBLP:conf/tldi/ChlipalaPH05}.
-
-For function application, if the expression in function position synthesizes an arrow type, the argument is analyzed against the synthesized argument type:
-\begin{equation}\label{rule:syn-ap}
-\inferrule{
-  \hsyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}\\
-  \hana{\hGamma}{\hexp_2}{\htau_2}
-}{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
-}
-\end{equation}
-
-Numbers synthesize type $\tnum$:
-\begin{equation}\label{rule:syn-num}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hnum{n}}{\tnum}
-}
-\end{equation}
-
-Addition operates like a function over numbers:
-\begin{equation}\label{rule:syn-plus}
-\inferrule{
-  \hana{\hGamma}{\hexp_1}{\tnum}\\
-  \hana{\hGamma}{\hexp_2}{\tnum}
-}{
-  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
-}
-\end{equation}
-
-The rules given so far are sufficient to type complete H-expressions. The remaining rules give H-expressions with holes a well-defined static semantics.
-
-The empty hole synthesizes the hole type:
-\begin{equation}\label{rule:syn-ehole}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hehole}{\tehole}
-}
-\end{equation}
-
-A non-empty hole contains an H-expression that is ``under construction''. The inner expression must synthesize some type, but the non-empty hole synthesizes only the hole type:
-\begin{equation}\label{rule:syn-hole}
-\inferrule{
-  \hsyn{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
-}
-\end{equation}
-The type compatibility judgement $\tcompat{\htau}{\htau'}$, which appeared as a premise in the subsumption rule, makes the hole type compatible with any other type:
-\begin{subequations}\label{rules:tcompat}
-\begin{equation}\label{rule:tcompat-hole}
-\inferrule{ }{
-  \tcompat{\htau}{\tehole}
-}
-\end{equation}
-The remaining rules, given in the appendix, establish that type compatibility is symmetric and reflexive (but not transitive.)
-% \begin{equation}\label{rule:tcompat-comm}
-% \inferrule{
-%   \tcompat{\htau}{\htau'}
-% }{
-%   \tcompat{\htau'}{\htau}
-% }
-% \end{equation}
-% \begin{equation}\label{rule:tcompat-num}
-% \inferrule{ }{
-%   \tcompat{\tnum}{\tnum}
-% }
-% \end{equation}
-% \begin{equation}\label{rule:tcompat-arr}
-% \inferrule{
-%   \tcompat{\htau_1}{\htau_1'}\\
-%   \tcompat{\htau_2}{\htau_2'}
-% }{
-%   \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-% }
-% \end{equation}
-\end{subequations}
-Consequently, by subsumption, we can derive that $\hana{id : \tarr{\tnum}{\tnum}}{\hhole{id}}{\tnum}$, as is necessary to synthesize a type for the H-expression on Line 14 of Fig.
-\ref{fig:first-example}.
-%In other words, this mechanism is essential if  users are to able to construct a program in anything but an ``outside in'' fashion.
-
-The final rule handles function applications where the expression in function position synthesizes a hole type, rather than an arrow type.
-We treat it as if it had instead synthesized $\tarr{\tehole}{\tehole}$:
-\begin{equation}\label{rule:syn-ap-2}
-\inferrule{
-  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
-  \hana{\hGamma}{\hexp_2}{\tehole}
-}{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}
-}
-\end{equation}
-
-\todo{redundant text w/ citation}
-The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}.
-Their system (which was not bidirectionally typed nor an editor model) also needed to define two rules for function application.
-In general, when a premise requires that a synthesized type be of a particular form, we need a special case where the synthesized hole type is treated instead as if it were the ``holey-est'' type of that form.\footnote{Alternatively, we might add a rule that allows expressions that synthesize hole type to then non-deterministically synthesize any other type, but maintaining determinism is useful in practice, so we avoid this approach.}
-
-\end{subequations}
-\subsection{Focus Model}\label{sec:cursors}
-\begin{figure}[t]
-\hspace{-3px}$\arraycolsep=3pt\begin{array}{lllllll}
-\mathsf{ZTyp} & \ztau & ::= &
-  %\zlsel{\htau} ~\vert~
-  \zwsel{\htau} ~\vert~
-  %\zrsel{\htau} ~\vert~
-  \tarr{\ztau}{\htau} ~\vert~
-  \tarr{\htau}{\ztau} \\
-\mathsf{ZExp} & \zexp & ::= &
-  %\zlsel{\hexp} ~\vert~
-  \zwsel{\hexp} ~\vert~
-  %\zrsel{\hexp} ~\vert~
-  \zexp : \htau ~\vert~
-  \hexp : \ztau ~\vert~
-  \hlam{x}{\zexp} ~\vert~
-  \hap{\zexp}{\hexp} ~\vert~
-  \hap{\hexp}{\zexp} ~\vert~
-  \hadd{\zexp}{\hexp} ~\vert~
-  \hadd{\hexp}{\zexp} ~\vert~
-  \hhole{\zexp}
-\end{array}$
-%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
-\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single focus.}
-\label{fig:zexp-syntax}
-\end{figure}
-
-In order to identify a single subtree of an H-type or H-expression as the current focus of action, we apply Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.
-The syntax of Z-types, $\ztau$, and Z-expressions, $\zexp$, is given in Figure \ref{fig:zexp-syntax}.
-The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that is the current focus.
-All other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the focus will be found.
-All other sub-terms are H-types or H-expressions.
-Taken together, every syntactically well-formed Z-type and Z-expression contains exactly one focused H-type or H-expression.
-
-We write $\removeSel{\ztau}$ for the H-type constructed by removing the focus marker from the Z-type $\ztau$. This straightforward metafunction is defined as follows:
-\begin{align*}
-%\removeSel{(\zlsel{\htau})} & = \htau\\
-\removeSel{(\zwsel{\htau})} & = \htau\\
-%\removeSel{(\zrsel{\htau})} & = \htau\\
-\removeSel{(\tarr{\ztau}{\htau})} & = \tarr{\removeSel{\ztau}}{\htau}\\
-\removeSel{(\tarr{\htau}{\ztau})} & = \tarr{\htau}{\removeSel{\ztau}}
-\end{align*}
-
-Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by removing the focus marker from the Z-expression $\zexp$.
-The definition of this metafunction is analagous, so we leave it in the appendix for concision.
-% \begin{align*}
-% %\removeSel{(\zlsel{\hexp})} & = \hexp\\
-% \removeSel{(\zwsel{\hexp})} & = \hexp\\
-% %\removeSel{(\zrsel{\hexp})} & = \hexp\\
-% \removeSel{(\zexp : \htau)} & = \removeSel{\zexp} : \htau\\
-% \removeSel{(\hexp : \ztau)} & = \hexp : \removeSel{\ztau}\\
-% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\
-% \removeSel{(\hap{\zexp}{\hexp})} & = \hap{\removeSel{\zexp}}{\hexp}\\
-% \removeSel{(\hap{\hexp}{\zexp})} & = \hap{\hexp}{\removeSel{\zexp}}\\
-% \removeSel{(\hadd{\zexp}{\hexp})} & = \hadd{\removeSel{\zexp}}{\hexp}\\
-% \removeSel{(\hadd{\hexp}{\zexp})} & = \hadd{\hexp}{\removeSel{\zexp}}\\
-% \removeSel{\hhole{\zexp}} &= \hhole{\removeSel{\zexp}}
-% \end{align*}
-
-\subsection{Action Semantics}\label{sec:actions}
-\begin{figure}[t]
-\hspace{-3px}$\arraycolsep=3pt\begin{array}{llcllll}
-\mathsf{Action} & \alpha & ::= &
-  \aMove{\delta} ~\vert~
-  %\aSelect{\delta} ~\vert~
-  \aDel ~\vert~
-  %\aReplace{\htau} ~\vert~
-  %\aReplace{\hexp} ~\vert~
-  \aConstruct{\varphi} ~\vert~
-  \aFinish\\
-\mathsf{Direction} & \delta & ::= &
-  \dChild ~\vert~
-  \dParent ~\vert~
-  \dNext ~\vert~
-  \dPrev\\
-\mathsf{Shape} & \varphi & ::= &
-  \farr ~\vert~
-  \fnum \\
-& & \vert &
-  \fasc ~\vert~
-  \fvar{x} ~\vert~
-  \flam{x} ~\vert~
-  \fap ~\vert~
-  \farg ~\vert~
-  \fnumlit{n} ~\vert~
-  \fplus
-\end{array}$
-%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
-\caption{Syntax of actions.}
-\label{fig:action-syntax}
-\vspace{-8px}
-\end{figure}
-
-The syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, or \emph{shapes}, $\varphi$, is given in Figure \ref{fig:action-syntax}.
-Actions are performed on Z-types and Z-expressions according to the \emph{action semantics} of Hazelnut, which is organized around three judgements:\todo{fix spacing}
-\[\arraycolsep=10pt\begin{array}{ll}
-%\textbf{Judgement Form} & \textbf{Description}\\
-\performTyp{\ztau}{\alpha}{\ztau'} & \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
-\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'} & \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
-& \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
-\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'} & \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
-& \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
-& \text{against $\htau$}
-\end{array}\]
-
-As suggested by the descriptions above, the action semantics maintains the following \emph{action sensibility} theorem:
-\begin{theorem}[Action Sensibility] Both of the following hold:
-\label{thrm:actsafe}
-\begin{enumerate}
-\item If $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
-  $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ then
-  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
-\item If $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
-  $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ then
-  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
-\end{enumerate}
-\end{theorem}
-In words, every action leaves the program in a semantically well-defined state.
-More specifically, the first clause of Theorem \ref{thrm:actsafe} establishes that actions performed on expressions that synthesize a type can only produce expressions that also synthesize some (possibly different) type.
-The second clause establishes that actions performed on expressions in analytic position (e.g.
-those under type ascriptions or in argument position, see above) can only produce expressions that can also be analyzed against the expected type.% Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.
-
-It is also useful to maintain a \emph{deterministic} action semantics, i.e. every well-defined action should produce a unique Z-type or Z-expression. Formally, this is stated as follows:
-\begin{theorem}[Action Determinism] All of the following hold:
-\label{thrm:actdet}
-\begin{enumerate}
-\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
-\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
-  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
-  $\zexp' = \zexp''$ and $\htau' = \htau''$.
-% \item If all of
-
-%   \begin{quote}
-%     \begin{enumerate}
-%     \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
-%     \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
-%     \item $\tcompat{\htau}{\htau'}$, and
-%     \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
-%       $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
-%     \end{enumerate}
-%   \end{quote}
-%   hold, then $\zexp' = \zexp''$.
-\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
-  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
-  \zexp''$.
-\end{enumerate}
-\end{theorem}
-
-In order to maintain determinism, we will need to supplement the definition of type compatibility above with a definition for \emph{type incompatibility}, $\tincompat{\htau}{\htau'}$.
-The key rule establishes that arrow types are incompatible with the $\tnum$ type:
-\begin{subequations}
-  % \begin{equation}
-  %   \inferrule{
-  %     \tincompat{\htau}{\htau'}
-  %   }{
-  %     \tincompat{\htau'}{\htau}
-  %   }
-  % \end{equation}
-  \begin{equation}
-    \inferrule{ }{
-      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
-    }
-  \end{equation}
-  % \begin{equation}
-  %   \inferrule{
-  %     \tincompat{\htau_1}{\htau_1'}
-  %   }{
-  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-  %   }
-  % \end{equation}
-  % \begin{equation}
-  %   \inferrule{
-  %     \tincompat{\htau_2}{\htau_2'}
-  %   }{
-  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-  %   }
-  % \end{equation}
-\end{subequations}
-The remaining rules, given in the appendix, establish that type incompatibility is symmetric and covariant.
-\subsubsection{Subsumption}
-
-The action semantics includes a subsumption rule much like the one from the underlying semantics of H-expressions:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
-    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
-    \tcompat{\htau}{\htau''}
-  }{
-    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
-  }
-\end{equation}
-In other words, if the expression synthesizes a type, then we defer to the synthetic action performance judgement, as long as it produces an expression that synthesizes a type compatible with the type provided for analysis.
-It is easy to see that this satisfies Theorem 1 by applying the IH and subsumption.
-
-\subsubsection{Relative Movement} Movement actions change the focus but do not change the underlying H-type or H-expression (so action sensibility is easy to show for these rules as well.)
-
-The rules for relative movement within Z-types are given below and should be self-explanatory:
-\begin{subequations}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\tarr{\htau_1}{\htau_2}}
-    }{
-      \aMove{\dChild}
-    }{
-      \tarr{\zwsel{\htau_1}}{\htau_2}
-    }
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \tarr{\zwsel{\htau_1}}{\htau_2}
-    }{
-      \aMove{\dParent}
-    }{
-      \zwsel{\tarr{\htau_1}{\htau_2}}
-    }
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \tarr{{\htau_1}}{\zwsel{\htau_2}}
-    }{
-      \aMove{\dParent}
-    }{
-      \zwsel{\tarr{\htau_1}{\htau_2}}
-    }
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \tarr{\zwsel{\htau_1}}{{\htau_2}}
-    }{
-      \aMove{\dNext}
-    }{
-      {\tarr{\htau_1}{\zwsel{\htau_2}}}
-    }
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \tarr{{\htau_1}}{\zwsel{\htau_2}}
-    }{
-      \aMove{\dPrev}
-    }{
-      {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
-    }
-  }
-\end{equation}
-% \begin{equation}
-% \inferrule{
-%   \performTyp{
-%     \ztau
-%   }{
-%     \aMove{\delta}
-%   }{
-%     \ztau'
-%   }
-% }{
-%   \performTyp{
-%     \tarr{\ztau}{\htau}
-%   }{
-%     \aMove{\delta}
-%   }{
-%     \tarr{\ztau'}{\htau}
-%   }
-% }
-% \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{
-%       \ztau
-%     }{
-%       \aMove{\delta}
-%     }{
-%       \ztau'
-%     }
-%   }{
-%     \performTyp{
-%       \tarr{\htau}{\ztau}
-%     }{
-%       \aMove{\delta}
-%     }{
-%       \tarr{\htau}{\ztau}
-%     }
-%   }
-% \end{equation}
-\end{subequations}
-% The final two rules above recurse into the zipper structure.
-
-The rules for relative movement within Z-expressions are similar.
-Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:\todo{remove prevSib}
-\begin{subequations}
-\begin{equation}
-\inferrule{
-  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
-}{
-  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
-}
-\end{equation}
-\begin{equation}
-  \inferrule{
-  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
-}{
-  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
-}
-\end{equation}
-\end{subequations}
-For concision, we show only the rules for ascription here:
-\begin{subequations}
-  \begin{equation}
-    \label{r:movefirstchild}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\hexp : \htau}
-    }{
-      \aMove{\dChild}
-    }{
-      \zwsel{\hexp} : \htau
-    }
-  }
-\end{equation}
-\begin{equation}
-  \label{r:moveparent}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\hexp} : \htau
-    }{
-      \aMove{\dParent}
-    }{
-      \zwsel{\hexp : \htau}
-    }
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \hexp : \zwsel{\htau}
-    }{
-      \aMove{\dParent}
-    }{
-      \zwsel{\hexp : \htau}
-    }
-  }
-\end{equation}
-\begin{equation}
-  \label{r:movenextsib}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\hexp} : \htau
-    }{
-      \aMove{\dNext}
-    }{
-      \hexp : \zwsel{\htau}
-    }
-  }
-\end{equation}
-\begin{equation}
-  \label{r:moveprevsib}
-  \inferrule{ }{
-    \performTyp{
-      \hexp : \zwsel{\htau}
-    }{
-      \aMove{\dPrev}
-    }{
-      \zwsel{\hexp} : \htau
-    }
-  }
-\end{equation}
-\begin{equation}
-\inferrule{
-  \performTyp{
-    \zexp
-  }{
-    \aMove{\delta}
-  }{
-    \zexp'
-  }
-}{
-  \performTyp{
-    \zexp : \htau
-  }{
-    \aMove{\delta}
-  }{
-    \zexp' : \htau
-  }
-}
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \performTyp{
-      \ztau
-    }{
-      \aMove{\delta}
-    }{
-      \ztau'
-    }
-  }{
-    \performTyp{
-      \hexp : \ztau
-    }{
-      \aMove{\delta}
-    }{
-      \hexp : \ztau'
-    }
-  }
-\end{equation}
-and this for the example\todo{revise}
-\begin{equation}\label{r:movefirstchild-lam}
-\inferrule{ }{
-  \performMove{
-    \zwsel{\hlam{x}{\hexp}}
-  }{
-    \aMove{\dChild}
-  }{
-    \hlam{x}{\zwsel{\hexp}}
-  }
-}
-\end{equation}
-
-\begin{equation}\label{r:moveparent-ap2}
-  \inferrule{ }{
-    \performMove{
-      \hap{{\hexp_1}}{\zwsel{\hexp_2}}
-    }{
-      \aMove{\dParent}
-    }{
-      \zwsel{\hap{\hexp_1}{\hexp_2}}
-    }
-  }
-\end{equation}
-
-\begin{equation}\label{r:moveparent-hole}
-  \inferrule{ }{
-    \performMove{
-      \hhole{\zwsel{\hexp}}
-    }{
-      \aMove{\dParent}
-    }{
-      \zwsel{\hhole{\hexp}}
-    }
-  }
-\end{equation}
-\end{subequations}
-
-\paragraph{Reachability}
-
-We give the following theorem to demonstrate that the rules for movement do
-indeed capture the process of moving the focus to any editable position
-within a term. Intuitively, given two zippered expressions that are the
-same up to the erasure of focus, its proof gives a list of composable
-actions that, when applied to the first term produces the second.
-
-As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
-every action in such a list, it is possible consider moving to an ill-typed
-term. Therefore the statement here matches the mutually recursive structure
-of the bidirectional typing judgements.
-
-\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
-\label{thrm:reach}
-Both of the following hold:
-\begin{enumerate}
-  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-   = \zexp'$.
-
-  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-   = \zexp'$.
-\end{enumerate}
-\end{theorem}
-
-\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.
-
-Again, the rule for Z-types is self-explanatory:
-\begin{subequations}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\htau}
-    }{
-      \aDel
-    }{
-      \zwsel{\tehole}
-    }
-  }
-\end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{\ztau}{\aDel}{\ztau'}
-%   }{
-%     \performTyp{\tarr{\ztau}{\htau}}{\aDel}{\tarr{\ztau'}{\htau}}
-%   }
-% \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{\ztau}{\aDel}{\ztau'}
-%   }{
-%     \performTyp{\tarr{\htau}{\ztau}}{\aDel}{\tarr{\htau}{\ztau'}}
-%   }
-% \end{equation}
-\end{subequations}
-
-Deletion within a Z-expression is similarly straightforward:
-\begin{subequations}
-\begin{equation}
-  \inferrule{ }{
-    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
-  }
-\end{equation}
-%\end{subequations}
-% The base case turns into a hole:
-%\begin{subequations}
-% \begin{equation}
-% \inferrule{ }{
-%   \performDel{\zwsel{\hexp}}{\hehole}
-% }
-% \end{equation}
-% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
-% \begin{equation}
-%   \inferrule{
-%     \performDel{\zexp}{\zexp'}
-%   }{
-%     \performDel{\zexp : \htau}{\zexp' : \htau}
-%   }
-% \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{\ztau}{\aDel}{\ztau'}
-%   }{
-%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
-%   }
-% \end{equation}
-\end{subequations}
-
-\paragraph{Deletability}\todo{do we care?}
-The meaning of the deletion action is to remove the entire term in focus,
-and its definition does not depend on the structure of the term being
-deleted. Therefore, there is no theorem to state that acts as a checksum on
-the rules analogous to Theorem \ref{thrm:reach} or \ref{thrm:construct},
-since we could not have forgotten any rules corresponding to the structure
-of the terms.
-
-\subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.
-
-Again, let us begin with type actions. The $\aConstruct{\farr}$ action constructs an arrow type. The focused H-type becomes the argument type, and the focus is placed on an empty return type hole:
-\begin{subequations}
-  \begin{equation}
-    \label{r:contarr}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\htau}
-    }{
-      \aConstruct{\farr}
-    }{
-      \tarr{\htau}{\zwsel{\tehole}}
-    }
-  }
-\end{equation}
-
-The $\aConstruct{\fnum}$ action replaces an empty Z-type hole with the $\tnum$ type:
-  \begin{equation}
-    \label{r:contnum}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\tehole}
-    }{
-      \aConstruct{\fnum}
-    }{
-      \zwsel{\tnum}
-    }
-  }
-\end{equation}
-
-% Construction proceeds recursively down the zipper:
-%   \begin{equation}
-%     \label{r:contarrL}
-%   \inferrule{
-%     \performTyp{\ztau}{\aConstruct{\varphi}}{\ztau'}
-%   }{
-%     \performTyp{
-%       \tarr{\ztau}{\htau}
-%     }{
-%       \aConstruct{\varphi}
-%     }{
-%       \tarr{\ztau'}{\htau}
-%     }
-%   }
-% \end{equation}
-%   \begin{equation}
-%     \label{r:contarrR}
-%   \inferrule{
-%     \performTyp{\ztau}{\aConstruct{\varphi}}{\ztau'}
-%   }{
-%     \performTyp{
-%       \tarr{\htau}{\ztau}
-%     }{
-%       \aConstruct{\varphi}
-%     }{
-%       \tarr{\htau}{\ztau'}
-%     }
-%   }
-% \end{equation}
-\end{subequations}
-
-\begin{subequations}
-
-Moving on to expression actions, we start to see more interesting rules.
-The $\aConstruct{\fasc}$ action operates differently depending on whether the focused expression synthesizes a type or is being analyzed against a type.
-In the first case, the ascribed type is the synthesized type:
-\begin{equation}
-  \label{r:constructasc}
-  \inferrule{ }{
-    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}{\htau}
-  }
-\end{equation}
-In the second case, the ascribed type is the type provided for analysis:
-\begin{equation}
-  \inferrule{ }{
-    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}
-  }
-\end{equation}
-
-The $\aConstruct{\fvar{x}}$ action places the variable $x$ into the focused empty hole.
-If that hole is being asked to synthesize a type, then the result of the action synthesizes the type assigned to $x$ in the context:
-\begin{equation}
-  \label{r:conevar}
-  \inferrule{ }{
-    \performSyn{\hGamma, x : \htau}{\zwsel{\hehole}}{\tehole}{\aConstruct{\fvar{x}}}{\zwsel{x}}{\htau}
-  }
-\end{equation}
-If the focused empty hole is being analyzed against a type that is inconsistent with the type assigned to $x$ by the context, $x$ is placed inside a hole:
-\begin{equation}
- \label{r:conevar2}
-  \inferrule{
-    \tincompat{\htau}{\htau'}
-  }{
-    \performAna{\hGamma, x : \htau'}{\zwsel{\hehole}}{\htau}{\aConstruct{\fvar{x}}}{\hhole{\zwsel{x}}}
-  }
-\end{equation}
-The rule above featured in the example in Section \ref{sec:example}.
-
-Notice that no rule was necessary for the case where the hole was being analyzed against a type compatible with the variable's type, because this case is handled by the action subsumption rule.
-
-The $\aConstruct{\flam{x}}$ action places a lambda term binding $x$ into an empty hole.
-If the focused empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the focus in the argument type position:
-\begin{equation}
-  \label{r:conelamhole}
-  \inferrule{ }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hehole}}
-      {\tehole}
-      {\aConstruct{\flam{x}}}
-      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
-      {\tarr{\tehole}{\tehole}}
-  }
-\end{equation}
-The type ascription is necessary because lambda expressions do not synthesize a type. If the focused empty hole is being analyzed against an arrow type, then no ascription is necessary:
-\begin{equation}
-  \inferrule{ }{
-    \performAna
-      {\hGamma}
-      {\zwsel{\hehole}}
-      {\tarr{\htau_1}{\htau_2}}
-      {\aConstruct{\flam{x}}}
-      {\hlam{x}{\zwsel{\hehole}}}
-  }
-\end{equation}
-
-If the focused empty hole is being analyzed against a type that is
-incompatible with any arrow type, expressed in the premise as an arrow with
-two holes, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
-is inserted inside a hole, to maintain Theorem \ref{thrm:actsafe}:
-\begin{equation}
-  \inferrule{
-    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
-  }{
-    \performAna
-      {\hGamma}
-      {\zwsel{\hehole}}
-      {\htau}
-      {\aConstruct{\flam{x}}}
-      {\hhole{
-        \hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}
-      }}
-  }
-\end{equation}
-
-The $\aConstruct{\fap}$ action applies the expression in focus to a hole. If the focused expression synthesizes a function type, then the rule is straightforward:
-\begin{equation}
-  \label{r:coneapfn}
-  \inferrule{ }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\tarr{\htau_1}{\htau_2}}
-      {\aConstruct{\fap}}
-      {\hap{\hexp}{\zwsel{\hehole}}}
-      {\htau_2}
-  }
-\end{equation}
-
-If the focused expression synthesizes a hole type, then we can treat it as if it synthesized the $\tarr{\tehole}{\tehole}$ type, exactly as described in Sec. \ref{sec:holes}:
-\begin{equation}
-  \inferrule{ }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\tehole}
-      {\aConstruct{\fap}}
-      {\hap{\hexp}{\zwsel{\hehole}}}
-      {\tehole}
-  }
-\end{equation}
-
-Finally, if the focused expression synthesizes a type that is incompatible with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
-\begin{equation}
-  \inferrule{
-    \tincompat{\htau}{\tarr{\tehole}{\tehole}}
-  }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\htau}
-      {\aConstruct{\fap}}
-      {\hap{\hhole{\hexp}}{\zwsel{\hehole}}}
-      {\tehole}
-  }
-\end{equation}
-
-The $\aConstruct{\farg}$ action places the focused expression instead in the argument position of an application.
-Because the function position is always an empty hole in this situation, we only need a single rule:
-\begin{equation}
-  \inferrule{ }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\htau}
-      {\aConstruct{\farg}}
-      {\hap{\zwsel{\hehole}}{\hexp}}
-      {\tehole}
-  }
-\end{equation}
-
-The $\aConstruct{\fnumlit{n}}$ action places the number expression $\hnum{n}$ into an empty hole. If the focused hole is being asked to synthesize a type, then the rule is straightforward:
-\begin{equation}
-  \label{r:conenumnum}
-  \inferrule{ }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hehole}}
-      {\tehole}
-      {\aConstruct{\fnumlit{n}}}
-      {\zwsel{\hnum{n}}}
-      {\tnum}
-  }
-\end{equation}
-If the focused hole is being analyzed against a type that is incompatible with $\tnum$, then we must place the number expression inside a hole:
-\begin{equation}
-  \inferrule{
-    \tincompat{\htau}{\tnum}
-  }{
-    \performAna
-      {\hGamma}
-      {\zwsel{\hehole}}
-      {\htau}
-      {\aConstruct{\fnumlit{n}}}
-      {\hhole{\zwsel{\hnum{n}}}}
-  }
-\end{equation}
-
-Finally, the $\aConstruct{\fplus}$ action constructs a plus expression with the focused expression as its first argument.
-If the focused expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
-\begin{equation}
-  \inferrule{
-    \tcompat{\htau}{\tnum}
-  }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\htau}
-      {\aConstruct{\fplus}}
-      {\hadd{\hexp}{\zwsel{\hehole}}}
-      {\tnum}
-  }
-\end{equation}
-
-Otherwise, we must place the focused expression inside a hole:
-\begin{equation}
-  \inferrule{
-    \tincompat{\htau}{\tnum}
-  }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\htau}
-      {\aConstruct{\fplus}}
-      {\hadd{\hhole{\hexp}}{\zwsel{\hehole}}}
-      {\tnum}
-  }
-\end{equation}
-\end{subequations}
-Notice that we do not have an action that explicitly wraps an expression in a non-empty hole.
-These arise implicitly when an action that would not na\"ively satisfy Theorem \ref{thrm:actsafe} is performed (see Figure \ref{fig:first-example}.)
-
-\subsubsection{Finishing}
-The final action we will consider in Hazelnut is $\aFinish$, which finishes the focused non-empty hole.
-
-If the focused non-empty hole appears in synthetic position, then it can always be finished:
-\begin{subequations}
-  \begin{equation}
-    \label{r:finishana}
-  \inferrule{
-    \hsyn{\hGamma}{\hexp}{\htau'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hhole{\hexp}}}
-      {\tehole}
-      {\aFinish}
-      {\zwsel{\hexp}}
-      {\htau'}
-  }
-\end{equation}
-
-If the focused non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the wrapped expression is consistent with the type the hole is being analyzed against.
-This amounts to analyzing those contents against the provided type (by subsumption):
-\begin{equation}
-  \inferrule{
-    \hana{\hGamma}{\hexp}{\htau}
-  }{
-    \performAna
-      {\hGamma}
-      {\zwsel{\hhole{\hexp}}}
-      {\htau}
-      {\aFinish}
-      {\zwsel{\hexp}}
-  }
-\end{equation}
-\end{subequations}
-
-\paragraph{Finishability}\todo{do we care?}
-As with deletion, since the meaning of the finish action does not depend on
-the structure of the term being finished other than requiring that it must
-be one of the two hole forms, there is no checksum theorem to state for
-this action.
-
-\subsubsection{Zipper Cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression.
-We also need to define the recursive cases, which propagate the action into the subtree where the focus appears.
-These rules follow the structure of the corresponding rules in the statics of H-expressions.
-
-\begin{subequations}
-For example, when the focus is in the expression position of an ascription, we use the analytic action performance judgement:
-\begin{equation}
-\inferrule{
-  \performAna
-    {\hGamma}
-    {\zexp}
-    {\htau}
-    {\alpha}
-    {\zexp'}
-}{
-  \performSyn
-    {\hGamma}
-    {\zexp : \htau}
-    {\htau}
-    {\alpha}
-    {\zexp' : \htau}
-    {\htau}
-}
-\end{equation}
-
-When the focus is in the type position of an ascription, we must re-check the ascribed expression because the type might have changed (in practice, one would optimize this check to only occur if the type actually was changed):
-\begin{equation}
-\inferrule{
-  \performTyp{\ztau}{\alpha}{\ztau'}\\
-  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
-}{
-  \performSyn
-    {\hGamma}
-    {\hexp : \ztau}
-    {\removeSel{\ztau}}
-    {\alpha}
-    {\hexp : \ztau'}
-    {\removeSel{\ztau'}}
-}
-\end{equation}
-
-If the focus is in the body of a lambda expression, then we must use the analytic action performance rule:
-\begin{equation}
-\inferrule{
-  \performAna
-    {\hGamma, x : \htau_1}
-    {\zexp}
-    {\htau_2}
-    {\alpha}
-    {\zexp'}
-}{
-  \performAna
-    {\hGamma}
-    {\hlam{x}{\zexp}}
-    {\tarr{\htau_1}{\htau_2}}
-    {\alpha}
-    {\hlam{x}{\zexp'}}
-}
-\end{equation}
-
-There are two rules that handle the case where the focus is in the function position of an application, corresponding to the two application rules in the statics.
-Each involves rechecking the argument against the new function type:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau_2}
-      {\alpha}
-      {\zexp'}
-      {\tarr{\htau_3}{\htau_4}}\\
-    \hana{\hGamma}{\hexp}{\htau_3}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\zexp}{\hexp}}
-      {\htau_1}
-      {\alpha}
-      {\hap{\zexp'}{\hexp}}
-      {\htau_4}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau_2}
-      {\alpha}
-      {\zexp'}
-      {\tehole}\\
-    \hana{\hGamma}{\hexp}{\tehole}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\zexp}{\hexp}}
-      {\htau_1}
-      {\alpha}
-      {\hap{\zexp'}{\hexp}}
-      {\tehole}
-  }
-\end{equation}
-
-Similarly, there are two rules that handle the case where the focus is in the argument position:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\hexp}{\tarr{\htau_2}{\htau}}\\
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\htau_2}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\hexp}{\zexp}}
-      {\htau}
-      {\alpha}
-      {\hap{\hexp}{\zexp'}}
-      {\htau}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\hexp}{\tehole}\\
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\tehole}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\hexp}{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\hap{\hexp}{\zexp'}}
-      {\tehole}
-  }
-\end{equation}
-
-The rules for the addition operator follow from the statics directly:
-\begin{equation}
-  \inferrule{
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\tnum}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hadd{\zexp}{\hexp}}
-      {\tnum}
-      {\alpha}
-      {\hadd{\zexp'}{\hexp}}
-      {\tnum}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\tnum}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hadd{\hexp}{\zexp}}
-      {\tnum}
-      {\alpha}
-      {\hadd{\hexp}{\zexp'}}
-      {\tnum}
-  }
-\end{equation}
-
-Finally, if the focus is inside a non-empty hole, we special case the situation where the action results in a doubly-nested empty hole, $\hhole{\hehole}$, to eliminate the nesting (given our current action semantics, only the delete action can cause this form to arise and the form $\hhole{\hhole{\zexp}}$ cannot arise):
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau}
-      {\alpha}
-      {\zexp'}
-      {\htau'}\\
-    \zexp' \neq \zwsel{\hehole}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hhole{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\hhole{\zexp'}}
-      {\tehole}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau}
-      {\alpha}
-      {\zwsel{\hehole}}
-      {\tehole}\\
-  }{
-    \performSyn
-      {\hGamma}
-      {\hhole{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\zwsel{\hehole}}
-      {\tehole}
-  }
-\end{equation}
-
-\paragraph{Constructibility}\todo{why is this with the zipper rules not the
-    construct rules?}
-
-We give the following theorem to demonstrate that the rules for the action
-$\aConstruct{\varphi}$ do indeed capture the process of building any term
-of the form $\varphi$. Intuitively, its proof gives a list of composable
-actions that, when applied one sequentially to an empty hole in focus,
-build any well-typed expression.
-
-As with Theorem \ref{thrm:reach}, the statement of this theorem must match
-the mutually recursive structure of the bidirectional typing judgements.
-
-\begin{theorem}[Constructability]\todo{why $\htau$ and not $\ztau$?}
-\label{thrm:construct}
-Both of the following hold:
-\begin{enumerate}
-  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
-      L \in \mathit{list} ~\mathtt{action}$ such that
-      $\mathit{iterate}~ L~ \zwsel{\hhole{}} = \zexp'$ and
-      $\removeSel{\zexp'} = \hexp$
-
-  \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
-      L \in \mathit{list} ~\mathtt{action}$ such that $\mathit{iterate}~
-      L~ \zwsel{\hhole{}} = \zexp'$ and $\removeSel{\zexp'} = \hexp$
-\end{enumerate}
-\end{theorem}
-
-Because actions are defined on the zippered form of expressions, we argue
-that we can construct a term that is the same as the target up to the
-erasure of focus rather than on the nose. Since any form inside a zippered
-term may be in focus, there are linearly many such terms for a given
-$\hexp$---Theorem \ref{thrm:construct} can be composed with
-Theorem \ref{thrm:reach} to reach any one that might be desired in
-particular.
-
-\end{subequations}
+\subsection{Mechanization}
+\label{sec:mech}\label{sec:mt}
+\input{agda}
 
@@
 
-\section{Dynamics of Hazelnut}\label{sec:dynamics}
+% \section{Dynamics of Hazelnut}\label{sec:dynamics}
 
-\subsection{Complete Expressions}
-... canonical forms and type safety with e complete in it ...
+% \subsection{Complete Expressions}
+% ... canonical forms and type safety with e complete in it ...
 
-\subsection{Holes as TODOs}
-... add an e err and define new type safety ...
+% \subsection{Holes as TODOs}
+% ... add an e err and define new type safety ...
 
-\subsection{Indeterminate Expressions}
-\input{dynamics}
+% \subsection{Indeterminate Expressions}
+% \input{dynamics}
 
-\clearpage
+% \clearpage
 
@@
 
-\section{Mechanization}
-\label{sec:mech}\label{sec:mt}
-\input{agda}
+\input{sumtypes}
 

### 2016-07-05 intro.tex
tweaks to sec 2 and 3

@@
 These complications have motivated a long line of research into \emph{structure editors}, i.e.
-program editors where the edit state is a  program structure, rather than text.
+program editors where every edit state is a program structure.
 % Eliminating text eliminates the possibility of syntax errors.
@@
 The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed.
-As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics.
+As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
 %Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful. 
@@
 This paper develops a principled solution to this problem.
-We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}, respectively.
+We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}.
 Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e.
 well-typed) H-expression with a single superimposed \emph{cursor}.
-We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because we follow Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
+We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because our encoding follows Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
 a function argument), only actions consistent with that type are permitted.
 
-Na\"ively, enforcing this injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner.
-For example, the programmer might need to construct the outer function application form before identifying the intended function.
+Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner.
+For example, the programmer would often need to construct the outer function application form before identifying the intended function.
 To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type.
 This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}.
-In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remains under construction.
+In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remain under construction.
 %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}.
@@
 \begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
-\item We begin in Section
-    \ref{sec:example} with two example edit sequences to develop the reader's intuitions.  
-\item We then give a detailed overview of Hazelnut's semantics and metatheory, which have been mechanized using the Agda proof assistant, in Section \ref{sec:hazel}.
+\item We begin in Sec. 
+    \ref{sec:example} with two examples of edit sequences to develop the reader's intuitions.  
+\item We then give a detailed overview of Hazelnut's semantics and metatheory, which have been mechanized using the Agda proof assistant, in Sec.  \ref{sec:hazel}.
 \item Hazelnut is a {foundational} calculus, i.e.
 a calculus that language and editor designers are expected to extend with higher level constructs.
-We show how Hazelnut's rich mechanized metatheory guides the development of an extension of Hazelnut with simple sum types in Section \ref{sec:extending}.
-\item In Section \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
+We extend Hazelnut with simple sum types in Sec.
+\ref{sec:extending} to demonstrate how Hazelnut's rich mechanized metatheory guides such a development.
+\item In Sec.
+\ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
 Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
@@
 
-\item In Section \ref{sec:rw}, we summarize related work.
-In particular, much of the technical machinery needed to handle type holes coincides with that developed in the study of gradual type systems.
-Similarly, one interpretation of expression holes is as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work.
+\item In Sec.
+\ref{sec:rw}, we summarize related work.
+In particular, much of the technical machinery needed to handle type holes coincides with machinery developed in the study of gradual type systems.
+Similarly, expression holes can be interpreted as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work.
 %This suggests a principled design for an ``edit and resume'' feature.
 
-\item We conclude in Section \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
+\item We conclude in Sec.  \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
 \end{itemize} 

### 2016-07-05 formally.tex
missing formally.tex

--- 2016-04-27 syntax-popl17.tex
@@
-%\documentclass[12pt]{article}
-\documentclass[preprint,9pt]{sigplanconf}
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-\usepackage{amsmath}
-\usepackage{amssymb} 
-\usepackage{amsthm}
-\usepackage{ stmaryrd }
-\usepackage{mathpartir}
-\usepackage{stmaryrd}
-\usepackage{wasysym}
-\usepackage{extarrows}
-\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-\usepackage{mathpazo}
-\usepackage{colortab}
-\usepackage{url}
-\usepackage{todonotes}
-\sloppy
-\def \TirNameStyle #1{\small\rulename{#1}}
-\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}
-
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-\newtheorem{definition}{Definition}
-\newenvironment{proof-sketch}{\noindent{\emph{Proof Sketch.}}}{\qed}
-\makeatletter
-
-\renewcommand\topfraction{0.85}
-\renewcommand\bottomfraction{0.85}
-\renewcommand\textfraction{0.1}
-\renewcommand\floatpagefraction{0.85}
-
-\AtBeginDocument{%
- \abovedisplayskip=3pt
- \abovedisplayshortskip=0pt
- \belowdisplayskip=2pt
- \belowdisplayshortskip=0pt
-}
-
-\setlength{\floatsep}{10pt}
-\setlength{\textfloatsep}{12pt}
-
-\usepackage[compact]{titlesec}  
-\titlespacing*{\section}{0pt}{3pt}{3pt}
-\titlespacing*{\subsection}{0pt}{4pt}{3pt}
-\titlespacing*{\subsubsection}{0pt}{4pt}{3pt}
-\titlespacing*{\paragraph}{0pt}{3pt}{5pt}
-\setlength{\skip\footins}{3pt}
-
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-
-\usepackage{enumitem}
-
-\input{macros}
-\begin{document}
-
-\conferenceinfo{-}{-} 
-\copyrightyear{-} 
-\copyrightdata{[to be supplied]}
-
-%\titlebanner{}        % These are ignored unless
-\preprintfooter{Draft}   % 'preprint' option specified.
-
-\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus}
-%\subtitle{Modular Type Constructors}
-
-\authorinfo{\vspace{-2px}}{}{}
-%\authorinfo{~}{~}{~}
-%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
-%         {Carnegie Mellon University}
-%         {\{comar, aldrich\}@cs.cmu.edu}
-
-\maketitle
-\begin{abstract}
-Programs are rich inductive structures, but programmers typically construct and manipulate them indirectly, through flat textual representations.
-This indirection  comes at a cost: programmers must understand the  subtleties of parsing, and it can require several text editor actions to make a single syntactically and semantically well-defined change.
-During these sequences of edit actions, or when the programmer makes a mistake, programmers and programming tools must contend with malformed text or semantically ill-defined programs.
-
-\emph{Structure editors} promise to alleviate these burdens by exposing only edit actions that  produce sensible changes to the program structure.
-Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}.
-They also focus primarily on syntactic well-formedness, so programs can still be left semantically ill-defined as they are being constructed.
-In this paper, we report on our ongoing efforts to develop Hazelnut, a minimal  structure editor defined in a principled type-theoretic style where all edit actions leave the program in both a syntactically and semantically well-defined state.
-Uniquely, Hazelnut does not force the programmer to construct the program in a strictly ``outside-in'' fashion.
-Formally, Hazelnut is a bidirectionally typed lambda calculus extended with
-1) \emph{holes} (which mark subterms that are being  constructed from the inside out);
-2) a \emph{focus model}; and 3) a bidirectional \emph{action model} equipped with a useful \emph{action sensibility} theorem.
-\end{abstract}
-
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
-%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
-%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%extensible languages; module systems; type abstraction; typed compilation; type-level computation
-
-\section{Introduction}\label{sec:introduction}
-\input{intro}
-
-\section{Programming in Hazelnut}\label{sec:example}
-\input{example}
-
-\section{Hazelnut, Formally}
-\label{sec:hazel}
-Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
-\begin{itemize}
-\item \textbf{H-types} and \textbf{H-expressions} (Sec.
-\ref{sec:holes}), which are terms with \emph{holes}.
-Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
-\item \textbf{Z-types} and \textbf{Z-expressions} (Sec.
-\ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
-\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.
-
-Whenever an action is performed on a well-typed expression, it produces another well-typed expression in a \emph{sensible} manner.
-More specifically, the action semantics satisfies a crucial \emph{sensibility theorem}, stated in Sec.
-\ref{sec:actions}.
+% !TEX root = hazelnut-popl17.tex
+
+The previous section introduced Hazelnut by example. In this section, we systematically introduce the following  structures:
+\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
+\item \textbf{H-expressions} and \textbf{H-types} (Sec.
+\ref{sec:holes}), i.e.
+expressions and types with \emph{holes}.
+H-types classify H-expressions according to Hazelnut's \textbf{bidirectionally typed static semantics}.
+\item \textbf{Z-expressions} and \textbf{Z-types} (Sec.
+\ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-expressions and H-types, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-expression (resp.
+Z-type) corresponds to an H-expression (resp.
+H-type) by \emph{cursor erasure}.
+\item \textbf{Actions} (Sec.
+\ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectionally typed action semantics}.
+The action semantics enjoys a rich metatheory.
+Of particular note, the \emph{sensibility theorem} establishes that every edit state is well-typed after cursor erasure.
 \end{itemize}
 
-In our overview of the semantics below, we will reproduce only the most interesting rules, and in some cases we will do so ``out of order.'' The appendix (and our Agda formalization, see Sec.
-\ref{sec:mech}) defines the complete collection of rules in their dependency order.
+Our overview below omits certain ``uninteresting'' details.
+The supplement includes the complete collection of rules, in definitional order.
+These rules, along with the proofs of the metatheorems, have been mechanized using Agda \cite{norell:thesis}, also in the supplement.
+We will summarize this effort in Sec.
+\ref{sec:mech}.% The supplement contains the Agda sources and additional technical details.
+
 \subsection{Holes}\label{sec:holes}
+
+The syntax of H-types and H-expressions is given in Figure \ref{fig:hexp-syntax}.
+Most of the forms correspond directly to those of the simply-typed lambda calculus extended with type $\tnum$.
+The number expression corresponding to the number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$.
+In addition to these standard forms, \emph{empty holes} are drawn $\hehole$ and \emph{non-empty H-expression holes} are drawn $\hhole{\hexp}$.
+In our simple calculus, all well-formed type expressions are valid types, so we do not need non-empty H-type holes.%Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.
+
 \begin{figure}[t]
@@
 
-The syntax of H-types and H-expressions is given in Figure \ref{fig:hexp-syntax}.
-Most of the forms correspond directly to those of the simply-typed lambda calculus extended with type $\tnum$.
-The number expression corresponding to the number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$.
-In addition to these standard forms, \emph{empty holes} are drawn $\hehole$ and \emph{non-empty H-expression holes} are drawn $\hhole{\hexp}$.
-In our simple calculus, all well-formed type expressions are valid types, so we do not need non-empty H-type holes.%Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.
-
 We refer to terms that do not contain subterms of hole form as \emph{complete}.
@@
 \end{equation}
-\begin{equation}
+\begin{equation}\label{rule:move-parent-arr-right}
   \inferrule{ }{
@@
 \end{equation}
-\begin{equation}
+\begin{equation}\label{rule:move-nextSib-arr}
   \inferrule{ }{
@@
 \end{equation}
-\begin{equation}
+\begin{equation}\label{rule:move-parent-asc-right}
   \inferrule{ }{
@@
 \end{equation}
+and this for the example\todo{revise}
+\begin{equation}\label{r:movefirstchild-lam}
+\inferrule{ }{
+  \performMove{
+    \zwsel{\hlam{x}{\hexp}}
+  }{
+    \aMove{\dChild}
+  }{
+    \hlam{x}{\zwsel{\hexp}}
+  }
+}
+\end{equation}
+
+\begin{equation}\label{r:moveparent-ap2}
+  \inferrule{ }{
+    \performMove{
+      \hap{{\hexp_1}}{\zwsel{\hexp_2}}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hap{\hexp_1}{\hexp_2}}
+    }
+  }
+\end{equation}
+
+\begin{equation}\label{r:moveparent-hole}
+  \inferrule{ }{
+    \performMove{
+      \hhole{\zwsel{\hexp}}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hhole{\hexp}}
+    }
+  }
+\end{equation}
 \end{subequations}
+
+\paragraph{Reachability}
+
+We give the following theorem to demonstrate that the rules for movement do
+indeed capture the process of moving the focus to any editable position
+within a term. Intuitively, given two zippered expressions that are the
+same up to the erasure of focus, its proof gives a list of composable
+actions that, when applied to the first term produces the second.
+
+As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
+every action in such a list, it is possible consider moving to an ill-typed
+term. Therefore the statement here matches the mutually recursive structure
+of the bidirectional typing judgements.
+
+\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
+\label{thrm:reach}
+Both of the following hold:
+\begin{enumerate}
+  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
+   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
+   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
+   = \zexp'$.
+
+  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
+   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
+   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
+   = \zexp'$.
+\end{enumerate}
+\end{theorem}
+
 \subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.
@@
 If the focused expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
-\begin{equation}
+\begin{equation}\label{rule:construct-plus-compat}
   \inferrule{
@@
   \begin{equation}
-    \label{r:finishana}
   \inferrule{
@@
 This amounts to analyzing those contents against the provided type (by subsumption):
-\begin{equation}
+\begin{equation}\label{r:finishana}
   \inferrule{
@@
 
+\paragraph{Finishability}\todo{do we care?}
+As with deletion, since the meaning of the finish action does not depend on
+the structure of the term being finished other than requiring that it must
+be one of the two hole forms, there is no checksum theorem to state for
+this action.
+
 \subsubsection{Zipper Cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression.
@@
 
+\paragraph{Constructibility}\todo{why is this with the zipper rules not the
+    construct rules?}
+
+We give the following theorem to demonstrate that the rules for the action
+$\aConstruct{\varphi}$ do indeed capture the process of building any term
+of the form $\varphi$. Intuitively, its proof gives a list of composable
+actions that, when applied one sequentially to an empty hole in focus,
+build any well-typed expression.
+
+As with Theorem \ref{thrm:reach}, the statement of this theorem must match
+the mutually recursive structure of the bidirectional typing judgements.
+
+\begin{theorem}[Constructability]\todo{why $\htau$ and not $\ztau$?}
+\label{thrm:construct}
+Both of the following hold:
+\begin{enumerate}
+  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
+      L \in \mathit{list} ~\mathtt{action}$ such that
+      $\mathit{iterate}~ L~ \zwsel{\hhole{}} = \zexp'$ and
+      $\removeSel{\zexp'} = \hexp$
+
+  \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
+      L \in \mathit{list} ~\mathtt{action}$ such that $\mathit{iterate}~
+      L~ \zwsel{\hhole{}} = \zexp'$ and $\removeSel{\zexp'} = \hexp$
+\end{enumerate}
+\end{theorem}
+
+Because actions are defined on the zippered form of expressions, we argue
+that we can construct a term that is the same as the target up to the
+erasure of focus rather than on the nose. Since any form inside a zippered
+term may be in focus, there are linearly many such terms for a given
+$\hexp$---Theorem \ref{thrm:construct} can be composed with
+Theorem \ref{thrm:reach} to reach any one that might be desired in
+particular.
 \end{subequations}
-
-\section{Mechanization}
-\label{sec:mech}\label{sec:mt}
-\input{agda}
-
-\section{Implementation}
-\label{sec:impl}
-\input{implementation}
-
-\section{Related Work}\label{sec:rw}
-%\subsection{Structure Editors}
-
-Structured editing has been recognized as a way to avoid the possibility of syntax errors for decades.  An early example is the
-The Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981}, first published in 1981.
-The synthesizer generator~\cite{Reps:1984:SG:390010.808247} allows the user to create an attribute-grammar specification that then can be used to generate a structured editor.
-CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language.
-Barista \cite{ko_barista:_2006} is a modern take on the same basic concept.%These early systems were developed  of the systems are rooted in the type-theoretic tradition.
-
-Novice programmers have been a common target for structure editors. For example,
-GNOME\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates.
-Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.
-Touchdevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students.
-Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students.
-These are largely drag-and-drop user interfaces with a limited action model and an unclear semantics.
-
-Not all structure editors are for educational purposes. For example,
-mbeddr \cite{voelter_mbeddr:_2012} is an extensible C-based Programming Language and IDE (nominally, for programming embedded systems.)
-mbeddr is build on top of the commercial JetBrains MPS framework for constructing structure editors.
-Another popular approach is to bring elements of structured editing into a traditional editor.
-Codelets \cite{oney_codelets:_2012} uses structured editing to add interactive documentation and examples in an editor. 
-Our previous work on Graphite~\cite{Omar:2012:ACC:2337223.2337324} allows developers to associate structured editing interfaces called  \emph{palettes} with types.
-Graphite is integrated into a text-based program editor (Eclipse.)
-
-Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g.
-Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes.
-These  systems are also, to our knowledge, not formally well-defined but rather exist only as part of system implementations.
-
-Perhaps the systems most similar in spirit to Hazelnut are Lamdu~\cite{lamdu} and Unison~\cite{unison}.
-Like Hazelnut, these are both statically typed functional language editors.
-In both cases, the language is similar to Haskell.
-In Lamdu, the editor uses structure editing to enable Live Programming, where the code is always being executed as it is being written.
-
-Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified.
-For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model.
-Unison is also a rather large body of Haskell code, though its action model appears superficially more similar to ours.
-We maintain what we believe to be a stronger action sensibility invariant than Unison (i.e.
-in Unison, one must construct expressions from the outside-in.) These systems are rich sources of interesting ideas, however -- there is room enough for many different approaches in this (re-)emerging space.
-
-%Drag-and-drop / for novices: lots of examples, e.g. Alice and others
-%
-%Contemporary: Lamdu, MPS/Mbeddr, TouchDevelop
-%
-%Hybrid: Cyrus' active code completion paper
-
-%\subsection{Refactoring Models}
-%(Michael, can you fill this section out?)
-
-%\subsection{Formal Editor Models}
-%Need to do a search to see what else has been done...
-
-\section{Discussion \& Conclusion}
-\label{sec:future}
-This paper presented Hazelnut, a type theoretic structure editor calculus.
-Our aim is to take a principled approach to its design by formally specifying its semantics, providing strong metatheoretic guarantees, mechanizing its semantics and metatheory in Agda and implementing it using the concepts of  functional reactive programming.
-As of this submission, we have achieved reasonable confidence in the formal system presented above, and have transitioned our focus toward the mechanization and implementation efforts.
-By the time of presentation, we anticipate having complete or nearly complete versions of these.
-
-\subsection{Future Work}
-Hazelnut is, obviously, a very limited language at its core.
-So the most obvious avenue for future work is to increase the expressive power of this language.
-Our plan is to simultaneously maintain a mechanization and implementation (following, for example, Standard ML) as we proceed, ultimately producing the first large-scale, formally verified bidirectionally typed language codesigned with a type-aware editor.
-It may be that certain language features are unnecessary given a sufficiently advanced type-aware structure editor (e.g.
-SML's \texttt{open}?), while other features may only be practical with editor support.
-We intend to use Hazelnut and derivative systems thereof as a platform for rigorously exploring such questions.
-
-There are various aspects of the editor model that we have not yet formalized.
-For example, our action model does not consider how actions are actually entered using, for example, key combinations or chords.
-It also did not provide any specific model of how available actions will be determined for presentation to the user.
-In practice, we would want also to rank available actions in some reasonable manner (perhaps based on usage data gathered from other users or code repositories.)
-
-Another research direction is in exploring how types can be used to control the presentation of expressions in the editor.
-For example, following our approach in a textual setting on \emph{type-specific languages} (TSLs), it should be possible to have the type that an expression is being analyzed against define alternative display forms and interaction modes \cite{TSLs}.
-
-Finally, we did not consider any aspects of \emph{collaborative programming}, such as a packaging system, a differencing algorithm for use in a source control system, support for multiple simultaneous focii for different users, and so on.
-These are all interesting avenues for future work.
-
-On the theoretical side, the notion of having one of many possible holes in a term in focus has a very strong intuitive connection
-  to the proof theoretic notion of focusing \cite{Simmons11tr}. Beyond just
-  the name, both seem to involve, in some sense, a search through the space of possible
-  ways to finish a derivation. We intend to explore this connection to see
-  if it's coincidental or more meaningful and welcome insights in this regard.
-
-We already discussed a connection to gradual typing \cite{Siek06a}.
-We hope to explore this connection more thoroughly.
-In particular, it may be possible to better support exploratory and live programming by allowing even programs with holes in them to execute as long as those holes are only in the type portions, by deferring to the semantics given in work on gradual typing.
-
-It may also be possible to give a dynamics to incomplete expressions.
-Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}.
-In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
-$\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
-explored an equational theory for terms with holes, but intend to once our
-formalization effort is more mature.
-
-\begin{quote}
-In any case, these are but steps toward more graphical program-description
-systems, for we will not forever stay confined to mere strings of symbols.
-
---- Marvin Minsky, Turing Award lecture
-\end{quote}% We recommend abbrvnat bibliography style.
-\clearpage
-\bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\bibliography{../research}
-%\softraggedright
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{document}

### 2016-07-05 formally.tex
added rules/macro for type matching at arrow

@@
 The supplement includes the complete collection of rules, in definitional order.
-These rules, along with the proofs of the metatheorems, have been mechanized using Agda \cite{norell:thesis}, also in the supplement.
+These rules, along with the proofs of the metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement.
 We will summarize this effort in Sec.
@@
 
-The syntax of H-types and H-expressions is given in Figure \ref{fig:hexp-syntax}.
-Most of the forms correspond directly to those of the simply-typed lambda calculus extended with type $\tnum$.
-The number expression corresponding to the number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$.
-In addition to these standard forms, \emph{empty holes} are drawn $\hehole$ and \emph{non-empty H-expression holes} are drawn $\hhole{\hexp}$.
-In our simple calculus, all well-formed type expressions are valid types, so we do not need non-empty H-type holes.%Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.
+Figure \ref{fig:hexp-syntax} defines the syntax of H-types and H-expressions.
+Most forms correspond directly to those of the simply typed lambda calculus (STLC) extended with a single base type, $\tnum$, of numbers (cf.
+\cite{pfpl}.) The number expression corresponding to the mathematical number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$.
+The form $\hexp : \htau$ is an explicit \emph{type ascription}.
+
+In addition to these standard forms, \emph{type holes} and \emph{empty expression holes} are both drawn $\hehole$ and \emph{non-empty expression holes} are drawn $\hhole{\hexp}$.
+In our simple calculus, all well-formed H-types are valid classifiers of expressions, so we do not need non-empty type holes.
+Types and expressions that do not contain holes are \emph{complete types} and \emph{complete expressions}, respectively.
+Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete.
+The rules are omitted here.
+We will not define a dynamics for Hazelnut explicitly, but the dynamics for complete H-expressions would be entirely standard (in Sec.
+\ref{sec:future} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers.
+%Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.
 
@@
 $\arraycolsep=4pt\begin{array}{lllllll}
-\mathsf{HTyp} & \tau,\htau & ::= &
+\mathsf{HTyp} & \htau & ::= &
   \tarr{\htau}{\htau} ~\vert~
@@
   \tehole\\
-\mathsf{HExp} & e,\hexp & ::= &
+\mathsf{HExp} & \hexp & ::= &
   \hexp : \htau ~\vert~
@@
 
-We refer to terms that do not contain subterms of hole form as \emph{complete}.
-Informally, we will use metavariables $\tau$ and $e$ rather than $\htau$ and $\hexp$ for complete H-types and H-expressions, respectively.
-Formally, we can derive $\hcomplete{\tau}$ when $\tau$ is a complete H-type, and $\hcomplete{e}$ when $e$ is a complete H-expression.
-We omit the straightforward definitions of these judgements for concision.
-The dynamics of Hazelnut, which we need not detail here, is defined only  over complete H-expressions (i.e.
-we can only ``run'' a complete program, though see Sec.
-\ref{sec:future}.)\todo{get rid of e/edot distinction}\todo{change this to refer to dynamics section}
-
-The statics of Hazelnut is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100}, i.e. around the following mutually defined typing judgements:
+The statics of Hazelnut is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,bidi-tutorial} around the following mutually defined typing judgements:
 \[\arraycolsep=15pt\begin{array}{ll}
@@
 \end{array}\]
-where typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$.
+where typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner.
 Derivations of the type analysis judgement establish that $\hexp$ can appear where an expression of type $\htau$ is expected.
-Derivations of the type synthesis judgement determine a type that can be assigned to $\hexp$ even in positions where an expected type is not known (e.g.
+Derivations of the type synthesis judgement infer a type from $\hexp$, which is necessary in positions where an expected type is not available (e.g.
 at the top level.) Algorithmically, the type is an ``input'' of the type analysis judgement, but an ``output'' of the type synthesis judgement.
@@
 type ascriptions are unnecessary when an expression is being analyzed against a known type.
-Making a judgemental distinction between these two notions will be essential for giving a sensible action semantics to our system (Sec. \ref{sec:actions}.)
+Making a judgemental distinction between these two notions will be essential in our action semantics (Sec. \ref{sec:actions}.)
 
@@
 
+\begin{figure}
+\noindent\fbox{$\tcompat{\htau}{\htau'}$}
+% \begin{subequations}%\label{rules:tcompat}
+% \begin{equation}%\label{rule:tcompat-comm}
+% \inferrule
+% %[TCSym]
+% {
+%   \tcompat{\htau}{\htau'}
+% }{
+%   \tcompat{\htau'}{\htau}
+% }
+% \end{equation}
+\begin{mathpar}
+\inferrule{ }{
+  \tcompat{\tehole}{\htau}
+}
+
+% \end{equation}
+% \begin{equation}%\label{rule:tcompat-hole}
+\inferrule{ }{
+  \tcompat{\htau}{\tehole}
+}
+
+\inferrule{ }{
+  \tcompat{\htau}{\htau}
+}
+
+\inferrule{
+  \tcompat{\htau_1}{\htau_1'}\\
+  \tcompat{\htau_2}{\htau_2'}
+}{
+  \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
+}
+\end{mathpar}
+% \begin{equation}%\label{rule:tcompat-num}
+% \end{equation}
+% \begin{equation}%\label{rule:tcompat-arr}
+% \end{equation}
+% \end{subequations}
+\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}
+\begin{mathpar}
+% \begin{subequations}
+% \begin{equation}
+\inferrule{ }{
+  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
+}
+
+% \end{equation}
+% \begin{equation}
+\inferrule{ }{
+  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
+}
+% \end{equation}
+% \end{subequations}
+\end{mathpar}
+% \noindent\fbox{$\tincompat{\htau}{\htau'}$}
+% \begin{subequations}
+%   % \begin{equation}
+%   %   \inferrule{
+%   %     \tincompat{\htau}{\htau'}
+%   %   }{
+%   %     \tincompat{\htau'}{\htau}
+%   %   }
+%   % \end{equation}
+%   \begin{equation}
+%     \inferrule{ }{
+%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tnum}
+%     }
+%   \end{equation}
+%   \begin{equation}
+%     \inferrule{ }{
+%       \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
+%     }
+%   \end{equation}
+%   \begin{equation}
+%     \inferrule{
+%       \tincompat{\htau_1}{\htau_1'}
+%     }{
+%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
+%     }
+%   \end{equation}
+%   \begin{equation}
+%     \inferrule{
+%       \tincompat{\htau_2}{\htau_2'}
+%     }{
+%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
+%     }
+%   \end{equation}
+% \end{subequations}
+\caption{Type consistency and type matching at arrow.}
+\label{fig:type-consistency}
+\end{figure}
+
 \begin{subequations}\label{rules:syn-ana}
-Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any \emph{compatible} type.
-This is expressed by the \emph{subsumption rule}:
+Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any other \emph{consistent} type, according to the following \emph{subsumption rule}:
 \begin{equation}\label{rule:ana-subsume}
@@
 \end{equation}
-The \emph{H-type compatibility judgement}, $\tcompat{\htau}{\htau'}$, reduces to syntactic equality for complete H-types.
-For incomplete H-types, the rules are given after we discuss the semantics of holes below.
-
+The \emph{H-type consistency relation}, $\tcompat{\htau}{\htau'}$, reduces to equality for complete H-types.
+We discuss its purpose with regard to incomplete H-types when we discuss the semantics of holes below.
 First, let us briefly review the standard constructs.
-Type ascription allows the user to state  a type for the ascribed expression to be analyzed against:
+
+Type ascription allows the user to explicitly state  a type for the ascribed expression to be analyzed against:
 \begin{equation}\label{rule:syn-asc}
@@
 
-A variable synthesizes the type that the context assigns to it:
+A variable synthesizes the hypothesized type:
 \begin{equation}\label{rule:syn-var}
@@
 
-Functions are not themselves annotated with types, so they can only appear in analytic position:
+Lambda abstractions are not themselves annotated, so they can appear only in analytic position:
 \begin{equation}\label{rule:syn-lam}
 \inferrule{
+  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
   \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
 }{
-  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
-}
-\end{equation}
-It would be straightforward to also add a ``half-annotated'' lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out of our calculus \cite{DBLP:conf/tldi/ChlipalaPH05}.
-
-For function application, if the expression in function position synthesizes an arrow type, the argument is analyzed against the synthesized argument type:
+  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
+}
+\end{equation}
+It is possible to also define a ``half-annotated'' lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.
+
+When applying a function, if the expression in function position synthesizes an arrow type, the argument is analyzed against the synthesized argument type:
+% \begin{equation}\label{rule:syn-ap}
+% \inferrule{
+%   \hsyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}\\
+%   \hana{\hGamma}{\hexp_2}{\htau_2}
+% }{
+%   \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
+% }
+% \end{equation}
 \begin{equation}\label{rule:syn-ap}
 \inferrule{
-  \hsyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}\\
+  \hsyn{\hGamma}{\hexp_1}{\htau}\\
+  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
   \hana{\hGamma}{\hexp_2}{\htau_2}
 }{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
+  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
 }
@@
 
-A non-empty hole contains an H-expression that is ``under construction''. The inner expression must synthesize some type, but the non-empty hole synthesizes only the hole type:
+A non-empty hole contains an H-expression that is ``under construction'', as described in Sec.
+\ref{sec:example}.
+The  expression inside the hole must synthesize some type, but, like the empty hole, the non-empty hole synthesizes only the hole type:
 \begin{equation}\label{rule:syn-hole}
@@
 \end{equation}
-The type compatibility judgement $\tcompat{\htau}{\htau'}$, which appeared as a premise in the subsumption rule, makes the hole type compatible with any other type:
+The type consistency judgement $\tcompat{\htau}{\htau'}$, which appeared as a premise in the subsumption rule, makes the hole type compatible with any other type:
 \begin{subequations}\label{rules:tcompat}
@@
 \end{equation}
-The remaining rules, given in the appendix, establish that type compatibility is symmetric and reflexive (but not transitive.)
+\end{subequations}
+The remaining rules, given in the appendix, establish that type consistency is symmetric and reflexive (but not transitive.)
+
 % \begin{equation}\label{rule:tcompat-comm}
@@
 % \end{equation}
-\end{subequations}
-Consequently, by subsumption, we can derive that $\hana{id : \tarr{\tnum}{\tnum}}{\hhole{id}}{\tnum}$, as is necessary to synthesize a type for the H-expression on Line 14 of Fig.
+
+Taken we can derive that $\hana{id : \tarr{\tnum}{\tnum}}{\hhole{id}}{\tnum}$, as is necessary to synthesize a type for the H-expression on Line 14 of Fig.
 \ref{fig:first-example}.
@@
 
-The final rule handles function applications where the expression in function position synthesizes a hole type, rather than an arrow type.
+% The final rule handles function applications where the expression in function position synthesizes a hole type, rather than an arrow type.
 We treat it as if it had instead synthesized $\tarr{\tehole}{\tehole}$:
-\begin{equation}\label{rule:syn-ap-2}
-\inferrule{
-  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
-  \hana{\hGamma}{\hexp_2}{\tehole}
-}{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}
-}
-\end{equation}
+% \begin{equation}\label{rule:syn-ap-2}
+% \inferrule{
+%   \hsyn{\hGamma}{\hexp_1}{\tehole}\\
+%   \hana{\hGamma}{\hexp_2}{\tehole}
+% }{
+%   \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}
+% }
+% \end{equation}
 

### 2016-07-05 formally.tex
added statics figures

@@
 In addition to these standard forms, \emph{type holes} and \emph{empty expression holes} are both drawn $\hehole$ and \emph{non-empty expression holes} are drawn $\hhole{\hexp}$.
-In our simple calculus, all well-formed H-types are valid classifiers of expressions, so we do not need non-empty type holes.
+We do not need non-empty type holes because every H-type is a valid classifier of H-expressions.
 Types and expressions that do not contain holes are \emph{complete types} and \emph{complete expressions}, respectively.
-Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete.
-The rules are omitted here.
-We will not define a dynamics for Hazelnut explicitly, but the dynamics for complete H-expressions would be entirely standard (in Sec.
+Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete (see supplement.) We are not concerned here with defining a dynamic semantics for Hazelnut, but the dynamics for complete H-expressions would be entirely standard (in Sec.
 \ref{sec:future} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers.
@@
 \end{figure}
-
-The statics of Hazelnut is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,bidi-tutorial} around the following mutually defined typing judgements:
-\[\arraycolsep=15pt\begin{array}{ll}
-%\textbf{Judgement Form} & \textbf{Description}\\
-\hana{\hGamma}{\hexp}{\htau} & \text{$\hexp$ analyzes against $\htau$}\\
-\hsyn{\hGamma}{\hexp}{\htau} & \text{$\hexp$ synthesizes $\htau$}
-\end{array}\]
-where typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner.
-Derivations of the type analysis judgement establish that $\hexp$ can appear where an expression of type $\htau$ is expected.
-Derivations of the type synthesis judgement infer a type from $\hexp$, which is necessary in positions where an expected type is not available (e.g.
-at the top level.) Algorithmically, the type is an ``input'' of the type analysis judgement, but an ``output'' of the type synthesis judgement.
-%The rules describe a \emph{local type inference} scheme, i.e.
-type ascriptions are unnecessary when an expression is being analyzed against a known type.
-Making a judgemental distinction between these two notions will be essential in our action semantics (Sec. \ref{sec:actions}.)
-
- %We use the metavariable $\Gamma$ for \emph{complete typing contexts}, i.e. typing contexts where each hypothesis mentions only complete types.
-
 \begin{figure}
-\noindent\fbox{$\tcompat{\htau}{\htau'}$}
+\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\tau$ and $\tau'$ are consistent}
 % \begin{subequations}%\label{rules:tcompat}
@@
 % \end{subequations}
-\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}
+\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
 \begin{mathpar}
@@
 % \end{subequations}
-\caption{Type consistency and type matching at arrow.}
+\caption{H-type consistency and H-type matching at arrow.}
 \label{fig:type-consistency}
 \end{figure}
+
+\begin{figure}
+\fbox{$\hana{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ analyzes against $\htau$}
+\begin{subequations}
+\begin{equation}\label{rule:ana-subsume}
+\inferrule{
+  \hsyn{\hGamma}{\hexp}{\htau'}\\
+  \tcompat{\htau}{\htau'}
+}{
+  \hana{\hGamma}{\hexp}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-lam}
+\inferrule{
+  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
+  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
+}{
+  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
+}
+\end{equation}
+\end{subequations}
+\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
+\begin{subequations}
+\begin{equation}\label{rule:syn-asc}
+\inferrule{
+  \hana{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hexp : \htau}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-var}
+\inferrule{ }{
+  \hsyn{\hGamma, x : \htau}{x}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-ap}
+\inferrule{
+  \hsyn{\hGamma}{\hexp_1}{\htau}\\
+  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
+  \hana{\hGamma}{\hexp_2}{\htau_2}
+}{
+  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-num}
+\inferrule{ }{
+  \hsyn{\hGamma}{\hnum{n}}{\tnum}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-plus}
+\inferrule{
+  \hana{\hGamma}{\hexp_1}{\tnum}\\
+  \hana{\hGamma}{\hexp_2}{\tnum}
+}{
+  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-ehole}
+\inferrule{ }{
+  \hsyn{\hGamma}{\hehole}{\tehole}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-hole}
+\inferrule{
+  \hsyn{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
+}
+\end{equation}
+\end{subequations}
+\caption{Analysis and synthesis.}
+\label{fig:ana-syn}
+\end{figure}
+
+The statics of Hazelnut is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,bidi-tutorial} around the two mutually defined judgements in Figure \ref{fig:ana-syn}.
+Typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner.
+Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, establish that $\hexp$ can appear where an expression of type $\htau$ is expected.
+Derivations of the type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, infer a type from $\hexp$, which is necessary in positions where an expected type is not available (e.g.
+at the top level.) Algorithmically, the type is an ``input'' of the type analysis judgement, but an ``output'' of the type synthesis judgement.
+%The rules describe a \emph{local type inference} scheme, i.e.
+type ascriptions are unnecessary when an expression is being analyzed against a known type.
+Making a judgemental distinction between these two notions will be essential in our action semantics (Sec. \ref{sec:actions}.)
+
+ %We use the metavariable $\Gamma$ for \emph{complete typing contexts}, i.e. typing contexts where each hypothesis mentions only complete types.
 
@@
 \end{equation}
-The \emph{H-type consistency relation}, $\tcompat{\htau}{\htau'}$, reduces to equality for complete H-types.
-We discuss its purpose with regard to incomplete H-types when we discuss the semantics of holes below.
-First, let us briefly review the standard constructs.
+
+Figure \ref{fig:type-consistency} defines the \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$ as a reflexive, symmetric, non-transitive relation between H-types.
+It coincides with equality for complete H-types.
+Two incomplete H-types are consistent if they differ only up to the replacement of branches with holes.
+The type hole is  compatible with every type.
+This notion of consistency coincides with the notion of consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ type \cite{Siek06a}.
 
 Type ascription allows the user to explicitly state  a type for the ascribed expression to be analyzed against:
-\begin{equation}\label{rule:syn-asc}
-\inferrule{
-  \hana{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hexp : \htau}{\htau}
-}
-\end{equation}
-
 A variable synthesizes the hypothesized type:
-\begin{equation}\label{rule:syn-var}
-\inferrule{ }{
-  \hsyn{\hGamma, x : \htau}{x}{\htau}
-}
-\end{equation}
 
@@
 % \end{equation}
-\begin{equation}\label{rule:syn-ap}
-\inferrule{
-  \hsyn{\hGamma}{\hexp_1}{\htau}\\
-  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
-  \hana{\hGamma}{\hexp_2}{\htau_2}
-}{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
-}
-\end{equation}
 
 Numbers synthesize type $\tnum$:
-\begin{equation}\label{rule:syn-num}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hnum{n}}{\tnum}
-}
-\end{equation}
 
 Addition operates like a function over numbers:
-\begin{equation}\label{rule:syn-plus}
-\inferrule{
-  \hana{\hGamma}{\hexp_1}{\tnum}\\
-  \hana{\hGamma}{\hexp_2}{\tnum}
-}{
-  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
-}
-\end{equation}
 
@@
 The empty hole synthesizes the hole type:
-\begin{equation}\label{rule:syn-ehole}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hehole}{\tehole}
-}
-\end{equation}
-
 A non-empty hole contains an H-expression that is ``under construction'', as described in Sec.
@@
 The  expression inside the hole must synthesize some type, but, like the empty hole, the non-empty hole synthesizes only the hole type:
-\begin{equation}\label{rule:syn-hole}
-\inferrule{
-  \hsyn{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
-}
-\end{equation}
 The type consistency judgement $\tcompat{\htau}{\htau'}$, which appeared as a premise in the subsumption rule, makes the hole type compatible with any other type:

### 2016-07-05 formally.tex
finished 3.1 except for a few minor todos

@@
 % \end{subequations}
-\caption{H-type consistency and H-type matching at arrow.}
+\caption{H-type consistency and matched arrow types.}
 \label{fig:type-consistency}
@@
 Typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner.
+We identify the context up to exchange and adopt the usual identification convention for structures that differ only up to alpha-equivalence.
+
 Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, establish that $\hexp$ can appear where an expression of type $\htau$ is expected.
@@
 \begin{subequations}\label{rules:syn-ana}
-Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any other \emph{consistent} type, according to the following \emph{subsumption rule}:
-\begin{equation}\label{rule:ana-subsume}
-\inferrule{
-  \hsyn{\hGamma}{\hexp}{\htau'}\\
-  \tcompat{\htau}{\htau'}
-}{
-  \hana{\hGamma}{\hexp}{\htau}
-}
-\end{equation}
-
-Figure \ref{fig:type-consistency} defines the \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$ as a reflexive, symmetric, non-transitive relation between H-types.
+Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any other \emph{consistent} type, according to the \emph{subsumption rule}, Rule (\ref{rule:ana-subsume}).
+
+Figure \ref{fig:type-consistency} defines the \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$ as a reflexive and symmetric (but not transitive) relation between H-types.
 It coincides with equality for complete H-types.
-Two incomplete H-types are consistent if they differ only up to the replacement of branches with holes.
-The type hole is  compatible with every type.
-This notion of consistency coincides with the notion of consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ type \cite{Siek06a}.
-
-Type ascription allows the user to explicitly state  a type for the ascribed expression to be analyzed against:
-A variable synthesizes the hypothesized type:
-
-Lambda abstractions are not themselves annotated, so they can appear only in analytic position:
-\begin{equation}\label{rule:syn-lam}
-\inferrule{
-  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
+Two incomplete H-types are consistent if they differ only where a hole appears in either type.
+The type hole is therefore compatible with every type.
+This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ type \cite{Siek06a}.\todo{fwd reference related work}
+
+Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions, $\hlam{x}{\hexp}$.
+There is no type synthesis rule that applies to this form, so lambda abstractions can appear only in analytic position, i.e.
+where an expected type is known.\footnote{It is possible to also define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.} Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
+\begin{equation*}
+\inferrule{
   \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
 }{
-  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
-}
-\end{equation}
-It is possible to also define a ``half-annotated'' lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.
-
-When applying a function, if the expression in function position synthesizes an arrow type, the argument is analyzed against the synthesized argument type:
-% \begin{equation}\label{rule:syn-ap}
-% \inferrule{
-%   \hsyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}\\
-%   \hana{\hGamma}{\hexp_2}{\htau_2}
-% }{
-%   \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
-% }
-% \end{equation}
-
-Numbers synthesize type $\tnum$:
-
-Addition operates like a function over numbers:
-
-The rules given so far are sufficient to type complete H-expressions. The remaining rules give H-expressions with holes a well-defined static semantics.
-
-The empty hole synthesizes the hole type:
+  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
+}
+\end{equation*}
+The problem is that this standard rule alone leaves us unable to analyze lambda abstractions against the type hole, because the type hole is not immediately of the form $\tarr{\htau_1}{\htau_2}$.
+There are two possible solutions to this problem.
+One solution would be to define a second rule specifically for this case:
+\begin{equation*}
+\inferrule{
+  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
+}{
+  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
+}
+\end{equation*}
+Intead, we combine these two possible rules into a single rule through the simple auxiliary \emph{matched arrow type} judgement, $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure \ref{fig:type-consistency}.
+This judgement leaves arrow types alone and assigns the arrow type $\tarr{\tehole}{\tehole}$ to the type hole.
+It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement\todo{is admissible the right word}.
+Encouragingly, this judgement also arises in the study of gradual type systems\todo{first paper?}.
+
+Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of ascription form, $\hexp : \htau$.
+This allows the user to explicitly state  a type for the ascribed expression to be analyzed against.
+
+Rule (\ref{rule:syn-var}) is the standard rule for variables.
+
+Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function application.
+It again makes use of the matched function type judgement to combine what would otherwise need to be two rules for function application -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$ synthesizes the type hole.
+Indeed, Siek and Taha needed two application rules for the same fundamental reason \cite{Siek06a}.
+Subsequent work on gradual typing introduced this notion of type matching\todo{cite}.
+
+Rule (\ref{rule:syn-num}) states that number literals synthesize the $\tnum$ type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves like a function over numbers.
+
+The rules given so far are sufficient to type complete H-expressions. The two remaining rules give H-expressions with holes a well-defined static semantics.
+
+Rule (\ref{rule:syn-ehole}) states that the empty expression hole synthesizes the type hole.
+
 A non-empty hole contains an H-expression that is ``under construction'', as described in Sec.
 \ref{sec:example}.
-The  expression inside the hole must synthesize some type, but, like the empty hole, the non-empty hole synthesizes only the hole type:
-The type consistency judgement $\tcompat{\htau}{\htau'}$, which appeared as a premise in the subsumption rule, makes the hole type compatible with any other type:
-\begin{subequations}\label{rules:tcompat}
-\begin{equation}\label{rule:tcompat-hole}
-\inferrule{ }{
-  \tcompat{\htau}{\tehole}
-}
-\end{equation}
-\end{subequations}
-The remaining rules, given in the appendix, establish that type consistency is symmetric and reflexive (but not transitive.)
+According to Rule (\ref{rule:syn-hole}), the  expression inside the hole must synthesize some type, but, like the empty hole, the non-empty hole synthesizes only the hole type.
 
@@
 
-Taken we can derive that $\hana{id : \tarr{\tnum}{\tnum}}{\hhole{id}}{\tnum}$, as is necessary to synthesize a type for the H-expression on Line 14 of Fig.
-\ref{fig:first-example}.
+Given these rules, it is instructive to derive the following (by subsumption):  
+\[\hana{incr : \tarr{\tnum}{\tnum}}{\hhole{incr}}{\tnum}\]
+% This is the key premise necessary to synthesize the $\tnum$ type for the H-expression that results from cursor erasure -- defined next -- of the Z-expression on Line 16 of Fig.
+\ref{fig:second-example}.
 %In other words, this mechanism is essential if  users are to able to construct a program in anything but an ``outside in'' fashion.
@@
 
-\todo{redundant text w/ citation}
-The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}.
+% \todo{redundant text w/ citation}
+% The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}.
 Their system (which was not bidirectionally typed nor an editor model) also needed to define two rules for function application.
@@
 \begin{figure}[t]
-\hspace{-3px}$\arraycolsep=3pt\begin{array}{lllllll}
+\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
 \mathsf{ZTyp} & \ztau & ::= &

### 2016-07-05 formally.tex
tweaked colors
tweaks

@@
 \begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
-\item \textbf{H-expressions} and \textbf{H-types} (Sec.
-\ref{sec:holes}), i.e.
-expressions and types with \emph{holes}.
-H-types classify H-expressions according to Hazelnut's \textbf{bidirectionally typed static semantics}.
+\item \textbf{H-types} and \textbf{H-expressions} (Sec.
+\ref{sec:holes}), which are types and expressions with {holes}.
+H-types classify H-expressions according to Hazelnut's \textbf{bidirectional static semantics}.
 \item \textbf{Z-expressions} and \textbf{Z-types} (Sec.
@@
 \item \textbf{Actions} (Sec.
-\ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectionally typed action semantics}.
+\ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectional action semantics}.
 The action semantics enjoys a rich metatheory.
@@
 
-\subsection{Holes}\label{sec:holes}
+\subsection{H-types and H-expressions}\label{sec:holes}
 
@@
 
-Rule (\ref{rule:syn-num}) states that number literals synthesize the $\tnum$ type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves like a function over numbers.
-
-The rules given so far are sufficient to type complete H-expressions. The two remaining rules give H-expressions with holes a well-defined static semantics.
+Rule (\ref{rule:syn-num}) states that numbers synthesize the $\tnum$ type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves like a function over numbers.
+
+The rules described so far are sufficient to type complete H-expressions. The two remaining rules give H-expressions with holes a well-defined static semantics.
 
@@
 \ref{sec:example}.
-According to Rule (\ref{rule:syn-hole}), the  expression inside the hole must synthesize some type, but, like the empty hole, the non-empty hole synthesizes only the hole type.
+According to Rule (\ref{rule:syn-hole}), this inner  expression must synthesize some type, but, like the empty hole,  non-empty holes synthesize the type hole.
 
@@
 \end{subequations}
-\subsection{Focus Model}\label{sec:cursors}
+\subsection{Z-expressions and Z-types}\label{sec:cursors}
 \begin{figure}[t]

### 2016-07-05 formally.tex
sec 3.2 done

@@
 H-types classify H-expressions according to Hazelnut's \textbf{bidirectional static semantics}.
-\item \textbf{Z-expressions} and \textbf{Z-types} (Sec.
-\ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-expressions and H-types, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-expression (resp.
-Z-type) corresponds to an H-expression (resp.
-H-type) by \emph{cursor erasure}.
+\item \textbf{Z-types} and \textbf{Z-expressions} (Sec.
+\ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-types and H-expressions, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-type (resp.
+Z-expression) corresponds to an H-type (resp.
+H-expression) by \emph{cursor erasure}.
 \item \textbf{Actions} (Sec.
@@
 
-Figure \ref{fig:hexp-syntax} defines the syntax of H-types and H-expressions.
+Figure \ref{fig:hexp-syntax} defines the syntax of H-types, $\htau$, and H-expressions, $\hexp$.
 Most forms correspond directly to those of the simply typed lambda calculus (STLC) extended with a single base type, $\tnum$, of numbers (cf.
@@
 We do not need non-empty type holes because every H-type is a valid classifier of H-expressions.
-Types and expressions that do not contain holes are \emph{complete types} and \emph{complete expressions}, respectively.
+
+Types and expressions that contain no holes are \emph{complete types} and \emph{complete expressions}, respectively.
 Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete (see supplement.) We are not concerned here with defining a dynamic semantics for Hazelnut, but the dynamics for complete H-expressions would be entirely standard (in Sec.
@@
 
-The statics of Hazelnut is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,bidi-tutorial} around the two mutually defined judgements in Figure \ref{fig:ana-syn}.
+Hazelnut's static semantics is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,bidi-tutorial} around the two mutually defined judgements in Figure \ref{fig:ana-syn}.
 Typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner.
-We identify the context up to exchange and adopt the usual identification convention for structures that differ only up to alpha-equivalence.
+We identify the context up to exchange and adopt the standard identification convention for formal structures that differ only up to alpha-equivalence.
 
@@
 
-Figure \ref{fig:type-consistency} defines the \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$ as a reflexive and symmetric (but not transitive) relation between H-types.
-It coincides with equality for complete H-types.
-Two incomplete H-types are consistent if they differ only where a hole appears in either type.
+The \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$, that appears as a premise in the subsumption rule is a reflexive and symmetric (but not transitive) relation between H-types defined judgementally in Figure \ref{fig:type-consistency}.
+This relation coincides with equality for complete H-types.
+Two incomplete H-types are consistent if they differ only at positions where a hole appears in either type.
 The type hole is therefore compatible with every type.
-This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ type \cite{Siek06a}.\todo{fwd reference related work}
+This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ (i.e.
+unknown) type \cite{Siek06a}.\todo{fwd reference related work}
 
@@
 The problem is that this standard rule alone leaves us unable to analyze lambda abstractions against the type hole, because the type hole is not immediately of the form $\tarr{\htau_1}{\htau_2}$.
-There are two possible solutions to this problem.
+There are two plausible solutions to this problem.
 One solution would be to define a second rule specifically for this case:
@@
 Intead, we combine these two possible rules into a single rule through the simple auxiliary \emph{matched arrow type} judgement, $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure \ref{fig:type-consistency}.
-This judgement leaves arrow types alone and assigns the arrow type $\tarr{\tehole}{\tehole}$ to the type hole.
+This judgement leaves arrow types alone and assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$.
 It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement\todo{is admissible the right word}.
@@
 Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function application.
-It again makes use of the matched function type judgement to combine what would otherwise need to be two rules for function application -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$ synthesizes the type hole.
+It also makes use of the matched function type judgement to combine what would otherwise need to be two rules for function application -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$ synthesizes the type hole.
 Indeed, Siek and Taha needed two application rules for the same fundamental reason \cite{Siek06a}.
@@
 \end{subequations}
-\subsection{Z-expressions and Z-types}\label{sec:cursors}
+\subsection{Z-types and Z-expressions}\label{sec:cursors}
+\newcommand{\cvert}{{\hspace{1.8px}{\vert}\hspace{1.8px}}}
 \begin{figure}[t]
@@
   %\zlsel{\htau} ~\vert~
-  \zwsel{\htau} ~\vert~
+  \zwsel{\htau} \cvert
   %\zrsel{\htau} ~\vert~
-  \tarr{\ztau}{\htau} ~\vert~
+  \tarr{\ztau}{\htau} \cvert
   \tarr{\htau}{\ztau} \\
@@
   %\zlsel{\hexp} ~\vert~
-  \zwsel{\hexp} ~\vert~
+  \zwsel{\hexp} \cvert
   %\zrsel{\hexp} ~\vert~
-  \zexp : \htau ~\vert~
-  \hexp : \ztau ~\vert~
-  \hlam{x}{\zexp} ~\vert~
-  \hap{\zexp}{\hexp} ~\vert~
-  \hap{\hexp}{\zexp} ~\vert~
-  \hadd{\zexp}{\hexp} ~\vert~
-  \hadd{\hexp}{\zexp} ~\vert~
+  \zexp : \htau \cvert
+  \hexp : \ztau \cvert
+  \hlam{x}{\zexp} \cvert
+  \hap{\zexp}{\hexp} \cvert
+  \hap{\hexp}{\zexp} \cvert
+  \hadd{\zexp}{\hexp} \cvert
+  \hadd{\hexp}{\zexp} \cvert
   \hhole{\zexp}
@@
 %\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
-\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single focus.}
+\caption{Syntax of Z-types and Z-expressions.}
 \label{fig:zexp-syntax}
@@
 
-In order to identify a single subtree of an H-type or H-expression as the current focus of action, we apply Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.
-The syntax of Z-types, $\ztau$, and Z-expressions, $\zexp$, is given in Figure \ref{fig:zexp-syntax}.
-The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that is the current focus.
-All other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the focus will be found.
-All other sub-terms are H-types or H-expressions.
-Taken together, every syntactically well-formed Z-type and Z-expression contains exactly one focused H-type or H-expression.
-
-We write $\removeSel{\ztau}$ for the H-type constructed by removing the focus marker from the Z-type $\ztau$. This straightforward metafunction is defined as follows:
+Hazelnut's action semantics operates over Z-types, $\ztau$, and Z-expressions, $\zexp$.
+Figure \ref{fig:zexp-syntax} defines the syntax of Z-types and Z-expressions.
+A Z-type (resp.
+Z-expression) represents an H-type (resp.
+H-expression) with a single superimposed \emph{cursor}.
+
+The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on.
+All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the cursor will be found.
+Any other subterm is ``dotted'', i.e.
+it is either an H-type or H-expression.
+Taken together, every Z-type and Z-expression contains exactly one cursor by construction.
+This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)
+
+We write $\removeSel{\ztau}$ for the H-type constructed by erasing the cursor from $\ztau$, which we refer to as the \emph{cursor erasure} of $\ztau$.
+This straightforward metafunction is defined as follows:
 \begin{align*}
@@
 
-Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by removing the focus marker from the Z-expression $\zexp$.
-The definition of this metafunction is analagous, so we leave it in the appendix for concision.
+Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by erasing the cursor from the Z-expression $\zexp$, i.e.
+the \emph{cursor erasure} of $\zexp$.
+The definition of this metafunction is analogous, so we omit it for concision (see supplement.)
 % \begin{align*}
@@
 % \removeSel{(\hexp : \ztau)} & = \hexp : \removeSel{\ztau}\\
-% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\
+% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\ogoo
 % \removeSel{(\hap{\zexp}{\hexp})} & = \hap{\removeSel{\zexp}}{\hexp}\\

### 2016-07-06 formally.tex
progress on sec 3

@@
 \subsection{Z-types and Z-expressions}\label{sec:cursors}
-\newcommand{\cvert}{{\hspace{1.8px}{\vert}\hspace{1.8px}}}
+\newcommand{\cvert}{{\,{\vert}\,}}
 \begin{figure}[t]
@@
 The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on.
-All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the cursor will be found.
-Any other subterm is ``dotted'', i.e.
+All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the focus will be found.
+Any other sub-term is ``dotted'', i.e.
 it is either an H-type or H-expression.
-Taken together, every Z-type and Z-expression contains exactly one cursor by construction.
+Taken together, every Z-type and Z-expression contains exactly one focused H-type or H-expression by construction.
 This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)
@@
 Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by erasing the cursor from the Z-expression $\zexp$, i.e.
-the \emph{cursor erasure} of $\zexp$.
+the cursor erasure of $\zexp$.
 The definition of this metafunction is analogous, so we omit it for concision (see supplement.)
@@
 
-\subsection{Action Semantics}\label{sec:actions}
+\subsection{Actions}\label{sec:actions}
+
+We now arrive at the heart of Hazelnut: its \emph{bidirectional action semantics}.  
+Figure \ref{fig:action-syntax} defines the syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, and \emph{shapes}, $\psi$.
+
+Expression actions are governed by two mutually defined judgements, 1) the \emph{synthetic action judgement}:
+\[
+\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}
+\]
+and 2) \emph{the analytic action judgement}:
+\[
+\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
+\]
+
+In some Z-expressions, the cursor is in a type ascription, so we also need a \emph{type action judgement}, pronounced ``performing $\alpha$ on $\ztau$ results in $\ztau'$'':
+\[
+\performTyp{\ztau}{\alpha}{\ztau'}
+\]
+
+Before giving the rules for these judgements, let us state two key metatheorems.
+These metatheorems deeply inform the design of the rules.
+Their proofs, which are mechanized in Agda, are by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules, starting in Sec.
+\ref{sec:movement}.
+
+% The type action judgement appears as a premise in the rules that handle Z-expressions where the cursor is inside a . 
+% \[\begin{array}{ll}
+% %\textbf{Judgement Form} & \textbf{Description}\\
+% & 
+% \end{array}\]
+% \[
+% \begin{array}{l}
+% \fbox{$\performTyp{\ztau}{\alpha}{\ztau'}$} ~~ \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
+% \fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$} \\
+% \quad \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
+% \quad \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
+% \fbox{$\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$} \\
+% \quad \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
+% \quad \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
+% \quad \text{against $\htau$}
+% \end{array}\]
+\subsubsection{Sensibility}
+The expression action judgements obey the following \emph{action sensibility} theorem:
+\begin{theorem}[Action Sensibility]\label{thrm:actsafe} Both of the following hold:\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ 
+   then
+  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
+\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
+  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
+\end{enumerate}
+\end{theorem}
+\noindent In other words, if an edit state (i.e.
+a Z-expression) is statically meaningful, i.e.
+its cursor erasure is well-typed, then performing an action on it leaves the resulting edit state statically meaningful.
+In particular, the first clause of Theorem \ref{thrm:actsafe} establishes that when an action is performed on an edit state whose cursor erasure synthesizes an H-type, the result is an edit state whose cursor erasure also synthesizes some (possibly different) H-type.
+The second clause establishes that when an action is performed using the analytic action judgement on an edit state whose cursor erasure analyzes against some H-type, the result is a Z-expression whose cursor erasure also analyzes against the same H-type.
+% Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.
+
+The top-level sequence of actions in Figures \ref{fig:first-example} and \ref{fig:second-example} begin with an empty hole, so the derivation of each edit step must conclude with the synthetic action judgement.
+The analytic action judgement and the type action judgement appear as premises in the derivations of some of these steps.
+
 \begin{figure}[t]
@@
   %\aReplace{\hexp} ~\vert~
-  \aConstruct{\varphi} ~\vert~
+  \aConstruct{\psi} ~\vert~
   \aFinish\\
-\mathsf{Direction} & \delta & ::= &
+\mathsf{Dir} & \delta & ::= &
   \dChild ~\vert~
   \dParent ~\vert~
-  \dNext ~\vert~
-  \dPrev\\
-\mathsf{Shape} & \varphi & ::= &
+  \dNext \\ % ~\vert~
+  % \dPrev\\
+\mathsf{Shape} & \psi & ::= &
   \farr ~\vert~
@@
   \fnumlit{n} ~\vert~
-  \fplus
+  \fplus\\
+& & \vert & 
+  {\color{gray}\fnehole}
 \end{array}$
@@
 \label{fig:action-syntax}
-\vspace{-8px}
 \end{figure}
 
-The syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, or \emph{shapes}, $\varphi$, is given in Figure \ref{fig:action-syntax}.
-Actions are performed on Z-types and Z-expressions according to the \emph{action semantics} of Hazelnut, which is organized around three judgements:\todo{fix spacing}
-\[\arraycolsep=10pt\begin{array}{ll}
-%\textbf{Judgement Form} & \textbf{Description}\\
-\performTyp{\ztau}{\alpha}{\ztau'} & \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
-\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'} & \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
-& \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
-\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'} & \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
-& \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
-& \text{against $\htau$}
-\end{array}\]
-
-As suggested by the descriptions above, the action semantics maintains the following \emph{action sensibility} theorem:
-\begin{theorem}[Action Sensibility] Both of the following hold:
-\label{thrm:actsafe}
-\begin{enumerate}
-\item If $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
-  $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ then
-  $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
-\item If $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
-  $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ then
-  $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
-\end{enumerate}
-\end{theorem}
-In words, every action leaves the program in a semantically well-defined state.
-More specifically, the first clause of Theorem \ref{thrm:actsafe} establishes that actions performed on expressions that synthesize a type can only produce expressions that also synthesize some (possibly different) type.
-The second clause establishes that actions performed on expressions in analytic position (e.g.
-those under type ascriptions or in argument position, see above) can only produce expressions that can also be analyzed against the expected type.% Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.
-
-It is also useful to maintain a \emph{deterministic} action semantics, i.e. every well-defined action should produce a unique Z-type or Z-expression. Formally, this is stated as follows:
+\subsubsection{Determinism}
+Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
 \begin{theorem}[Action Determinism] All of the following hold:
 \label{thrm:actdet}
-\begin{enumerate}
+\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
 \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
@@
 
-In order to maintain determinism, we will need to supplement the definition of type compatibility above with a definition for \emph{type incompatibility}, $\tincompat{\htau}{\htau'}$.
-The key rule establishes that arrow types are incompatible with the $\tnum$ type:
-\begin{subequations}
+In order to maintain determinism, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$.
+One can define this notion either direrctly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement.
+The supplement does the latter.
+The key rule establishes that arrow types are inconsistent with $\tnum$:
+% \begin{subequations}
   % \begin{equation}
@@
   % \end{equation}
-  \begin{equation}
+  \begin{equation*}
     \inferrule{ }{
@@
     }
-  \end{equation}
+  \end{equation*}
   % \begin{equation}
@@
   % \end{equation}
-\end{subequations}
-The remaining rules, given in the appendix, establish that type incompatibility is symmetric and covariant.
-\subsubsection{Subsumption}
-
-The action semantics includes a subsumption rule much like the one from the underlying semantics of H-expressions:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
-    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
-    \tcompat{\htau}{\htau''}
-  }{
-    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
-  }
-\end{equation}
-In other words, if the expression synthesizes a type, then we defer to the synthetic action performance judgement, as long as it produces an expression that synthesizes a type compatible with the type provided for analysis.
-It is easy to see that this satisfies Theorem 1 by applying the IH and subsumption.
-
-\subsubsection{Relative Movement} Movement actions change the focus but do not change the underlying H-type or H-expression (so action sensibility is easy to show for these rules as well.)
-
-The rules for relative movement within Z-types are given below and should be self-explanatory:
+% \end{subequations}
+% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
+The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.\todo{big if true!!!}
+
+\subsubsection{Relative Movement}\label{sec:movement} 
+The rules below define relative movement within Z-types. They should be self-explanatory:
 \begin{subequations}
@@
 \end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \tarr{{\htau_1}}{\zwsel{\htau_2}}
-    }{
-      \aMove{\dPrev}
-    }{
-      {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
-    }
-  }
-\end{equation}
+% \begin{equation}
+%   \inferrule{ }{
+%     \performTyp{
+%       \tarr{{\htau_1}}{\zwsel{\htau_2}}
+%     }{
+%       \aMove{\dPrev}
+%     }{
+%       {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
+%     }
+%   }
+% \end{equation}
 % \begin{equation}
@@
 \end{subequations}
+Two more rules are needed to recurse into the zipper structure. We define these zipper rules in an action-independent manner in Sec. \ref{sec:zipper-cases}. 
 % The final two rules above recurse into the zipper structure.
 
-The rules for relative movement within Z-expressions are similar.
-Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:\todo{remove prevSib}
+The rules for relative movement within Z-expressions are similarly straightforward.
+Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:\todo{remove prevSib}
 \begin{subequations}
@@
 \end{subequations}
-For concision, we show only the rules for ascription here:
+% \begin{figure}
+% \judgbox{\performMove{\zexp}{\aMove{\delta}}{\zexp'}}{(Additional movement-action rules:)}
+% \begin{displaymath}
+% \begin{array}{@{}rcl}
+%   %\multicolumn{3}{l}{\textbf{Movement Actions for Sum Type Forms}:}
+%   %\\
+%   \TABperformMove
+%       {\zwsel{\hexp : \htau}}
+%       {\aMove{\dChild}}
+%       {      \zwsel{\hexp} : \htau}
+%   \\
+%   \TABperformMove
+%       {
+%       \zwsel{\hexp} : \htau
+%     }{
+%       \aMove{\dParent}
+%     }{
+%       \zwsel{\hexp : \htau}
+%     }
+%   \\
+%   \TABperformMove
+%     {
+%       \hexp : \zwsel{\htau}
+%     }{
+%       \aMove{\dParent}
+%     }{
+%       \zwsel{\hexp : \htau}
+%     }
+%   \\
+%   \TABperformMove
+%       {
+%       \zwsel{\hexp} : \htau
+%     }{
+%       \aMove{\dNext}
+%     }{
+%       \hexp : \zwsel{\htau}
+%     }
+%   \\[2mm]
+%   \TABperformMove
+%       {      {\tsum{{\htau_1}}{{\zwsel{\htau_2}}}}}
+%       {\aMove{\dParent}}
+%       {\zwsel{\tsum{{\htau_1}}{\htau_2}}}
+%   \\[2mm]
+%   \TABperformMove
+%       {\zwsel{\hinj{i}{\hexp}}}
+%       {\aMove{\dChild}}
+%       {\hinj{i}{\zwsel{\hexp}}}      
+%   \\
+%   \TABperformMove
+%       {\hinj{i}{\zwsel{\hexp}}}      
+%       {\aMove{\dParent}}
+%       {\zwsel{\hinj{i}{\hexp}}}
+%   \\[2mm]
+%   \TABperformMove
+%       {\zwsel{\hcase{\hexp}{x}{\hexp_1}{y}{\hexp_2}}}
+%       {\aMove{\dChild}}
+%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
+%   \\
+%   \TABperformMove
+%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
+%       {\aMove{\dNext}}
+%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
+%   \\
+%   \TABperformMove
+%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
+%       {\aMove{\dNext}}
+%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
+%   \\
+%   \TABperformMove
+%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
+%       {\aMove{\dPrev}}
+%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
+%   \\
+%   \TABperformMove
+%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
+%       {\aMove{\dPrev}}
+%       {      {\hcase{{\zwsel{\hexp}}}{x}{{\hexp_1}}{y}{{\hexp_2}}}}
+%   \\
+%   \TABperformMove
+%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
+%       {\aMove{\dParent}}
+%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
+%   \\
+%   \TABperformMove
+%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
+%       {\aMove{\dParent}}
+%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
+%   \\
+%   \TABperformMove
+%       {      {\hcase{{\hexp}}{x}{\hexp_1}{y}{\zwsel{\hexp_2}}}}
+%       {\aMove{\dParent}}
+%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
+% \end{array}
+% \end{displaymath}
+% \caption{Movement action semantics for sum type forms.}
+% \end{figure}
+The expression movement judgement is defined as follows.
+
+\paragraph{Ascription}
 \begin{subequations}
@@
 \end{equation}
-\begin{equation}
-  \label{r:moveprevsib}
-  \inferrule{ }{
-    \performTyp{
-      \hexp : \zwsel{\htau}
-    }{
-      \aMove{\dPrev}
-    }{
-      \zwsel{\hexp} : \htau
-    }
-  }
-\end{equation}
-\begin{equation}
-\inferrule{
-  \performTyp{
-    \zexp
-  }{
-    \aMove{\delta}
-  }{
-    \zexp'
-  }
-}{
-  \performTyp{
-    \zexp : \htau
-  }{
-    \aMove{\delta}
-  }{
-    \zexp' : \htau
-  }
-}
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \performTyp{
-      \ztau
-    }{
-      \aMove{\delta}
-    }{
-      \ztau'
-    }
-  }{
-    \performTyp{
-      \hexp : \ztau
-    }{
-      \aMove{\delta}
-    }{
-      \hexp : \ztau'
-    }
-  }
-\end{equation}
-and this for the example\todo{revise}
+% \begin{equation}
+%   \label{r:moveprevsib}
+%   \inferrule{ }{
+%     \performTyp{
+%       \hexp : \zwsel{\htau}
+%     }{
+%       \aMove{\dPrev}
+%     }{
+%       \zwsel{\hexp} : \htau
+%     }
+%   }
+% \end{equation}
+% \begin{equation}
+% \inferrule{
+%   \performTyp{
+%     \zexp
+%   }{
+%     \aMove{\delta}
+%   }{
+%     \zexp'
+%   }
+% }{
+%   \performTyp{
+%     \zexp : \htau
+%   }{
+%     \aMove{\delta}
+%   }{
+%     \zexp' : \htau
+%   }
+% }
+% \end{equation}
+% \begin{equation}
+%   \inferrule{
+%     \performTyp{
+%       \ztau
+%     }{
+%       \aMove{\delta}
+%     }{
+%       \ztau'
+%     }
+%   }{
+%     \performTyp{
+%       \hexp : \ztau
+%     }{
+%       \aMove{\delta}
+%     }{
+%       \hexp : \ztau'
+%     }
+%   }
+% \end{equation}
+
+\paragraph{Lambda}
 \begin{equation}\label{r:movefirstchild-lam}
@@
 \end{equation}
-
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \hlam{x}{\zwsel{\hexp}}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hlam{x}{\hexp}}
+    }
+  }
+\end{equation}
+\paragraph{Application}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \zwsel{\hap{\hexp_1}{\hexp_2}}
+    }{
+      \aMove{\dChild}
+    }{
+      \hap{\zwsel{\hexp_1}}{\hexp_2}
+    }
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \hap{\zwsel{\hexp_1}}{\hexp_2}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hap{\hexp_1}{\hexp_2}}
+    }
+  }
+\end{equation}
 \begin{equation}\label{r:moveparent-ap2}
@@
 \end{equation}
-
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \hap{\zwsel{\hexp_1}}{\hexp_2}
+    }{
+      \aMove{\dNext}
+    }{
+      \hap{\hexp_1}{\zwsel{\hexp_2}}
+    }
+  }
+\end{equation}
+% \begin{equation}
+%   \inferrule{ }{
+%     \performMove{
+%       \hap{\hexp_1}{\zwsel{\hexp_2}}
+%     }{
+%       \aMove{\dPrev}
+%     }{
+%       \hap{\zwsel{\hexp_1}}{\hexp_2}
+%     }
+%   }
+% \end{equation}
+
+\paragraph{Plus}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \zwsel{\hadd{\hexp_1}{\hexp_2}}
+    }{
+      \aMove{\dChild}
+    }{
+      \hadd{\zwsel{\hexp_1}}{\hexp_2}
+    }
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \hadd{\zwsel{\hexp_1}}{\hexp_2}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hadd{\hexp_1}{\hexp_2}}
+    }
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \hadd{{\hexp_1}}{\zwsel{\hexp_2}}
+    }{
+      \aMove{\dParent}
+    }{
+      \zwsel{\hadd{\hexp_1}{\hexp_2}}
+    }
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \hadd{\zwsel{\hexp_1}}{\hexp_2}
+    }{
+      \aMove{\dNext}
+    }{
+      \hadd{\hexp_1}{\zwsel{\hexp_2}}
+    }
+  }
+\end{equation}
+% \begin{equation}
+%   \inferrule{ }{
+%     \performMove{
+%       \hadd{\hexp_1}{\zwsel{\hexp_2}}
+%     }{
+%       \aMove{\dPrev}
+%     }{
+%       \hadd{\zwsel{\hexp_1}}{\hexp_2}
+%     }
+%   }
+% \end{equation}
+
+\paragraph{Non-Empty Hole}
+\begin{equation}
+\inferrule{ }{
+  \performMove{
+    \zwsel{\hhole{\hexp}}
+  }{
+    \aMove{\dChild}
+  }{
+    \hhole{\zwsel{\hexp}}
+  }
+}
+\end{equation}
 \begin{equation}\label{r:moveparent-hole}
@@
 \end{equation}
+Again, additional rules are needed to recurse into the zipper structure, but we will define these zipper rules in an action-independent manner in Sec. \ref{sec:zipper-cases}. 
 \end{subequations}
 
-\paragraph{Reachability}
-
-We give the following theorem to demonstrate that the rules for movement do
-indeed capture the process of moving the focus to any editable position
-within a term. Intuitively, given two zippered expressions that are the
-same up to the erasure of focus, its proof gives a list of composable
-actions that, when applied to the first term produces the second.
-
-As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
-every action in such a list, it is possible consider moving to an ill-typed
-term. Therefore the statement here matches the mutually recursive structure
-of the bidirectional typing judgements.
-
-\begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
-\label{thrm:reach}
-Both of the following hold:
-\begin{enumerate}
-  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-   $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-   = \zexp'$.
-
-  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-   = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-   ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-   = \zexp'$.
+The rules above are numerous and fairly uninteresting.
+That makes them quite hazardous -- the editor designer might make a mistake absent-mindedly.
+One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.\todo{make sure ian proved this }
+\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} \todo{make lemmas number with theorems} 
+\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+\item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then $\removeSel{\ztau}=\removeSel{\ztau'}$.
+\item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then $\removeSel{\zexp}=\removeSel{\zexp'}$.
 \end{enumerate}
-\end{theorem}
+\end{lemma}
+\noindent Lemma \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.
+
+A more thorough check would be to establish \emph{reachability}, i.e.
+that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within an H-expression.\todo{pending ian progress}
+
+% \paragraph{Reachability}
+
+% We give the following theorem to demonstrate that the rules for movement do
+% indeed capture the process of moving the focus to any editable position
+% within a term. Intuitively, given two zippered expressions that are the
+% same up to the erasure of focus, its proof gives a list of composable
+% actions that, when applied to the first term produces the second.
+
+% As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
+% every action in such a list, it is possible consider moving to an ill-typed
+% term. Therefore the statement here matches the mutually recursive structure
+% of the bidirectional typing judgements.
+
+% \begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
+% \label{thrm:reach}
+% Both of the following hold:
+% \begin{enumerate}
+%   \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+%    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
+%    = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
+%    ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
+%    = \zexp'$.
+
+%   \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+%    $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
+%    = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
+%    ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
+%    = \zexp'$.
+% \end{enumerate}
+% \end{theorem}
 
@@
 
-\subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.
+\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ into the program at or around the focus.
 
@@
 %   \inferrule{
-%     \performTyp{\ztau}{\aConstruct{\varphi}}{\ztau'}
+%     \performTyp{\ztau}{\aConstruct{\psi}}{\ztau'}
 %   }{
@@
 %     }{
-%       \aConstruct{\varphi}
+%       \aConstruct{\psi}
 %     }{
@@
 
-\paragraph{Finishability}\todo{do we care?}
-As with deletion, since the meaning of the finish action does not depend on
-the structure of the term being finished other than requiring that it must
-be one of the two hole forms, there is no checksum theorem to state for
-this action.
-
-\subsubsection{Zipper Cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression.
-We also need to define the recursive cases, which propagate the action into the subtree where the focus appears.
-These rules follow the structure of the corresponding rules in the statics of H-expressions.
-
-\begin{subequations}
-For example, when the focus is in the expression position of an ascription, we use the analytic action performance judgement:
-\begin{equation}
-\inferrule{
-  \performAna
-    {\hGamma}
-    {\zexp}
-    {\htau}
-    {\alpha}
-    {\zexp'}
-}{
-  \performSyn
-    {\hGamma}
-    {\zexp : \htau}
-    {\htau}
-    {\alpha}
-    {\zexp' : \htau}
-    {\htau}
-}
-\end{equation}
-
-When the focus is in the type position of an ascription, we must re-check the ascribed expression because the type might have changed (in practice, one would optimize this check to only occur if the type actually was changed):
-\begin{equation}
-\inferrule{
-  \performTyp{\ztau}{\alpha}{\ztau'}\\
-  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
-}{
-  \performSyn
-    {\hGamma}
-    {\hexp : \ztau}
-    {\removeSel{\ztau}}
-    {\alpha}
-    {\hexp : \ztau'}
-    {\removeSel{\ztau'}}
-}
-\end{equation}
-
-If the focus is in the body of a lambda expression, then we must use the analytic action performance rule:
-\begin{equation}
-\inferrule{
-  \performAna
-    {\hGamma, x : \htau_1}
-    {\zexp}
-    {\htau_2}
-    {\alpha}
-    {\zexp'}
-}{
-  \performAna
-    {\hGamma}
-    {\hlam{x}{\zexp}}
-    {\tarr{\htau_1}{\htau_2}}
-    {\alpha}
-    {\hlam{x}{\zexp'}}
-}
-\end{equation}
-
-There are two rules that handle the case where the focus is in the function position of an application, corresponding to the two application rules in the statics.
-Each involves rechecking the argument against the new function type:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau_2}
-      {\alpha}
-      {\zexp'}
-      {\tarr{\htau_3}{\htau_4}}\\
-    \hana{\hGamma}{\hexp}{\htau_3}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\zexp}{\hexp}}
-      {\htau_1}
-      {\alpha}
-      {\hap{\zexp'}{\hexp}}
-      {\htau_4}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau_2}
-      {\alpha}
-      {\zexp'}
-      {\tehole}\\
-    \hana{\hGamma}{\hexp}{\tehole}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\zexp}{\hexp}}
-      {\htau_1}
-      {\alpha}
-      {\hap{\zexp'}{\hexp}}
-      {\tehole}
-  }
-\end{equation}
-
-Similarly, there are two rules that handle the case where the focus is in the argument position:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\hexp}{\tarr{\htau_2}{\htau}}\\
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\htau_2}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\hexp}{\zexp}}
-      {\htau}
-      {\alpha}
-      {\hap{\hexp}{\zexp'}}
-      {\htau}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\hexp}{\tehole}\\
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\tehole}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\hexp}{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\hap{\hexp}{\zexp'}}
-      {\tehole}
-  }
-\end{equation}
-
-The rules for the addition operator follow from the statics directly:
-\begin{equation}
-  \inferrule{
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\tnum}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hadd{\zexp}{\hexp}}
-      {\tnum}
-      {\alpha}
-      {\hadd{\zexp'}{\hexp}}
-      {\tnum}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\tnum}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hadd{\hexp}{\zexp}}
-      {\tnum}
-      {\alpha}
-      {\hadd{\hexp}{\zexp'}}
-      {\tnum}
-  }
-\end{equation}
-
-Finally, if the focus is inside a non-empty hole, we special case the situation where the action results in a doubly-nested empty hole, $\hhole{\hehole}$, to eliminate the nesting (given our current action semantics, only the delete action can cause this form to arise and the form $\hhole{\hhole{\zexp}}$ cannot arise):
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau}
-      {\alpha}
-      {\zexp'}
-      {\htau'}\\
-    \zexp' \neq \zwsel{\hehole}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hhole{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\hhole{\zexp'}}
-      {\tehole}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau}
-      {\alpha}
-      {\zwsel{\hehole}}
-      {\tehole}\\
-  }{
-    \performSyn
-      {\hGamma}
-      {\hhole{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\zwsel{\hehole}}
-      {\tehole}
-  }
-\end{equation}
-
 \paragraph{Constructibility}\todo{why is this with the zipper rules not the
@@
 We give the following theorem to demonstrate that the rules for the action
-$\aConstruct{\varphi}$ do indeed capture the process of building any term
-of the form $\varphi$. Intuitively, its proof gives a list of composable
+$\aConstruct{\psi}$ do indeed capture the process of building any term
+of the form $\psi$. Intuitively, its proof gives a list of composable
 actions that, when applied one sequentially to an empty hole in focus,
@@
 particular.
+
+\paragraph{Finishability}\todo{do we care?}
+As with deletion, since the meaning of the finish action does not depend on
+the structure of the term being finished other than requiring that it must
+be one of the two hole forms, there is no checksum theorem to state for
+this action.
+
+\subsubsection{Zipper Cases}\label{sec:zipper-cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression.
+We also need to define the recursive cases, which propagate the action into the subtree where the focus appears.
+These rules follow the structure of the corresponding rules in the statics of H-expressions.
+
+\begin{subequations}
+For example, when the focus is in the expression position of an ascription, we use the analytic action performance judgement:
+\begin{equation}
+\inferrule{
+  \performAna
+    {\hGamma}
+    {\zexp}
+    {\htau}
+    {\alpha}
+    {\zexp'}
+}{
+  \performSyn
+    {\hGamma}
+    {\zexp : \htau}
+    {\htau}
+    {\alpha}
+    {\zexp' : \htau}
+    {\htau}
+}
+\end{equation}
+
+When the focus is in the type position of an ascription, we must re-check the ascribed expression because the type might have changed (in practice, one would optimize this check to only occur if the type actually was changed):
+\begin{equation}
+\inferrule{
+  \performTyp{\ztau}{\alpha}{\ztau'}\\
+  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
+}{
+  \performSyn
+    {\hGamma}
+    {\hexp : \ztau}
+    {\removeSel{\ztau}}
+    {\alpha}
+    {\hexp : \ztau'}
+    {\removeSel{\ztau'}}
+}
+\end{equation}
+
+If the focus is in the body of a lambda expression, then we must use the analytic action performance rule:
+\begin{equation}
+\inferrule{
+  \performAna
+    {\hGamma, x : \htau_1}
+    {\zexp}
+    {\htau_2}
+    {\alpha}
+    {\zexp'}
+}{
+  \performAna
+    {\hGamma}
+    {\hlam{x}{\zexp}}
+    {\tarr{\htau_1}{\htau_2}}
+    {\alpha}
+    {\hlam{x}{\zexp'}}
+}
+\end{equation}
+
+There are two rules that handle the case where the focus is in the function position of an application, corresponding to the two application rules in the statics.
+Each involves rechecking the argument against the new function type:
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
+    \performSyn
+      {\hGamma}
+      {\zexp}
+      {\htau_2}
+      {\alpha}
+      {\zexp'}
+      {\tarr{\htau_3}{\htau_4}}\\
+    \hana{\hGamma}{\hexp}{\htau_3}
+  }{
+    \performSyn
+      {\hGamma}
+      {\hap{\zexp}{\hexp}}
+      {\htau_1}
+      {\alpha}
+      {\hap{\zexp'}{\hexp}}
+      {\htau_4}
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
+    \performSyn
+      {\hGamma}
+      {\zexp}
+      {\htau_2}
+      {\alpha}
+      {\zexp'}
+      {\tehole}\\
+    \hana{\hGamma}{\hexp}{\tehole}
+  }{
+    \performSyn
+      {\hGamma}
+      {\hap{\zexp}{\hexp}}
+      {\htau_1}
+      {\alpha}
+      {\hap{\zexp'}{\hexp}}
+      {\tehole}
+  }
+\end{equation}
+
+Similarly, there are two rules that handle the case where the focus is in the argument position:
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\hexp}{\tarr{\htau_2}{\htau}}\\
+    \performAna
+      {\hGamma}
+      {\zexp}
+      {\htau_2}
+      {\alpha}
+      {\zexp'}
+  }{
+    \performSyn
+      {\hGamma}
+      {\hap{\hexp}{\zexp}}
+      {\htau}
+      {\alpha}
+      {\hap{\hexp}{\zexp'}}
+      {\htau}
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\hexp}{\tehole}\\
+    \performAna
+      {\hGamma}
+      {\zexp}
+      {\tehole}
+      {\alpha}
+      {\zexp'}
+  }{
+    \performSyn
+      {\hGamma}
+      {\hap{\hexp}{\zexp}}
+      {\tehole}
+      {\alpha}
+      {\hap{\hexp}{\zexp'}}
+      {\tehole}
+  }
+\end{equation}
+
+The rules for the addition operator follow from the statics directly:
+\begin{equation}
+  \inferrule{
+    \performAna
+      {\hGamma}
+      {\zexp}
+      {\tnum}
+      {\alpha}
+      {\zexp'}
+  }{
+    \performSyn
+      {\hGamma}
+      {\hadd{\zexp}{\hexp}}
+      {\tnum}
+      {\alpha}
+      {\hadd{\zexp'}{\hexp}}
+      {\tnum}
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{
+    \performAna
+      {\hGamma}
+      {\zexp}
+      {\tnum}
+      {\alpha}
+      {\zexp'}
+  }{
+    \performSyn
+      {\hGamma}
+      {\hadd{\hexp}{\zexp}}
+      {\tnum}
+      {\alpha}
+      {\hadd{\hexp}{\zexp'}}
+      {\tnum}
+  }
+\end{equation}
+
+Finally, if the focus is inside a non-empty hole, we special case the situation where the action results in a doubly-nested empty hole, $\hhole{\hehole}$, to eliminate the nesting (given our current action semantics, only the delete action can cause this form to arise and the form $\hhole{\hhole{\zexp}}$ cannot arise):
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
+    \performSyn
+      {\hGamma}
+      {\zexp}
+      {\htau}
+      {\alpha}
+      {\zexp'}
+      {\htau'}\\
+    \zexp' \neq \zwsel{\hehole}
+  }{
+    \performSyn
+      {\hGamma}
+      {\hhole{\zexp}}
+      {\tehole}
+      {\alpha}
+      {\hhole{\zexp'}}
+      {\tehole}
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
+    \performSyn
+      {\hGamma}
+      {\zexp}
+      {\htau}
+      {\alpha}
+      {\zwsel{\hehole}}
+      {\tehole}\\
+  }{
+    \performSyn
+      {\hGamma}
+      {\hhole{\zexp}}
+      {\tehole}
+      {\alpha}
+      {\zwsel{\hehole}}
+      {\tehole}
+  }
+\end{equation}
+
+\subsubsection{Action Subsumption}\label{sec:action-subsumption}
+
+The bidirectional action semantics includes a subsumption rule similar to the subsumption rule of the bidirectional static semantics:
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
+    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
+    \tcompat{\htau}{\htau''}\\\\
+    \alpha \neq \aConstruct{\fasc}\\
+    \alpha \neq \aConstruct{\flam{x}}
+  }{
+    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
+  }
+\end{equation}
+In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action performance judgement, as long as it produces an expression that synthesizes a type compatible with the type provided for analysis.
+It is easy to see that this satisfies Theorem 1 by applying the IH and the typing subsumption rule.
+
+We need to specify two exceptions in this rule, because the analytic action semantics handles these cases explicitly, in a manner that does not coincide with how the synthetic action semantics would handle these cases if allowed to do so via action subsumption.
+We will see why this is the case in Sec.
+\ref{sec:construction}.
+Algorithmically, this captures the notion that subsumption should be the case of last resort when pattern matching sequentially.
+
 \end{subequations}

### 2016-07-06 hazelnut-popl17.tex
reformatted sum type extension as two figures: the main figure, and movement
finished 3.1 except for a few minor todos
tweaked colors
tweaks
sec 3.2 done
writing about hazelnut

@@
-\def\OPTIONConf{0}%
+\def\OPTIONConf{1}%
 %\documentclass[12pt]{article}
@@
 \definecolor{light-gray}{gray}{0.9}
+\usepackage{soul}
+\setulcolor{red}
 \usepackage{mathpazo}
@@
 \titlespacing*{\paragraph}{0pt}{4pt}{2pt}
-\setlength{\skip\footins}{3pt}
+\setlength{\skip\footins}{3pt plus 1px minus 5px}
 
@@
 
-Hazelnut is a foundational type-theoretic account of typed structure editing, rather than an end-user tool itself.
+\todo{rephrase}Hazelnut is a foundational type-theoretic account of typed structure editing, rather than an end-user tool itself.
 To that end, we describe how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to the calculus.
@@
 \section{Extending Hazelnut}\label{sec:extending}
-\todo{write this section} Add sum types.
-
-H-expression
-
-Z-expression + Erasure rules
-
-Statics
-
-Movement actions
-
-Construct actions
-
+%% \todo{write this section} Add sum types.
+%% H-expression
+%% Z-expression + Erasure rules
+%% Statics
+%% Movement actions
+%% Construct actions
 \input{sumtypes}
@@
 %\subsection{Structure Editors}
+
+\paragraph{Gradual typing: Has Matched Relation}
+%
+A. Rastogi, A. Chaudhuri, and B. Hosmer. The ins and outs of gradual type inference.
+\\
+Definition 3.1 is similar to our ``has matched'' relations:
+\\
+\url{https://www.cs.umd.edu/~avik/papers/iogti.pdf}
+
+This paper defines ``has matched'' relation for arrow, just like us:
+\\
+\url{http://cimini.info/publications/Gradualizer_Draft.pdf}
+\\
+They credit this technique to the following paper:
+\\
+\url{http://www.cs.ubc.ca/~rxg/ptsgp.pdf}
+\\
+But, the Rastogi paper came earlier.
 

### 2016-07-06 formally.tex

@@
 The rules above are numerous and fairly uninteresting.
-That makes them quite hazardous -- the editor designer might make a mistake absent-mindedly.
-One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.\todo{make sure ian proved this }
+That makes them quite hazardous -- we might make a mistake absent-mindedly.
+One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.
 \begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} \todo{make lemmas number with theorems} 
@@
 
-A more thorough check would be to establish \emph{reachability}, i.e.
-that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within an H-expression.\todo{pending ian progress}
+Another check is to establish \emph{reachability}, i.e.
+that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within a well-typed H-expression.
+
+This requires developing machinery for sequences of actions.
+There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define auxiliary syntax for sequences of actions, $\bar{\alpha}$, and ``multi\-step'' action judgements.
+To keep the core of the action semantics small, we took the latter approach in Figure \ref{fig:multistep}.
+
+\begin{figure}
+$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\\
+\fbox{$\performTyp{\ztau}{\bar{\alpha}}{\ztau'}$}
+\begin{mathpar}
+\inferrule{ }{
+    \performTyp{\ztau}{\cdot}{\ztau}
+}
+
+\inferrule{
+  \performTyp{\ztau}{\alpha}{\ztau'}\\
+  \performTyp{\ztau'}{\bar{\alpha}}{\ztau''}
+}{
+  \performTyp{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
+}
+\end{mathpar}
+\fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
+\begin{mathpar}
+\inferrule{ }{
+  \performSyn{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
+}
+
+\inferrule{
+  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
+  \performSyn{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
+}{
+  \performSyn{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
+}
+\end{mathpar}
+\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
+\begin{mathpar}
+\inferrule{ }{
+  \performAna{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
+}
+
+\inferrule{
+  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
+  \performAna{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
+}{
+  \performAna{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
+}
+\end{mathpar}
+\caption{Multistep Action Performance}
+\label{fig:multistep}
+\end{figure}
+
+With these definitions, we can state reachability as follows:
+
+\begin{theorem}[Reachability] ~
+\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+\item If $\removeSel{\ztau}=\removeSel{\ztau'}$ then there exists some $\bar\alpha$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performTyp{\ztau}{\bar\alpha}{\ztau'}$.
+\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$.
+\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zexp'}{\htau}$.
+\end{enumerate}
+\end{theorem}
+
+To prove Reachability, we break it into two lemmas. One...\todo{write this}
 
@@
 
-\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.
-
-Again, the rule for Z-types is self-explanatory:
+\subsubsection{Deletion} The $\aDel$ action replaces the cursor with an empty hole.
+
+The type action rule is self-explanatory:
 \begin{subequations}
@@
 
-\paragraph{Deletability}\todo{do we care?}
-The meaning of the deletion action is to remove the entire term in focus,
-and its definition does not depend on the structure of the term being
-deleted. Therefore, there is no theorem to state that acts as a checksum on
-the rules analogous to Theorem \ref{thrm:reach} or \ref{thrm:construct},
-since we could not have forgotten any rules corresponding to the structure
-of the terms.
-
-\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ into the program at or around the focus.
+\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ into the program at or around the cursor.
 

### 2016-07-06 formally.tex

@@
 One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.
-\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} \todo{make lemmas number with theorems} 
+\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure}
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
@@
 
+A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.\todo{define this in supplement}
+
 With these definitions, we can state reachability as follows:
 
-\begin{theorem}[Reachability] ~
+\begin{theorem}[Reachability]\label{thrm:reachability} ~
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
@@
 
-To prove Reachability, we break it into two lemmas. One...\todo{write this}
+The simplest way to prove Theorem \ref{thrm:reachability} is to  break it into two lemmas.
+Lemma \ref{lemma:reach-up} establishes that you can always move the cursor to the outermost position in an expression.
+This serves as a check on our $\aMove{\dParent}$ rules.
+\begin{lemma}[Reach Up]\label{lemma:reach-up} ~
+\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
+\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$.
+\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
+\end{enumerate}
+\end{lemma}
+Lemma \ref{lemma:reach-down} establishes that you can always move the cursor from the outermost position to any other position.
+This serves as a check on our $\aMove{\dChild}$ and $\aMove{\dNext}$ rules.
+\begin{lemma}[Reach Down]\label{lemma:reach-down} ~
+\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
+\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
+\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ nand $\performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
+\end{enumerate}
+\end{lemma}
+Theorem \ref{thrm:reachability} follows by straightforward composition of these two lemmas. (The witness to the existential is not, of course, the shortest path between any two cursor locations.)
 

### 2016-07-06 formally.tex
progress on construction and subsumption

@@
 Their proofs, which are mechanized in Agda, are by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules, starting in Sec.
-\ref{sec:movement}.
+\ref{sec:action-subsumption}.
 
@@
 % The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
-The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.\todo{big if true!!!}
+The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.
+
+\subsubsection{Action Subsumption}\label{sec:action-subsumption}
+
+The bidirectional action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
+    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
+    \tcompat{\htau}{\htau''}\\\\
+    \alpha \notin \mathcal{A}
+  }{
+    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
+  }
+\end{equation}
+In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.
+It is easy to see that this satisfies Theorem 1 by applying the IH and the Rule (\ref{rule:ana-subsume}).
+
+We need to make some exceptions in this rule to maintain determinism, because the analytic action semantics handles certain actions explicitly, in a manner that does not coincide with how the synthetic action semantics would handle these cases if allowed to do so via action subsumption.
+We will see why this is the case in Sec.
+\ref{sec:construction}.
+We do so by defining a finite set of actions $\mathcal{A}$ that do not go through subsumption:
+\[
+\mathcal{A} \overset{\tiny\textit{def}}{\Delta} \{\aConstruct{\fasc}, \aConstruct{\flam{x}}\}
+\]
+
+Algorithmically, this captures the notion that subsumption should be the case of last resort.
 
@@
 The rules for relative movement within Z-expressions are similarly straightforward.
-Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:\todo{remove prevSib}
+Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:\todo{remove prevSib in supplement too}
 \begin{subequations}
@@
 One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.
-\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure}
+
+\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} ~
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px] 
@@
 
-Another check is to establish \emph{reachability}, i.e.
+Another useful check is to establish \emph{reachability}, i.e.
 that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within a well-typed H-expression.
 
-This requires developing machinery for sequences of actions.
-There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define auxiliary syntax for sequences of actions, $\bar{\alpha}$, and ``multi\-step'' action judgements.
-To keep the core of the action semantics small, we took the latter approach in Figure \ref{fig:multistep}.
+This requires developing machinery for reasoning about sequences of actions.
+There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define a syntax for lists of actions, $\bar{\alpha}$, together with iterated action judgements.
+To keep the core of the action semantics small, we take the latter approach in Figure \ref{fig:multistep}.\todo{put stars on these?}
 
 \begin{figure}
-$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\\
+$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\vspace{4px}\\
 \fbox{$\performTyp{\ztau}{\bar{\alpha}}{\ztau'}$}
-\begin{mathpar}
+\vspace{-10px}\begin{mathpar}
 \inferrule{ }{
@@
 \fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
-\begin{mathpar}
+\vspace{-10px}\begin{mathpar}
 \inferrule{ }{
@@
 \fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
-\begin{mathpar}
+\vspace{-10px}\begin{mathpar}
 \inferrule{ }{
@@
 \end{mathpar}
-\caption{Multistep Action Performance}
+\caption{Iterated Action Judgements}
 \label{fig:multistep}
@@
 
-\subsubsection{Deletion} The $\aDel$ action replaces the cursor with an empty hole.
-
-The type action rule is self-explanatory:
-\begin{subequations}
+\subsubsection{Deletion} The $\aDel$ action inserts an empty hole at the cursor, deleting what was there before.
+
+The type action rule for $\aDel$ is self-explanatory:
 \begin{equation}
@@
 % \end{equation}
-\end{subequations}
 
@@
 
-\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ into the program at or around the cursor.
-
-Again, let us begin with type actions. The $\aConstruct{\farr}$ action constructs an arrow type. The focused H-type becomes the argument type, and the focus is placed on an empty return type hole:
+\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ at the cursor.
+
+Again, let us begin with type actions.
+The $\aConstruct{\farr}$ action constructs an arrow type.
+The H-type under the cursor becomes the argument type, and the cursor is placed on an empty return type hole:
 \begin{subequations}
@@
 
-The $\aConstruct{\fnum}$ action replaces an empty Z-type hole with the $\tnum$ type:
+The $\aConstruct{\fnum}$ action replaces an empty type hole under the cursor with the $\tnum$ type:
   \begin{equation}
@@
 
-Moving on to expression actions, we start to see more interesting rules.
-The $\aConstruct{\fasc}$ action operates differently depending on whether the focused expression synthesizes a type or is being analyzed against a type.
-In the first case, the ascribed type is the synthesized type:
+Moving on to expression actions, we start to see some more interesting rules.
+The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type.In the first case, the ascribed type is the synthesized type:
 \begin{equation}
@@
 \end{equation}
-In the second case, the ascribed type is the type provided for analysis:
+In the second case, the ascribed type is the provided type:
 \begin{equation}
@@
 \end{equation}
-
-The $\aConstruct{\fvar{x}}$ action places the variable $x$ into the focused empty hole.
-If that hole is being asked to synthesize a type, then the result of the action synthesizes the type assigned to $x$ in the context:
+This rule should supercede the subsumption rule, so we must list $\aConstruct{\fasc}$ in \todo{do not subsume set}.
+
+The $\aConstruct{\fvar{x}}$ action places the variable $x$ into an empty hole. If that hole is being asked to synthesize a type, then the result synthesizes the hypothesized type:
 \begin{equation}
@@
 \end{equation}
-If the focused empty hole is being analyzed against a type that is inconsistent with the type assigned to $x$ by the context, $x$ is placed inside a hole:
+If the hole is being analyzed against a type that is consistent with the hypothesized type, then the action semantics goes through the {action subsumption rule} described in Sec.
+\ref{sec:action-subsumption}.
+If the hole is being analyzed against a type that is inconsistent with the hypothesized type, $x$ is placed inside a hole:
 \begin{equation}
@@
 \end{equation}
-The rule above featured in the example in Section \ref{sec:example}.
-
-Notice that no rule was necessary for the case where the hole was being analyzed against a type compatible with the variable's type, because this case is handled by the action subsumption rule.
+The rule above featured on Line 15 of Figure \ref{fig:second-example}.
 
@@
 
-\subsubsection{Action Subsumption}\label{sec:action-subsumption}
-
-The bidirectional action semantics includes a subsumption rule similar to the subsumption rule of the bidirectional static semantics:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
-    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
-    \tcompat{\htau}{\htau''}\\\\
-    \alpha \neq \aConstruct{\fasc}\\
-    \alpha \neq \aConstruct{\flam{x}}
-  }{
-    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
-  }
-\end{equation}
-In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action performance judgement, as long as it produces an expression that synthesizes a type compatible with the type provided for analysis.
-It is easy to see that this satisfies Theorem 1 by applying the IH and the typing subsumption rule.
-
-We need to specify two exceptions in this rule, because the analytic action semantics handles these cases explicitly, in a manner that does not coincide with how the synthetic action semantics would handle these cases if allowed to do so via action subsumption.
-We will see why this is the case in Sec.
-\ref{sec:construction}.
-Algorithmically, this captures the notion that subsumption should be the case of last resort when pattern matching sequentially.
-
 \end{subequations}

### 2016-07-06 formally.tex
fixed def=
tweaked text in 3.3.3

@@
 \ref{sec:construction}.
-We do so by defining a finite set of actions $\mathcal{A}$ that do not go through subsumption:
+We do so by defining a finite set of actions $\mathcal{A}$ that do not need to go through subsumption:
 \[
-\mathcal{A} \overset{\tiny\textit{def}}{\Delta} \{\aConstruct{\fasc}, \aConstruct{\flam{x}}\}
+\mathcal{A} \overset{\tiny\textit{def}}{=} \{\aConstruct{\fasc}, \aConstruct{\flam{x}}\}
 \]
-
 Algorithmically, this captures the notion that subsumption should be the case of last resort.

### 2016-07-06 formally.tex
fixed subsumption

--- 2016-07-06 formally.tex
@@
 The bidirectional action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
-\begin{equation}
+\begin{equation}\label{rule:action-subsume}
   \inferrule{
@@
     \tcompat{\htau}{\htau''}\\\\
-    \alpha \notin \mathcal{A}
+    \alpha~\mathsf{subsumes}
   }{
@@
 In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.
-It is easy to see that this satisfies Theorem 1 by applying the IH and the Rule (\ref{rule:ana-subsume}).
-
-We need to make some exceptions in this rule to maintain determinism, because the analytic action semantics handles certain actions explicitly, in a manner that does not coincide with how the synthetic action semantics would handle these cases if allowed to do so via action subsumption.
-We will see why this is the case in Sec.
-\ref{sec:construction}.
-We do so by defining a finite set of actions $\mathcal{A}$ that do not go through subsumption:
-\[
-\mathcal{A} \overset{\tiny\textit{def}}{\Delta} \{\aConstruct{\fasc}, \aConstruct{\flam{x}}\}
-\]
-
+
+We need to make some exceptions in this rule to maintain determinism, because certain construction actions, described in Sec.
+\ref{sec:construction}, behave slightly differently depending on whether a type is known.
+We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
+\begin{mathpar}
+\inferrule{
+  \alpha \neq \aConstruct{\fasc}\\
+  (\forall x) \alpha \neq \aConstruct{\flam{x}} 
+}{
+  \alpha~\mathsf{subsumes}
+}
+\end{mathpar}
 Algorithmically, this captures the notion that subsumption should be the case of last resort.
+
+The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e. Rule (\ref{rule:ana-subsume}).
 
@@
 Moving on to expression actions, we start to see some more interesting rules.
-The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type.In the first case, the ascribed type is the synthesized type:
+The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type.
+In the first case, the ascribed type is the synthesized type:
 \begin{equation}
@@
 \end{equation}
-This rule should supercede the subsumption rule, so we must list $\aConstruct{\fasc}$ in \todo{do not subsume set}.
+This rule should supercede the subsumption rule, so we must list $\aConstruct{\fasc}$ in $\mathcal{A}$.
 

### 2016-07-06 formally.tex

@@
 
-The top-level sequence of actions in Figures \ref{fig:first-example} and \ref{fig:second-example} begin with an empty hole, so the derivation of each edit step must conclude with the synthetic action judgement.
-The analytic action judgement and the type action judgement appear as premises in the derivations of some of these steps.
-
 \begin{figure}[t]
@@
 In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.
+The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e.
+Rule (\ref{rule:ana-subsume}).
 
@@
 We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
-\begin{mathpar}
+\begin{equation}\label{rule:subsumes}
 \inferrule{
@@
 }
-\end{mathpar}
+\end{equation}
 Algorithmically, this captures the notion that subsumption should be the case of last resort.
-
-The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e. Rule (\ref{rule:ana-subsume}).
 
@@
 
-\paragraph{Lambda}
+\paragraph{Lambda}\vspace{-3px}
 \begin{equation}\label{r:movefirstchild-lam}
@@
 \end{equation}
-\paragraph{Application}
+\paragraph{Application}\vspace{-5px}
 \begin{equation}
@@
 
-\begin{figure}
-$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\vspace{4px}\\
-\fbox{$\performTyp{\ztau}{\bar{\alpha}}{\ztau'}$}
-\vspace{-10px}\begin{mathpar}
-\inferrule{ }{
-    \performTyp{\ztau}{\cdot}{\ztau}
-}
-
-\inferrule{
-  \performTyp{\ztau}{\alpha}{\ztau'}\\
-  \performTyp{\ztau'}{\bar{\alpha}}{\ztau''}
-}{
-  \performTyp{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
-}
-\end{mathpar}
-\fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
-\vspace{-10px}\begin{mathpar}
-\inferrule{ }{
-  \performSyn{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
-}
-
-\inferrule{
-  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
-  \performSyn{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
-}{
-  \performSyn{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
-}
-\end{mathpar}
-\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
-\vspace{-10px}\begin{mathpar}
-\inferrule{ }{
-  \performAna{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
-}
-
-\inferrule{
-  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
-  \performAna{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
-}{
-  \performAna{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
-}
-\end{mathpar}
-\caption{Iterated Action Judgements}
-\label{fig:multistep}
-\end{figure}
-
 A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.\todo{define this in supplement}
@@
 \end{theorem}
+
+\begin{figure}
+$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\vspace{4px}\\
+\fbox{$\performTyp{\ztau}{\bar{\alpha}}{\ztau'}$}
+\vspace{-10px}\begin{mathpar}
+\inferrule{ }{
+    \performTyp{\ztau}{\cdot}{\ztau}
+}
+
+\inferrule{
+  \performTyp{\ztau}{\alpha}{\ztau'}\\
+  \performTyp{\ztau'}{\bar{\alpha}}{\ztau''}
+}{
+  \performTyp{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
+}
+\end{mathpar}
+\fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
+\vspace{-10px}\begin{mathpar}
+\inferrule{ }{
+  \performSyn{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
+}
+
+\inferrule{
+  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
+  \performSyn{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
+}{
+  \performSyn{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
+}
+\end{mathpar}
+\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
+\vspace{-10px}\begin{mathpar}
+\inferrule{ }{
+  \performAna{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
+}
+
+\inferrule{
+  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
+  \performAna{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
+}{
+  \performAna{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
+}
+\end{mathpar}
+\caption{Iterated Action Judgements}
+\label{fig:multistep}
+\end{figure}
 
@@
 
-Again, let us begin with type actions.
-The $\aConstruct{\farr}$ action constructs an arrow type.
-The H-type under the cursor becomes the argument type, and the cursor is placed on an empty return type hole:
+\paragraph{Types} The $\aConstruct{\farr}$ action constructs an arrow type. The H-type under the cursor becomes the argument type, and the cursor is placed on an empty return type hole:
 \begin{subequations}
@@
 
-Moving on to expression actions, we start to see some more interesting rules.
+\paragraph{Ascription} Moving on to expression actions, we start to see some more interesting rules.
 The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type.
-In the first case, the ascribed type is the synthesized type:
+In the first case, the synthesized type appears in the ascription:
 \begin{equation}
@@
 \end{equation}
-In the second case, the ascribed type is the provided type:
+In the second case, the type provided for analysis appears in the ascription:
 \begin{equation}
@@
 \end{equation}
-This rule should supercede the subsumption rule, so we must list $\aConstruct{\fasc}$ in $\mathcal{A}$.
-
-The $\aConstruct{\fvar{x}}$ action places the variable $x$ into an empty hole. If that hole is being asked to synthesize a type, then the result synthesizes the hypothesized type:
+This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).
+
+\paragraph{Variables} The $\aConstruct{\fvar{x}}$ action places the variable $x$ into an empty hole.
+If that hole is being asked to synthesize a type, then the result synthesizes the hypothesized type:
 \begin{equation}
@@
 
-The $\aConstruct{\flam{x}}$ action places a lambda term binding $x$ into an empty hole.
-If the focused empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the focus in the argument type position:
+\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole.
+If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
 \begin{equation}
@@
 \end{equation}
-The type ascription is necessary because lambda expressions do not synthesize a type. If the focused empty hole is being analyzed against an arrow type, then no ascription is necessary:
-\begin{equation}
-  \inferrule{ }{
+The type ascription is necessary because lambda expressions do not synthesize a type.
+If the focused empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
+\begin{equation}
+  \inferrule{
+    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
+  }{
     \performAna
@@
       {\zwsel{\hehole}}
-      {\tarr{\htau_1}{\htau_2}}
+      {\htau}
       {\aConstruct{\flam{x}}}
@@
 \end{equation}
-
-If the focused empty hole is being analyzed against a type that is
-incompatible with any arrow type, expressed in the premise as an arrow with
-two holes, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
-is inserted inside a hole, to maintain Theorem \ref{thrm:actsafe}:
+Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).
+
+Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
+is inserted inside a hole, which defers the type inconsistency as previously discussed:\todo{supp}
 \begin{equation}
@@
 
-The $\aConstruct{\fap}$ action applies the expression in focus to a hole. If the focused expression synthesizes a function type, then the rule is straightforward:
+\paragraph{Application} The $\aConstruct{\fap}$ action applies the expression under the cursor. The following rule handles the case where the synthesized type has matched function type:\todo{supp}
 \begin{equation}
   \label{r:coneapfn}
-  \inferrule{ }{
+  \inferrule{
+    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
+  }{
     \performSyn
@@
       {\zwsel{\hexp}}
-      {\tarr{\htau_1}{\htau_2}}
+      {\htau}
       {\aConstruct{\fap}}
@@
 \end{equation}
-
-If the focused expression synthesizes a hole type, then we can treat it as if it synthesized the $\tarr{\tehole}{\tehole}$ type, exactly as described in Sec. \ref{sec:holes}:
-\begin{equation}
-  \inferrule{ }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\tehole}
-      {\aConstruct{\fap}}
-      {\hap{\hexp}{\zwsel{\hehole}}}
-      {\tehole}
-  }
-\end{equation}
-
-Finally, if the focused expression synthesizes a type that is incompatible with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
+If the expression under the cursor synthesizes a type that is inconsistent with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
 \begin{equation}
@@
 
-The $\aConstruct{\farg}$ action places the focused expression instead in the argument position of an application.
-Because the function position is always an empty hole in this situation, we only need a single rule:
+The $\aConstruct{\farg}$ action instead places the expression under the cursor in the argument position of an application form.
+Because the function position is always an empty hole in this situation, we need only a single rule:
 \begin{equation}
@@
 
-The $\aConstruct{\fnumlit{n}}$ action places the number expression $\hnum{n}$ into an empty hole. If the focused hole is being asked to synthesize a type, then the rule is straightforward:
+\paragraph{Numbers} The $\aConstruct{\fnumlit{n}}$ action replaces an empty hole with the number expression $\hnum{n}$.
+If the empty hole is being asked to synthesize a type, then the rule is straightforward:
 \begin{equation}
@@
 \end{equation}
-If the focused hole is being analyzed against a type that is incompatible with $\tnum$, then we must place the number expression inside a hole:
+If the empty hole is being analyzed against a type that is incompatible with $\tnum$, then we must place the number expression inside the hole:
 \begin{equation}
@@
 
-Finally, the $\aConstruct{\fplus}$ action constructs a plus expression with the focused expression as its first argument.
-If the focused expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
+The $\aConstruct{\fplus}$ action constructs a plus expression with the expression under the cursor as its first argument.
+If that expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
 \begin{equation}\label{rule:construct-plus-compat}
@@
 \end{equation}
-
-Otherwise, we must place the focused expression inside a hole:
+Otherwise, we must place that expression inside a hole:
 \begin{equation}
@@
 \end{equation}
-\end{subequations}
-Notice that we do not have an action that explicitly wraps an expression in a non-empty hole.
-These arise implicitly when an action that would not na\"ively satisfy Theorem \ref{thrm:actsafe} is performed (see Figure \ref{fig:first-example}.)
+
+\paragraph{Non-Empty Holes} The final shape is $\fnehole$. This explicitly places the expression under the cursor inside a hole:
+\begin{equation}
+\inferrule{ }{
+  \performSyn
+    {\hGamma}
+    {\zwsel{\hexp}}
+    {\htau}
+    {\aConstruct{\fnehole}}
+    {\hhole{\zwsel{\hexp}}}
+    {\tehole}
+}
+\end{equation}\end{subequations}
+
+The $\fnehole$ shape is grayed out in Figure \ref{fig:action-syntax} because we do not expect the programmer to perform it explicitly -- other actions automatically insert holes when a type inconsistency would arise.
+Its inclusion is mainly to make it easier to state another ``checksum'' theorem: \emph{constructability}
+
+\paragraph{Constructability}
+To check that we have defined ``enough'' construct actions, we need to establish that we can start from an empty hole and arrive at any well-typed expression with, for simplicity, the cursor on the outside (Lemma \ref{lemma:reach-down} allows us to then move the cursor anywhere else.) As with reachability, we rely on the iterated action judgements defined in Figure \ref{fig:multistep}.
+\begin{theorem}[Constructability] ~
+\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+\item For every $\htau$ there exists $\bar\alpha$ such that $\performTyp{\zwsel{\tehole}}{\bar\alpha}{\zwsel{\htau}}$.
+\item If $\hana{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that $\performAna{\hGamma}{\zwsel{\hhole{}}}{\htau}{\bar\alpha}{\zwsel{\hexp}}$.
+\item If $\hsyn{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that $\performSyn{\hGamma}{\zwsel{\hhole{}}}{\tehole}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
+\end{enumerate}
+\end{theorem}
+
+Witnesses...\todo{talk about this}
 
 \subsubsection{Finishing}
-The final action we will consider in Hazelnut is $\aFinish$, which finishes the focused non-empty hole.
-
-If the focused non-empty hole appears in synthetic position, then it can always be finished:
+The final action we need to consider is $\aFinish$, which finishes the non-empty hole under the cursor.
+
+If the non-empty hole appears in synthetic position, then it can always be finished:
 \begin{subequations}
@@
 
-If the focused non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the wrapped expression is consistent with the type the hole is being analyzed against.
-This amounts to analyzing those contents against the provided type (by subsumption):
+\todo{make sure there is no more compatibility or focus words}If the non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the enveloped expression is consistent with the type that the hole is being analyzed against.
+This amounts to analyzing the enveloped expression against the provided type (by subsumption):
 \begin{equation}\label{r:finishana}
@@
 
-\paragraph{Constructibility}\todo{why is this with the zipper rules not the
-    construct rules?}
-
-We give the following theorem to demonstrate that the rules for the action
-$\aConstruct{\psi}$ do indeed capture the process of building any term
-of the form $\psi$. Intuitively, its proof gives a list of composable
-actions that, when applied one sequentially to an empty hole in focus,
-build any well-typed expression.
-
-As with Theorem \ref{thrm:reach}, the statement of this theorem must match
-the mutually recursive structure of the bidirectional typing judgements.
-
-\begin{theorem}[Constructability]\todo{why $\htau$ and not $\ztau$?}
-\label{thrm:construct}
-Both of the following hold:
-\begin{enumerate}
-  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
-      L \in \mathit{list} ~\mathtt{action}$ such that
-      $\mathit{iterate}~ L~ \zwsel{\hhole{}} = \zexp'$ and
-      $\removeSel{\zexp'} = \hexp$
-
-  \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\exists \zexp'$ $\exists
-      L \in \mathit{list} ~\mathtt{action}$ such that $\mathit{iterate}~
-      L~ \zwsel{\hhole{}} = \zexp'$ and $\removeSel{\zexp'} = \hexp$
-\end{enumerate}
-\end{theorem}
-
-Because actions are defined on the zippered form of expressions, we argue
-that we can construct a term that is the same as the target up to the
-erasure of focus rather than on the nose. Since any form inside a zippered
-term may be in focus, there are linearly many such terms for a given
-$\hexp$---Theorem \ref{thrm:construct} can be composed with
-Theorem \ref{thrm:reach} to reach any one that might be desired in
-particular.
-
-\paragraph{Finishability}\todo{do we care?}
-As with deletion, since the meaning of the finish action does not depend on
-the structure of the term being finished other than requiring that it must
-be one of the two hole forms, there is no checksum theorem to state for
-this action.
-
-\subsubsection{Zipper Cases}\label{sec:zipper-cases} The rules given so far handle the base cases, where the action has ``reached'' the focused expression.
-We also need to define the recursive cases, which propagate the action into the subtree where the focus appears.
-These rules follow the structure of the corresponding rules in the statics of H-expressions.
+\subsubsection{Zipper Cases}\label{sec:zipper-cases} The rules defined so far handle the base cases, i.e.
+the cases where the action has ``reached'' the expression under the cursor.
+We also need to define the recursive cases, which propagate the action into the subtree where the cursor appears, as encoded by the zipper structure.
+These rules essentially follow the structure of the corresponding rules in the statics.
 
 \begin{subequations}
-For example, when the focus is in the expression position of an ascription, we use the analytic action performance judgement:
-\begin{equation}
-\inferrule{
-  \performAna
-    {\hGamma}
-    {\zexp}
-    {\htau}
-    {\alpha}
-    {\zexp'}
-}{
-  \performSyn
-    {\hGamma}
-    {\zexp : \htau}
-    {\htau}
-    {\alpha}
-    {\zexp' : \htau}
-    {\htau}
-}
-\end{equation}
-
-When the focus is in the type position of an ascription, we must re-check the ascribed expression because the type might have changed (in practice, one would optimize this check to only occur if the type actually was changed):
-\begin{equation}
-\inferrule{
-  \performTyp{\ztau}{\alpha}{\ztau'}\\
-  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
-}{
-  \performSyn
-    {\hGamma}
-    {\hexp : \ztau}
-    {\removeSel{\ztau}}
-    {\alpha}
-    {\hexp : \ztau'}
-    {\removeSel{\ztau'}}
-}
-\end{equation}
-
-If the focus is in the body of a lambda expression, then we must use the analytic action performance rule:
-\begin{equation}
-\inferrule{
+For example, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):\todo{supp}
+\begin{equation}
+\inferrule{
+  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
   \performAna
@@
     {\hlam{x}{\zexp}}
-    {\tarr{\htau_1}{\htau_2}}
+    {\htau}
     {\alpha}
@@
 
-There are two rules that handle the case where the focus is in the function position of an application, corresponding to the two application rules in the statics.
-Each involves rechecking the argument against the new function type:
+When the cursor is in the expression position of an ascription, we use the analytic  judgement, mirroring Rule (\ref{rule:syn-asc}):
+\begin{equation}
+\inferrule{
+  \performAna
+    {\hGamma}
+    {\zexp}
+    {\htau}
+    {\alpha}
+    {\zexp'}
+}{
+  \performSyn
+    {\hGamma}
+    {\zexp : \htau}
+    {\htau}
+    {\alpha}
+    {\zexp' : \htau}
+    {\htau}
+}
+\end{equation}
+
+When the focus is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
+\begin{equation}
+\inferrule{
+  \performTyp{\ztau}{\alpha}{\ztau'}\\
+  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
+}{
+  \performSyn
+    {\hGamma}
+    {\hexp : \ztau}
+    {\removeSel{\ztau}}
+    {\alpha}
+    {\hexp : \ztau'}
+    {\removeSel{\ztau'}}
+}
+\end{equation}
+
+When the cursor is in the function position of an application, the rule mirrors Rule (\ref{rule:syn-ap}):
 \begin{equation}
@@
       {\zexp'}
-      {\tarr{\htau_3}{\htau_4}}\\
-    \hana{\hGamma}{\hexp}{\htau_3}
+      {\htau_3}\\\\
+    \arrmatch{\htau_3}{\tarr{\htau_4}{\htau_5}}\\
+    \hana{\hGamma}{\hexp}{\htau_4}
   }{
@@
       {\hap{\zexp'}{\hexp}}
-      {\htau_4}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
-    \performSyn
+      {\htau_5}
+  }
+\end{equation}
+% \begin{equation}
+%   \inferrule{
+%     \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
+%     \performSyn
+%       {\hGamma}
+%       {\zexp}
+%       {\htau_2}
+%       {\alpha}
+%       {\zexp'}
+%       {\tehole}\\
+%     \hana{\hGamma}{\hexp}{\tehole}
+%   }{
+%     \performSyn
+%       {\hGamma}
+%       {\hap{\zexp}{\hexp}}
+%       {\htau_1}
+%       {\alpha}
+%       {\hap{\zexp'}{\hexp}}
+%       {\tehole}
+%   }
+% \end{equation}
+
+The situation is similar when the cursor is in argument position:
+\begin{equation}
+  \inferrule{
+    \hsyn{\hGamma}{\hexp}{\htau_2}\\
+    \arrmatch{\htau_2}{\tarr{\htau_3}{\htau_4}}\\
+    \performAna
       {\hGamma}
       {\zexp}
-      {\htau_2}
+      {\htau_3}
       {\alpha}
       {\zexp'}
-      {\tehole}\\
-    \hana{\hGamma}{\hexp}{\tehole}
   }{
@@
       {\hGamma}
-      {\hap{\zexp}{\hexp}}
+      {\hap{\hexp}{\zexp}}
       {\htau_1}
       {\alpha}
-      {\hap{\zexp'}{\hexp}}
-      {\tehole}
-  }
-\end{equation}
-
-Similarly, there are two rules that handle the case where the focus is in the argument position:
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\hexp}{\tarr{\htau_2}{\htau}}\\
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\htau_2}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\hexp}{\zexp}}
-      {\htau}
-      {\alpha}
       {\hap{\hexp}{\zexp'}}
-      {\htau}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\hexp}{\tehole}\\
-    \performAna
-      {\hGamma}
-      {\zexp}
-      {\tehole}
-      {\alpha}
-      {\zexp'}
-  }{
-    \performSyn
-      {\hGamma}
-      {\hap{\hexp}{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\hap{\hexp}{\zexp'}}
-      {\tehole}
-  }
-\end{equation}
-
-The rules for the addition operator follow from the statics directly:
+      {\htau_4}
+  }
+\end{equation}
+% \begin{equation}
+%   \inferrule{
+%     \hsyn{\hGamma}{\hexp}{\tehole}\\
+%     \performAna
+%       {\hGamma}
+%       {\zexp}
+%       {\tehole}
+%       {\alpha}
+%       {\zexp'}
+%   }{
+%     \performSyn
+%       {\hGamma}
+%       {\hap{\hexp}{\zexp}}
+%       {\tehole}
+%       {\alpha}
+%       {\hap{\hexp}{\zexp'}}
+%       {\tehole}
+%   }
+% \end{equation}
+
+The rules for the addition operator mirror Rule (\ref{rule:syn-plus}):
 \begin{equation}
@@
 
-Finally, if the focus is inside a non-empty hole, we special case the situation where the action results in a doubly-nested empty hole, $\hhole{\hehole}$, to eliminate the nesting (given our current action semantics, only the delete action can cause this form to arise and the form $\hhole{\hhole{\zexp}}$ cannot arise):
+Finally, if the cursor is inside a non-empty hole, the relevant zipper rule mirrors Rule (\ref{rule:syn-ehole}):
 \begin{equation}
@@
       {\htau'}\\
-    \zexp' \neq \zwsel{\hehole}
   }{
@@
 \end{equation}
-\begin{equation}
-  \inferrule{
-    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
-    \performSyn
-      {\hGamma}
-      {\zexp}
-      {\htau}
-      {\alpha}
-      {\zwsel{\hehole}}
-      {\tehole}\\
-  }{
-    \performSyn
-      {\hGamma}
-      {\hhole{\zexp}}
-      {\tehole}
-      {\alpha}
-      {\zwsel{\hehole}}
-      {\tehole}
-  }
-\end{equation}
-
+% \begin{equation}
+%   \inferrule{
+%     \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
+%     \performSyn
+%       {\hGamma}
+%       {\zexp}
+%       {\htau}
+%       {\alpha}
+%       {\zwsel{\hehole}}
+%       {\tehole}\\
+%   }{
+%     \performSyn
+%       {\hGamma}
+%       {\hhole{\zexp}}
+%       {\tehole}
+%       {\alpha}
+%       {\zwsel{\hehole}}
+%       {\tehole}
+%   }
+% \end{equation}
+
+Theorem \ref{thrm:actsafe} directly checks the correctness of these rules.
+Moreover, the zipper rules arise ubiquitously in derivations of edit steps, so the proofs of the other ``check'' theorems, e.g.
+Reachability and Constructability, serve as a check that none of these rules have been missed.
 \end{subequations}

### 2016-07-07 agda.tex
tightened up mechanization section

@@
 % !TEX root = hazelnut-popl17.tex
-In the previous section, we gave an overview of the most important rules in
-the semantics of Hazelnut, and stated the important theorems. In a few
-cases, we sketched out why these theorems will hold.
-
-In order to formally verify that our design meets the stated objectives, we
-are preparing a formalization of the grammar, the judgements given above
-and the metatheory in the proof assistant Agda \cite{norell:thesis}. We
+So far, we have given an overview of the most important judgements and rules in
+the semantics of Hazelnut, and stated the important metatheorems. In a few
+cases, we have informally sketched out why these metatheorems will hold. In order to formally establish that our design meets our stated objectives, we
+have mechanized the semantics and metatheory of Hazelnut using the Agda proof assistant \cite{norell:thesis}. We 
 refer readers unfamiliar with Agda to the Agda Wiki, hosted
@@
 
-Our formalization of Hazelnut is under development at
-\url{http://github.com/hazelgrove/agda-tfp16}. At the time of submission, we 
-have completed much of the initial groundwork, but the proofs of 
-Theorem \ref{thrm:actsafe} and Theorem \ref{thrm:actdet} (and a few  
-lemmas that we elided here) are under construction.
+This development is available in the supplemental material.
+At the time of submission, all of the rules as well as the proofs of Theorem \ref{thrm:actsafe} and Theorem \ref{thrm:actdet} (including the proofs a number of technical lemmas that we elided here) are complete.\todo{rest of it}
 
@@
 includes a more detailed discussion of the technical representation
-decisions made. The core idea of our formalization is to encode each judgement as a
-dependent type. The rules of the judgements become the constructors of the
-type, and derivations of theorems values of the type. This is a rich
+decisions that we made. The main idea is standard: we encode each judgement as a
+dependent type. The rules defining the judgements become the constructors of this
+type, and derivations are terms of these type. This is a rich
 setting that allows proofs to take advantage of pattern matching on the
-shape of derivations, closely matching on-paper proofs of similar
-properties.
-
-The formalization differs from the calculus defined here in a few small 
-ways. Most interestingly, instead of giving separate movement rules for each 
-form of Z-expression, we abstract
-over different corresponding forms that happen to have the same arity, e.g. 
-additions and applications. Formalizing this intuition
-reduces the number of cases we need to consider 
-somewhat, but more importantly allows us to write a slightly more
-general calculus -- it will be easier to extend Hazelnut with more 
-interesting language features with this generic infrastructure in place.
-
-Because the metatheory in this paper is largely concerned with the statics
-of Hazelnut rather than its dynamics, we adopt Barendregt's convention for
-bound variables and avoid substitution entirely \cite{urban}. Future
-iterations will need a more mature technique for reasoning about
-binding---likely de Bruijn indices or abstract binding
-trees \cite{lh09unibind,Pouillard11}.
+shape of derivations, closely matching standard on-paper proofs.
+No proof automation was used, because the proof structure itself is likely to be interesting to researchers who plan to build upon our work.
+% The formalization differs from the calculus defined here in a few small 
+% ways. Most interestingly, instead of giving separate movement rules for each 
+% form of Z-expression, we abstract
+% over different corresponding forms that happen to have the same arity, e.g. 
+% additions and applications. Formalizing this intuition
+% reduces the number of cases we need to consider 
+% somewhat, but more importantly allows us to write a slightly more
+% general calculus -- it will be easier to extend Hazelnut with more 
+% interesting language features with this generic infrastructure in place.
+We adopt Barendregt's convention for
+bound variables \cite{urban}. Hazelnut's semantics does not need substitution, so we do not need to adopt more sophisticated encodings (e.g. \cite{lh09unibind,Pouillard11}.)

### 2016-07-07 formally.tex
progress on 4
fixed subsumes

@@
     \tcompat{\htau}{\htau''}\\\\
-    \alpha~\mathsf{subsumes}
+    \zexp, \alpha~\mathsf{subsumes}
   }{
@@
 We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
+\begin{subequations}
 \begin{equation}\label{rule:subsumes}
@@
 }{
-  \alpha~\mathsf{subsumes}
-}
-\end{equation}
+  \zexp, \alpha~\mathsf{subsumes}
+}
+\end{equation}
+\begin{equation}\label{rule:subsumes}
+\inferrule{
+  \zexp \neq \zwsel{\hexp}
+}{
+  \zexp, \alpha~\mathsf{subsumes}
+}
+\end{equation}
+\end{subequations}
 Algorithmically, this captures the notion that subsumption should be the case of last resort.
@@
 To check that we have defined ``enough'' construct actions, we need to establish that we can start from an empty hole and arrive at any well-typed expression with, for simplicity, the cursor on the outside (Lemma \ref{lemma:reach-down} allows us to then move the cursor anywhere else.) As with reachability, we rely on the iterated action judgements defined in Figure \ref{fig:multistep}.
-\begin{theorem}[Constructability] ~
+\begin{theorem}[Constructability]\label{thrm:constructability} ~
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
@@
 We also need to define the recursive cases, which propagate the action into the subtree where the cursor appears, as encoded by the zipper structure.
-These rules essentially follow the structure of the corresponding rules in the statics.
+For types, the zipper rules are straightforward:
+\begin{subequations}
+\begin{equation}
+    %\label{r:contarrL}
+  \inferrule{
+    \performTyp{\ztau}{\alpha}{\ztau'}
+  }{
+    \performTyp{
+      \tarr{\ztau}{\htau}
+    }{
+      \alpha
+    }{
+      \tarr{\ztau'}{\htau}
+    }
+  }
+\end{equation}
+  \begin{equation}
+    %\label{r:contarrR}
+  \inferrule{
+    \performTyp{\ztau}{\alpha}{\ztau'}
+  }{
+    \performTyp{
+      \tarr{\htau}{\ztau}
+    }{
+      \alpha
+    }{
+      \tarr{\htau}{\ztau'}
+    }
+  }
+\end{equation}
+\end{subequations}
+For expressions, the zipper rules essentially follow the structure of the corresponding rules in the statics.
 

### 2016-07-07 formally.tex
final sec 4

@@
 
-Before giving the rules for these judgements, let us state two key metatheorems.
-These metatheorems deeply inform the design of the rules.
-Their proofs, which are mechanized in Agda, are by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules, starting in Sec.
-\ref{sec:action-subsumption}.
-
 % The type action judgement appears as a premise in the rules that handle Z-expressions where the cursor is inside a . 
@@
 \subsubsection{Sensibility}
-The expression action judgements obey the following \emph{action sensibility} theorem:
+
+Before giving the rules for these judgements, let us state the key metatheorem: \emph{sensibility}.
+This metatheorem deeply inform the design of the rules, given starting in Sec.
+\ref{sec:action-subsumption}.
+Its proof, which is mechanized in Agda, is by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules.
 \begin{theorem}[Action Sensibility]\label{thrm:actsafe} Both of the following hold:\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
@@
 
-\subsubsection{Determinism}
-Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
-\begin{theorem}[Action Determinism] All of the following hold:
-\label{thrm:actdet}
-\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
-\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
-  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
-  $\zexp' = \zexp''$ and $\htau' = \htau''$.
-% \item If all of
+% \subsubsection{Determinism}
+% Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
+% \begin{theorem}[Action Determinism] All of the following hold:
+% \label{thrm:actdet}
+% \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+% \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
+% \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
+%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
+%   $\zexp' = \zexp''$ and $\htau' = \htau''$.
+% % \item If all of
 
@@
 %   hold, then $\zexp' = \zexp''$.
-\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
-  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
-  \zexp''$.
-\end{enumerate}
-\end{theorem}
-
-In order to maintain determinism, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$.
+% % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
+% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
+% %   \zexp''$.
+% % \end{enumerate}
+% \end{theorem}
+
+\subsubsection{Type Inconsistency}
+In order to define the action semantics, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$.
 One can define this notion either direrctly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement.
@@
     \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
-    \tcompat{\htau}{\htau''}\\\\
-    \zexp, \alpha~\mathsf{subsumes}
+    \tcompat{\htau}{\htau''}%\\\\
+    % \zexp, \alpha~\mathsf{subsumes}
   }{
@@
 
-We need to make some exceptions in this rule to maintain determinism, because certain construction actions, described in Sec.
+% We need to make some exceptions in this rule to maintain determinism, because certain construction actions, described in Sec.
 \ref{sec:construction}, behave slightly differently depending on whether a type is known.
 We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
-\begin{subequations}
-\begin{equation}\label{rule:subsumes}
-\inferrule{
-  \alpha \neq \aConstruct{\fasc}\\
-  (\forall x) \alpha \neq \aConstruct{\flam{x}} 
-}{
-  \zexp, \alpha~\mathsf{subsumes}
-}
-\end{equation}
-\begin{equation}\label{rule:subsumes}
-\inferrule{
-  \zexp \neq \zwsel{\hexp}
-}{
-  \zexp, \alpha~\mathsf{subsumes}
-}
-\end{equation}
-\end{subequations}
-Algorithmically, this captures the notion that subsumption should be the case of last resort.
+% \begin{subequations}
+% \begin{equation}\label{rule:subsumes}
+% \inferrule{
+%   \alpha \neq \aConstruct{\fasc}\\
+%   (\forall x) \alpha \neq \aConstruct{\flam{x}} 
+% }{
+%   \zexp, \alpha~\mathsf{subsumes}
+% }
+% \end{equation}
+% \begin{equation}\label{rule:subsumes}
+% \inferrule{
+%   \zexp \neq \zwsel{\hexp}
+% }{
+%   \zexp, \alpha~\mathsf{subsumes}
+% }
+% \end{equation}
+% \end{subequations}
+% Algorithmically, this captures the notion that subsumption should be the case of last resort.
 
@@
 \end{equation}
-This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).
+% This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).
 
@@
 If the focused empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
-\begin{equation}
+\begin{equation}\label{rule:performAna-lam-1}
   \inferrule{
@@
 \end{equation}
-Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).
+% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).
 
@@
 is inserted inside a hole, which defers the type inconsistency as previously discussed:\todo{supp}
-\begin{equation}
+\begin{equation}\label{rule:performAna-construct-lam-2}
   \inferrule{
@@
 
-Witnesses...\todo{talk about this}
-
 \subsubsection{Finishing}

### 2016-07-07 agda.tex
more progress

@@
 So far, we have given an overview of the most important judgements and rules in
-the semantics of Hazelnut, and stated the important metatheorems. In a few
+the semantics of Hazelnut, and stated the critical metatheorem, Sensibility, and several auxiliary ``checksum'' metatheorems. In a few
 cases, we have informally sketched out why these metatheorems will hold. In order to formally establish that our design meets our stated objectives, we
-have mechanized the semantics and metatheory of Hazelnut using the Agda proof assistant \cite{norell:thesis}. We 
+have mechanized the semantics of Hazelnut using the Agda proof assistant \cite{norell:thesis}. We 
 refer readers unfamiliar with Agda to the Agda Wiki, hosted
@@
 This development is available in the supplemental material.
-At the time of submission, all of the rules as well as the proofs of Theorem \ref{thrm:actsafe} and Theorem \ref{thrm:actdet} (including the proofs a number of technical lemmas that we elided here) are complete.\todo{rest of it}
+At the time of submission, the full set of rules is available and the proofs of Theorem \ref{thrm:actsafe}, Theorem \ref{lemma:movement-erasure} and a number of technical lemmas that we elided or mentioned informally in the text are complete.
+We also have nearly all of the cases of Reachability and Constructability complete (and see no major barrier to the remaining cases -- they are simply tedious.) \todo{include link to repos in non-anonymous supplementals}\todo{talk about determinism either here or there}
 

### 2016-07-07 formally.tex
more progress

@@
 \subsubsection{Type Inconsistency}
-In order to define the action semantics, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$.
-One can define this notion either direrctly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement.
+In some of the rules below, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$.
+One can define this notion either directly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement.
 The supplement does the latter.
@@
 
-The bidirectional action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
+The action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
 \begin{equation}\label{rule:action-subsume}
@@
 Rule (\ref{rule:ana-subsume}).
+Algorithmically, subsumption should be the rule of last resort.
 
@@
 
-\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} ~
+\begin{theorem}[Movement Erasure Invariance]\label{lemma:movement-erasure} ~
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px] 
@@
 \end{enumerate}
-\end{lemma}
-\noindent Lemma \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.
+\end{theorem}
+\noindent Theorem \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.
 
@@
 \end{lemma}
-Theorem \ref{thrm:reachability} follows by straightforward composition of these two lemmas. (The witness to the existential is not, of course, the shortest path between any two cursor locations.)
+Theorem \ref{thrm:reachability} follows by straightforward composition of these two lemmas. %(The witness to the existential is not, of course, the shortest path between any two cursor locations.)
 
@@
 
-\paragraph{Ascription} Moving on to expression actions, we start to see some more interesting rules.
-The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type.
+\paragraph{Ascription} The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type.
 In the first case, the synthesized type appears in the ascription:
@@
 The type ascription is necessary because lambda expressions do not synthesize a type.
-If the focused empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
+If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
 \begin{equation}\label{rule:performAna-lam-1}
@@
 \begin{subequations}
-For example, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):\todo{supp}
+In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):\todo{supp}
 \begin{equation}
@@
 When the focus is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
-\begin{equation}
+\begin{equation}\label{rule:zipper-asc}
 \inferrule{

### 2016-07-07 implementation.tex
added new impl figure; more closely matches terminology and syntax from paper
minor figure tweak: user moves in time
minor: bigger figure
tweaked title linebreak spacing; updated colors and stroke for hazel
final sec 4
more progress

--- 2016-04-27 implementation.tex
@@
+% !TEX root = hazelnut-popl17.tex
 \subsection{Implementation Concepts}
-
-\begin{figure}
-\centering
-\includegraphics[width=4in]{Implementation_Diagram}
-\caption{Implementation Concepts. Each Action--Zipper Structure--View combination is considered to appear ``instantaneously'' on the timeline.}
-\label{fig:FRP}
-\end{figure}
+Central to any implementation of Hazelnut is a stream of edit states whose cursor erasures synthesize types according to the synthetic action judgement under an empty context, $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$.
+The middle row of Figure \ref{fig:impl-overview} diagrams this stream of edit states.
+By Theorem \ref{thrm:actsafe}, the editor does not need to typecheck each edit state (though in some of the rules, e.g.
+Rule (\ref{rule:zipper-asc}) which handles the situation where the type in a type ascription changes, portions of the program do need to be typechecked.) For example, the reader is encouraged to re-examine the examples in Figure \ref{fig:first-example} and \ref{fig:second-example} -- the cursor erasure of each edit state synthesizes a type.
 
@@
 
-A substantially simpler system that we developed while exploring the ideas that led to Hazelnut can be found at the following URL:
-\url{http://www.cs.cmu.edu/~comar/nestedpairs/}.
+\begin{figure}
+\centering
+\includegraphics[width=0.90\columnwidth]{impl-overview2}
+\caption{...}
+\label{fig:impl-overview}
+\end{figure}

### 2016-07-07 agda.tex
clarified missing thms

@@
 At the time of submission, the full set of rules is available and the proofs of Theorem \ref{thrm:actsafe}, Theorem \ref{lemma:movement-erasure} and a number of technical lemmas that we elided or mentioned informally in the text are complete.
-We also have nearly all of the cases of Reachability and Constructability complete (and see no major barrier to the remaining cases -- they are simply tedious.) \todo{include link to repos in non-anonymous supplementals}\todo{talk about determinism either here or there}
+We also have nearly all of the cases of Reachability and Constructability complete, and we see no major barrier to the remaining cases -- we prioritized developing a clear exposition of our contributions after proving representative cases and finding that they were straightforward.
+The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.\todo{include link to repos in non-anonymous supplementals}\todo{talk about determinism either here or there}
 

### 2016-07-07 implementation.tex
clarified missing thms

@@
 Rule (\ref{rule:zipper-asc}) which handles the situation where the type in a type ascription changes, portions of the program do need to be typechecked.) For example, the reader is encouraged to re-examine the examples in Figure \ref{fig:first-example} and \ref{fig:second-example} -- the cursor erasure of each edit state synthesizes a type.
+
+The programmer examines a view generated from each edit state and produces actions in some implementation-defined manner (e.g.
+using a keyboard, mouse, touchscreen, voice interface, or neural implant.) Each new action causes a new abstract edit state to arise according to an implementation of the action semantics.
+This then causes a new to arise.
+This is a simple event-based functional reactive programming model \cite{Wan:2000:FRP:349299.349331}.
+
+If an action is not well-defined according to the action semantics, the implementation must reject it.
+In fact, an implementation is encouraged to present an ``action palette'' that either hides or visibly disables actions that are not well-defined (see below.)
 

### 2016-07-07 agda.tex
impl section revised

@@
 So far, we have given an overview of the most important judgements and rules in
-the semantics of Hazelnut, and stated the critical metatheorem, Sensibility, and several auxiliary ``checksum'' metatheorems. In a few
+the semantics of Hazelnut, and stated the critical metatheorem, Sensibility, and several auxiliary ``checks''. In a few
 cases, we have informally sketched out why these metatheorems will hold. In order to formally establish that our design meets our stated objectives, we
-have mechanized the semantics of Hazelnut using the Agda proof assistant \cite{norell:thesis}. We 
-refer readers unfamiliar with Agda to the Agda Wiki, hosted
-at \url{http://wiki.portal.chalmers.se/agda/}.
+have mechanized the semantics of Hazelnut using the Agda proof assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
+at \url{http://wiki.portal.chalmers.se/agda/}.)
 
@@
 We also have nearly all of the cases of Reachability and Constructability complete, and we see no major barrier to the remaining cases -- we prioritized developing a clear exposition of our contributions after proving representative cases and finding that they were straightforward.
-The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.\todo{include link to repos in non-anonymous supplementals}\todo{talk about determinism either here or there}
+The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.\todo{talk about determinism either here or there}
 

### 2016-07-07 implementation.tex
impl section revised

@@
 \subsection{Implementation Concepts}
-Central to any implementation of Hazelnut is a stream of edit states whose cursor erasures synthesize types according to the synthetic action judgement under an empty context, $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$.
+Central to any implementation of Hazelnut is a stream of edit states whose cursor erasures synthesize types under an empty context according to the synthetic action judgement, $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$.
 The middle row of Figure \ref{fig:impl-overview} diagrams this stream of edit states.
@@
 using a keyboard, mouse, touchscreen, voice interface, or neural implant.) Each new action causes a new abstract edit state to arise according to an implementation of the action semantics.
-This then causes a new to arise.
+This then causes a new view to arise.
 This is a simple event-based functional reactive programming model \cite{Wan:2000:FRP:349299.349331}.
 
-If an action is not well-defined according to the action semantics, the implementation must reject it.
-In fact, an implementation is encouraged to present an ``action palette'' that either hides or visibly disables actions that are not well-defined (see below.)
-
-The key question that must be answered for any implementation strategy is: how do we model a stream of actions from a user?
-Let us assume that these actions are chosen (using some input device, preferably, a keyboard) from some ``palette'' that never presents the user with actions that are not semantically well-defined, according to the action semantics defined earlier.
-%In a traditional editor, the input from the user is a stream of characters, and there are no guarantees that at any point that the program is syntactically well-formed, so the designer leaves editing as an .
-%In contrast, in a structure editor, the input from the user is a stream of operations.
-As such, each new action will ``atomically'' generate a new Z-expression. 
-This insight leads us to conclude that a natural way to implement this editor would be using event-based Functional Reactive Programming~\cite{Wan:2000:FRP:349299.349331} (FRP).
-Figure~\ref{fig:FRP} illustrates the concept of an FRP-based implementation of a  structure editor organized like Hazelnut.
-The input from the user is a stream of actions.  Each action results in a change to the underlying abstract model (i.e., a new Z-expression is created after each action.)
-Each model change results in an updated \emph{view} which is then presented to the user.  The user can then consider this new view when they choose a new action as input.
-
-\subsection{HZ}
-We explore the concepts presented in the paper in HZ, our implementation of Hazelnut.
-In order to reach a wide audience, we decided to implement HZ in the web browser.
-In order to take advantage of all the benefits of FRP, we chose to implement HZ using OCaml\footnote{https://ocaml.org/}, the \texttt{js\_of\_ocaml} compiler\footnote{http://ocsigen.org/js\_of\_ocaml/} and the OCaml React library\footnote{http://erratique.ch/software/react}.
-
-At the time of the writing of this paper, our implementation of HZ includes encodings of Z-expression as presented in this paper.
-We consider this ZExp to be our model. 
-HZ renders the model as a string embedded in HTML.
-Currently we support only the delete action.
-Other actions are currently under development.
-We anticipate having a substantially more functional implementation by the time this work is presented (our focus thusfar has been on the metatheory.)
-The work-in-progress code as well as directions for how to compile and run it can be found here: \url{https://github.com/hazelgrove/impl-tfp16}.
+If an action is not well-defined according to Hazelnut's action semantics, the implementation must reject it.
+In fact, the implementation is encouraged to present an ``action palette'' that either hides or visibly disables actions that are not well-defined (see below.)
 
@@
 \centering
-\includegraphics[width=0.90\columnwidth]{impl-overview2}
-\caption{...}
+\includegraphics[width=\columnwidth]{impl-overview2}
+\caption{Implementation Concepts}
 \label{fig:impl-overview}
 \end{figure}
+
+\subsection{HZ}
+We have developed a simple reference implementation of Hazelnut.
+In order to reach a wide audience, we decided to implement HZ in the web browser.
+In order to take advantage of a mature implementation of the FRP model, we chose to implement HZ using OCaml\footnote{https://ocaml.org/}, the \texttt{js\_of\_ocaml} compiler and associated libraries\footnote{http://ocsigen.org/js\_of\_ocaml/} and the OCaml React library\footnote{http://erratique.ch/software/react}.
+
+At the time of the writing, 
+HZ's view computation renders the model as a string embedded into HTML.
+The action palette is a collection of buttons and text boxes, which are disabled when an action is not well-defined or an invalid action argument is entered.
+We determine this by simply attempting to run the action internally and catching the exception that is raised when an action is undefined.
+This, of course, is not meant to score marks for usability or performance.
+Instead, as a simple reference implementation, we expect its source code to be of use to others who are interested in layering a more fluid user interface atop the core semantics (which have been implemented in a style that follows the presentation in this paper closely), or extensions thereof.

### 2016-07-07 agda.tex
final popl paper + supplement

@@
 We also have nearly all of the cases of Reachability and Constructability complete, and we see no major barrier to the remaining cases -- we prioritized developing a clear exposition of our contributions after proving representative cases and finding that they were straightforward.
-The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.\todo{talk about determinism either here or there}
+The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.
 
@@
 bound variables \cite{urban}. Hazelnut's semantics does not need substitution, so we do not need to adopt more sophisticated encodings (e.g. \cite{lh09unibind,Pouillard11}.)
+
+\subsubsection{Determinism}
+A useful property would be \emph{action determinism}, i.e. that performing an action produces a unique result. Formally, this would be established as follows:
+% \begin{theorem}[Action Determinism] All of the following hold:
+% \label{thrm:actdet}
+\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+\item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
+\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
+  $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
+  $\zexp' = \zexp''$ and $\htau' = \htau''$.
+% \item If all of
+
+  % \begin{quote}
+  %   \begin{enumerate}
+  %   \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
+  %   \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
+  %   \item $\tcompat{\htau}{\htau'}$, and
+  %   \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
+  %     $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
+  %   \end{enumerate}
+  % \end{quote}
+  % hold, then $\zexp' = \zexp''$.
+\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
+  $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
+  \zexp''$.
+\end{enumerate}
+% \end{theorem}
+
+This is not a theorem of the system as described in this paper.
+The reason is quite subtle: the semantics of two construction actions, $\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the analytic case than they do in the synthetic case -- the difference has to do with what the generated ascription contains.
+Both cases can technically ``fire'' due to action subsumption.
+The mechanization includes a ``proof'' of determinism for all cases but these two.
+Resolving this ambiguity formally requires some rather inelegant machinery to exclude these particular base cases from action subsumption, so we are content to admonish implementors who apply action subsumption when another analytic rule would do.

### 2016-07-07 formally.tex
final popl paper + supplement

@@
 The supplement includes the complete collection of rules, in definitional order.
-These rules, along with the proofs of the metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement.
+These rules, along with the proofs of two key metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement.
 We will summarize this effort in Sec.
@@
 Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete (see supplement.) We are not concerned here with defining a dynamic semantics for Hazelnut, but the dynamics for complete H-expressions would be entirely standard (in Sec.
-\ref{sec:future} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers.
+\ref{sec:rw} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers.
 %Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.
@@
 Two incomplete H-types are consistent if they differ only at positions where a hole appears in either type.
-The type hole is therefore compatible with every type.
+The type hole is therefore consistent with every type.
 This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ (i.e.
-unknown) type \cite{Siek06a}.\todo{fwd reference related work}
+unknown) type \cite{Siek06a}.
 
@@
 This judgement leaves arrow types alone and assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$.
-It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement\todo{is admissible the right word}.
-Encouragingly, this judgement also arises in the study of gradual type systems\todo{first paper?}.
+It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement.
+Encouragingly, this judgement also arises in the study of gradual type systems \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.
 
@@
 Indeed, Siek and Taha needed two application rules for the same fundamental reason \cite{Siek06a}.
-Subsequent work on gradual typing introduced this notion of type matching\todo{cite}.
+Later work on gradual typing introduced this notion of type matching to resolve this redundancy.
 
@@
 The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on.
-All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the focus will be found.
+All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the cursor will be found.
 Any other sub-term is ``dotted'', i.e.
 it is either an H-type or H-expression.
-Taken together, every Z-type and Z-expression contains exactly one focused H-type or H-expression by construction.
+Taken together, every Z-type and Z-expression contains exactly one selected H-type or H-expression by construction.
 This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)
@@
 The rules for relative movement within Z-expressions are similarly straightforward.
-Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:\todo{remove prevSib in supplement too}
+Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:
 \begin{subequations}
@@
 There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define a syntax for lists of actions, $\bar{\alpha}$, together with iterated action judgements.
-To keep the core of the action semantics small, we take the latter approach in Figure \ref{fig:multistep}.\todo{put stars on these?}
-
-A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.\todo{define this in supplement}
+To keep the core of the action semantics small, we take the latter approach in Figure \ref{fig:multistep}.
+
+A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.
 
@@
 \end{equation}
-The type ascription is necessary because lambda expressions do not synthesize a type.
-If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
+The type ascription is necessary because lambda expressions do not synthesize a type. If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:
 \begin{equation}\label{rule:performAna-lam-1}
@@
 Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
-is inserted inside a hole, which defers the type inconsistency as previously discussed:\todo{supp}
+is inserted inside a hole, which defers the type inconsistency as previously discussed:
 \begin{equation}\label{rule:performAna-construct-lam-2}
@@
 
-\paragraph{Application} The $\aConstruct{\fap}$ action applies the expression under the cursor. The following rule handles the case where the synthesized type has matched function type:\todo{supp}
+\paragraph{Application} The $\aConstruct{\fap}$ action applies the expression under the cursor. The following rule handles the case where the synthesized type has matched function type:
 \begin{equation}
@@
 \end{equation}
-If the empty hole is being analyzed against a type that is incompatible with $\tnum$, then we must place the number expression inside the hole:
+If the empty hole is being analyzed against a type that is inconsistent with $\tnum$, then we must place the number expression inside the hole:
 \begin{equation}
@@
 
-\todo{make sure there is no more compatibility or focus words}If the non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the enveloped expression is consistent with the type that the hole is being analyzed against.
+If the non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the enveloped expression is consistent with the type that the hole is being analyzed against.
 This amounts to analyzing the enveloped expression against the provided type (by subsumption):
@@
 \begin{subequations}
-In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):\todo{supp}
+In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):
 \begin{equation}
@@
 
-When the focus is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
+When the cursor is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
 \begin{equation}\label{rule:zipper-asc}

### 2016-07-07 hazelnut-popl17.tex
Lemmas and theorems use the same counter
reachability
progress on construction and subsumption
tweaked title linebreak spacing; updated colors and stroke for hazel
final sec 4
final popl paper + supplement

--- 2016-07-05 hazelnut-popl17.tex
@@
 \newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
+\newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{corollary}{Corollary}
@@
 \usepackage{enumitem}
+\makeatletter
+\def\thm@space@setup{%
+  \thm@preskip=4px plus 2px minus 2px
+  \thm@postskip=\thm@preskip % or whatever, if you don't want them to be equal
+}
+\makeatother
 
@@
  Calculus}
+% \title{Hazelnut:\\A Bidirectionally Typed Structure Editor Calculus}
 
@@
 
-This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{focus} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: it's {edit actions} operate over statically meaningful terms.
+This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: it's {edit actions} operate over statically meaningful (i.e.
+well-typed) terms.
 Na\"ively, this prohibition on ill-typed edit states would force the programmer to construct terms in a rigid ``outside-in'' manner.
@@
 
-\clearpage
+% \clearpage
 
@@
 
-\section{Related Work}\label{sec:rw}
+\section{Related Work and Discussion}\label{sec:rw}
 %\subsection{Structure Editors}
-
-\todo{TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}}
-
-Structured editing has been recognized as a way to avoid the possibility of syntax errors for decades.  An early example is the
-The Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981}, first published in 1981.
-The synthesizer generator~\cite{Reps:1984:SG:390010.808247} allows the user to create an attribute-grammar specification that then can be used to generate a structured editor.
-CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language.
-Barista \cite{ko_barista:_2006} is a modern take on the same basic concept.%These early systems were developed  of the systems are rooted in the type-theoretic tradition.
-
-Novice programmers have been a common target for structure editors. For example,
-GNOME\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates.
+\subsection{Structure Editors}
+Syntactic structure editors have a long history -- the Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in 1981.
+Novice programmers have been a common target for structure editors.
+For example,
+GNOME~\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates. 
+Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students.
 Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.
-Touchdevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students.
-Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students.
-These are largely drag-and-drop user interfaces with a limited action model and an unclear semantics.
+TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students.
+An implementation of Hazelnut might be useful in teaching students about the typed lambda calculus, though that has not been our explicit aim with this work.
 
 Not all structure editors are for educational purposes. For example,
-mbeddr \cite{voelter_mbeddr:_2012} is an extensible C-based Programming Language and IDE (nominally, for programming embedded systems.)
-mbeddr is build on top of the commercial JetBrains MPS framework for constructing structure editors.
-Another popular approach is to bring elements of structured editing into a traditional editor.
-Codelets \cite{oney_codelets:_2012} uses structured editing to add interactive documentation and examples in an editor.
-Our previous work on Graphite~\cite{Omar:2012:ACC:2337223.2337324} allows developers to associate structured editing interfaces called  \emph{palettes} with types.
-Graphite is integrated into a text-based program editor (Eclipse.)
+mbeddr \cite{voelter_mbeddr:_2012} is a structure editor for a C-based programming language (nominally, for programming embedded systems.)  
+Lamdu~\cite{lamdu}, like Hazelnut, is a structure editor for a statically typed functional language. It is designed for use by professional programmers.
 
-Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g.
+The examples given so far either define a language with a trivial static semantics, or do not attempt to maintain well-typedness as an edit invariant.
+This can pose problems, for reasons discussed in the Introduction.
+One apparent exception is Unison~\cite{unison}, a structure editor for a typed functional language similar to Haskell.
+Like Hazelnut, it seems to define some notion of well-typedness for expressions with holes (though there is no technical documentation on virtually any aspect of its design.) Unlike Hazelnut, it does not have a notion analagous to Hazelnut's notion of a non-empty hole.
+As such, programmers must construct programs in a rigid outside-in manner, as discussed in Sec.
+\ref{sec:example}.
+Another system with the same problem is Polymorphic Blocks, a block-based user interface where the structure of block connectors encodes a type \cite{DBLP:conf/chi/LernerFG15}.
+
+We fundamentally differ from these projects in our design philosophy: we consider it essential to start by building type theoretic foundations, which are independent of nearly all decisions about the user interface.
+In contrast, these editors have developed innovative user interfaces (e.g.
+see the discussion in \cite{DBLP:conf/sle/VolterSBK14}) but lack a principled foundational calculus.
+In this respect, we follow the philosophical approach taken by languages that are rooted in the type theoretic tradition and have gone to great effort to develop a clear metatheory, like Standard ML \cite{mthm97-for-dart,Harper00atype-theoretic}.
+In the future, we hope that these lines of research will merge to produce a human-usable typed structure editor with sound formal foundations.
+Our contribution, then, is in defining and analyzing the theoretical properties of a small foundational calculus that could be extended to achieve this vision.
+
+Some structure editor generators do operate on formal or semi-formal definitions of an underlying language.
+For example, the Synthesizer Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an attribute grammar-based language implementation that then can be used to generate a structured editor.
+CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language.
+Barista is a programmatic toolkit for building structure editors \cite{ko_barista:_2006}.
+mbeddr is build on top of the commercial JetBrains MPS framework for constructing structure editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}.
+These systems do not give a semantics to the edit states of the structure editor itself, or maintain non-trivial edit invariants, as Hazelnut does.
+%These early systems were developed  of the systems are rooted in the type-theoretic tradition.
+
+Related to structure editors are value editors, which operate directly on simple values (but not generally expressions or functions) of a programming language.
+For example, Eros is a typed value editor based in Haskell \cite{DBLP:conf/icfp/Elliott07}.
+
+\subsection{Gradual Type Systems}
+A significant contribution of this paper is the discovery of a clear technical relationship between typed structure editing and gradual typing.
+In particular, the machinery necessary to give a reasonable semantics to type holes -- i.e.
+type consistency and type matching -- coincides with that developed in the study of gradual type systems for functional languages.
+The pioneering work of Siek and Taha \cite{Siek06a} introduced type consistency.
+Subsequent work developed the concept of type matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}.
+In retrospect, this relationship is perhaps unsurprising: gradual typing is, notionally, also motivated by the idea of iterated development of a program where every intermediate state is well-defined in some sense.
+%Typed structure editing endows this intuitive notion with technical force.
+
+Recent work has discovered a systematic procedure for generating a ``gradual version'' of a standard type system \cite{DBLP:conf/popl/CiminiS16}.
+This system, called the Gradualizer, operates on a logic program that specifies a simple type assignment system with some additional annotations to generate a corresponding specification of a gradual type system.
+The authors leave support for working with bidirectional type systems as future work.
+This suggests the possibility of an analagous ``Editorializer'' that generates a specification of a typed structure editor from a simple language definition.
+Our exposition in Sec.
+\ref{sec:extending} certainly suggests that many of the necessary definitions follow seemingly mechanically from the definition of the static semantics, and the relationship with gradual typing suggests that many of the technical details of this transformation may already exist in the Gradualizer.
+One possibility we have explored informally is to use Agda's reflection features to implement such a system.
+
+An aspect of gradual typing that we did not touch on directly here is its concern with  assigning a dynamics to programs where type information is not known, by inserting dynamic type casts.
+This would correspond to assigning a dynamics to Hazelnut expressions with type holes such that a run-time error occurs when a type hole is found to be unfillable through evaluation.
+
+\subsection{Type Holes as Unification Variables}
+Another possible interpretation of type holes is as explicit unification variables.
+In other words, we might define the dynamics of Hazelnut such that a program cannot be run if some automated type inference procedure cannot statically fill in any type holes that arise.
+For a simple calculus, e.g.
+the STLC upon which Hazelnut is based, type inference is known to be decidable \cite{damas1982principal}.
+In more complex settings, e.g.
+in a dependently typed language, a partial decision procedure may still be useful in this regard, both at edit-time and (just prior to) run-time.
+Indeed, text editor modes for proof assistants, e.g.
+for Agda, attempt to do exactly this for indicated ``type holes'' (and do not always succeed.)
+
+\subsection{Exceptions}
+Expression holes, too, could dynamically be interpreted in several ways.
+One straightforward approach would be to dynamically raise an exception if evaluation encounters an expression hole.
+Indeed, placing \textt{raise Unimplemented} or similar in portions of an expression that are under construction is a common practice across programming languages.
+
+\subsection{Type-Directed Program Synthesis}
+Some text editor modes, e.g.
+those for proof assistants like Agda \cite{norell:thesis} and Idris \cite{brady2013idris}, support a more explicit hole-based programming model where indicated expression holes are treated as sites where the system can be asked to automatically generate an expression of an appropriate type.
+These systems are not statically well-defined themselves (though see below.)
+
+The Graphite system used an informal model of typed holes in Java to allow library providers to associate interactive code generation interfaces with types \cite{Omar:2012:ACC:2337223.2337324}.
+%RThis system was limited in that these interfaces could not themselves manipulate terms.
+
+More generally, the topic of type-directed program synthesis an active area of research, e.g.
+\cite{DBLP:conf/pldi/OseraZ15}.
+By maintaining static well-definedness throughout the editing process, Hazelnut provides researchers interested in editor-integrated type-directed program synthesis  with a formal foundation upon which to build.% such systems could be integrated directly into editors.
+
+\subsection{Contextual Modal Type Theory}
+Expression holes can also be understood by invoking the notion of a \emph{metavariable} as found in contextual modal type theory (CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}.
+In particular, expression holes have types and are surrounded by contexts, just as metavariables in CMTT are associated with types and contexts.
+This helps to clarify the logical meaning of a typing derivation in Hazelnut -- it conveys well-typedness relative to an (implicit) modal context that extracts each expression hole's type and context.
+The modal context must be emptied -- i.e.
+the expression holes must be instantiated with expressions of the proper type in the proper context -- before the expression can be considered complete.
+This corresponds to the notion of modal necessity in contextual modal logic.
+
+Making the modal context explicit in our semantics is not technically useful given our stated purpose -- interactive program editing is not merely hole filling in Hazelnut (i.e.
+the cursor need not be on a hole).
+Moreover, the hole's type and context become apparent as our action semantics traverses the zipper structure on each step.
+Some interactive proof assistants, however, support a tactic model based directly on hole filling, so the connection to CMTT and similar systems is more useful.
+For example,  Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT and aspects of Idris' editor support \cite{brady2013idris} are based on McBride's OLEG \cite{mcbride2000dependently} and Lee and Friedman have explored a lambda calculus with contexts for a similar purpose \cite{DBLP:conf/icfp/LeeF96}.
+
+One interesting avenue of future work is to elaborate expression holes to CMTT's closures, i.e.
+CMTT terms of the form $\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable associated with each hole and $\text{id}(\Gamma)$ is the explicit identity substitution.
+This would allow us to evaluate expressions with holes such that the closure ``accumulates'' substitutions explicitly.
+When evaluation gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped with a notion of progress under a non-empty modal context), it would then be possible for the programmer to choose holes from amongst the visible holes (which may have been duplicated) to edit in their original context.
+Once finished, the CMTT hole instantiation operation, together with a metatheorem that establishes that reduction commutes with instantiation, would enable an ``edit and resume'' feature with a clear formal basis.
+This notion of reduction commuting with instantiation has also been studied in other calculi \cite{DBLP:journals/entcs/Sands97}.
+Being able to edit a running program also has connections to less formal work on ``live programming'' interfaces \cite{burckhardt2013s,lamdu}.
+%, e.g.
+
+\subsection{Evaluation Strategies: A High-Dimensional Space}
+To summarize, we have discussed three different evaluation strategies in the presence of type holes:
+\begin{enumerate}[noitemsep]
+\item ...as preventing evaluation (the standard approach.)
+\item ...as unknown types, in the gradual sense.
+\item ...as unification variables.
+\end{enumerate}
+In addition, we have discussed four different evaluation strategies in the presence of expression holes:
+\begin{enumerate}[noitemsep]
+\item ...as preventing evaluation (the standard approach.)
+\item ...as causing exceptions.
+\item ...as sites for automatic program synthesis.
+\item ...as the closures of CMTT.
+\end{enumerate}
+
+Every combination of these choices could well be considered in the design of a full-scale programming system in the spirit of Hazelnut.
+Indeed, the user might be given several options from among these combinations, depending on their usage scenario.
+Many of these warrant further inquiry.
+
+% \todo{Eclipse parsing partial programs // other heuristics}
+
+% \paragraph{Gradual typing: Has Matched Relation}
+% %
+% A. Rastogi, A. Chaudhuri, and B. Hosmer. The ins and outs of gradual type inference.
+% \\
+% Definition 3.1 is similar to our ``has matched'' relations:
+% \\
+% \url{https://www.cs.umd.edu/~avik/papers/iogti.pdf}
+
+% This paper defines ``has matched'' relation for arrow, just like us:
+% \\
+% \url{http://cimini.info/publications/Gradualizer_Draft.pdf}
+% \\
+% They credit this technique to the following paper:
+% \\
+% \url{http://www.cs.ubc.ca/~rxg/ptsgp.pdf}
+% \\
+% But, the Rastogi paper came earlier.
+
+% \todo{TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}}
+
+% Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g.
 Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes.
@@
 
-Perhaps the systems most similar in spirit to Hazelnut are Lamdu~\cite{lamdu} and Unison~\cite{unison}.
-Like Hazelnut, these are both statically typed functional language editors.
-In both cases, the language is similar to Haskell.
-In Lamdu, the editor uses structure editing to enable Live Programming, where the code is always being executed as it is being written.
-
-Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified.
+% Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified.
 For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model.
@@
 
-\todo{TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}}
+% \todo{TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}}
 
-\todo{TODO: cite} \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}
+% \todo{TODO: cite} \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}
 
@@
 
-\section{Discussion \& Conclusion}
+\section{Conclusion}
 \label{sec:future}
-\todo{revise}This paper presented Hazelnut, a type theoretic structure editor calculus.
-Our aim is to take a principled approach to its design by formally specifying its semantics, providing strong metatheoretic guarantees, mechanizing its semantics and metatheory in Agda and implementing it using the concepts of  functional reactive programming.
-As of this submission, we have achieved reasonable confidence in the formal system presented above, and have transitioned our focus toward the mechanization and implementation efforts.
-By the time of presentation, we anticipate having complete or nearly complete versions of these.
+This paper presented Hazelnut, a type theoretic structure editor calculus.
+Our aim was to take a principled approach to its design by formally defining its static semantics as well as its action semantics and developing a rich metatheory.
+Moreover, we have mechanized substantial portions of the metatheory, including the crucial Sensibility theorem that establishes that every edit state is statically meaningful.
 
-In addition to eliminating malformed edit states, structure editors also promise to increase the speed of development.
-However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
-whether a keyboard or a mouse is in use.\todo{integrate this}
+In addition to simplifying the job of an editor designer, typed structure editors also promise to increase the speed of development by eliminating redundant syntax and supporting higher-level primitive actions.
+However, we did not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
+whether a keyboard or a mouse is in use.
+Indeed, we consider it a virtue of this work that such implementation details do not enter into our design.
 
-By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
-By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?
+% By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
+% By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?
 In progress, unfinished, incomple?} and yet still in a well defined type state.
-This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}
+% This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}
 
 \subsection{Future Work}
+\subsubsection{Richer Languages}
 Hazelnut is, obviously, a very limited language at its core.
-So the most obvious avenue for future work is to increase the expressive power of this language.
-Our plan is to simultaneously maintain a mechanization and implementation (following, for example, Standard ML) as we proceed, ultimately producing the first large-scale, formally verified bidirectionally typed language codesigned with a type-aware editor.
+So the most obvious avenue for future work is to increase the expressive power of this language by extension.
+Our plan is to simultaneously maintain a mechanization and implementation (following, for example, Standard ML) as we proceed, ultimately producing the first large-scale, formally verified bidirectionally typed structure editor.
+
+It is interesting to note that the demarcation between the language and the editor is fuzzy (indeed, non-existent) in Hazelnut.
+There may well be interesting opportunities in language design when the language is being codesigned with a typed structure editor.
 It may be that certain language features are unnecessary given a sufficiently advanced type-aware structure editor (e.g.
@@
 
+\subsubsection{Editor Services}
 There are various aspects of the editor model that we have not yet formalized.
 For example, our action model does not consider how actions are actually entered using, for example, key combinations or chords.
-It also did not provide any specific model of how available actions will be determined for presentation to the user.
-In practice, we would want also to rank available actions in some reasonable manner (perhaps based on usage data gathered from other users or code repositories.)
+In practice, we would want also to rank available actions in some reasonable manner (perhaps based on usage data gathered from other users or code repositories.) Designing a rigorous typed probability model over actions and H-expressions is one avenue of research that we have started to explore, with intriguing initial results.
 
+\subsubsection{Programmable Actions}
+Our language of actions is intentionally primitive.
+However, even now it acts much like a simple imperative command language.
+This suggests future expansion to, for example, a true tactic language.
+Alternatively, it may be more useful to develop the notion of an \emph{action macro}, whereby functional programs could themselves be lifted to the level of actions to compute non-trivial compound actions.
+
+\subsubsection{Views}
 Another research direction is in exploring how types can be used to control the presentation of expressions in the editor.
-For example, following our approach in a textual setting on \emph{type-specific languages} (TSLs), it should be possible to have the type that an expression is being analyzed against define alternative display forms and interaction modes \cite{TSLs}.
+For example, following an approach developed in a textual setting of developing \emph{type-specific languages} (TSLs), it should be possible to have the type that an expression is being analyzed against define alternative display forms and interaction modes \cite{TSLs}.
 
+It should also be possible to develop a semantics of semantic comments, i.e.
+comments that mention semantic structures.
+These would be subject to the same operations, e.g.
+renaming, as other structures, helping to address the problem of comments becoming inconsistent with code.
+
+\subsubsection{Collaborative Programming}
 Finally, we did not consider any aspects of \emph{collaborative programming}, such as a packaging system, a differencing algorithm for use in a source control system, support for multiple simultaneous focii for different users, and so on.
@@
 
-On the theoretical side, the notion of having one of many possible holes in a term in focus has a very strong intuitive connection
-  to the proof theoretic notion of focusing \cite{Simmons11tr}. Beyond just
-  the name, both seem to involve, in some sense, a search through the space of possible
-  ways to finish a derivation. We intend to explore this connection to see
-  if it's coincidental or more meaningful and welcome insights in this regard.
+\subsubsection{Empirical Evaluation}
+Although we make few empirical claims in this paper, it is ultimately an empirical question as to whether structure editors, and typed structure editors, are practical.
+We hope to conduct user studies once a richer semantics has been developed.
 
-We already discussed a connection to gradual typing \cite{Siek06a}.
+\subsubsection{More Theory}
+Connections with gradual type systems and CMTT, discussed in the previous section, seem likely to continue to be revealing.
+
+The notion of having one of many possible locations within a term under a cursor has a very strong intuitive connection to the proof theoretic notion of focusing \cite{Simmons11tr}.
+Building closer connections with proof theory (and category theory) is likely to be a fruitful avenue of further inquiry.
+
+% We already discussed a connection to gradual typing \cite{Siek06a}.
 We hope to explore this connection more thoroughly.
@@
 
-It may also be possible to give a dynamics to incomplete expressions.
+% It may also be possible to give a dynamics to incomplete expressions.
 Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}.
 In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
-$\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
-explored an equational theory for terms with holes, but intend to once our
-formalization effort is more mature.\todo{fix bib}
+% $\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
+% explored an equational theory for terms with holes, but intend to once our
+% formalization effort is more mature.\todo{fix bib}
 
-\todo{new action form that makes actions extensible given proofs of
-admissiblity of a derived form, like prevSib}
+% \todo{new action form that makes actions extensible given proofs of
+% admissiblity of a derived form, like prevSib}
 

### 2016-07-07 implementation.tex
final popl paper + supplement

@@
 In order to reach a wide audience, we decided to implement HZ in the web browser.
-In order to take advantage of a mature implementation of the FRP model, we chose to implement HZ using OCaml\footnote{https://ocaml.org/}, the \texttt{js\_of\_ocaml} compiler and associated libraries\footnote{http://ocsigen.org/js\_of\_ocaml/} and the OCaml React library\footnote{http://erratique.ch/software/react}.
+In order to take advantage of a mature implementation of the FRP model, we chose to implement HZ using OCaml\footnote{https://ocaml.org/}, the \texttt{js\_of\_ocaml} compiler and associated libraries \cite{DBLP:conf/ml/Balat06}\footnote{http://ocsigen.org/js\_of\_ocaml/} and the OCaml React library\footnote{http://erratique.ch/software/react}.
 
 At the time of the writing, 
-HZ's view computation renders the model as a string embedded into HTML.
+HZ's view computation renders the model as a string embedded into HTML that approximates our typeset notation in this paper.
 The action palette is a collection of buttons and text boxes, which are disabled when an action is not well-defined or an invalid action argument is entered.
-We determine this by simply attempting to run the action internally and catching the exception that is raised when an action is undefined.
-This, of course, is not meant to score marks for usability or performance.
+We determine this by simply attempting to perform the action internally and catching the exception that is raised when an action is undefined.
+This design, of course, is not meant to score marks for usability or performance.
 Instead, as a simple reference implementation, we expect its source code to be of use to others who are interested in layering a more fluid user interface atop the core semantics (which have been implemented in a style that follows the presentation in this paper closely), or extensions thereof.

### 2016-07-08 hazelnut-popl17.tex

@@
 Barista is a programmatic toolkit for building structure editors \cite{ko_barista:_2006}.
-mbeddr is build on top of the commercial JetBrains MPS framework for constructing structure editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}.
+mbeddr is built on top of the commercial JetBrains MPS framework for constructing structure editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}.
 These systems do not give a semantics to the edit states of the structure editor itself, or maintain non-trivial edit invariants, as Hazelnut does.

### 2016-07-13 formally.tex
added missing rule numbers + cleaned up bibliography

@@
 Our overview below omits certain ``uninteresting'' details.
-The supplement includes the complete collection of rules, in definitional order.
+The supplement includesfffadfasfaasdff the complete collection of rules, in definitional order.
 These rules, along with the proofs of two key metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement.
@@
 \subsection{H-types and H-expressions}\label{sec:holes}
-
 Figure \ref{fig:hexp-syntax} defines the syntax of H-types, $\htau$, and H-expressions, $\hexp$.
@@
 % \end{equation}
+\begin{subequations}\label{eqns:consistency}
 \begin{mathpar}
+% \begin{equation}
 \inferrule{ }{
   \tcompat{\tehole}{\htau}
-}
-
-% \end{equation}
+}%~\text{(1a)}
+% \end{equation}
+% \end{minipage}
+% \begin{minipage}{0.15\linewidth}
 % \begin{equation}%\label{rule:tcompat-hole}
+~~~~~~~~
 \inferrule{ }{
   \tcompat{\htau}{\tehole}
-}
-
+}%~\text{(1b)}
+% \end{equation}
+% \end{minipage}
+% \begin{minipage}{0.15\linewidth}
+% \begin{equation}
+~~~~~~~~
 \inferrule{ }{
   \tcompat{\htau}{\htau}
-}
-
+}%~\text{(1c)}
+% \end{equation}
+% \end{minipage}
+% \begin{minipage}{0.15\linewidth}
+% \begin{equation}
+~~~~~~~~
 \inferrule{
@@
   \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-}
+}~~~~~~~~\text{(\ref*{eqns:consistency}a-d)}
+% \end{equation}
+% \end{minipage}
 \end{mathpar}
+\end{subequations}
 % \begin{equation}%\label{rule:tcompat-num}
@@
 \fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
-\begin{mathpar}
-% \begin{subequations}
-% \begin{equation}
+\begin{subequations}
+\begin{minipage}{0.5\linewidth}
+\begin{equation}
 \inferrule{ }{
@@
 }
-
-% \end{equation}
-% \begin{equation}
+\end{equation}
+\end{minipage}
+\begin{minipage}{0.5\linewidth}
+\begin{equation}
 \inferrule{ }{
@@
 }
-% \end{equation}
+\end{equation}
+\end{minipage}
 % \end{subequations}
-\end{mathpar}
+% \end{mathpar}
 % \noindent\fbox{$\tincompat{\htau}{\htau'}$}
@@
 %   \end{equation}
-% \end{subequations}
+\end{subequations}
 \caption{H-type consistency and matched arrow types.}
@@
 Before giving the rules for these judgements, let us state the key metatheorem: \emph{sensibility}.
-This metatheorem deeply inform the design of the rules, given starting in Sec.
+This metatheorem deeply informs the design of the rules, given starting in Sec.
 \ref{sec:action-subsumption}.

### 2016-07-13 formally.tex
corrected typo
added stars to iterated action judgements and gave them rule numbers

@@
 Our overview below omits certain ``uninteresting'' details.
-The supplement includesfffadfasfaasdff the complete collection of rules, in definitional order.
+The supplement includes the complete collection of rules, in definitional order.
 These rules, along with the proofs of two key metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement.
@@
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\removeSel{\ztau}=\removeSel{\ztau'}$ then there exists some $\bar\alpha$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performTyp{\ztau}{\bar\alpha}{\ztau'}$.
-\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$.
-\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zexp'}{\htau}$.
+\item If $\removeSel{\ztau}=\removeSel{\ztau'}$ then there exists some $\bar\alpha$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performTypI{\ztau}{\bar\alpha}{\ztau'}$.
+\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$.
+\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zexp'}{\htau}$.
 \end{enumerate}
@@
 $\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\vspace{4px}\\
-\fbox{$\performTyp{\ztau}{\bar{\alpha}}{\ztau'}$}
-\vspace{-10px}\begin{mathpar}
-\inferrule{ }{
-    \performTyp{\ztau}{\cdot}{\ztau}
-}
-
+\fbox{$\performTypI{\ztau}{\bar{\alpha}}{\ztau'}$}
+
+\vspace{-10px}\begin{subequations}
+\begin{minipage}{0.35\linewidth}
+\begin{equation}
+\inferrule{ }{
+    \performTypI{\ztau}{\cdot}{\ztau}
+}
+\end{equation}
+\end{minipage}
+\begin{minipage}{0.65\linewidth}
+\begin{equation}
 \inferrule{
   \performTyp{\ztau}{\alpha}{\ztau'}\\
-  \performTyp{\ztau'}{\bar{\alpha}}{\ztau''}
-}{
-  \performTyp{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
-}
+  \performTypI{\ztau'}{\bar{\alpha}}{\ztau''}
+}{
+  \performTypI{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
+}
+\end{equation}
+\end{minipage}
+\end{subequations}
+
+\fbox{$\performSynI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
+\vspace{-10px}
+
+\begin{subequations}\label{rules:iterated-syn}
+\hspace{-12px}
+% \begin{minipage}{0.88\linewidth}
+\begin{mathpar}
+\inferrule{ }{
+  \performSynI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
+}
+~~~~~~
+\inferrule{
+  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
+  \performSynI{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
+}{
+  \performSynI{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
+}
+~~~~~
+\text{(\ref*{rules:iterated-syn}a-b)}
 \end{mathpar}
-\fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
-\vspace{-10px}\begin{mathpar}
-\inferrule{ }{
-  \performSyn{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
-}
-
-\inferrule{
-  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
-  \performSyn{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
-}{
-  \performSyn{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
-}
+% \end{minipage}
+% \begin{minipage}{0.12\linewidth}
+% (11a-b)
+% \end{minipage}
+\end{subequations}
+
+\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
+\vspace{-12px}
+\begin{subequations}\label{rules:iterated-ana}
+\begin{mathpar}
+~~~~~~~\inferrule{ }{
+  \performAnaI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
+}
+~~~~~~~~~~~~~~
+\inferrule{
+  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
+  \performAnaI{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
+}{
+  \performAnaI{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
+}
+~~~~~~~~~~
+\text{(\ref*{rules:iterated-ana}a-b)}
 \end{mathpar}
-\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
-\vspace{-10px}\begin{mathpar}
-\inferrule{ }{
-  \performAna{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
-}
-
-\inferrule{
-  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
-  \performAna{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
-}{
-  \performAna{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
-}
-\end{mathpar}
+\end{subequations}
 \caption{Iterated Action Judgements}
@@
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
-\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$.
-\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
+\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTypI{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
+\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$.
+\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
 \end{enumerate}
@@
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
-\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
-\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ nand $\performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
+\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTypI{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
+\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
+\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ nand $\performSynI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
 \end{enumerate}
@@
 \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item For every $\htau$ there exists $\bar\alpha$ such that $\performTyp{\zwsel{\tehole}}{\bar\alpha}{\zwsel{\htau}}$.
-\item If $\hana{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that $\performAna{\hGamma}{\zwsel{\hhole{}}}{\htau}{\bar\alpha}{\zwsel{\hexp}}$.
-\item If $\hsyn{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that $\performSyn{\hGamma}{\zwsel{\hhole{}}}{\tehole}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
+\item For every $\htau$ there exists $\bar\alpha$ such that $\performTypI{\zwsel{\tehole}}{\bar\alpha}{\zwsel{\htau}}$.
+\item If $\hana{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that: $$\performAnaI{\hGamma}{\zwsel{\hhole{}}}{\htau}{\bar\alpha}{\zwsel{\hexp}}$$
+\item If $\hsyn{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that: $$\performSynI{\hGamma}{\zwsel{\hhole{}}}{\tehole}{\bar\alpha}{\zwsel{\hexp}}{\htau}$$
 \end{enumerate}

### 2016-07-13 hazelnut-popl17.tex
added missing rule numbers + cleaned up bibliography

@@
 %\documentclass[12pt]{article}
+\RequirePackage{etex}
 \documentclass[preprint,9pt]{sigplanconf}
@@
 \usepackage{wasysym}
+\usepackage{multicol}
 \usepackage{extarrows}
@@
 \begin{quote}
-In any case, these are but steps toward more graphical program-description
-systems, for we will not forever stay confined to mere strings of symbols.
+\emph{In any case, these are but steps toward more graphical program-description
+systems, for we will not forever stay confined to mere strings of symbols.}
 
---- Marvin Minsky, Turing Award lecture
-\end{quote}% We recommend abbrvnat bibliography style.
+--- Marvin Minsky, Turing Award lecture \cite{DBLP:journals/jacm/Minsky70}
+\end{quote}
+
 \clearpage
+% We recommend abbrvnat bibliography style.
 \bibliographystyle{abbrvnat}
@@
 % The bibliography should be embedded for final submission.
-
 \bibliography{bibliography}

### 2016-09-08 hazelnut-popl17.tex
making a TR version
fixed mis-labeling of rules
added vspace back to author list to make things line up nicer /twitch
its / its correction

@@
 \def\OPTIONConf{1}%
+%\newif\iftr\trfalse
+\newif\iftr\trtrue
 %\documentclass[12pt]{article}
@@
 \usepackage{rulelinks}
+
+%\usepackage{balance} %% Balance last page's two columns
 
@@
 %\authorinfo{\vspace{-2px}}{}{}
-\authorinfo{~}{~}{\vspace{-10px}}
+%\authorinfo{~}{~}{\vspace{-10px}}
 %\authorinfo{Cyrus Omar \and Jonathan Aldrich}
@@
 %         {\{comar, aldrich\}@cs.cmu.edu}
+\iftr
+\authorinfo{
+        Cyrus Omar$^1$
+        \and Ian Voysey$^1$
+        \and Michael Hilton$^2$
+        \and Jonathan Aldrich$^1$
+        \and Matthew A. Hammer$^3$
+}
+{
+ $^1$~Carnegie Mellon University
+ \and
+ $^2$~Oregon State University
+ \and
+ $^3$~University of Colorado Boulder\vspace{-10px}
+}
+{
+%\email{comar@cs.cmu.edu}\\
+%\texttt{http://www.cs.cmu.edu/\homedir comar/}
+%\email{hiltonm@eecs.oregonstate.edu}\\
+}
+\else
+\authorinfo{~}{~}{\vspace{-10px}}
+\fi
+
+%
+%\authorrunning{Omar et al.} % abbreviated author list (for running head)
+%
+%%%% list of authors for the TOC (use if author list has to be modified)
+%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
+%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
+%
+%\email{Matthew.Hammer@colorado.edu}
 
@@
 
-This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: it's {edit actions} operate over statically meaningful (i.e.
+This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: its {edit actions} operate over statically meaningful (i.e.
 well-typed) terms.
@@
 
+\iftr
+\clearpage
+%\onecolumn
+\appendix
+\input{appendix}
+\else
+% No Appendix Here
+\fi
+
 \end{document}

### 2016-09-18 hazelnut-popl17-rebuttal.txt
rebuttal

We thank all of the reviewers for their thoughtful questions, comments and suggestions.  In this response, we first address the “questions for authors to address” in Reviews B and C.  Then, after the break marked below, we address other issues raised in Reviews B and D.

Review B

“Rule(11c)--Are you guaranteed in this case that x is already in the context? Does this mean that you have to introduce the binding location for a variable before you can refer to the variable?  Is that overly restrictive in practice? Why not allow holes in the context?”

Yes, the variable x must be in the context, due to the sensibility invariant (though the type of a variable in the context can be any H-type, including the hole type.) It would certainly be interesting to extend the language of H-expressions to assign meaning to terms where variables occur free. This would then require tweaking the rules in the action semantics that introduce bindings (e.g. those for construct lam x) to ensure that the introduced binding is at a type consistent with its use. We will mention this as future work!

Review C

“Please address my question about the orthogonality of the navigation system and the type system above.”

It is indeed the case that movement actions are type-independent (Sec 3.3.4), and that they could straightforwardly be generated automatically from a term signature, as suggested in Sec. 6.2. Such a system would need to come equipped with a metatheorem that guaranteed that the generated movement action rules collectively obey the Movement Erasure Invariance and Reachability theorems established by this paper.

With regard to a “cursor free version” of Hazelnut, note a crucial distinction between reduction and action performance: one eventually wants reduction to occur everywhere it is possible, but each action should not eventually be performed everywhere it may be possible. The notion of a locus of action is therefore essential to the notion of a program editor. It would be possible to define this notion using some structure other than a zipper. For example, we could define the notion of a path into an H-expression. With such a model, the movement actions would not touch the H-expression. However, this could complicate the specification of the other actions. We would be happy to discuss our choice of cursor representation more thoroughly in the final paper.

”To what extent can one precompile or partially evaluate in the presence of holes?”

Section 6.6 discussed the possibility of defining a dynamic (operational) semantics for incomplete H-expressions based on CMTT. Evaluation could proceed until a hole appeared in an elimination position. Compilation relies on a theory of evaluation, so it could be considered after developing this theory. It would also be interesting to explore the intersection of incremental computation and our action semantics -- many of the computations necessary for compiling a term could be reused when compiling the term that arises from performing an action on it.

-----------------------------------------------------
Below, we address more questions and issues raised in the reviews (this section goes beyond 500 words):

Review B

“Proofs of reachability and constructabilty have not yet been fully mechanized”

As promised in Sec. 3.4 of the submission, the last remaining cases of these theorems have been mechanized in the repository linked in the non-anonymous supplement. No fundamentally new insight was necessary. (The sum type extension has also been mechanized.)

“Do you need actions for renaming bound variables?”

It would be possible to define such actions primitively, but we envision higher-level actions like these as being “derived actions” in the sense that they can be defined in terms of the “core” action language that this paper defines. In particular, renaming a bound variable could be “internally” accomplished by deletion and reconstruction. In Sec. 7.1.3, we discuss “action macros” as future work -- they could be used to implement refactoring features like this. We will clarify this point.

“Why does the construct arrow make the type at the cursor location into the argument type?  Why not the result type?  Why not have two commands for introducing function types, one where the cursor into the argument and a second into the return?  This approach would be consistent with what you did for function application.” 
and
“Similarly for plus.  Why does the construct plus command put the argument in the left hand side?  Because of the symmetry of addition?”

Good points! The decision to leave the expression under the cursor on the left is formally an arbitrary one. We made the choice based on our intuition that programs are typically written “left to right”, i.e. in a textual language one typically writes the left side of an arrow type or addition before writing down “->” or “+” and then the right side. It would be possible to include a variant of these actions that made the opposite decision, by analogy to the construct arg action. Alternatively, some or all of these could be derived actions as described above. We will take care to indicate which aspects of our calculus are based on informal intuitions like this in the final version.

“The deletion rule for Z-expressions in analysis changes not just the expression being deleted but also the inferred type.  Does this “action outside the hole” cause any difficulties?”

The deletion rule in analytic position does change the type itself, but the new type -- being the hole type -- is consistent with every type, so there is no difficulty (indeed, if there were, the corresponding case of Sensibility would fail to go through.)

“I don’t follow what you mean by “the resulting cursor erasure.” It seems like that would be type \tau.’ (the type inferred for the cursor eraser), but you are actually checking that \tau. is consistent with \tau.’’ not \tau.’.  What am I missing?  (Sorry if that was difficult to follow!)”

By “the resulting cursor erasure” we mean “the cursor erasure of the edit state resulting from performing synthetic action \alpha” (we will use this wording in the final version). For Sensibility to hold, we need that the type provided for analysis, \tau., is consistent with the type synthesized from the resulting edit state, \tau.’’ (the type synthesized from the initial edit state, \tau.’, is irrelevant.)

Review D

“How does this scale to polymorphism?”

Higher-rank polymorphism in a bidirectional system is possible, e.g. see Dunfield and Krishnaswami 2013. We see no fundamental impediment to building a structure editor in the style of Hazelnut that supports that mechanism. The introduction of type variables would require lifting to the type level some of the mechanisms related to binding and non-empty holes that were developed at the expression level in this paper.

“Must the source language type system be bidirectional?”

We agree that this question is worthy of more thorough discussion in the paper!

Replacing the bidirectional type system with one that performs HM-style type inference would be tricky, because type reconstruction typically assumes a complete expression. In intermediate edit states, then, one would not always have enough information to definitively determine the types of variables nor the expected type of an expression. At best, then, we could attempt to constrain the action semantics to allow only those actions consistent with presently known constraints. However, later edits would then have non-local implications with respect to well-typedness. This could possibly be addressed by a “whole-program” pass on each edit that put newly ill-typed sub-terms into holes. However, from a user perspective, this would require error messages at each such hole. The problem of coming up with usable error messages when type inference fails is notoriously difficult.

In contrast, in a bidirectional system the locality of inference precludes “spooky errors at a distance”. Instead, the interaction is a sort of local dialog between the programmer and the system involving simple, familiar concepts -- types with holes -- rather than sets of constraints.

“Must the source language be designed with structure editing in mind?”

Sec. 6.2 discusses related work on the Gradualizer, which covers how to go from a non-gradually-typed language to one that is gradual. The transition from a language without holes to one with holes should be similar, i.e. it should not generally require anticipation by the language designer, and could conceivably be automated to a significant extent.

(That said, we are also excited by the possibilities of _intentionally_ codesigning a language with its structure editor as discussed in Sec. 7.1.1.)

“Is it strictly necessary to have a universal subsumption rule that works for any expression?  This seems to cause you trouble with determinism in 3.4.1.  You may be better off proving this as an admissible rule of the entire system rather than including it as a rule itself.  That would lead to some restrictions on a constructor-by-constructor basis, but that seems more fundamentally palatable to me.”

We would not be able prove that the subsumption rule as stated is admissible in a deterministic variant of our system because its inclusion is what breaks determinism, as you inferred.

We have explored an alternative definition of subsumption that does maintain determinism by explicitly excluding the two situations (mentioned in the paper) where both a synthetic and an analytic rule can fire. This is likely to be more economical than attempting to positively characterize the situations where subsumption should be permitted.

Another approach that we are exploring in our mechanization is to define a predicate over action performance derivations that excludes those that apply subsumption prematurely. Only those derivations that satisfy this predicate are the subject of the determinism theorem.

We believe that a clear description of the issue is quite valuable, so we do not plan to attempt to settle on any one of these solutions in the final paper. Note that in practice, the question of whether to apply subsumption is straightforward -- it should only be applied as a rule of last resort (see our implementation.)

“why not have a synthesis rule for lambda that gives x the hole type and synthesizes the type of the body?  What goes wrong?”

This would be a valid extension of the calculus, but a more conventional approach would be to introduce “half annotated” lambdas as described by Chlipala et al. in their paper “Strict Bidirectional Type Checking” (TLDI 2005). We left these out for simplicity.

“lambdas are the only constructor in your theory with no synthesis rule (even injections for sums have them)”

Correction: there is no type synthesis rule for injections. The synthetic action rule for injections inserts an ascription, like the synthetic action rule for lambdas.

“Is the assumption that the typing derivations are kept around by the editor and then modified?  Or is it just an optimization to only re-check if certain edit actions are applied?”

Only those action rules that have a typing premise need to invoke the typechecker, and then only on subtrees of the program (see our implementation). Keeping around the types of subtrees would allow for further optimization even beyond this.

### 2016-09-24 formally.tex
parens for type consistency
adding parenthesis for the inductive rule in the typesetting of the ~
judgement (both in the appendix and the main text). this resolves
possible operator precedence questions before they arise (this came up
a couple times in feedback i got for a practice talk)

@@
 }{
-  \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
+  \tcompat{(\tarr{\htau_1}{\htau_2})}{(\tarr{\htau_1'}{\htau_2'})}
 }~~~~~~~~\text{(\ref*{eqns:consistency}a-d)}
@@
 
-Hazelnut's static semantics is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,bidi-tutorial} around the two mutually defined judgements in Figure \ref{fig:ana-syn}.
+Hazelnut's static semantics is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial} around the two mutually defined judgements in Figure \ref{fig:ana-syn}.
 Typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner.

### 2016-09-24 hazelnut-popl17.tex
working on #6

@@
 \documentclass[preprint,9pt]{sigplanconf}
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
 \usepackage[T1]{fontenc}
@@
 
-%\usepackage{balance} %% Balance last page's two columns
-
 \newtheorem{theorem}{Theorem}
@@
 
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
 % known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
+\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}
 
@@
 \usepackage{xcolor}
-%\usepackage{adjustbox}
 
@@
 
-%\titlebanner{}        % These are ignored unless
 \preprintfooter{Draft}   % 'preprint' option specified.
@@
  Calculus}
-% \title{Hazelnut:\\A Bidirectionally Typed Structure Editor Calculus}
-
-%\authorinfo{\vspace{-2px}}{}{}
-%\authorinfo{~}{~}{\vspace{-10px}}
-%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
-%         {Carnegie Mellon University}
-%         {\{comar, aldrich\}@cs.cmu.edu}
 \iftr
@@
 
-%
-%\authorrunning{Omar et al.} % abbreviated author list (for running head)
-%
-%%%% list of authors for the TOC (use if author list has to be modified)
-%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
-%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
-%
-%\email{Matthew.Hammer@colorado.edu}
-
 \maketitle
 \begin{abstract}
-% Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker.
-Not all text survives this journey.
-In particular, text that arises transiently, or when the programmer makes a mistake, is often malformed or ill-typed.
-% Not all text that arises during the programming process survives the journey through a parser and typechecker.
-In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
-% Contending with malformed program text or well-formed but meaningless syntax trees is difficult for programmers and their tools alike.
 
-%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
-% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}.
-They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
-% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.
+\emph{Structure editors} allow programmers to edit the tree structure of
+a program directly. This can have cognitive benefits, particularly for
+novice and end-user programmers (as evidenced by the popularity of
+structure editors like Scratch.) It also simplifies matters for tool
+designers, because they do not need to contend with malformed program text.
 
-\emph{Structure editors} allow programmers to edit the tree structure of a program directly.
-This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they do not need to contend with malformed program text.
+This paper defines Hazelnut, a {structure editor} based on a small
+bidirectionally typed lambda calculus extended with \emph{holes} and
+a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond
+syntactic well-formedness: its {edit actions} operate over statically
+meaningful (i.e. well-typed) terms.  Na\"ively, this prohibition on
+ill-typed edit states would force the programmer to construct terms in a
+rigid ``outside-in'' manner. To avoid this problem, the {action semantics}
+automatically places terms assigned a type that is inconsistent with the
+expected type {inside} a {hole}. This safely defers the type consistency
+check until the term inside the hole is \emph{finished}.
 
-This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: its {edit actions} operate over statically meaningful (i.e.
-well-typed) terms.
-Na\"ively, this prohibition on ill-typed edit states would force the programmer to construct terms in a rigid ``outside-in'' manner.
-To avoid this problem, the {action semantics} automatically places terms assigned a type that is inconsistent with the expected type {inside} a {hole}.
-This safely defers the type consistency check until the term inside the hole is \emph{finished}.
+Hazelnut is a foundational type-theoretic account of typed structure
+editing, rather than an end-user tool itself. To that end, we describe how
+Hazelnut's rich metatheory, which we have mechanized in Agda, guides the
+definition of an extension to the calculus. We also discuss various
+plausible evaluation strategies for terms with holes, and in so doing
+reveal connections with gradual typing and contextual modal type theory
+(the Curry-Howard interpretation of contextual modal logic.) Finally, we
+discuss how Hazelnut's semantics lends itself to implementation as a
+functional reactive program. Our reference implementation is written
+using \lstinline{js_of_ocaml}.
 
-Hazelnut is a foundational type-theoretic account of typed structure editing, rather than an end-user tool itself.
-To that end, we describe how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to the calculus.
-We also discuss various plausible evaluation strategies for terms with holes, and in so doing reveal connections with gradual typing and contextual modal type theory (the Curry-Howard interpretation of contextual modal logic.) Finally, we  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program.
-Our reference implementation is written using \lstinline{js_of_ocaml}.
-
-%Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
 \end{abstract}
@@
 
-% \clearpage
-
-% \section{Dynamics of Hazelnut}\label{sec:dynamics}
-
-% \subsection{Complete Expressions}
-% ... canonical forms and type safety with e complete in it ...
-
-% \subsection{Holes as TODOs}
-% ... add an e err and define new type safety ...
-
-% \subsection{Indeterminate Expressions}
-% \input{dynamics}
-
-% \clearpage
-
 \section{Extending Hazelnut}\label{sec:extending}
-%% \todo{write this section} Add sum types.
-%% H-expression
-%% Z-expression + Erasure rules
-%% Statics
-%% Movement actions
-%% Construct actions
 \input{sumtypes}
@@
 \section{Related Work and Discussion}\label{sec:rw}
-%\subsection{Structure Editors}
 \subsection{Structure Editors}
-Syntactic structure editors have a long history -- the Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in 1981.
-Novice programmers have been a common target for structure editors.
-For example,
-GNOME~\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates.
-Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students.
-Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.
-TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students.
-An implementation of Hazelnut might be useful in teaching students about the typed lambda calculus, though that has not been our explicit aim with this work.
+Syntactic structure editors have a long history -- the Cornell Program
+Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in
+1981. Novice programmers have been a common target for structure
+editors. For example, GNOME~\cite{garlan_gnome:_1984} was developed to
+teach programming to undergraduates.
+Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language
+with an integrated structure editor for teaching novice CS undergraduate
+students. Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure
+editor targeted at children ages 8 to 16.
+TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure
+editor for programming on touch-based devices, and is used to teach high
+school students. An implementation of Hazelnut might be useful in teaching
+students about the typed lambda calculus, though that has not been our
+explicit aim with this work.
 
 Not all structure editors are for educational purposes. For example,
-mbeddr \cite{voelter_mbeddr:_2012} is a structure editor for a C-based programming language (nominally, for programming embedded systems.)
-Lamdu~\cite{lamdu}, like Hazelnut, is a structure editor for a statically typed functional language. It is designed for use by professional programmers.
+mbeddr \cite{voelter_mbeddr:_2012} is a structure editor for a C-based
+programming language (nominally, for programming embedded systems.)
+Lamdu~\cite{lamdu}, like Hazelnut, is a structure editor for a statically
+typed functional language. It is designed for use by professional
+programmers.
 
-The examples given so far either define a language with a trivial static semantics, or do not attempt to maintain well-typedness as an edit invariant.
-This can pose problems, for reasons discussed in the Introduction.
-One apparent exception is Unison~\cite{unison}, a structure editor for a typed functional language similar to Haskell.
-Like Hazelnut, it seems to define some notion of well-typedness for expressions with holes (though there is no technical documentation on virtually any aspect of its design.) Unlike Hazelnut, it does not have a notion analagous to Hazelnut's notion of a non-empty hole.
-As such, programmers must construct programs in a rigid outside-in manner, as discussed in Sec.
-\ref{sec:example}.
-Another system with the same problem is Polymorphic Blocks, a block-based user interface where the structure of block connectors encodes a type \cite{DBLP:conf/chi/LernerFG15}.
+The examples given so far either define a language with a trivial static
+semantics, or do not attempt to maintain well-typedness as an edit
+invariant. This can pose problems, for reasons discussed in the
+Introduction. One apparent exception is Unison~\cite{unison}, a structure
+editor for a typed functional language similar to Haskell. Like Hazelnut,
+it seems to define some notion of well-typedness for expressions with holes
+(though there is no technical documentation on virtually any aspect of its
+design.) Unlike Hazelnut, it does not have a notion analagous to Hazelnut's
+notion of a non-empty hole. As such, programmers must construct programs in
+a rigid outside-in manner, as discussed in Sec. \ref{sec:example}. Another
+system with the same problem is Polymorphic Blocks, a block-based user
+interface where the structure of block connectors encodes a
+type \cite{DBLP:conf/chi/LernerFG15}.
 
-We fundamentally differ from these projects in our design philosophy: we consider it essential to start by building type theoretic foundations, which are independent of nearly all decisions about the user interface.
-In contrast, these editors have developed innovative user interfaces (e.g.
-see the discussion in \cite{DBLP:conf/sle/VolterSBK14}) but lack a principled foundational calculus.
-In this respect, we follow the philosophical approach taken by languages that are rooted in the type theoretic tradition and have gone to great effort to develop a clear metatheory, like Standard ML \cite{mthm97-for-dart,Harper00atype-theoretic}.
-In the future, we hope that these lines of research will merge to produce a human-usable typed structure editor with sound formal foundations.
-Our contribution, then, is in defining and analyzing the theoretical properties of a small foundational calculus that could be extended to achieve this vision.
+We fundamentally differ from these projects in our design philosophy: we
+consider it essential to start by building type theoretic foundations,
+which are independent of nearly all decisions about the user interface. In
+contrast, these editors have developed innovative user interfaces (e.g. see
+the discussion in \cite{DBLP:conf/sle/VolterSBK14}) but lack a principled
+foundational calculus. In this respect, we follow the philosophical
+approach taken by languages that are rooted in the type theoretic tradition
+and have gone to great effort to develop a clear metatheory, like Standard
+ML \cite{mthm97-for-dart,Harper00atype-theoretic}.  In the future, we hope
+that these lines of research will merge to produce a human-usable typed
+structure editor with sound formal foundations. Our contribution, then, is
+in defining and analyzing the theoretical properties of a small
+foundational calculus that could be extended to achieve this vision.
 
-Some structure editor generators do operate on formal or semi-formal definitions of an underlying language.
-For example, the Synthesizer Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an attribute grammar-based language implementation that then can be used to generate a structured editor.
-CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language.
-Barista is a programmatic toolkit for building structure editors \cite{ko_barista:_2006}.
-mbeddr is built on top of the commercial JetBrains MPS framework for constructing structure editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}.
-These systems do not give a semantics to the edit states of the structure editor itself, or maintain non-trivial edit invariants, as Hazelnut does.
-%These early systems were developed  of the systems are rooted in the type-theoretic tradition.
+Some structure editor generators do operate on formal or semi-formal
+definitions of an underlying language. For example, the Synthesizer
+Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an
+attribute grammar-based language implementation that then can be used to
+generate a structured editor. CENTAUR~\cite{Borras:1988:CS:64140.65005}
+produces a language specific environment from a user defined formal
+specification of a language. Barista is a programmatic toolkit for building
+structure editors \cite{ko_barista:_2006}. mbeddr is built on top of the
+commercial JetBrains MPS framework for constructing structure
+editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}. These
+systems do not give a semantics to the edit states of the structure editor
+itself, or maintain non-trivial edit invariants, as Hazelnut does.
 
-Related to structure editors are value editors, which operate directly on simple values (but not generally expressions or functions) of a programming language.
-For example, Eros is a typed value editor based in Haskell \cite{DBLP:conf/icfp/Elliott07}.
+Related to structure editors are value editors, which operate directly on
+simple values (but not generally expressions or functions) of a programming
+language. For example, Eros is a typed value editor based in
+Haskell \cite{DBLP:conf/icfp/Elliott07}.
 
 \subsection{Gradual Type Systems}
-A significant contribution of this paper is the discovery of a clear technical relationship between typed structure editing and gradual typing.
-In particular, the machinery necessary to give a reasonable semantics to type holes -- i.e.
-type consistency and type matching -- coincides with that developed in the study of gradual type systems for functional languages.
-The pioneering work of Siek and Taha \cite{Siek06a} introduced type consistency.
-Subsequent work developed the concept of type matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}.
-In retrospect, this relationship is perhaps unsurprising: gradual typing is, notionally, also motivated by the idea of iterated development of a program where every intermediate state is well-defined in some sense.
+A significant contribution of this paper is the discovery of a clear
+technical relationship between typed structure editing and gradual
+typing. In particular, the machinery necessary to give a reasonable
+semantics to type holes -- i.e. type consistency and type matching --
+coincides with that developed in the study of gradual type systems for
+functional languages. The pioneering work of Siek and Taha \cite{Siek06a}
+introduced type consistency. Subsequent work developed the concept of type
+matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}. In
+retrospect, this relationship is perhaps unsurprising: gradual typing is,
+notionally, also motivated by the idea of iterated development of a program
+where every intermediate state is well-defined in some sense.
 %Typed structure editing endows this intuitive notion with technical force.
 
-Recent work has discovered a systematic procedure for generating a ``gradual version'' of a standard type system \cite{DBLP:conf/popl/CiminiS16}.
-This system, called the Gradualizer, operates on a logic program that specifies a simple type assignment system with some additional annotations to generate a corresponding specification of a gradual type system.
-The authors leave support for working with bidirectional type systems as future work.
-This suggests the possibility of an analagous ``Editorializer'' that generates a specification of a typed structure editor from a simple language definition.
-Our exposition in Sec.
-\ref{sec:extending} certainly suggests that many of the necessary definitions follow seemingly mechanically from the definition of the static semantics, and the relationship with gradual typing suggests that many of the technical details of this transformation may already exist in the Gradualizer.
-One possibility we have explored informally is to use Agda's reflection features to implement such a system.
+Recent work has discovered a systematic procedure for generating a
+``gradual version'' of a standard type
+system \cite{DBLP:conf/popl/CiminiS16}. This system, called the
+Gradualizer, operates on a logic program that specifies a simple type
+assignment system with some additional annotations to generate a
+corresponding specification of a gradual type system. The authors leave
+support for working with bidirectional type systems as future work. This
+suggests the possibility of an analagous ``Editorializer'' that generates a
+specification of a typed structure editor from a simple language
+definition. Our exposition in Sec. \ref{sec:extending} certainly suggests
+that many of the necessary definitions follow seemingly mechanically from
+the definition of the static semantics, and the relationship with gradual
+typing suggests that many of the technical details of this transformation
+may already exist in the Gradualizer. One possibility we have explored
+informally is to use Agda's reflection features to implement such a system.
 
-An aspect of gradual typing that we did not touch on directly here is its concern with  assigning a dynamics to programs where type information is not known, by inserting dynamic type casts.
-This would correspond to assigning a dynamics to Hazelnut expressions with type holes such that a run-time error occurs when a type hole is found to be unfillable through evaluation.
+An aspect of gradual typing that we did not touch on directly here is its
+concern with assigning a dynamics to programs where type information is not
+known, by inserting dynamic type casts. This would correspond to assigning
+a dynamics to Hazelnut expressions with type holes such that a run-time
+error occurs when a type hole is found to be unfillable through evaluation.
 
 \subsection{Type Holes as Unification Variables}
-Another possible interpretation of type holes is as explicit unification variables.
-In other words, we might define the dynamics of Hazelnut such that a program cannot be run if some automated type inference procedure cannot statically fill in any type holes that arise.
-For a simple calculus, e.g.
-the STLC upon which Hazelnut is based, type inference is known to be decidable \cite{damas1982principal}.
-In more complex settings, e.g.
-in a dependently typed language, a partial decision procedure may still be useful in this regard, both at edit-time and (just prior to) run-time.
-Indeed, text editor modes for proof assistants, e.g.
-for Agda, attempt to do exactly this for indicated ``type holes'' (and do not always succeed.)
+Another possible interpretation of type holes is as explicit unification
+variables. In other words, we might define the dynamics of Hazelnut such
+that a program cannot be run if some automated type inference procedure
+cannot statically fill in any type holes that arise. For a simple calculus,
+e.g. the STLC upon which Hazelnut is based, type inference is known to be
+decidable \cite{damas1982principal}. In more complex settings, e.g. in a
+dependently typed language, a partial decision procedure may still be
+useful in this regard, both at edit-time and (just prior to)
+run-time. Indeed, text editor modes for proof assistants, e.g. for Agda,
+attempt to do exactly this for indicated ``type holes'' (and do not always
+succeed.)
 
 \subsection{Exceptions}
-Expression holes, too, could dynamically be interpreted in several ways.
-One straightforward approach would be to dynamically raise an exception if evaluation encounters an expression hole.
-Indeed, placing \textt{raise Unimplemented} or similar in portions of an expression that are under construction is a common practice across programming languages.
+Expression holes, too, could dynamically be interpreted in several
+ways. One straightforward approach would be to dynamically raise an
+exception if evaluation encounters an expression hole. Indeed,
+placing \textt{raise Unimplemented} or similar in portions of an expression
+that are under construction is a common practice across programming
+languages.
 
 \subsection{Type-Directed Program Synthesis}
-Some text editor modes, e.g.
-those for proof assistants like Agda \cite{norell:thesis} and Idris \cite{brady2013idris}, support a more explicit hole-based programming model where indicated expression holes are treated as sites where the system can be asked to automatically generate an expression of an appropriate type.
-These systems are not statically well-defined themselves (though see below.)
+Some text editor modes, e.g. those for proof assistants like
+Agda \cite{norell:thesis} and Idris \cite{brady2013idris}, support a more
+explicit hole-based programming model where indicated expression holes are
+treated as sites where the system can be asked to automatically generate an
+expression of an appropriate type. These systems are not statically
+well-defined themselves (though see below.)
 
-The Graphite system used an informal model of typed holes in Java to allow library providers to associate interactive code generation interfaces with types \cite{Omar:2012:ACC:2337223.2337324}.
-%RThis system was limited in that these interfaces could not themselves manipulate terms.
+The Graphite system used an informal model of typed holes in Java to allow
+library providers to associate interactive code generation interfaces with
+types \cite{Omar:2012:ACC:2337223.2337324}.
 
-More generally, the topic of type-directed program synthesis an active area of research, e.g.
-\cite{DBLP:conf/pldi/OseraZ15}.
-By maintaining static well-definedness throughout the editing process, Hazelnut provides researchers interested in editor-integrated type-directed program synthesis  with a formal foundation upon which to build.% such systems could be integrated directly into editors.
+More generally, the topic of type-directed program synthesis an active area
+of research, e.g. \cite{DBLP:conf/pldi/OseraZ15}. By maintaining static
+well-definedness throughout the editing process, Hazelnut provides
+researchers interested in editor-integrated type-directed program synthesis
+with a formal foundation upon which to build.
 
 \subsection{Contextual Modal Type Theory}
-Expression holes can also be understood by invoking the notion of a \emph{metavariable} as found in contextual modal type theory (CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}.
-In particular, expression holes have types and are surrounded by contexts, just as metavariables in CMTT are associated with types and contexts.
-This helps to clarify the logical meaning of a typing derivation in Hazelnut -- it conveys well-typedness relative to an (implicit) modal context that extracts each expression hole's type and context.
-The modal context must be emptied -- i.e.
-the expression holes must be instantiated with expressions of the proper type in the proper context -- before the expression can be considered complete.
-This corresponds to the notion of modal necessity in contextual modal logic.
+Expression holes can also be understood by invoking the notion of
+a \emph{metavariable} as found in contextual modal type theory
+(CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}. In particular, expression
+holes have types and are surrounded by contexts, just as metavariables in
+CMTT are associated with types and contexts. This helps to clarify the
+logical meaning of a typing derivation in Hazelnut -- it conveys
+well-typedness relative to an (implicit) modal context that extracts each
+expression hole's type and context. The modal context must be emptied --
+i.e. the expression holes must be instantiated with expressions of the
+proper type in the proper context -- before the expression can be
+considered complete. This corresponds to the notion of modal necessity in
+contextual modal logic.
 
-Making the modal context explicit in our semantics is not technically useful given our stated purpose -- interactive program editing is not merely hole filling in Hazelnut (i.e.
-the cursor need not be on a hole).
-Moreover, the hole's type and context become apparent as our action semantics traverses the zipper structure on each step.
-Some interactive proof assistants, however, support a tactic model based directly on hole filling, so the connection to CMTT and similar systems is more useful.
-For example,  Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT and aspects of Idris' editor support \cite{brady2013idris} are based on McBride's OLEG \cite{mcbride2000dependently} and Lee and Friedman have explored a lambda calculus with contexts for a similar purpose \cite{DBLP:conf/icfp/LeeF96}.
+Making the modal context explicit in our semantics is not technically
+useful given our stated purpose -- interactive program editing is not
+merely hole filling in Hazelnut (i.e. the cursor need not be on a
+hole). Moreover, the hole's type and context become apparent as our action
+semantics traverses the zipper structure on each step. Some interactive
+proof assistants, however, support a tactic model based directly on hole
+filling, so the connection to CMTT and similar systems is more useful. For
+example, Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT
+and aspects of Idris' editor support \cite{brady2013idris} are based on
+McBride's OLEG \cite{mcbride2000dependently} and Lee and Friedman have
+explored a lambda calculus with contexts for a similar
+purpose \cite{DBLP:conf/icfp/LeeF96}.
 
-One interesting avenue of future work is to elaborate expression holes to CMTT's closures, i.e.
-CMTT terms of the form $\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable associated with each hole and $\text{id}(\Gamma)$ is the explicit identity substitution.
-This would allow us to evaluate expressions with holes such that the closure ``accumulates'' substitutions explicitly.
-When evaluation gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped with a notion of progress under a non-empty modal context), it would then be possible for the programmer to choose holes from amongst the visible holes (which may have been duplicated) to edit in their original context.
-Once finished, the CMTT hole instantiation operation, together with a metatheorem that establishes that reduction commutes with instantiation, would enable an ``edit and resume'' feature with a clear formal basis.
-This notion of reduction commuting with instantiation has also been studied in other calculi \cite{DBLP:journals/entcs/Sands97}.
-Being able to edit a running program also has connections to less formal work on ``live programming'' interfaces \cite{burckhardt2013s,lamdu}.
-%, e.g.
+One interesting avenue of future work is to elaborate expression holes to
+CMTT's closures, i.e. CMTT terms of the form
+$\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable
+associated with each hole and $\text{id}(\Gamma)$ is the explicit identity
+substitution. This would allow us to evaluate expressions with holes such
+that the closure ``accumulates'' substitutions explicitly. When evaluation
+gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped
+with a notion of progress under a non-empty modal context), it would then
+be possible for the programmer to choose holes from amongst the visible
+holes (which may have been duplicated) to edit in their original
+context. Once finished, the CMTT hole instantiation operation, together
+with a metatheorem that establishes that reduction commutes with
+instantiation, would enable an ``edit and resume'' feature with a clear
+formal basis. This notion of reduction commuting with instantiation has
+also been studied in other
+calculi \cite{DBLP:journals/entcs/Sands97}. Being able to edit a running
+program also has connections to less formal work on ``live programming''
+interfaces \cite{burckhardt2013s,lamdu}.
 
 \subsection{Evaluation Strategies: A High-Dimensional Space}
-To summarize, we have discussed three different evaluation strategies in the presence of type holes:
+To summarize, we have discussed three different evaluation strategies in
+the presence of type holes:
 \begin{enumerate}[noitemsep]
@@
 \end{enumerate}
-In addition, we have discussed four different evaluation strategies in the presence of expression holes:
+In addition, we have discussed four different evaluation strategies in the
+presence of expression holes:
 \begin{enumerate}[noitemsep]
@@
 
-Every combination of these choices could well be considered in the design of a full-scale programming system in the spirit of Hazelnut.
-Indeed, the user might be given several options from among these combinations, depending on their usage scenario.
-Many of these warrant further inquiry.
+Every combination of these choices could well be considered in the design
+of a full-scale programming system in the spirit of Hazelnut. Indeed, the
+user might be given several options from among these combinations,
+depending on their usage scenario. Many of these warrant further inquiry.
 
@@
 \label{sec:future}
-This paper presented Hazelnut, a type theoretic structure editor calculus.
-Our aim was to take a principled approach to its design by formally defining its static semantics as well as its action semantics and developing a rich metatheory.
-Moreover, we have mechanized substantial portions of the metatheory, including the crucial Sensibility theorem that establishes that every edit state is statically meaningful.
+This paper presented Hazelnut, a type theoretic structure editor
+calculus. Our aim was to take a principled approach to its design by
+formally defining its static semantics as well as its action semantics and
+developing a rich metatheory. Moreover, we have mechanized substantial
+portions of the metatheory, including the crucial Sensibility theorem that
+establishes that every edit state is statically meaningful.
 
-In addition to simplifying the job of an editor designer, typed structure editors also promise to increase the speed of development by eliminating redundant syntax and supporting higher-level primitive actions.
-However, we did not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g.
-whether a keyboard or a mouse is in use.
-Indeed, we consider it a virtue of this work that such implementation details do not enter into our design.
+In addition to simplifying the job of an editor designer, typed structure
+editors also promise to increase the speed of development by eliminating
+redundant syntax and supporting higher-level primitive actions. However, we
+did not discuss such ``edit costs'' here, because they depend on particular
+implementation details, e.g. whether a keyboard or a mouse is in
+use. Indeed, we consider it a virtue of this work that such implementation
+details do not enter into our design.
 
-% By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.
-% By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?
-In progress, unfinished, incomple?} and yet still in a well defined type state.
-% This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}
+% By keeping the program in both a structurally and semantically
+% well-defined state at all times, Hazelnut allows users to avoid premature
+% commitment~\cite{green1996usability}.  By inserting a hole, the user can
+% leave certain parts of the program unfinished, \todo{What is the best
+% word to define this concept?  In progress, unfinished, incomple?} and yet
+% still in a well defined type state.  This also enables progressive
+% evaluation~\cite{green1996usability}, because unfinished solutions are
+% also are well-defined at all times, thus enabling tools to provide
+% evaluation of unfinished solutions.\todo{integrate this}
 
@@
 \subsubsection{Richer Languages}
-Hazelnut is, obviously, a very limited language at its core.
-So the most obvious avenue for future work is to increase the expressive power of this language by extension.
-Our plan is to simultaneously maintain a mechanization and implementation (following, for example, Standard ML) as we proceed, ultimately producing the first large-scale, formally verified bidirectionally typed structure editor.
+Hazelnut is, obviously, a very limited language at its core. So the most
+obvious avenue for future work is to increase the expressive power of this
+language by extension. Our plan is to simultaneously maintain a
+mechanization and implementation (following, for example, Standard ML) as
+we proceed, ultimately producing the first large-scale, formally verified
+bidirectionally typed structure editor.
 
-It is interesting to note that the demarcation between the language and the editor is fuzzy (indeed, non-existent) in Hazelnut.
-There may well be interesting opportunities in language design when the language is being codesigned with a typed structure editor.
-It may be that certain language features are unnecessary given a sufficiently advanced type-aware structure editor (e.g.
-SML's \texttt{open}?), while other features may only be practical with editor support.
-We intend to use Hazelnut and derivative systems thereof as a platform for rigorously exploring such questions.
+It is interesting to note that the demarcation between the language and the
+editor is fuzzy (indeed, non-existent) in Hazelnut. There may well be
+interesting opportunities in language design when the language is being
+codesigned with a typed structure editor. It may be that certain language
+features are unnecessary given a sufficiently advanced type-aware structure
+editor (e.g. SML's \texttt{open}?), while other features may only be
+practical with editor support. We intend to use Hazelnut and derivative
+systems thereof as a platform for rigorously exploring such questions.
 
 \subsubsection{Editor Services}
-There are various aspects of the editor model that we have not yet formalized.
-For example, our action model does not consider how actions are actually entered using, for example, key combinations or chords.
-In practice, we would want also to rank available actions in some reasonable manner (perhaps based on usage data gathered from other users or code repositories.) Designing a rigorous typed probability model over actions and H-expressions is one avenue of research that we have started to explore, with intriguing initial results.
+There are various aspects of the editor model that we have not yet
+formalized. For example, our action model does not consider how actions are
+actually entered using, for example, key combinations or chords. In
+practice, we would want also to rank available actions in some reasonable
+manner (perhaps based on usage data gathered from other users or code
+repositories.) Designing a rigorous typed probability model over actions
+and H-expressions is one avenue of research that we have started to
+explore, with intriguing initial results.
 
 \subsubsection{Programmable Actions}
-Our language of actions is intentionally primitive.
-However, even now it acts much like a simple imperative command language.
-This suggests future expansion to, for example, a true tactic language.
-Alternatively, it may be more useful to develop the notion of an \emph{action macro}, whereby functional programs could themselves be lifted to the level of actions to compute non-trivial compound actions.
+Our language of actions is intentionally primitive. However, even now it
+acts much like a simple imperative command language. This suggests future
+expansion to, for example, a true tactic language. Alternatively, it may be
+more useful to develop the notion of an \emph{action macro}, whereby
+functional programs could themselves be lifted to the level of actions to
+compute non-trivial compound actions.
 
 \subsubsection{Views}
-Another research direction is in exploring how types can be used to control the presentation of expressions in the editor.
-For example, following an approach developed in a textual setting of developing \emph{type-specific languages} (TSLs), it should be possible to have the type that an expression is being analyzed against define alternative display forms and interaction modes \cite{TSLs}.
+Another research direction is in exploring how types can be used to control
+the presentation of expressions in the editor. For example, following an
+approach developed in a textual setting of developing \emph{type-specific
+languages} (TSLs), it should be possible to have the type that an
+expression is being analyzed against define alternative display forms and
+interaction modes \cite{TSLs}.
 
-It should also be possible to develop a semantics of semantic comments, i.e.
-comments that mention semantic structures.
-These would be subject to the same operations, e.g.
-renaming, as other structures, helping to address the problem of comments becoming inconsistent with code.
+It should also be possible to develop a semantics of semantic comments,
+i.e. comments that mention semantic structures. These would be subject to
+the same operations, e.g. renaming, as other structures, helping to address
+the problem of comments becoming inconsistent with code.
 
 \subsubsection{Collaborative Programming}
-Finally, we did not consider any aspects of \emph{collaborative programming}, such as a packaging system, a differencing algorithm for use in a source control system, support for multiple simultaneous focii for different users, and so on.
-These are all interesting avenues for future work.
+Finally, we did not consider any aspects of \emph{collaborative
+programming}, such as a packaging system, a differencing algorithm for use
+in a source control system, support for multiple simultaneous focii for
+different users, and so on. These are all interesting avenues for future
+work.
 
 \subsubsection{Empirical Evaluation}
-Although we make few empirical claims in this paper, it is ultimately an empirical question as to whether structure editors, and typed structure editors, are practical.
-We hope to conduct user studies once a richer semantics has been developed.
+Although we make few empirical claims in this paper, it is ultimately an
+empirical question as to whether structure editors, and typed structure
+editors, are practical. We hope to conduct user studies once a richer
+semantics has been developed.
 
 \subsubsection{More Theory}
-Connections with gradual type systems and CMTT, discussed in the previous section, seem likely to continue to be revealing.
+Connections with gradual type systems and CMTT, discussed in the previous
+section, seem likely to continue to be revealing.
 
-The notion of having one of many possible locations within a term under a cursor has a very strong intuitive connection to the proof theoretic notion of focusing \cite{Simmons11tr}.
-Building closer connections with proof theory (and category theory) is likely to be a fruitful avenue of further inquiry.
+The notion of having one of many possible locations within a term under a
+cursor has a very strong intuitive connection to the proof theoretic notion
+of focusing \cite{Simmons11tr}. Building closer connections with proof
+theory (and category theory) is likely to be a fruitful avenue of further
+inquiry.
 
-% We already discussed a connection to gradual typing \cite{Siek06a}.
-We hope to explore this connection more thoroughly.
-In particular, it may be possible to better support exploratory and live programming by allowing even programs with holes in them to execute as long as those holes are only in the type portions, by deferring to the semantics given in work on gradual typing.
+% We already discussed a connection to gradual typing \cite{Siek06a}. We
+% hope to explore this connection more thoroughly. In particular, it may be
+% possible to better support exploratory and live programming by allowing
+% even programs with holes in them to execute as long as those holes are
+% only in the type portions, by deferring to the semantics given in work on
+% gradual typing.
 
-% It may also be possible to give a dynamics to incomplete expressions.
-Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}.
-In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
-% $\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
-% explored an equational theory for terms with holes, but intend to once our
-% formalization effort is more mature.\todo{fix bib}
+% It may also be possible to give a dynamics to incomplete
+% expressions. Prior work on staged evaluation suggests that there may be a
+% connection to modal logic, viewing holes as quantifying over all possible
+% terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In
+% developing a dynamic semantics, we will also need to handle terms like
+% $\hhole{\hehole}$ and $\hhole{\hhole{\hexp}}$. In our semantics given
+% here, we eliminated them as they came up in a somewhat \emph{ad hoc}
+% manner. We have not yet explored an equational theory for terms with
+% holes, but intend to once our formalization effort is more
+% mature.\todo{fix bib}
 
@@
 \clearpage
-% We recommend abbrvnat bibliography style.
+
 \bibliographystyle{abbrvnat}
-
 % The bibliography should be embedded for final submission.
 \bibliography{bibliography}
-%\softraggedright
-%P. Q. Smith, and X. Y. Jones. ...reference text...
 

### 2016-09-24 example.tex
finished 3.1 except for a few minor todos
working on #6

@@
 % !TEX root = hazelnut-popl17.tex
-The reader is encouraged to follow along with the examples that we will discuss in this section using the implementation.
+The reader is encouraged to follow along with the examples that we will
+discuss in this section using the implementation.
 
-%
 \subsection{Example 1: Constructing the Increment Function}
 
-Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from the empty hole via the indicated sequence of {actions}.
-We will introduce Hazelnut's formal syntax and define the referenced rules in Sec.
-\ref{sec:hazel}.
-First, let us build some high-level intuitions.
+Figure~\ref{fig:first-example} shows an edit sequence that constructs the
+increment function, of type $\tarr{\tnum}{\tnum}$, starting from the empty
+hole via the indicated sequence of {actions}. We will introduce Hazelnut's
+formal syntax and define the referenced rules in
+Sec. \ref{sec:hazel}. First, let us build some high-level intuitions.
 
-%In the first task (Lines 1-9), the user constructs the identity function over numbers.
-In the second task (Lines 10-19), the user applies this function (assumed to be bound to a variable, $id$), to the number expression $\hnum{3}$.
-% Each of these tasks is carried out interactively, through the sequence of \emph{actions} shown in the  column labeled \textbf{Next Action}.
-For reference, we cite the relevant rules from Sec.
-\ref{sec:hazel} in the final column.
+The edit state in Hazelnut is a {Z-expression}, $\zexp$. Every Z-expression
+has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the
+{cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively. For
+example, on Line 1, the empty expression hole, $\hhole{}$, is under the
+cursor.
 
-% The second and third columns of the
-% table show the program as it is being constructed in two forms.
-The second column shows it as an \textbf{H-expression}, which is an expression that can contain \emph{holes}, delimited by $\llparenthesis$ and $\rrparenthesis$.
-The third column shows a corresponding \textbf{Z-expression}.
-Z-expressions are H-expressions with a single focus on some sub-term, delimited by $\triangleright$ and $\triangleleft$.
-The focus need not be on a hole.
-% on working on filling just one of the holes.
-% Each action produces a new Z-expression, but this may or may not correspond to a new H-expression (in particular, some actions only move the focus, without changing the structure of the term.)
-% to the hole in
-%focus in the Z-Expression to produce the next line, which may or may not
-%produce a substantively different H-Expression.
+Actions act relative to the cursor. The first action that we perform is
+$\aConstruct{\flam{x}}$, which instantiates the hole with a lambda
+abstraction binding the variable $x$. This results in the Z-expression on
+Line 2, consisting of a lambda abstraction with an arrow type
+ascription. The cursor is placed on the argument type hole. Type holes are
+typeset like empty expression holes for visual consistency.
 
-The edit state in Hazelnut is a {Z-expression}, $\zexp$.
-Every Z-expression has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the {cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively.
-For example, on Line 1, the empty expression hole, $\hhole{}$, is under the cursor.
-
-Actions act relative to the cursor.
-The first action that we perform is $\aConstruct{\flam{x}}$, which instantiates the hole with a lambda abstraction binding the variable $x$.
-This results in the Z-expression on Line 2, consisting of a lambda abstraction with an arrow type ascription.
-The cursor is placed on the argument type hole.
-Type holes are typeset like empty expression holes for visual consistency.
-
-The actions on Lines 2-4 complete the type ascription.
-In particular, the $\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and the $\aMove{\dNext}$ action moves the cursor to the next hole.
+The actions on Lines 2-4 complete the type ascription. In particular, the
+$\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and
+the $\aMove{\dNext}$ action moves the cursor to the next hole.
 
@@
 \# & \textbf{Z-Expression} &
-%\textbf{H-Expression} & 
-% \textbf{Type} & 
 \textbf{Next Action} & \textbf{Rule}
@@
 \zwsel{\hhole{}} &
-% \hhole{} &
-% \tehole 
-% &
 \aConstruct{\flam{x}} &
@@
 \hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} &
-% \hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
-% \tarr{\tehole}{\tehole} &
 \aConstruct{\fnum{}} &
@@
 \hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
-% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
-% \tarr{\tnum}{\tehole} & 
 \aMove{\dNext{}} &
@@
 &
-% \text{\textquotedbl}&
-% \tarr{\tnum}{\tehole} & 
 \aConstruct{\fnum{}} &
@@
 \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
-% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
-% \tarr{\tnum}{\tnum} &
 \aMove{\dParent{}} &
@@
 &
-% \text{\textquotedbl}&
-% \tarr{\tnum}{\tnum} & 
 \aMove{\dParent{}} &
@@
 \\ 8 &
-% &
 \zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
-% \tarr{\tnum}{\tnum} &
-% &
 \aMove{\dChild{}} &
@@
 \\ 9 &
-% &
 \hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
-% \tarr{\tnum}{\tnum} & 
 \aConstruct{\fvar{x}} &
@@
 \\ 10 &
-% \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
-% &
 \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
-% \tarr{\tnum}{\tnum} & 
 {\aConstruct{\fplus}}
@@
 
-The actions on Lines 5-8 move the cursor to the function body.
-For simplicity, Hazelnut defines only a small collection of primitive movement actions.
-In practice, an editor would also define compound movement actions, e.g.
-an action that moves the cursor directly to the next remaining hole, in terms of these primitive movement actions.
+The actions on Lines 5-8 move the cursor to the function body. For
+simplicity, Hazelnut defines only a small collection of primitive movement
+actions. In practice, an editor would also define compound movement
+actions, e.g. an action that moves the cursor directly to the next
+remaining hole, in terms of these primitive movement actions.
 
-Lines 9-11 complete the function body by first constructing the variable $x$, then  constructing the plus form, and finally constructing the number $\hnum{1}$.
-Notice that we did not need to construct the function body in an ``outside-in'' manner, i.e.
-we constructed $x$ before constructing the outer plus form.
-Luckily, the transient function bodies, $x$ and $x + \hhole{}$, could be checked against type $\tnum$ (as we will detail in Sec.
-\ref{sec:holes}.)
+Lines 9-11 complete the function body by first constructing the variable
+$x$, then constructing the plus form, and finally constructing the number
+$\hnum{1}$. Notice that we did not need to construct the function body in
+an ``outside-in'' manner, i.e. we constructed $x$ before constructing the
+outer plus form. Luckily, the transient function bodies, $x$ and $x +
+\hhole{}$, could be checked against type $\tnum$ (as we will detail in
+Sec. \ref{sec:holes}.)
 
@@
 \colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
-% \vspace{-6px}
 $\arraycolsep=4px
 \begin{array}{|r||l|l||l|l|}
-% \\
 \hline
 \# & \textbf{Z-Expression} &
-%\textbf{H-Expression} & 
-% \textbf{Type} & 
 \textbf{Next Action} & \textbf{Rule}
@@
 \zwsel{\hhole{}} &
-% \hhole{} &
-% \tehole 
-% &
 \aConstruct{\fvar{incr}} \hphantom{~\,\,~~}&
@@
 
-Figure \ref{fig:second-example} shows an edit sequence that constructs the expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the increment function from Figure \ref{fig:first-example}.
+Figure \ref{fig:second-example} shows an edit sequence that constructs the
+expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the
+increment function from Figure \ref{fig:first-example}.
 
-We begin on Line 13 by constructing the variable $incr$.
-Line 14 then constructs the application form with $incr$ in function position, leaving the cursor on a hole in the argument position.
-Notice that we did not need to construct the outer application form before identifying the function being applied.
-Again, the transient edit state happens to be well-typed, so we needed to make no special allowances for this order of actions.
+We begin on Line 13 by constructing the variable $incr$. Line 14 then
+constructs the application form with $incr$ in function position, leaving
+the cursor on a hole in the argument position. Notice that we did not need
+to construct the outer application form before identifying the function
+being applied. Again, the transient edit state happens to be well-typed, so
+we needed to make no special allowances for this order of actions.
 
-We now need to apply $incr$ again, so we perform the same action on Line 15 as we did on Line 13, i.e.
-$\aConstruct{\fvar{incr}}$.
-In a syntactic structure editor, performing such an action would result in the following edit state:
+We now need to apply $incr$ again, so we perform the same action on Line 15
+as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic
+structure editor, performing such an action would result in the following
+edit state:
 \[
@@
 \]
-This edit state is ill-typed (after \emph{cursor erasure}): the argument of $incr$ must be of type $\tnum$ but here it is of type $\tarr{\tnum}{\tnum}$.
-Hazelnut cannot allow such an edit state to arise.
+This edit state is ill-typed (after \emph{cursor erasure}): the argument of
+$incr$ must be of type $\tnum$ but here it is of type
+$\tarr{\tnum}{\tnum}$. Hazelnut cannot allow such an edit state to arise.
 
-The programmer could alternatively have performed the $\aConstruct{\fap}$ action on Line 15.
-This would result in the following edit state, which is well-typed according to the static semantics that we will define in the next section:
+The programmer could alternatively have performed the $\aConstruct{\fap}$
+action on Line 15. This would result in the following edit state, which is
+well-typed according to the static semantics that we will define in the
+next section:
 \[
@@
 \]
-The problem is that the programmer is not able to identify the intended function before constructing the outer function application form.
-This stands in contrast to Lines 13-14.
-% This is qualitatively unnatural.
+The problem is that the programmer is not able to identify the intended
+function before constructing the function application form. This stands in
+contrast to Lines 13-14.
 
-Hazelnut's action semantics addresses this problem: rather than disallowing the $\aConstruct{\fvar{incr}}$ action on Line 15, it leaves $incr$ inside a hole:
+Hazelnut's action semantics addresses this problem: rather than disallowing
+the $\aConstruct{\fvar{incr}}$ action on Line 15, it leaves $incr$ inside a
+hole:
 \[
@@
 \]
-This defers the type consistency check, exactly as an empty hole in the same position does.
-One way to think about non-empty holes is as an internalization of the ``red underline'' that text or syntactic structure editors display to indicate a type inconsistency.
-By internalizing this concept, the presence of a type inconsistency does not leave the entire program formally meaningless.
+This defers the type consistency check, exactly as an empty hole in the
+same position does. One way to think about non-empty holes is as an
+internalization of the ``squiggly underline'' that text or syntactic
+structure editors display to indicate a type inconsistency. By
+internalizing this concept, the presence of a type inconsistency does not
+leave the entire program formally meaningless.
 
-The expression inside a non-empty hole must itself be well-typed, so the programmer can continue to edit it.
-Lines 16-17 proceed to apply the inner mention of $incr$ to a number literal, $\hnum{3}$.
-Finally, Lines 18-19 move the cursor to the non-empty hole and Line 20 performs the $\aFinish$ action.
-The $\aFinish$ action removes the hole if the type of the expression inside the hole is consistent with the expected type, as it now is.
-This results in the final edit state on Line 20, as desired.
-In practice, the editor might automatically perform the $\aFinish$ action as soon as it becomes possible, but for simplicity, Hazelnut formally requires that it be performed explicitly.
-
-% So far, editing has proceeded in an essentially type-directed, outside-in fashion -- the user first specified the type of the function, then produced a body of that type by the action on Line 8.
-Lines 10-12 similarly begin in a type-directed manner with the user giving an explicit type ascription, indicating that the expression that they are constructing will have type $\tnum$.
-
-% However, on Line 13, the user performs the $\aConstruct{\fvar{id}}$ action.
-Notice that $id$ has type $\tarr{\tnum}{\tnum}$, which is not consistent with the type $\tnum$ given in the ascription.
-Na\"ively, this would produce a type error, leaving the program in a well-formed but semantically undefined state.
-One way to avoid this state is to simply not make this action available in the program configuration on Line 12.
-This is inflexible, forcing an outside-in approach to program construction (i.e.
-the user would need to construct the function application form before constructing the variable $id$.) Instead, Hazelnut permits this action, but places the variable $id$ inside a hole.
-This defers the consistency check that would normally occur: a hole can be checked against any type, as long as its contents have some type.
-The cursor is placed inside the hole.
-The user then proceeds to apply $id$ to the number expression $\hnum{3}$.
-At this point, the expression inside the hole has a type consistent with the ascription, so the user can \emph{finish} the hole.
-In our simple formalism, this requires moving the cursor to the hole (in practice, the system might find the nearest parent of hole form.) The result is the complete, well-typed program shown on Line 19 (notice that \emph{complete} is distinct from \emph{closed} -- the variable $id$ is free on Line 19, so this is not a closed program.)
-
-%% The third column~(\textbf{Next Action}) lists the first user action:
-%% Constructing a lambda abstraction using variable~$x$.
-%% %
-%% The final column~(\textbf{Semantics}) indicates the semantic rule for this
-%% action, Rule (\ref{r:conelamhole}), which gives general semantics for
-%% introducing lambda terms into holes.
-%% %
-%% In Section~\ref{sec:hazel}, we list this rule, and the other rules used in
-%% this final column. In total, these rules give a formal semantics to the
-%% user actions, which relate each line's Z-Expression to the Z-Expression on
-%% the subsequent line.
-
-%% In addition to introducing the lambda term, and its variable, the
-%% first user action~$\aConstruct{\flam{x}}$ also introduces a type
-%% ascription for this function, as an arrow type, with holes for the
-%% type of its domain and codomain.
-%% %
-%% The actions for Lines~2--5 consist of the user filling these holes
-%% with the basetype $\tnum{}$.
-%% %
-%% To do so, the user constructs the type constructor twice (Lines 2 and
-%% 4), and navigates between the holes with a move action (Line~3).
-%% %
-%% Generally, the move action~$\dNext$ moves the focus from one
-%% sub-structure to the next sibling sub-structure of the (common) parent
-%% structure; in this case, it moves from the domain type of the arrow
-%% type to the codomain of the arrow type.
-%% %
+The expression inside a non-empty hole must itself be well-typed, so the
+programmer can continue to edit it. Lines 16-17 proceed to apply the inner
+mention of $incr$ to a number literal, $\hnum{3}$. Finally, Lines 18-19
+move the cursor to the non-empty hole and Line 20 performs the $\aFinish$
+action. The $\aFinish$ action removes the hole if the type of the
+expression inside the hole is consistent with the expected type, as it now
+is. This results in the final edit state on Line 20, as desired. In
+practice, the editor might automatically perform the $\aFinish$ action as
+soon as it becomes possible, but for simplicity, Hazelnut formally requires
+that it be performed explicitly.

### 2016-09-24 formally.tex
working out #5.
working on #6

@@
 
-The previous section introduced Hazelnut by example. In this section, we systematically introduce the following  structures:
+The previous section introduced Hazelnut by example. In this section, we
+systematically introduce the following structures:
 \begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
-\item \textbf{H-types} and \textbf{H-expressions} (Sec.
-\ref{sec:holes}), which are types and expressions with {holes}.
-H-types classify H-expressions according to Hazelnut's \textbf{bidirectional static semantics}.
-\item \textbf{Z-types} and \textbf{Z-expressions} (Sec.
-\ref{sec:cursors}), which superimpose\- a \emph{cursor} onto H-types and H-expressions, respectively (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Every Z-type (resp.
-Z-expression) corresponds to an H-type (resp.
-H-expression) by \emph{cursor erasure}.
-\item \textbf{Actions} (Sec.
-\ref{sec:actions}), which act relative to the cursor according to Hazelnut's \textbf{bidirectional action semantics}.
-The action semantics enjoys a rich metatheory.
-Of particular note, the \emph{sensibility theorem} establishes that every edit state is well-typed after cursor erasure.
+\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}),
+  which are types and expressions with {holes}. H-types classify
+  H-expressions according to Hazelnut's \textbf{bidirectional static
+    semantics}.
+
+\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}),
+  which superimpose\- a \emph{cursor} onto H-types and H-expressions,
+  respectively (following Huet's \emph{zipper pattern}
+  \cite{JFP::Huet1997}.) Every Z-type (resp. Z-expression) corresponds to
+  an H-type (resp. H-expression) by \emph{cursor erasure}.
+
+\item \textbf{Actions} (Sec. \ref{sec:actions}), which act relative to the
+  cursor according to Hazelnut's \textbf{bidirectional action
+    semantics}. The action semantics enjoys a rich metatheory. Of
+  particular note, the \emph{sensibility theorem} establishes that every
+  edit state is well-typed after cursor erasure.
 \end{itemize}
 
-Our overview below omits certain ``uninteresting'' details.
-The supplement includes the complete collection of rules, in definitional order.
-These rules, along with the proofs of two key metatheorems, have been mechanized in Agda \cite{norell:thesis}, also in the supplement.
-We will summarize this effort in Sec.
-\ref{sec:mech}.% The supplement contains the Agda sources and additional technical details.
+Our overview below omits certain ``uninteresting'' details. The supplement
+includes the complete collection of rules, in definitional order. These
+rules, along with the proofs of two key metatheorems, have been mechanized
+in Agda \cite{norell:thesis}, also in the supplement. We will summarize
+this effort in Sec. \ref{sec:mech}.
 
 \subsection{H-types and H-expressions}\label{sec:holes}
-Figure \ref{fig:hexp-syntax} defines the syntax of H-types, $\htau$, and H-expressions, $\hexp$.
-Most forms correspond directly to those of the simply typed lambda calculus (STLC) extended with a single base type, $\tnum$, of numbers (cf.
-\cite{pfpl}.) The number expression corresponding to the mathematical number $n$ is drawn $\hnum{n}$, and for simplicity, we define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$.
-The form $\hexp : \htau$ is an explicit \emph{type ascription}.
-
-In addition to these standard forms, \emph{type holes} and \emph{empty expression holes} are both drawn $\hehole$ and \emph{non-empty expression holes} are drawn $\hhole{\hexp}$.
-We do not need non-empty type holes because every H-type is a valid classifier of H-expressions.
-
-Types and expressions that contain no holes are \emph{complete types} and \emph{complete expressions}, respectively.
-Formally, we can derive $\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when $\hexp$ is complete (see supplement.) We are not concerned here with defining a dynamic semantics for Hazelnut, but the dynamics for complete H-expressions would be entirely standard (in Sec.
-\ref{sec:rw} we discuss the multi-dimensional design space around the question of evaluating incomplete H-expressions.)% The complete types and expressions correspond immediately to the types and expressions of the STLC with numbers.
-%Holes mark subterms that are, notionally, ``under construction.'' We will see what this formally corresponds to in a moment.
+Figure \ref{fig:hexp-syntax} defines the syntax of H-types, $\htau$, and
+H-expressions, $\hexp$. Most forms correspond directly to those of the
+simply typed lambda calculus (STLC) extended with a single base type,
+$\tnum$, of numbers (cf. \cite{pfpl}.) The number expression corresponding
+to the mathematical number $n$ is drawn $\hnum{n}$, and for simplicity, we
+define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$. The form
+$\hexp : \htau$ is an explicit \emph{type ascription}.
+
+In addition to these standard forms, \emph{type holes} and \emph{empty
+  expression holes} are both drawn $\hehole$ and \emph{non-empty expression
+  holes} are drawn $\hhole{\hexp}$. We do not need non-empty type holes
+because every H-type is a valid classifier of H-expressions.
+
+Types and expressions that contain no holes are \emph{complete types} and
+\emph{complete expressions}, respectively. Formally, we can derive
+$\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when
+$\hexp$ is complete (see supplement.) We are not concerned here with
+defining a dynamic semantics for Hazelnut, but the dynamics for complete
+H-expressions would be entirely standard (in Sec. \ref{sec:rw} we discuss
+the multi-dimensional design space around the question of evaluating
+incomplete H-expressions.)
 
@@
 \end{array}$
-%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
 \caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
@@
 \noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\tau$ and $\tau'$ are consistent}
-% \begin{subequations}%\label{rules:tcompat}
-% \begin{equation}%\label{rule:tcompat-comm}
-% \inferrule
-% %[TCSym]
-% {
-%   \tcompat{\htau}{\htau'}
-% }{
-%   \tcompat{\htau'}{\htau}
-% }
-% \end{equation}
 \begin{subequations}\label{eqns:consistency}
 \begin{mathpar}
-% \begin{equation}
+
 \inferrule{ }{
   \tcompat{\tehole}{\htau}
-}%~\text{(1a)}
-% \end{equation}
-% \end{minipage}
-% \begin{minipage}{0.15\linewidth}
-% \begin{equation}%\label{rule:tcompat-hole}
+}
 ~~~~~~~~
@@
   \tcompat{\htau}{\tehole}
-}%~\text{(1b)}
-% \end{equation}
-% \end{minipage}
-% \begin{minipage}{0.15\linewidth}
-% \begin{equation}
+}
 ~~~~~~~~
@@
   \tcompat{\htau}{\htau}
-}%~\text{(1c)}
-% \end{equation}
-% \end{minipage}
-% \begin{minipage}{0.15\linewidth}
-% \begin{equation}
+}
 ~~~~~~~~
@@
 }~~~~~~~~\text{(\ref*{eqns:consistency}a-d)}
-% \end{equation}
-% \end{minipage}
 \end{mathpar}
 \end{subequations}
-% \begin{equation}%\label{rule:tcompat-num}
-% \end{equation}
-% \begin{equation}%\label{rule:tcompat-arr}
-% \end{equation}
-% \end{subequations}
 \fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
@@
 \end{minipage}
-% \end{subequations}
-% \end{mathpar}
-% \noindent\fbox{$\tincompat{\htau}{\htau'}$}
-% \begin{subequations}
-%   % \begin{equation}
-%   %   \inferrule{
-%   %     \tincompat{\htau}{\htau'}
-%   %   }{
-%   %     \tincompat{\htau'}{\htau}
-%   %   }
-%   % \end{equation}
-%   \begin{equation}
-%     \inferrule{ }{
-%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tnum}
-%     }
-%   \end{equation}
-%   \begin{equation}
-%     \inferrule{ }{
-%       \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
-%     }
-%   \end{equation}
-%   \begin{equation}
-%     \inferrule{
-%       \tincompat{\htau_1}{\htau_1'}
-%     }{
-%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-%     }
-%   \end{equation}
-%   \begin{equation}
-%     \inferrule{
-%       \tincompat{\htau_2}{\htau_2'}
-%     }{
-%       \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-%     }
-%   \end{equation}
 \end{subequations}
@@
 \begin{figure}
+\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
+\begin{subequations}
+\begin{equation}\label{rule:syn-asc}
+\inferrule{
+  \hana{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hexp : \htau}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-var}
+\inferrule{ }{
+  \hsyn{\hGamma, x : \htau}{x}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-ap}
+\inferrule{
+  \hsyn{\hGamma}{\hexp_1}{\htau}\\
+  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
+  \hana{\hGamma}{\hexp_2}{\htau_2}
+}{
+  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-num}
+\inferrule{ }{
+  \hsyn{\hGamma}{\hnum{n}}{\tnum}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-plus}
+\inferrule{
+  \hana{\hGamma}{\hexp_1}{\tnum}\\
+  \hana{\hGamma}{\hexp_2}{\tnum}
+}{
+  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-ehole}
+\inferrule{ }{
+  \hsyn{\hGamma}{\hehole}{\tehole}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-hole}
+\inferrule{
+  \hsyn{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
+}
+\end{equation}
+\end{subequations}
 \fbox{$\hana{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ analyzes against $\htau$}
@@
 \end{subequations}
-\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
-\begin{subequations}
-\begin{equation}\label{rule:syn-asc}
-\inferrule{
-  \hana{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hexp : \htau}{\htau}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-var}
-\inferrule{ }{
-  \hsyn{\hGamma, x : \htau}{x}{\htau}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-ap}
-\inferrule{
-  \hsyn{\hGamma}{\hexp_1}{\htau}\\
-  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
-  \hana{\hGamma}{\hexp_2}{\htau_2}
-}{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-num}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hnum{n}}{\tnum}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-plus}
-\inferrule{
-  \hana{\hGamma}{\hexp_1}{\tnum}\\
-  \hana{\hGamma}{\hexp_2}{\tnum}
-}{
-  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-ehole}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hehole}{\tehole}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-hole}
-\inferrule{
-  \hsyn{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
-}
-\end{equation}
-\end{subequations}
-\caption{Analysis and synthesis.}
+\caption{Synthesis and analysis.}
 \label{fig:ana-syn}
@@
 
-Hazelnut's static semantics is organized as a \emph{bidirectional type system} \cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial} around the two mutually defined judgements in Figure \ref{fig:ana-syn}.
-Typing contexts, $\hGamma$, map each variable $x \in \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner.
-We identify the context up to exchange and adopt the standard identification convention for formal structures that differ only up to alpha-equivalence.
-
-Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, establish that $\hexp$ can appear where an expression of type $\htau$ is expected.
-Derivations of the type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, infer a type from $\hexp$, which is necessary in positions where an expected type is not available (e.g.
-at the top level.) Algorithmically, the type is an ``input'' of the type analysis judgement, but an ``output'' of the type synthesis judgement.
-%The rules describe a \emph{local type inference} scheme, i.e.
-type ascriptions are unnecessary when an expression is being analyzed against a known type.
-Making a judgemental distinction between these two notions will be essential in our action semantics (Sec. \ref{sec:actions}.)
-
- %We use the metavariable $\Gamma$ for \emph{complete typing contexts}, i.e. typing contexts where each hypothesis mentions only complete types.
+Hazelnut's static semantics is organized as a \emph{bidirectional type
+  system}
+\cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial}
+around the two mutually defined judgements in Figure
+\ref{fig:ana-syn}. Typing contexts, $\hGamma$, map each variable $x \in
+\domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner. We
+identify the context up to exchange and adopt the standard identification
+convention for formal structures that differ only up to alpha-equivalence.
+
+Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$,
+establish that $\hexp$ can appear where an expression of type $\htau$ is
+expected. Derivations of the type synthesis judgement,
+$\hsyn{\hGamma}{\hexp}{\htau}$, infer a type from $\hexp$, which is
+necessary in positions where an expected type is not available (e.g. at the
+top level.) Algorithmically, the type is an ``input'' of the type analysis
+judgement, but an ``output'' of the type synthesis judgement.  Making a
+judgemental distinction between these two notions will be essential in our
+action semantics (Sec. \ref{sec:actions}.)
 
 \begin{subequations}\label{rules:syn-ana}
-Type synthesis is stronger than type analysis in that if an expression is able to synthesize a type, it can also be analyzed against that type, or any other \emph{consistent} type, according to the \emph{subsumption rule}, Rule (\ref{rule:ana-subsume}).
-
-The \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$, that appears as a premise in the subsumption rule is a reflexive and symmetric (but not transitive) relation between H-types defined judgementally in Figure \ref{fig:type-consistency}.
-This relation coincides with equality for complete H-types.
-Two incomplete H-types are consistent if they differ only at positions where a hole appears in either type.
-The type hole is therefore consistent with every type.
-This notion of H-type consistency coincides with the notion of type consistency that Siek and Taha discovered in their foundational work on gradual type systems, if we interpret the type hole as the $?$ (i.e.
-unknown) type \cite{Siek06a}.
-
-Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions, $\hlam{x}{\hexp}$.
-There is no type synthesis rule that applies to this form, so lambda abstractions can appear only in analytic position, i.e.
-where an expected type is known.\footnote{It is possible to also define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$, but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.} Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
+Type synthesis is stronger than type analysis in that if an expression is
+able to synthesize a type, it can also be analyzed against that type, or
+any other \emph{consistent} type, according to the \emph{subsumption rule},
+Rule (\ref{rule:ana-subsume}).
+
+The \emph{H-type consistency judgement}, $\tcompat{\htau}{\htau'}$, that
+appears as a premise in the subsumption rule is a reflexive and symmetric
+(but not transitive) relation between H-types defined judgementally in
+Figure \ref{fig:type-consistency}. This relation coincides with equality
+for complete H-types. Two incomplete H-types are consistent if they differ
+only at positions where a hole appears in either type. The type hole is
+therefore consistent with every type. This notion of H-type consistency
+coincides with the notion of type consistency that Siek and Taha discovered
+in their foundational work on gradual type systems, if we interpret the
+type hole as the $?$ (i.e. unknown) type \cite{Siek06a}.
+
+Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions,
+$\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this
+form, so lambda abstractions can appear only in analytic position,
+i.e. where an expected type is known.\footnote{It is possible to also
+  define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$,
+  but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.}
+Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
 \begin{equation*}
@@
 \end{equation*}
-The problem is that this standard rule alone leaves us unable to analyze lambda abstractions against the type hole, because the type hole is not immediately of the form $\tarr{\htau_1}{\htau_2}$.
-There are two plausible solutions to this problem.
-One solution would be to define a second rule specifically for this case:
+The problem is that this standard rule alone leaves us unable to analyze
+lambda abstractions against the type hole, because the type hole is not
+immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible
+solutions to this problem. One solution would be to define a second rule
+specifically for this case:
 \begin{equation*}
@@
 \end{equation*}
-Intead, we combine these two possible rules into a single rule through the simple auxiliary \emph{matched arrow type} judgement, $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure \ref{fig:type-consistency}.
-This judgement leaves arrow types alone and assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$.
-It is easy to see that the two rules above are admissible by appeal to Rule (\ref{rule:syn-lam}) and the matched arrow type judgement.
-Encouragingly, this judgement also arises in the study of gradual type systems \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.
-
-Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of ascription form, $\hexp : \htau$.
-This allows the user to explicitly state  a type for the ascribed expression to be analyzed against.
+Intead, we combine these two possible rules into a single rule through the
+simple auxiliary \emph{matched arrow type} judgement,
+$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure
+\ref{fig:type-consistency}. This judgement leaves arrow types alone and
+assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It
+is easy to see that the two rules above are admissible by appeal to Rule
+(\ref{rule:syn-lam}) and the matched arrow type judgement. Encouragingly,
+this judgement also arises in the study of gradual type systems
+\cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.
+
+Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of
+ascription form, $\hexp : \htau$. This allows the user to explicitly state
+a type for the ascribed expression to be analyzed against.
 
@@
 
-Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function application.
-It also makes use of the matched function type judgement to combine what would otherwise need to be two rules for function application -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$ synthesizes the type hole.
-Indeed, Siek and Taha needed two application rules for the same fundamental reason \cite{Siek06a}.
-Later work on gradual typing introduced this notion of type matching to resolve this redundancy.
-
-Rule (\ref{rule:syn-num}) states that numbers synthesize the $\tnum$ type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves like a function over numbers.
-
-The rules described so far are sufficient to type complete H-expressions. The two remaining rules give H-expressions with holes a well-defined static semantics.
-
-Rule (\ref{rule:syn-ehole}) states that the empty expression hole synthesizes the type hole.
-
-A non-empty hole contains an H-expression that is ``under construction'', as described in Sec.
-\ref{sec:example}.
-According to Rule (\ref{rule:syn-hole}), this inner  expression must synthesize some type, but, like the empty hole,  non-empty holes synthesize the type hole.
-
-% \begin{equation}\label{rule:tcompat-comm}
-% \inferrule{
-%   \tcompat{\htau}{\htau'}
-% }{
-%   \tcompat{\htau'}{\htau}
-% }
-% \end{equation}
-% \begin{equation}\label{rule:tcompat-num}
-% \inferrule{ }{
-%   \tcompat{\tnum}{\tnum}
-% }
-% \end{equation}
-% \begin{equation}\label{rule:tcompat-arr}
-% \inferrule{
-%   \tcompat{\htau_1}{\htau_1'}\\
-%   \tcompat{\htau_2}{\htau_2'}
-% }{
-%   \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-% }
-% \end{equation}
+Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function
+application. It also makes use of the matched function type judgement to
+combine what would otherwise need to be two rules for function application
+-- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$
+synthesizes the type hole. Indeed, Siek and Taha needed two application
+rules for the same fundamental reason \cite{Siek06a}. Later work on gradual
+typing introduced this notion of type matching to resolve this redundancy.
+
+Rule (\ref{rule:syn-num}) states that numbers synthesize the $\tnum$
+type. Rule (\ref{rule:syn-plus}) states that $\hexp_1 + \hexp_2$ behaves
+like a function over numbers.
+
+The rules described so far are sufficient to type complete
+H-expressions. The two remaining rules give H-expressions with holes a
+well-defined static semantics.
+
+Rule (\ref{rule:syn-ehole}) states that the empty expression hole
+synthesizes the type hole.
+
+A non-empty hole contains an H-expression that is ``under construction'',
+as described in Sec. \ref{sec:example}. According to Rule
+(\ref{rule:syn-hole}), this inner expression must synthesize some type,
+but, like the empty hole, non-empty holes synthesize the type hole.
 
@@
 \[\hana{incr : \tarr{\tnum}{\tnum}}{\hhole{incr}}{\tnum}\]
-% This is the key premise necessary to synthesize the $\tnum$ type for the H-expression that results from cursor erasure -- defined next -- of the Z-expression on Line 16 of Fig.
-\ref{fig:second-example}.
-%In other words, this mechanism is essential if  users are to able to construct a program in anything but an ``outside in'' fashion.
-
-% The final rule handles function applications where the expression in function position synthesizes a hole type, rather than an arrow type.
-We treat it as if it had instead synthesized $\tarr{\tehole}{\tehole}$:
-% \begin{equation}\label{rule:syn-ap-2}
-% \inferrule{
-%   \hsyn{\hGamma}{\hexp_1}{\tehole}\\
-%   \hana{\hGamma}{\hexp_2}{\tehole}
-% }{
-%   \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}
-% }
-% \end{equation}
-
-% \todo{redundant text w/ citation}
-% The hole type behaves much like the type $?$ in prior work by Siek and Taha on gradual types for functional languages \cite{Siek06a}.
-Their system (which was not bidirectionally typed nor an editor model) also needed to define two rules for function application.
-In general, when a premise requires that a synthesized type be of a particular form, we need a special case where the synthesized hole type is treated instead as if it were the ``holey-est'' type of that form.\footnote{Alternatively, we might add a rule that allows expressions that synthesize hole type to then non-deterministically synthesize any other type, but maintaining determinism is useful in practice, so we avoid this approach.}
 
@@
 \mathsf{ZTyp} & \ztau & ::= &
-  %\zlsel{\htau} ~\vert~
   \zwsel{\htau} \cvert
-  %\zrsel{\htau} ~\vert~
   \tarr{\ztau}{\htau} \cvert
@@
 \mathsf{ZExp} & \zexp & ::= &
-  %\zlsel{\hexp} ~\vert~
   \zwsel{\hexp} \cvert
-  %\zrsel{\hexp} ~\vert~
   \zexp : \htau \cvert
@@
 \end{array}$
-%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
 \caption{Syntax of Z-types and Z-expressions.}
@@
 
-Hazelnut's action semantics operates over Z-types, $\ztau$, and Z-expressions, $\zexp$.
-Figure \ref{fig:zexp-syntax} defines the syntax of Z-types and Z-expressions.
-A Z-type (resp.
-Z-expression) represents an H-type (resp.
-H-expression) with a single superimposed \emph{cursor}.
-
-The only base cases in these inductive grammars are $\zwsel{\htau}$ and $\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor is on.
-All of the other forms correspond to the recursive forms in the syntax of H-types and H-expressions, and contain exactly one ``hatted'' subterm that identifies the subtree where the cursor will be found.
-Any other sub-term is ``dotted'', i.e.
-it is either an H-type or H-expression.
-Taken together, every Z-type and Z-expression contains exactly one selected H-type or H-expression by construction.
-This can be understood as an instance of Huet's \emph{zipper pattern} \cite{JFP::Huet1997} (which, coincidentally, Huet encountered while implementing a structure editor.)
-
-We write $\removeSel{\ztau}$ for the H-type constructed by erasing the cursor from $\ztau$, which we refer to as the \emph{cursor erasure} of $\ztau$.
-This straightforward metafunction is defined as follows:
+Hazelnut's action semantics operates over Z-types, $\ztau$, and
+Z-expressions, $\zexp$. Figure \ref{fig:zexp-syntax} defines the syntax of
+Z-types and Z-expressions. A Z-type (resp. Z-expression) represents an
+H-type (resp. H-expression) with a single superimposed \emph{cursor}.
+
+The only base cases in these inductive grammars are $\zwsel{\htau}$ and
+$\zwsel{\hexp}$, which identify the H-type or H-expression that the cursor
+is on. All of the other forms correspond to the recursive forms in the
+syntax of H-types and H-expressions, and contain exactly one ``hatted''
+subterm that identifies the subtree where the cursor will be found. Any
+other sub-term is ``dotted'', i.e. it is either an H-type or
+H-expression. Taken together, every Z-type and Z-expression contains
+exactly one selected H-type or H-expression by construction. This can be
+understood as an instance of Huet's \emph{zipper pattern}
+\cite{JFP::Huet1997} (which, coincidentally, Huet encountered while
+implementing a structure editor.)
+
+We write $\removeSel{\ztau}$ for the H-type constructed by erasing the
+cursor from $\ztau$, which we refer to as the \emph{cursor erasure} of
+$\ztau$. This straightforward metafunction is defined as follows:
 \begin{align*}
-%\removeSel{(\zlsel{\htau})} & = \htau\\
 \removeSel{(\zwsel{\htau})} & = \htau\\
-%\removeSel{(\zrsel{\htau})} & = \htau\\
 \removeSel{(\tarr{\ztau}{\htau})} & = \tarr{\removeSel{\ztau}}{\htau}\\
@@
 
-Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by erasing the cursor from the Z-expression $\zexp$, i.e.
-the cursor erasure of $\zexp$.
-The definition of this metafunction is analogous, so we omit it for concision (see supplement.)
-% \begin{align*}
-% %\removeSel{(\zlsel{\hexp})} & = \hexp\\
-% \removeSel{(\zwsel{\hexp})} & = \hexp\\
-% %\removeSel{(\zrsel{\hexp})} & = \hexp\\
-% \removeSel{(\zexp : \htau)} & = \removeSel{\zexp} : \htau\\
-% \removeSel{(\hexp : \ztau)} & = \hexp : \removeSel{\ztau}\\
-% \removeSel{(\hlam{x}{\zexp})} & = \hlam{x}{\removeSel{\zexp}}\\ogoo
-% \removeSel{(\hap{\zexp}{\hexp})} & = \hap{\removeSel{\zexp}}{\hexp}\\
-% \removeSel{(\hap{\hexp}{\zexp})} & = \hap{\hexp}{\removeSel{\zexp}}\\
-% \removeSel{(\hadd{\zexp}{\hexp})} & = \hadd{\removeSel{\zexp}}{\hexp}\\
-% \removeSel{(\hadd{\hexp}{\zexp})} & = \hadd{\hexp}{\removeSel{\zexp}}\\
-% \removeSel{\hhole{\zexp}} &= \hhole{\removeSel{\zexp}}
-% \end{align*}
+Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by
+erasing the cursor from the Z-expression $\zexp$, i.e. the cursor erasure
+of $\zexp$. The definition of this metafunction is analogous, so we omit it
+for concision (see supplement.)
 
@@
 
-We now arrive at the heart of Hazelnut: its \emph{bidirectional action semantics}.  
-Figure \ref{fig:action-syntax} defines the syntax of \emph{actions}, $\alpha$, some of which involve \emph{directions}, $\delta$, and \emph{shapes}, $\psi$.
-
-Expression actions are governed by two mutually defined judgements, 1) the \emph{synthetic action judgement}:
+We now arrive at the heart of Hazelnut: its \emph{bidirectional action
+  semantics}.  Figure \ref{fig:action-syntax} defines the syntax of
+\emph{actions}, $\alpha$, some of which involve \emph{directions},
+$\delta$, and \emph{shapes}, $\psi$.
+
+Expression actions are governed by two mutually defined judgements, 1) the
+\emph{synthetic action judgement}:
 \[
@@
 
-In some Z-expressions, the cursor is in a type ascription, so we also need a \emph{type action judgement}, pronounced ``performing $\alpha$ on $\ztau$ results in $\ztau'$'':
+In some Z-expressions, the cursor is in a type ascription, so we also need
+a \emph{type action judgement}, pronounced ``performing $\alpha$ on $\ztau$
+results in $\ztau'$'':
 \[
@@
 
-% The type action judgement appears as a premise in the rules that handle Z-expressions where the cursor is inside a . 
-% \[\begin{array}{ll}
-% %\textbf{Judgement Form} & \textbf{Description}\\
-% & 
-% \end{array}\]
-% \[
-% \begin{array}{l}
-% \fbox{$\performTyp{\ztau}{\alpha}{\ztau'}$} ~~ \text{Performing $\alpha$ on $\ztau$ produces $\ztau'$}\\
-% \fbox{$\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$} \\
-% \quad \text{Performing $\alpha$ on $\zexp$ when $\removeSel{\zexp}$ synthesizes type $\htau$}\\
-% \quad \text{produces $\zexp'$ such that $\removeSel{\zexp'}$ synthesizes type $\htau'$}\\
-% \fbox{$\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$} \\
-% \quad \text{Performing $\alpha$ on $\zexp$ when analyzing $\removeSel{\zexp}$ against $\htau$}\\
-% \quad \text{produces $\zexp'$, such that $\removeSel{\zexp'}$ can also be analyzed}\\
-% \quad \text{against $\htau$}
-% \end{array}\]
 \subsubsection{Sensibility}
 
-Before giving the rules for these judgements, let us state the key metatheorem: \emph{sensibility}.
-This metatheorem deeply informs the design of the rules, given starting in Sec.
-\ref{sec:action-subsumption}.
-Its proof, which is mechanized in Agda, is by straightforward induction, so the reader is encouraged to think about the relevant cases of these proofs as we present the rules.
-\begin{theorem}[Action Sensibility]\label{thrm:actsafe} Both of the following hold:\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ 
-   then
+Before giving the rules for these judgements, let us state the key
+metatheorem: \emph{sensibility}. This metatheorem deeply informs the design
+of the rules, given starting in Sec. \ref{sec:action-subsumption}. Its
+proof, which is mechanized in Agda, is by straightforward induction, so the
+reader is encouraged to think about the relevant cases of these proofs as
+we present the rules.
+\begin{theorem}[Action Sensibility]
+  \label{thrm:actsafe} Both of the following hold:
+  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ then
     $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
@@
 \end{theorem}
-\noindent In other words, if an edit state (i.e.
-a Z-expression) is statically meaningful, i.e.
-its cursor erasure is well-typed, then performing an action on it leaves the resulting edit state statically meaningful.
-In particular, the first clause of Theorem \ref{thrm:actsafe} establishes that when an action is performed on an edit state whose cursor erasure synthesizes an H-type, the result is an edit state whose cursor erasure also synthesizes some (possibly different) H-type.
-The second clause establishes that when an action is performed using the analytic action judgement on an edit state whose cursor erasure analyzes against some H-type, the result is a Z-expression whose cursor erasure also analyzes against the same H-type.
-% Non-empty holes allow us to avoid top-down program construction becau but rather can construct fragments of the program inside a hole until ready to ``expose'' them to type analysis.
+\noindent In other words, if an edit state (i.e. a Z-expression) is
+statically meaningful, i.e. its cursor erasure is well-typed, then
+performing an action on it leaves the resulting edit state statically
+meaningful. In particular, the first clause of Theorem \ref{thrm:actsafe}
+establishes that when an action is performed on an edit state whose cursor
+erasure synthesizes an H-type, the result is an edit state whose cursor
+erasure also synthesizes some (possibly different) H-type. The second
+clause establishes that when an action is performed using the analytic
+action judgement on an edit state whose cursor erasure analyzes against
+some H-type, the result is a Z-expression whose cursor erasure also
+analyzes against the same H-type.
 
@@
   \aMove{\delta} ~\vert~
-  %\aSelect{\delta} ~\vert~
   \aDel ~\vert~
-  %\aReplace{\htau} ~\vert~
-  %\aReplace{\hexp} ~\vert~
   \aConstruct{\psi} ~\vert~
@@
   \dParent ~\vert~
-  \dNext \\ % ~\vert~
-  % \dPrev\\
+  \dNext \\
 \mathsf{Shape} & \psi & ::= &
@@
 \end{array}$
-%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
 \caption{Syntax of actions.}
@@
 
-% \subsubsection{Determinism}
-% Another useful property that we maintain is \emph{action determinism}, i.e. that performing an action should produce a unique result. Formally, this is established as follows:
-% \begin{theorem}[Action Determinism] All of the following hold:
-% \label{thrm:actdet}
-% \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-% \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
-% \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
-%   $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
-%   $\zexp' = \zexp''$ and $\htau' = \htau''$.
-% % \item If all of
-
-%   \begin{quote}
-%     \begin{enumerate}
-%     \item $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$, and
-%     \item $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$, and
-%     \item $\tcompat{\htau}{\htau'}$, and
-%     \item either $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ or
-%       $\performAna{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp''}$
-%     \end{enumerate}
-%   \end{quote}
-%   hold, then $\zexp' = \zexp''$.
-% % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
-% %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
-% %   \zexp''$.
-% % \end{enumerate}
-% \end{theorem}
-
 \subsubsection{Type Inconsistency}
-In some of the rules below, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$.
-One can define this notion either directly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement.
-The supplement does the latter.
-The key rule establishes that arrow types are inconsistent with $\tnum$:
-% \begin{subequations}
-  % \begin{equation}
-  %   \inferrule{
-  %     \tincompat{\htau}{\htau'}
-  %   }{
-  %     \tincompat{\htau'}{\htau}
-  %   }
-  % \end{equation}
+In some of the rules below, we will need to supplement our definition of
+type consistency from Figure \ref{fig:type-consistency} with a definition
+of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can
+define this notion either directly as the metatheoretic negation of type
+consistency, or as a separate inductively defined judgement. The supplement
+does the latter. The key rule establishes that arrow types are inconsistent
+with $\tnum$:
   \begin{equation*}
@@
   \end{equation*}
-  % \begin{equation}
-  %   \inferrule{
-  %     \tincompat{\htau_1}{\htau_1'}
-  %   }{
-  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-  %   }
-  % \end{equation}
-  % \begin{equation}
-  %   \inferrule{
-  %     \tincompat{\htau_2}{\htau_2'}
-  %   }{
-  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
-  %   }
-  % \end{equation}
-% \end{subequations}
-% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
-The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.
+The mechanization proves that this judgemental definition of type
+inconsistency is indeed the negation of type consistency.
 
@@
 
-The action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
+The action semantics includes a subsumption rule similar to the subsumption
+rule, Rule (\ref{rule:ana-subsume}), in the statics:
 \begin{equation}\label{rule:action-subsume}
@@
     \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
-    \tcompat{\htau}{\htau''}%\\\\
-    % \zexp, \alpha~\mathsf{subsumes}
+    \tcompat{\htau}{\htau''}
   }{
@@
 \end{equation}
-In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.
-The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e.
-Rule (\ref{rule:ana-subsume}).
-Algorithmically, subsumption should be the rule of last resort.
-
-% We need to make some exceptions in this rule to maintain determinism, because certain construction actions, described in Sec.
-\ref{sec:construction}, behave slightly differently depending on whether a type is known.
-We do so by defining a simple judgement $\alpha~\mathsf{subsumes}$ that excludes these action forms:\todo{put into supp}
-% \begin{subequations}
-% \begin{equation}\label{rule:subsumes}
-% \inferrule{
-%   \alpha \neq \aConstruct{\fasc}\\
-%   (\forall x) \alpha \neq \aConstruct{\flam{x}} 
-% }{
-%   \zexp, \alpha~\mathsf{subsumes}
-% }
-% \end{equation}
-% \begin{equation}\label{rule:subsumes}
-% \inferrule{
-%   \zexp \neq \zwsel{\hexp}
-% }{
-%   \zexp, \alpha~\mathsf{subsumes}
-% }
-% \end{equation}
-% \end{subequations}
-% Algorithmically, this captures the notion that subsumption should be the case of last resort.
+In other words, if the cursor erasure of the edit state synthesizes a type,
+then we defer to the synthetic action judgement, as long as the type of the
+resulting cursor erasure is consistent with the type provided for analysis.
+The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem
+\ref{thrm:actsafe} goes through by induction and static subsumption,
+i.e. Rule (\ref{rule:ana-subsume}). Algorithmically, subsumption should be
+the rule of last resort.
 
 \subsubsection{Relative Movement}\label{sec:movement}
-The rules below define relative movement within Z-types. They should be self-explanatory:
+The rules below define relative movement within Z-types. They should be
+self-explanatory:
 \begin{subequations}
@@
 \end{equation}
-% \begin{equation}
-%   \inferrule{ }{
-%     \performTyp{
-%       \tarr{{\htau_1}}{\zwsel{\htau_2}}
-%     }{
-%       \aMove{\dPrev}
-%     }{
-%       {\tarr{\zwsel{\htau_1}}{{\htau_2}}}
-%     }
-%   }
-% \end{equation}
-% \begin{equation}
-% \inferrule{
-%   \performTyp{
-%     \ztau
-%   }{
-%     \aMove{\delta}
-%   }{
-%     \ztau'
-%   }
-% }{
-%   \performTyp{
-%     \tarr{\ztau}{\htau}
-%   }{
-%     \aMove{\delta}
-%   }{
-%     \tarr{\ztau'}{\htau}
-%   }
-% }
-% \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{
-%       \ztau
-%     }{
-%       \aMove{\delta}
-%     }{
-%       \ztau'
-%     }
-%   }{
-%     \performTyp{
-%       \tarr{\htau}{\ztau}
-%     }{
-%       \aMove{\delta}
-%     }{
-%       \tarr{\htau}{\ztau}
-%     }
-%   }
-% \end{equation}
-\end{subequations}
-Two more rules are needed to recurse into the zipper structure. We define these zipper rules in an action-independent manner in Sec. \ref{sec:zipper-cases}. 
-% The final two rules above recurse into the zipper structure.
-
-The rules for relative movement within Z-expressions are similarly straightforward.
-Movement is type-independent, so we defer to an auxiliary expression movement judgement in both the analytic and synthetic case:
+\end{subequations}
+Two more rules are needed to recurse into the zipper structure. We define
+these zipper rules in an action-independent manner in
+Sec. \ref{sec:zipper-cases}.
+
+The rules for relative movement within Z-expressions are similarly
+straightforward. Movement is type-independent, so we defer to an auxiliary
+expression movement judgement in both the analytic and synthetic case:
 \begin{subequations}
@@
 \end{subequations}
-% \begin{figure}
-% \judgbox{\performMove{\zexp}{\aMove{\delta}}{\zexp'}}{(Additional movement-action rules:)}
-% \begin{displaymath}
-% \begin{array}{@{}rcl}
-%   %\multicolumn{3}{l}{\textbf{Movement Actions for Sum Type Forms}:}
-%   %\\
-%   \TABperformMove
-%       {\zwsel{\hexp : \htau}}
-%       {\aMove{\dChild}}
-%       {      \zwsel{\hexp} : \htau}
-%   \\
-%   \TABperformMove
-%       {
-%       \zwsel{\hexp} : \htau
-%     }{
-%       \aMove{\dParent}
-%     }{
-%       \zwsel{\hexp : \htau}
-%     }
-%   \\
-%   \TABperformMove
-%     {
-%       \hexp : \zwsel{\htau}
-%     }{
-%       \aMove{\dParent}
-%     }{
-%       \zwsel{\hexp : \htau}
-%     }
-%   \\
-%   \TABperformMove
-%       {
-%       \zwsel{\hexp} : \htau
-%     }{
-%       \aMove{\dNext}
-%     }{
-%       \hexp : \zwsel{\htau}
-%     }
-%   \\[2mm]
-%   \TABperformMove
-%       {      {\tsum{{\htau_1}}{{\zwsel{\htau_2}}}}}
-%       {\aMove{\dParent}}
-%       {\zwsel{\tsum{{\htau_1}}{\htau_2}}}
-%   \\[2mm]
-%   \TABperformMove
-%       {\zwsel{\hinj{i}{\hexp}}}
-%       {\aMove{\dChild}}
-%       {\hinj{i}{\zwsel{\hexp}}}      
-%   \\
-%   \TABperformMove
-%       {\hinj{i}{\zwsel{\hexp}}}      
-%       {\aMove{\dParent}}
-%       {\zwsel{\hinj{i}{\hexp}}}
-%   \\[2mm]
-%   \TABperformMove
-%       {\zwsel{\hcase{\hexp}{x}{\hexp_1}{y}{\hexp_2}}}
-%       {\aMove{\dChild}}
-%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
-%   \\
-%   \TABperformMove
-%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
-%       {\aMove{\dNext}}
-%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
-%   \\
-%   \TABperformMove
-%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
-%       {\aMove{\dNext}}
-%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
-%   \\
-%   \TABperformMove
-%       {      {\hcase{{\hexp}}{x}{{\hexp_1}}{y}{\zwsel{\hexp_2}}}}
-%       {\aMove{\dPrev}}
-%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
-%   \\
-%   \TABperformMove
-%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
-%       {\aMove{\dPrev}}
-%       {      {\hcase{{\zwsel{\hexp}}}{x}{{\hexp_1}}{y}{{\hexp_2}}}}
-%   \\
-%   \TABperformMove
-%       {      {\hcase{\zwsel{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
-%       {\aMove{\dParent}}
-%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
-%   \\
-%   \TABperformMove
-%       {      {\hcase{{\hexp}}{x}{\zwsel{\hexp_1}}{y}{\hexp_2}}}
-%       {\aMove{\dParent}}
-%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
-%   \\
-%   \TABperformMove
-%       {      {\hcase{{\hexp}}{x}{\hexp_1}{y}{\zwsel{\hexp_2}}}}
-%       {\aMove{\dParent}}
-%       {\zwsel{\hcase{{\hexp}}{x}{\hexp_1}{y}{\hexp_2}}}
-% \end{array}
-% \end{displaymath}
-% \caption{Movement action semantics for sum type forms.}
-% \end{figure}
 The expression movement judgement is defined as follows.
@@
 \end{equation}
-% \begin{equation}
-%   \label{r:moveprevsib}
-%   \inferrule{ }{
-%     \performTyp{
-%       \hexp : \zwsel{\htau}
-%     }{
-%       \aMove{\dPrev}
-%     }{
-%       \zwsel{\hexp} : \htau
-%     }
-%   }
-% \end{equation}
-% \begin{equation}
-% \inferrule{
-%   \performTyp{
-%     \zexp
-%   }{
-%     \aMove{\delta}
-%   }{
-%     \zexp'
-%   }
-% }{
-%   \performTyp{
-%     \zexp : \htau
-%   }{
-%     \aMove{\delta}
-%   }{
-%     \zexp' : \htau
-%   }
-% }
-% \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{
-%       \ztau
-%     }{
-%       \aMove{\delta}
-%     }{
-%       \ztau'
-%     }
-%   }{
-%     \performTyp{
-%       \hexp : \ztau
-%     }{
-%       \aMove{\delta}
-%     }{
-%       \hexp : \ztau'
-%     }
-%   }
-% \end{equation}
 
@@
 \end{equation}
-% \begin{equation}
-%   \inferrule{ }{
-%     \performMove{
-%       \hap{\hexp_1}{\zwsel{\hexp_2}}
-%     }{
-%       \aMove{\dPrev}
-%     }{
-%       \hap{\zwsel{\hexp_1}}{\hexp_2}
-%     }
-%   }
-% \end{equation}
 
@@
 \end{equation}
-% \begin{equation}
-%   \inferrule{ }{
-%     \performMove{
-%       \hadd{\hexp_1}{\zwsel{\hexp_2}}
-%     }{
-%       \aMove{\dPrev}
-%     }{
-%       \hadd{\zwsel{\hexp_1}}{\hexp_2}
-%     }
-%   }
-% \end{equation}
 
@@
 \end{equation}
-Again, additional rules are needed to recurse into the zipper structure, but we will define these zipper rules in an action-independent manner in Sec. \ref{sec:zipper-cases}. 
-\end{subequations}
-
-The rules above are numerous and fairly uninteresting.
-That makes them quite hazardous -- we might make a mistake absent-mindedly.
-One check against this is to prove a lemma that establishes that movement actions move the cursor but do not change the cursor erasure.
+Again, additional rules are needed to recurse into the zipper structure,
+but we will define these zipper rules in an action-independent manner in
+Sec. \ref{sec:zipper-cases}.
+\end{subequations}
+
+The rules above are numerous and fairly uninteresting. That makes them
+quite hazardous -- we might make a mistake absent-mindedly. One check
+against this is to prove a lemma that establishes that movement actions
+move the cursor but do not change the cursor erasure.
 
@@
   \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then $\removeSel{\ztau}=\removeSel{\ztau'}$.
-\item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then $\removeSel{\zexp}=\removeSel{\zexp'}$.
+  \item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then
+    $\removeSel{\ztau}=\removeSel{\ztau'}$.
+
+  \item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then
+    $\removeSel{\zexp}=\removeSel{\zexp'}$.
 \end{enumerate}
 \end{theorem}
-\noindent Theorem \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.
-
-Another useful check is to establish \emph{reachability}, i.e.
-that it is possible, through a sequence of movement actions, to move the cursor from any position to any other position within a well-typed H-expression.
-
-This requires developing machinery for reasoning about sequences of actions.
-There are two possibilities: we can either add a sequencing action, $\alpha; \alpha$, directly to the syntax of actions, or we can define a syntax for lists of actions, $\bar{\alpha}$, together with iterated action judgements.
-To keep the core of the action semantics small, we take the latter approach in Figure \ref{fig:multistep}.
-
-A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in the supplement, establishes that $\bar\alpha$ consists only of actions of the form $\aMove{\delta}$.
+\noindent Theorem \ref{lemma:movement-erasure} is useful also in that the
+relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its
+application.
+
+Another useful check is to establish \emph{reachability}, i.e. that it is
+possible, through a sequence of movement actions, to move the cursor from
+any position to any other position within a well-typed H-expression.
+
+This requires developing machinery for reasoning about sequences of
+actions. There are two possibilities: we can either add a sequencing
+action, $\alpha; \alpha$, directly to the syntax of actions, or we can
+define a syntax for lists of actions, $\bar{\alpha}$, together with
+iterated action judgements. To keep the core of the action semantics small,
+we take the latter approach in Figure \ref{fig:multistep}.
+
+A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in
+the supplement, establishes that $\bar\alpha$ consists only of actions of
+the form $\aMove{\delta}$.
 
@@
   \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\removeSel{\ztau}=\removeSel{\ztau'}$ then there exists some $\bar\alpha$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performTypI{\ztau}{\bar\alpha}{\ztau'}$.
-\item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$.
-\item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zexp'}{\htau}$.
+  \item If $\removeSel{\ztau}=\removeSel{\ztau'}$ then there exists some
+    $\bar\alpha$ such that $\bar{\alpha}~\mathsf{movements}$ and
+    $\performTypI{\ztau}{\bar\alpha}{\ztau'}$.
+
+  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some
+    $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and
+    $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zexp'}{\htau}$.
+
+  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\removeSel{\zexp}=\removeSel{\zexp'}$ then there exists some
+    $\bar{\alpha}$ such that $\bar{\alpha}~\mathsf{movements}$ and
+    $\performAnaI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$.
   \end{enumerate}
@@
 \hspace{-12px}
-% \begin{minipage}{0.88\linewidth}
 \begin{mathpar}
@@
 \end{mathpar}
-% \end{minipage}
-% \begin{minipage}{0.12\linewidth}
-% (11a-b)
-% \end{minipage}
 \end{subequations}
@@
 
-The simplest way to prove Theorem \ref{thrm:reachability} is to  break it into two lemmas.
-Lemma \ref{lemma:reach-up} establishes that you can always move the cursor to the outermost position in an expression.
-This serves as a check on our $\aMove{\dParent}$ rules.
+The simplest way to prove Theorem \ref{thrm:reachability} is to break it
+into two lemmas. Lemma \ref{lemma:reach-up} establishes that you can always
+move the cursor to the outermost position in an expression. This serves as
+a check on our $\aMove{\dParent}$ rules.
 \begin{lemma}[Reach Up]\label{lemma:reach-up} ~
   \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTypI{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
-\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$.
-\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
+  \item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$
+    such that $\bar\alpha~\mathsf{movements}$ and
+    $\performTypI{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
+
+  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$
+    then there exists some $\bar\alpha$ such that
+    $\bar\alpha~\mathsf{movements}$ and
+    $\performSynI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
+
+  \item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$
+    then there exists some $\bar\alpha$ such that
+    $\bar\alpha~\mathsf{movements}$ and
+    $\performAnaI{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$.
   \end{enumerate}
 \end{lemma}
-Lemma \ref{lemma:reach-down} establishes that you can always move the cursor from the outermost position to any other position.
-This serves as a check on our $\aMove{\dChild}$ and $\aMove{\dNext}$ rules.
+Lemma \ref{lemma:reach-down} establishes that you can always move the
+cursor from the outermost position to any other position. This serves as a
+check on our $\aMove{\dChild}$ and $\aMove{\dNext}$ rules.
 \begin{lemma}[Reach Down]\label{lemma:reach-down} ~
   \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTypI{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
-\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAnaI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
-\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ nand $\performSynI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
+  \item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$
+    such that $\bar\alpha~\mathsf{movements}$ and
+    $\performTypI{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
+
+  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$
+    then there exists some $\bar\alpha$ such that
+    $\bar\alpha~\mathsf{movements}$ nand
+    $\performSynI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
+
+  \item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$
+    then there exists some $\bar\alpha$ such that
+    $\bar\alpha~\mathsf{movements}$ and
+    $\performAnaI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
   \end{enumerate}
 \end{lemma}
-Theorem \ref{thrm:reachability} follows by straightforward composition of these two lemmas. %(The witness to the existential is not, of course, the shortest path between any two cursor locations.)
-
-% \paragraph{Reachability}
-
-% We give the following theorem to demonstrate that the rules for movement do
-% indeed capture the process of moving the focus to any editable position
-% within a term. Intuitively, given two zippered expressions that are the
-% same up to the erasure of focus, its proof gives a list of composable
-% actions that, when applied to the first term produces the second.
-
-% As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
-% every action in such a list, it is possible consider moving to an ill-typed
-% term. Therefore the statement here matches the mutually recursive structure
-% of the bidirectional typing judgements.
-
-% \begin{theorem}[Reachability]\todo{why $\htau$ and not $\ztau$?}
-% \label{thrm:reach}
-% Both of the following hold:
-% \begin{enumerate}
-%   \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-%    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-%    = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-%    ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-%    = \zexp'$.
-
-%   \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
-%    $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$ and $\removeSel{\zexp}
-%    = \removeSel{\zexp'}$ then $\exists L \in \mathit{list}
-%    ~\mathtt{action}$ such that $\mathit{iterate}~ L~ \zexp
-%    = \zexp'$.
-% \end{enumerate}
-% \end{theorem}
-
-\subsubsection{Deletion} The $\aDel$ action inserts an empty hole at the cursor, deleting what was there before.
+Theorem \ref{thrm:reachability} follows by straightforward composition of
+these two lemmas.
+
+\subsubsection{Deletion} The $\aDel$ action inserts an empty hole at the
+cursor, deleting what was there before.
 
@@
 \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{\ztau}{\aDel}{\ztau'}
-%   }{
-%     \performTyp{\tarr{\ztau}{\htau}}{\aDel}{\tarr{\ztau'}{\htau}}
-%   }
-% \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{\ztau}{\aDel}{\ztau'}
-%   }{
-%     \performTyp{\tarr{\htau}{\ztau}}{\aDel}{\tarr{\htau}{\ztau'}}
-%   }
-% \end{equation}
 
@@
 \end{equation}
-%\end{subequations}
-% The base case turns into a hole:
-%\begin{subequations}
-% \begin{equation}
-% \inferrule{ }{
-%   \performDel{\zwsel{\hexp}}{\hehole}
-% }
-% \end{equation}
-% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
-% \begin{equation}
-%   \inferrule{
-%     \performDel{\zexp}{\zexp'}
-%   }{
-%     \performDel{\zexp : \htau}{\zexp' : \htau}
-%   }
-% \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \performTyp{\ztau}{\aDel}{\ztau'}
-%   }{
-%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
-%   }
-% \end{equation}
-\end{subequations}
-
-\subsubsection{Construction}\label{sec:construction} The construction actions, $\aConstruct{\psi}$, are used to construct terms of a shape indicated by $\psi$ at the cursor.
-
-\paragraph{Types} The $\aConstruct{\farr}$ action constructs an arrow type. The H-type under the cursor becomes the argument type, and the cursor is placed on an empty return type hole:
+\end{subequations}
+
+\subsubsection{Construction}\label{sec:construction} The construction
+actions, $\aConstruct{\psi}$, are used to construct terms of a shape
+indicated by $\psi$ at the cursor.
+
+\paragraph{Types} The $\aConstruct{\farr}$ action constructs an arrow
+type. The H-type under the cursor becomes the argument type, and the cursor
+is placed on an empty return type hole:
 \begin{subequations}
@@
 
-The $\aConstruct{\fnum}$ action replaces an empty type hole under the cursor with the $\tnum$ type:
+The $\aConstruct{\fnum}$ action replaces an empty type hole under the
+cursor with the $\tnum$ type:
   \begin{equation}
@@
 \end{equation}
-
-% Construction proceeds recursively down the zipper:
-%   \begin{equation}
-%     \label{r:contarrL}
-%   \inferrule{
-%     \performTyp{\ztau}{\aConstruct{\psi}}{\ztau'}
-%   }{
-%     \performTyp{
-%       \tarr{\ztau}{\htau}
-%     }{
-%       \aConstruct{\psi}
-%     }{
-%       \tarr{\ztau'}{\htau}
-%     }
-%   }
-% \end{equation}
-%   \begin{equation}
-%     \label{r:contarrR}
-%   \inferrule{
-%     \performTyp{\ztau}{\aConstruct{\psi}}{\ztau'}
-%   }{
-%     \performTyp{
-%       \tarr{\htau}{\ztau}
-%     }{
-%       \aConstruct{\psi}
-%     }{
-%       \tarr{\htau}{\ztau'}
-%     }
-%   }
-% \end{equation}
 \end{subequations}
@@
 
-\paragraph{Ascription} The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type.
-In the first case, the synthesized type appears in the ascription:
+\paragraph{Ascription} The $\aConstruct{\fasc}$ action operates differently
+depending on whether the H-expression under the cursor synthesizes a type
+or is being analyzed against a type. In the first case, the synthesized
+type appears in the ascription:
 \begin{equation}
@@
 \end{equation}
-% This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).
-
-\paragraph{Variables} The $\aConstruct{\fvar{x}}$ action places the variable $x$ into an empty hole.
-If that hole is being asked to synthesize a type, then the result synthesizes the hypothesized type:
+
+\paragraph{Variables} The $\aConstruct{\fvar{x}}$ action places the
+variable $x$ into an empty hole. If that hole is being asked to synthesize
+a type, then the result synthesizes the hypothesized type:
 \begin{equation}
@@
 \end{equation}
-If the hole is being analyzed against a type that is consistent with the hypothesized type, then the action semantics goes through the {action subsumption rule} described in Sec.
-\ref{sec:action-subsumption}.
-If the hole is being analyzed against a type that is inconsistent with the hypothesized type, $x$ is placed inside a hole:
+If the hole is being analyzed against a type that is consistent with the
+hypothesized type, then the action semantics goes through the {action
+  subsumption rule} described in Sec. \ref{sec:action-subsumption}. If the
+hole is being analyzed against a type that is inconsistent with the
+hypothesized type, $x$ is placed inside a hole:
 \begin{equation}
@@
 
-\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole.
-If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
+\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda
+abstraction binding $x$ into an empty hole. If the empty hole is being
+asked to synthesize a type, then the result of the action is a lambda
+ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the
+argument type hole:
 \begin{equation}
@@
 \end{equation}
-The type ascription is necessary because lambda expressions do not synthesize a type. If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:
+The type ascription is necessary because lambda expressions do not
+synthesize a type. If the empty hole is being analyzed against a type with
+matched arrow type, then no ascription is necessary:
 \begin{equation}\label{rule:performAna-lam-1}
@@
 \end{equation}
-% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).
-
-Finally, if the empty hole is being analyzed against a type that has no matched arrow type, expressed in the premise as inconsistency with $\tarr{\tehole}{\tehole}$, then a lambda ascribed the type $\tarr{\tehole}{\tehole}$
-is inserted inside a hole, which defers the type inconsistency as previously discussed:
+
+Finally, if the empty hole is being analyzed against a type that has no
+matched arrow type, expressed in the premise as inconsistency with
+$\tarr{\tehole}{\tehole}$, then a lambda ascribed the type
+$\tarr{\tehole}{\tehole}$ is inserted inside a hole, which defers the type
+inconsistency as previously discussed:
 \begin{equation}\label{rule:performAna-construct-lam-2}
@@
 
-\paragraph{Application} The $\aConstruct{\fap}$ action applies the expression under the cursor. The following rule handles the case where the synthesized type has matched function type:
+\paragraph{Application} The $\aConstruct{\fap}$ action applies the
+expression under the cursor. The following rule handles the case where the
+synthesized type has matched function type:
 \begin{equation}
@@
 \end{equation}
-If the expression under the cursor synthesizes a type that is inconsistent with an arrow type, then we must place that expression inside a hole to maintain Theorem \ref{sec:holes}:
+If the expression under the cursor synthesizes a type that is inconsistent
+with an arrow type, then we must place that expression inside a hole to
+maintain Theorem \ref{sec:holes}:
 \begin{equation}
@@
 
-The $\aConstruct{\farg}$ action instead places the expression under the cursor in the argument position of an application form.
-Because the function position is always an empty hole in this situation, we need only a single rule:
+The $\aConstruct{\farg}$ action instead places the expression under the
+cursor in the argument position of an application form. Because the
+function position is always an empty hole in this situation, we need only a
+single rule:
 \begin{equation}
@@
 
-\paragraph{Numbers} The $\aConstruct{\fnumlit{n}}$ action replaces an empty hole with the number expression $\hnum{n}$.
-If the empty hole is being asked to synthesize a type, then the rule is straightforward:
+\paragraph{Numbers} The $\aConstruct{\fnumlit{n}}$ action replaces an empty
+hole with the number expression $\hnum{n}$. If the empty hole is being
+asked to synthesize a type, then the rule is straightforward:
 \begin{equation}
@@
 \end{equation}
-If the empty hole is being analyzed against a type that is inconsistent with $\tnum$, then we must place the number expression inside the hole:
+If the empty hole is being analyzed against a type that is inconsistent
+with $\tnum$, then we must place the number expression inside the hole:
 \begin{equation}
@@
 
-The $\aConstruct{\fplus}$ action constructs a plus expression with the expression under the cursor as its first argument.
-If that expression synthesizes a type consistent with $\tnum$, then the rule is straightforward:
+The $\aConstruct{\fplus}$ action constructs a plus expression with the
+expression under the cursor as its first argument. If that expression
+synthesizes a type consistent with $\tnum$, then the rule is
+straightforward:
 \begin{equation}\label{rule:construct-plus-compat}
@@
 
-\paragraph{Non-Empty Holes} The final shape is $\fnehole$. This explicitly places the expression under the cursor inside a hole:
+\paragraph{Non-Empty Holes} The final shape is $\fnehole$. This explicitly
+places the expression under the cursor inside a hole:
 \begin{equation}
@@
 
-The $\fnehole$ shape is grayed out in Figure \ref{fig:action-syntax} because we do not expect the programmer to perform it explicitly -- other actions automatically insert holes when a type inconsistency would arise.
-Its inclusion is mainly to make it easier to state another ``checksum'' theorem: \emph{constructability}
+The $\fnehole$ shape is grayed out in Figure \ref{fig:action-syntax}
+because we do not expect the programmer to perform it explicitly -- other
+actions automatically insert holes when a type inconsistency would
+arise. Its inclusion is mainly to make it easier to state another
+``checksum'' theorem: \emph{constructability}
 
@@
   \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-\item For every $\htau$ there exists $\bar\alpha$ such that $\performTypI{\zwsel{\tehole}}{\bar\alpha}{\zwsel{\htau}}$.
-\item If $\hana{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that: $$\performAnaI{\hGamma}{\zwsel{\hhole{}}}{\htau}{\bar\alpha}{\zwsel{\hexp}}$$
-\item If $\hsyn{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$ such that: $$\performSynI{\hGamma}{\zwsel{\hhole{}}}{\tehole}{\bar\alpha}{\zwsel{\hexp}}{\htau}$$
+  \item For every $\htau$ there exists $\bar\alpha$ such that
+    $\performTypI{\zwsel{\tehole}}{\bar\alpha}{\zwsel{\htau}}$.
+
+  \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$
+    such
+    that: $$\performSynI{\hGamma}{\zwsel{\hhole{}}}{\tehole}{\bar\alpha}{\zwsel{\hexp}}{\htau}$$
+
+  \item If $\hana{\hGamma}{\hexp}{\htau}$ then there exists $\bar\alpha$
+    such
+    that: $$\performAnaI{\hGamma}{\zwsel{\hhole{}}}{\htau}{\bar\alpha}{\zwsel{\hexp}}$$
   \end{enumerate}
@@
 \subsubsection{Finishing}
-The final action we need to consider is $\aFinish$, which finishes the non-empty hole under the cursor.
-
-If the non-empty hole appears in synthetic position, then it can always be finished:
+The final action we need to consider is $\aFinish$, which finishes the
+non-empty hole under the cursor.
+
+If the non-empty hole appears in synthetic position, then it can always be
+finished:
 \begin{subequations}
@@
 
-If the non-empty hole appears in analytic position, then it can only be finished if the type synthesized for the enveloped expression is consistent with the type that the hole is being analyzed against.
-This amounts to analyzing the enveloped expression against the provided type (by subsumption):
+If the non-empty hole appears in analytic position, then it can only be
+finished if the type synthesized for the enveloped expression is consistent
+with the type that the hole is being analyzed against. This amounts to
+analyzing the enveloped expression against the provided type (by
+subsumption):
 \begin{equation}\label{r:finishana}
@@
 
-\subsubsection{Zipper Cases}\label{sec:zipper-cases} The rules defined so far handle the base cases, i.e.
-the cases where the action has ``reached'' the expression under the cursor.
-We also need to define the recursive cases, which propagate the action into the subtree where the cursor appears, as encoded by the zipper structure.
-For types, the zipper rules are straightforward:
+\subsubsection{Zipper Cases}\label{sec:zipper-cases} The rules defined so
+far handle the base cases, i.e. the cases where the action has ``reached''
+the expression under the cursor. We also need to define the recursive
+cases, which propagate the action into the subtree where the cursor
+appears, as encoded by the zipper structure. For types, the zipper rules
+are straightforward:
 \begin{subequations}
 \begin{equation}
-    %\label{r:contarrL}
   \inferrule{
@@
   \begin{equation}
-    %\label{r:contarrR}
   \inferrule{
@@
 \end{subequations}
-For expressions, the zipper rules essentially follow the structure of the corresponding rules in the statics.
+For expressions, the zipper rules essentially follow the structure of the
+corresponding rules in the statics.
 
 \begin{subequations}
-In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):
+In particular, when the cursor is in the body of a lambda expression, the
+zipper case mirrors Rule (\ref{rule:syn-lam}):
 \begin{equation}
@@
 
-When the cursor is in the expression position of an ascription, we use the analytic  judgement, mirroring Rule (\ref{rule:syn-asc}):
+When the cursor is in the expression position of an ascription, we use the
+analytic  judgement, mirroring Rule (\ref{rule:syn-asc}):
 \begin{equation}
@@
 
-When the cursor is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
+When the cursor is in the type position of an ascription, we must re-check
+the ascribed expression because the cursor erasure might have changed (in
+practice, one would optimize this check to only occur if the cursor erasure
+did change):
 \begin{equation}\label{rule:zipper-asc}
@@
 
-When the cursor is in the function position of an application, the rule mirrors Rule (\ref{rule:syn-ap}):
+When the cursor is in the function position of an application, the rule
+mirrors Rule (\ref{rule:syn-ap}):
 \begin{equation}
@@
 \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \hsyn{\hGamma}{\removeSel{\zexp}}{\htau_2}\\
-%     \performSyn
-%       {\hGamma}
-%       {\zexp}
-%       {\htau_2}
-%       {\alpha}
-%       {\zexp'}
-%       {\tehole}\\
-%     \hana{\hGamma}{\hexp}{\tehole}
-%   }{
-%     \performSyn
-%       {\hGamma}
-%       {\hap{\zexp}{\hexp}}
-%       {\htau_1}
-%       {\alpha}
-%       {\hap{\zexp'}{\hexp}}
-%       {\tehole}
-%   }
-% \end{equation}
 
@@
 \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \hsyn{\hGamma}{\hexp}{\tehole}\\
-%     \performAna
-%       {\hGamma}
-%       {\zexp}
-%       {\tehole}
-%       {\alpha}
-%       {\zexp'}
-%   }{
-%     \performSyn
-%       {\hGamma}
-%       {\hap{\hexp}{\zexp}}
-%       {\tehole}
-%       {\alpha}
-%       {\hap{\hexp}{\zexp'}}
-%       {\tehole}
-%   }
-% \end{equation}
 
@@
 
-Finally, if the cursor is inside a non-empty hole, the relevant zipper rule mirrors Rule (\ref{rule:syn-ehole}):
+Finally, if the cursor is inside a non-empty hole, the relevant zipper rule
+mirrors Rule (\ref{rule:syn-ehole}):
 \begin{equation}
@@
 \end{equation}
-% \begin{equation}
-%   \inferrule{
-%     \hsyn{\hGamma}{\removeSel{\zexp}}{\htau}\\
-%     \performSyn
-%       {\hGamma}
-%       {\zexp}
-%       {\htau}
-%       {\alpha}
-%       {\zwsel{\hehole}}
-%       {\tehole}\\
-%   }{
-%     \performSyn
-%       {\hGamma}
-%       {\hhole{\zexp}}
-%       {\tehole}
-%       {\alpha}
-%       {\zwsel{\hehole}}
-%       {\tehole}
-%   }
-% \end{equation}
-
-Theorem \ref{thrm:actsafe} directly checks the correctness of these rules.
-Moreover, the zipper rules arise ubiquitously in derivations of edit steps, so the proofs of the other ``check'' theorems, e.g.
-Reachability and Constructability, serve as a check that none of these rules have been missed.
-\end{subequations}
+
+Theorem \ref{thrm:actsafe} directly checks the correctness of these
+rules. Moreover, the zipper rules arise ubiquitously in derivations of edit
+steps, so the proofs of the other ``check'' theorems, e.g. Reachability and
+Constructability, serve as a check that none of these rules have been
+missed.
+\end{subequations}

### 2016-09-24 implementation.tex
working on #6

@@
 \subsection{Implementation Concepts}
-Central to any implementation of Hazelnut is a stream of edit states whose cursor erasures synthesize types under an empty context according to the synthetic action judgement, $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$.
-The middle row of Figure \ref{fig:impl-overview} diagrams this stream of edit states.
-By Theorem \ref{thrm:actsafe}, the editor does not need to typecheck each edit state (though in some of the rules, e.g.
-Rule (\ref{rule:zipper-asc}) which handles the situation where the type in a type ascription changes, portions of the program do need to be typechecked.) For example, the reader is encouraged to re-examine the examples in Figure \ref{fig:first-example} and \ref{fig:second-example} -- the cursor erasure of each edit state synthesizes a type.
+Central to any implementation of Hazelnut is a stream of edit states whose
+cursor erasures synthesize types under an empty context according to the
+synthetic action judgement,
+$\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$. The middle
+row of Figure \ref{fig:impl-overview} diagrams this stream of edit
+states. By Theorem \ref{thrm:actsafe}, the editor does not need to
+typecheck each edit state (though in some of the rules, e.g. Rule
+(\ref{rule:zipper-asc}) which handles the situation where the type in a
+type ascription changes, portions of the program do need to be
+typechecked.) For example, the reader is encouraged to re-examine the
+examples in Figure \ref{fig:first-example} and \ref{fig:second-example} --
+the cursor erasure of each edit state synthesizes a type.
 
-The programmer examines a view generated from each edit state and produces actions in some implementation-defined manner (e.g.
-using a keyboard, mouse, touchscreen, voice interface, or neural implant.) Each new action causes a new abstract edit state to arise according to an implementation of the action semantics.
-This then causes a new view to arise.
-This is a simple event-based functional reactive programming model \cite{Wan:2000:FRP:349299.349331}.
+The programmer examines a view generated from each edit state and produces
+actions in some implementation-defined manner (e.g. using a keyboard,
+mouse, touchscreen, voice interface, or neural implant.) Each new action
+causes a new abstract edit state to arise according to an implementation of
+the action semantics. This then causes a new view to arise. This is a
+simple event-based functional reactive programming model
+\cite{Wan:2000:FRP:349299.349331}.
 
-If an action is not well-defined according to Hazelnut's action semantics, the implementation must reject it.
-In fact, the implementation is encouraged to present an ``action palette'' that either hides or visibly disables actions that are not well-defined (see below.)
+If an action is not well-defined according to Hazelnut's action semantics,
+the implementation must reject it. In fact, the implementation is
+encouraged to present an ``action palette'' that either hides or visibly
+disables actions that are not well-defined (see below.)
 
@@
 \subsection{HZ}
-We have developed a simple reference implementation of Hazelnut.
-In order to reach a wide audience, we decided to implement HZ in the web browser.
-In order to take advantage of a mature implementation of the FRP model, we chose to implement HZ using OCaml\footnote{https://ocaml.org/}, the \texttt{js\_of\_ocaml} compiler and associated libraries \cite{DBLP:conf/ml/Balat06}\footnote{http://ocsigen.org/js\_of\_ocaml/} and the OCaml React library\footnote{http://erratique.ch/software/react}.
+We have developed a simple reference implementation, HZ, of Hazelnut.  In
+order to reach a wide audience, we decided to implement HZ in the web
+browser.  In order to take advantage of a mature implementation of the FRP
+model, we chose to implement HZ using
+OCaml\footnote{\url{https://ocaml.org/}}, the \texttt{js\_of\_ocaml}
+compiler and associated libraries
+\cite{DBLP:conf/ml/Balat06}\footnote{\url{http://ocsigen.org/js\_of\_ocaml/}}
+and the OCaml React
+library\footnote{\url{http://erratique.ch/software/react}}.
 
-At the time of the writing, 
-HZ's view computation renders the model as a string embedded into HTML that approximates our typeset notation in this paper.
-The action palette is a collection of buttons and text boxes, which are disabled when an action is not well-defined or an invalid action argument is entered.
-We determine this by simply attempting to perform the action internally and catching the exception that is raised when an action is undefined.
-This design, of course, is not meant to score marks for usability or performance.
-Instead, as a simple reference implementation, we expect its source code to be of use to others who are interested in layering a more fluid user interface atop the core semantics (which have been implemented in a style that follows the presentation in this paper closely), or extensions thereof.
+At the time of the writing, HZ's view computation renders the model as a
+string embedded into HTML that approximates our typeset notation in this
+paper. The action palette is a collection of buttons and text boxes, which
+are disabled when an action is not well-defined or an invalid action
+argument is entered. We determine this by simply attempting to perform the
+action internally and catching the exception that is raised when an action
+is undefined. This design, of course, is not meant to score marks for
+usability or performance. Instead, as a simple reference implementation, we
+expect its source code to be of use to others who are interested in
+layering a more fluid user interface atop the core semantics (which have
+been implemented in a style that follows the presentation in this paper
+closely), or extensions thereof.

### 2016-09-24 intro.tex
commiting before deleting this file; it's my own cruft from the TFP work and isn't included anywhere

@@
 
-%% There are some benefits to this approach, to be sure, but the structural
-%% mismatch between programs and their textual representations also imposes
-%% various burdens.  For example, the primitive edit actions available in a
-%% text editor (e.g. inserting or deleting a character or word) do not
-%% always correspond to sensible structural transformations.
-
-% spj: describe the problem; state our contributions; STOP. one page max.
-
-% When constructing a program or proof in a language with rich type
-% structure, skilled programmers generally follow a \emph{type discipline}
-% where they first determine the type of the expression that they are
-% constructing in order to constrain the mental search space that they are
-% operating within.
-
-% For example, if the programmer knows that an expression of type
-% $\tarr{\tnum}{\tnum}$ is needed, then it is often the case (though, of course, not
-% necessarily the case) that the expression will take the form $$\hlam{\mathit{x}}{e}$$
-% for some variable $x$ and function body $e$. If the programmer chooses this
-% form, then after picking a suitable variable name, her focus will be on
-% constructing a suitable body, $e$. Following the type discipline, $e$ must
-% be of type $\tnum$, and so this process can begin anew.
-
-% The problem is that when using a text
-% editor to construct a program, it is easy, and indeed necessary, to deviate from this disciplined process.  Rather, text editors operate on sequences of
-% characters (i.e. \emph{text}.)
-
-%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
-% Programming languages, and therefore programs themselves, are rich mathematical structures.
-Programmers typically construct and manipulate well-typed expressions only indirectly, by editing text that is first parsed according to a textual syntax and then typechecked according to a static semantics.
-This indirection has practical benefits, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.
-
-% Every day, programmers use text-based tools to construct and manipulate programs.  
-% These programs are written in languages which defined by a textual syntax. 
-% While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.
-
-% It also complicates matters for tool designers because tools are often confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}.
-This may be because the programmer is in the midst of a sequence of
-% edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake.
-The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g.
-by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
-%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
-%Most programming languages define a textual syntax.
-This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs.
-While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.
+Programmers typically construct and manipulate well-typed expressions only
+indirectly, by editing text that is first parsed according to a textual
+syntax and then typechecked according to a static semantics. This
+indirection has practical benefits, to be sure -- text editors and other
+text-based tools benefit from decades of development effort -- but it also
+introduces some fundamental complexity into the programming process.
 
@@
     \ref{sec:example} with two examples of edit sequences to develop the reader's intuitions.
-\item We then give a detailed overview of Hazelnut's semantics and metatheory, which have been mechanized using the Agda proof assistant, in Sec.  \ref{sec:hazel}.
+\item We then give a detailed overview of Hazelnut's semantics and metatheory, and our mechanization in the Agda proof assistant, in Sec.  \ref{sec:hazel}.
 \item Hazelnut is a {foundational} calculus, i.e.
@@
 We extend Hazelnut with simple sum types in Sec.
-\ref{sec:extending} to demonstrate how Hazelnut's rich mechanized metatheory guides such a development.
+\ref{sec:extending} to demonstrate how Hazelnut's rich metatheory guides such a development.
 \item In Sec.

### 2016-09-24 intro.tex
commiting before deleting this file; it's my own cruft from the TFP work and isn't included anywhere
removed comments for the clean copy of intro, #6

@@
 
-First, it requires that programmers learn the subtleties of the textual syntax (e.g.
-the precedence of the various forms.) This can be particularly challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}.
-%For example, one study of novice programmers found syntax errors (e.g.
-unbalanced parentheses) to be the most common class of errors.
-% Too much detail?
+First, it requires that programmers learn the subtleties of the textual
+syntax (e.g. the precedence of the various forms.) This can be particularly
+challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}.
 
-The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers.
-In particular, program editors must contend with meaningless text on a regular basis.
-In a dataset analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of edit states were syntactically malformed \cite{6883030}.
-Some additional percentage of edit states were well-formed but ill-typed (the dataset did not contain enough information to determine the exact percentage.)
-Some of these meaningless edit states arose because the programmer was in the midst of a sequence of edits that left the edit state transiently meaningless, while others arose because the programmer made a logical mistake.
-Because the language definitions is silent about these edit states, it is difficult to design useful editor services, e.g.
-syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \cite{mens2004survey}.
-Editors must either disable these editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics, which can be misleading.
-%, or because the programmer has made a mistake.
-% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action
+The fact that not every sequence of characters corresponds to a meaningful
+program also complicates matters for tool designers. In particular, program
+editors must contend with meaningless text on a regular basis. In a dataset
+analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of
+edit states were syntactically malformed \cite{6883030}. Some additional
+percentage of edit states were well-formed but ill-typed (the dataset did
+not contain enough information to determine the exact percentage.)  Some of
+these meaningless edit states arose because the programmer was in the midst
+of a sequence of edits that left the edit state transiently meaningless,
+while others arose because the programmer made a logical mistake. Because
+the language definitions is silent about these edit states, it is difficult
+to design useful editor services, e.g. syntax
+highlighting~\cite{sarkar2015impact}, type-aware code
+completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324},
+and refactoring support \cite{mens2004survey}. Editors must either disable
+these editor services when they encounter meaningless edit states or
+develop \emph{ad hoc} heuristics, which can be misleading.
 
-% Many editors have developed \emph{ad hoc} workarounds for this problem, e.g.
-they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or continue past a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer.
-% This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.
+These complications have motivated a long line of research
+into \emph{structure editors}, i.e. program editors where every edit state
+is a program structure.
 
-% stuck editing a representation of the program instead of the structures
-% themselves. The editor does not restrict what the programmer may do: you
-% can delete characters that belong, insert ones that don't, forget things
-% that were needed, and so. There's nothing stopping us from accidentally
-% writing $$\lambda \mathit{x:num}.\mathit{(x,x)}$$ even though it's obvious
-% that building a pair can't hope to form a natural number.
+Most structure editors are \emph{syntactic structure editors}, i.e. the
+edit state is a syntax tree with \emph{holes} that stand for branches of
+the tree that have yet to be constructed, and the edit actions are
+context-free tree transformations. For example, Scratch is a syntactic
+structure editor that has achieved success as a tool for teaching children
+how to program \cite{Resnick:2009:SP:1592761.1592779}.
 
-% The type structure of the language makes this sort of error obvious: it's
-% not that you're adding characters that make your program incorrect, or even
-% malformed; you're adding characters that can't possibly create a structure
-% you want because of the type. Simply put, the primitive operations
-% available in text editors do not always correspond to sensible
-% transformations on the structure of the program.
-These complications have motivated a long line of research into \emph{structure editors}, i.e.
-program editors where every edit state is a program structure.
-% Eliminating text eliminates the possibility of syntax errors.
+The Scratch language has a trivial static semantics, but researchers have
+also designed syntactic structure editors for languages with a non-trivial
+static semantics. For example, \texttt{mbeddr} is an editor for a C-like
+language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an
+object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an
+editor for a functional language similar to Haskell \cite{lamdu}. Each of
+these editors presents an innovative user interface, but the non-trivial
+type and binding structure of the underlying language complicates its
+design. The reason is that a syntactic structure editor does not guarantee
+that every edit state is statically meaningful -- only that it is
+syntactically well-formed. As in textual program editors, these syntactic
+structure editors must either disable key editor services when they
+encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
 
-Most structure editors are \emph{syntactic structure editors}, i.e.
-the edit state is a syntax tree with \emph{holes} that stand for branches of the tree that have yet to be constructed, and the edit actions are context-free tree transformations.
-For example, Scratch is a syntactic structure editor that has achieved success as a tool for teaching children how to program \cite{Resnick:2009:SP:1592761.1592779}.
+This paper develops a principled solution to this problem. We introduce
+Hazelnut, a \emph{typed structure editor} based on a bidirectionally typed
+lambda calculus extended to assign static meaning to expressions and types
+with {holes}, which we call \textbf{H-expressions}
+and \textbf{H-types}. Hazelnut's formal \emph{action semantics} maintains
+the invariant that every edit state is a statically meaningful
+(i.e. well-typed) H-expression with a single superimposed \emph{cursor}. We
+call H-expressions and H-types with a cursor \textbf{Z-expressions}
+and \textbf{Z-types} (so prefixed because our encoding follows
+Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.)
 
-The Scratch language has a trivial static semantics, but researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics.
-For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}.
-Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design.
-The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed.
-As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
-%Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful.
-% Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.
-
-% \begin{itemize}
-% \item
-
-% First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e.
-``the implementation is the specification''.
-This makes it difficult to reason methodically about types and binding, so, in turn, it is difficult to design reliable editor services.
-
-% Even if a formal semantics and metatheory for the underlying language were forthcoming (e.g.
-following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}), it would go only part of the way towards addressing the challenges faced by structure editor designers.
-First, a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e.
-terms with holes.
-Moreover, a syntactic structure editor  -- so incorporating holes into the static semantics would in any case be only a minor salve.
-Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.
-
-This paper develops a principled solution to this problem.
-We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}.
-Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e.
-well-typed) H-expression with a single superimposed \emph{cursor}.
-We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because our encoding follows Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g.
-a function argument), only actions consistent with that type are permitted.
-
-Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner.
-For example, the programmer would often need to construct the outer function application form before identifying the intended function.
-To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type.
-This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}.
-In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remain under construction.
-%In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}.
-%Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)
-
-% By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state.
-%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics.
-Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.
-
-%However, these syntactically correct states can be semantically meaningless (i.e.
-undefined), because the language definitions generally only give meaning to complete, well-typed terms.
-(e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
-% This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
-% Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
-%Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g.
-Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .
+Na\"ively, enforcing an injunction on ill-typed edit states would force
+programmers to construct programs in a rigid ``outside-in'' manner. For
+example, the programmer would often need to construct the outer function
+application form before identifying the intended function. To address this
+problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole
+if the expression's type is inconsistent with the expected type. This
+meaningfully defers the type consistency check until the expression inside
+the hole is \emph{finished}. In other words, holes appear both at the
+leaves and at the internal nodes of an H-expression that remain under
+construction.
 
@@
 \begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
-\item We begin in Sec.
-    \ref{sec:example} with two examples of edit sequences to develop the reader's intuitions.
-\item We then give a detailed overview of Hazelnut's semantics and metatheory, and our mechanization in the Agda proof assistant, in Sec.  \ref{sec:hazel}.
-\item Hazelnut is a {foundational} calculus, i.e.
-a calculus that language and editor designers are expected to extend with higher level constructs.
-We extend Hazelnut with simple sum types in Sec.
-\ref{sec:extending} to demonstrate how Hazelnut's rich metatheory guides such a development.
-\item In Sec.
-\ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program.
-Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
+  \item We begin in Sec.  \ref{sec:example} with two examples of edit
+    sequences to develop the reader's intuitions.
 
-% \item However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
-% so doing, we discover interesting connections with gradual typing and contextual modal type theory.
-The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes.
-\todo{revise this one}
+  \item We then give a detailed overview of Hazelnut's semantics and
+  metatheory, and our mechanization in the Agda proof assistant, in
+  Sec.  \ref{sec:hazel}.
 
-\item In Sec.
-\ref{sec:rw}, we summarize related work.
-In particular, much of the technical machinery needed to handle type holes coincides with machinery developed in the study of gradual type systems.
-Similarly, expression holes can be interpreted as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work.
-%This suggests a principled design for an ``edit and resume'' feature.
+  \item Hazelnut is a {foundational} calculus, i.e. a calculus that
+  language and editor designers are expected to extend with higher level
+  constructs. We extend Hazelnut with simple sum types in
+  Sec.  \ref{sec:extending} to demonstrate how Hazelnut's rich metatheory
+  guides such a development.
 
-\item We conclude in Sec.  \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
+  \item In Sec.  \ref{sec:impl}, we briefly describe how Hazelnut's action
+  semantics lends itself to efficient implementation as a functional
+  reactive program. Our reference implementation is written
+  using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
+
+  \item In Sec.  \ref{sec:rw}, we summarize related work. In particular,
+  much of the technical machinery needed to handle type holes coincides
+  with machinery developed in the study of gradual type systems. Similarly,
+  expression holes can be interpreted as the closures of contextual modal
+  type theory, which, by its correspondence with contextual modal logic,
+  lays logical foundations beneath our work.
+
+  \item We conclude in Sec.  \ref{sec:future} by summarizing our vision of
+  a principled science of structure editor design rooted in type theory,
+  and suggest a number of future directions.
 \end{itemize}
-The supplemental material includes 1) a typeset listing of Hazelnut's rules in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our reference implementation of Hazelnut, both in source form and pre-compiled to JavaScript for use in a web browser.
+The supplemental material includes 1) a typeset listing of Hazelnut's rules
+in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our
+reference implementation of Hazelnut, both in source form and pre-compiled
+to JavaScript for use in a web browser.

### 2016-09-24 formally.tex
pass through for the easy syntactic errors pointed out in the reviews. this closes #8.

@@
 \end{equation*}
-Intead, we combine these two possible rules into a single rule through the
+Instead, we combine these two possible rules into a single rule through the
 simple auxiliary \emph{matched arrow type} judgement,

### 2016-09-24 hazelnut-popl17.tex
working on #6
pass through for the easy syntactic errors pointed out in the reviews. this closes #8.

@@
 \def\OPTIONConf{1}%
-%\newif\iftr\trfalse
 \newif\iftr\trtrue
-%\documentclass[12pt]{article}
 \RequirePackage{etex}
@@
 (though there is no technical documentation on virtually any aspect of its
-design.) Unlike Hazelnut, it does not have a notion analagous to Hazelnut's
+design.) Unlike Hazelnut, it does not have a notion analogous to Hazelnut's
 notion of a non-empty hole. As such, programmers must construct programs in
@@
 where every intermediate state is well-defined in some sense.
-%Typed structure editing endows this intuitive notion with technical force.
 
@@
 support for working with bidirectional type systems as future work. This
-suggests the possibility of an analagous ``Editorializer'' that generates a
+suggests the possibility of an analogous ``Editorializer'' that generates a
 specification of a typed structure editor from a simple language
@@
 
-% \todo{Eclipse parsing partial programs // other heuristics}
-
-% \paragraph{Gradual typing: Has Matched Relation}
-% %
-% A. Rastogi, A. Chaudhuri, and B. Hosmer. The ins and outs of gradual type inference.
-% \\
-% Definition 3.1 is similar to our ``has matched'' relations:
-% \\
-% \url{https://www.cs.umd.edu/~avik/papers/iogti.pdf}
-
-% This paper defines ``has matched'' relation for arrow, just like us:
-% \\
-% \url{http://cimini.info/publications/Gradualizer_Draft.pdf}
-% \\
-% They credit this technique to the following paper:
-% \\
-% \url{http://www.cs.ubc.ca/~rxg/ptsgp.pdf}
-% \\
-% But, the Rastogi paper came earlier.
-
-% \todo{TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}}
-
-% Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g.
-Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes.
-These  systems are also, to our knowledge, not formally well-defined but rather exist only as part of system implementations.
-
-% Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified.
-For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model.
-Unison is also a rather large body of Haskell code, though its action model appears superficially more similar to ours.
-We maintain what we believe to be a stronger action sensibility invariant than Unison (i.e.
-in Unison, one must construct expressions from the outside-in.) These systems are rich sources of interesting ideas, however -- there is room enough for many different approaches in this (re-)emerging space.
-
-% \todo{TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}}
-
-% \todo{TODO: cite} \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}
-
-%Drag-and-drop / for novices: lots of examples, e.g. Alice and others
-%
-%Contemporary: Lamdu, MPS/Mbeddr, TouchDevelop
-%
-%Hybrid: Cyrus' active code completion paper
-
-%\subsection{Refactoring Models}
-%(Michael, can you fill this section out?)
-
-%\subsection{Formal Editor Models}
-%Need to do a search to see what else has been done...
-
 \section{Conclusion}
@@
 details do not enter into our design.
-
-% By keeping the program in both a structurally and semantically
-% well-defined state at all times, Hazelnut allows users to avoid premature
-% commitment~\cite{green1996usability}.  By inserting a hole, the user can
-% leave certain parts of the program unfinished, \todo{What is the best
-% word to define this concept?  In progress, unfinished, incomple?} and yet
-% still in a well defined type state.  This also enables progressive
-% evaluation~\cite{green1996usability}, because unfinished solutions are
-% also are well-defined at all times, thus enabling tools to provide
-% evaluation of unfinished solutions.\todo{integrate this}
 
@@
 
-% We already discussed a connection to gradual typing \cite{Siek06a}. We
-% hope to explore this connection more thoroughly. In particular, it may be
-% possible to better support exploratory and live programming by allowing
-% even programs with holes in them to execute as long as those holes are
-% only in the type portions, by deferring to the semantics given in work on
-% gradual typing.
-
-% It may also be possible to give a dynamics to incomplete
-% expressions. Prior work on staged evaluation suggests that there may be a
-% connection to modal logic, viewing holes as quantifying over all possible
-% terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In
-% developing a dynamic semantics, we will also need to handle terms like
-% $\hhole{\hehole}$ and $\hhole{\hhole{\hexp}}$. In our semantics given
-% here, we eliminated them as they came up in a somewhat \emph{ad hoc}
-% manner. We have not yet explored an equational theory for terms with
-% holes, but intend to once our formalization effort is more
-% mature.\todo{fix bib}
-
-% \todo{new action form that makes actions extensible given proofs of
-% admissiblity of a derived form, like prevSib}
-
 \begin{quote}

### 2016-09-27 example.tex
removing nextSib in favor of the smaller set of actions "move parent" and "move child n". this makes the math in the paper consistent with the agda formalization, which got this update earlier tonight.
i beleive, but should be checked, that i didn't mess up any of the line
number in the example. i think i also caught the text that talked about
nextsib, mostly by deleting the macro and seeing what broke.

someone else should review this commit and make sure that is correct.

we should also consider adding text how you can get next and prev sib as
derived actions, maybe in the section where we t

@@
 
-The actions on Lines 2-4 complete the type ascription. In particular, the
+The actions on Lines 2-5 complete the type ascription. In particular, the
 $\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and
-the $\aMove{\dNext}$ action moves the cursor to the next hole.
+composition of the $\aMove{\dParent}$ and $\aMove{\dChildn{2}}$ actions
+move the cursor to the next hole.
 
@@
 \hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
-\aMove{\dNext{}} &
-\text{({\ref{rule:move-nextSib-arr}})}
+\aMove{\dParent} &
+\text{({\ref{rule:move-parent-arr-left}})}
 \\ 4 &
+\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\hhole{}}} &
+\aMove{\dChildn{2}} &
+\text{({\ref{rule:move-arr-c2}})}
+\\ 5 &
 \hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
@@
 \text{(\ref{r:contnum})}
-\\ 5 &
+\\ 6 &
 \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
@@
 \text{(\ref{rule:move-parent-arr-right})}
-\\ 6 &
+\\ 7 &
 \hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
@@
 \text{(\ref{rule:move-parent-asc-right})}
-\\ 7 &
+\\ 8 &
 \zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} &
-\aMove{\dChild} &
-\text{(\ref{r:movefirstchild})}
-\\ 8 &
+\aMove{\dChildn{1}} &
+\text{(\ref{r:movefirstchild-asc})}
+\\ 9 &
 \zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
-\aMove{\dChild{}} &
+\aMove{\dChildn{1}} &
 \text{(\ref{r:movefirstchild-lam})}
-\\ 9 &
+\\ 10 &
 \hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
@@
 \text{(\ref{r:conevar})}
-\\ 10 &
+\\ 11 &
 \hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
@@
 \text{(\ref{rule:construct-plus-compat})}
-\\ 11 &
+\\ 12 &
 \hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
@@
 \text{(\ref{r:conenumnum})}
-\\ 12 &
+\\ 13 &
 \hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} &
@@
 
-The actions on Lines 5-8 move the cursor to the function body. For
+The actions on Lines 6-9 move the cursor to the function body. For
 simplicity, Hazelnut defines only a small collection of primitive movement
@@
 
-Lines 9-11 complete the function body by first constructing the variable
+Lines 10-12 complete the function body by first constructing the variable
 $x$, then constructing the plus form, and finally constructing the number
@@
 \hline
-13 &
+14 &
 \zwsel{\hhole{}} &
@@
 \text{(\ref{r:conevar})}
-\\ 14 &
+\\ 15 &
 \zwsel{incr} &
@@
 \text{(\ref{r:coneapfn})}
-\\ 15 &
+\\ 16 &
 incr(\zwsel{\hhole{}}) &
@@
 \text{(\ref{r:conevar2})}
-\\ 16 &
+\\ 17 &
 incr(\hhole{\zwsel{incr}}) &
@@
 \text{(\ref{r:coneapfn})}
-\\ 17 &
+\\ 18 &
 incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} &
@@
 \text{(\ref{r:conenumnum})}
-\\ 18 &
+\\ 19 &
 incr(\hhole{incr(\zwsel{\hnum{3}})}) &
@@
 \text{(\ref{r:moveparent-ap2})}
-\\ 19 &
+\\ 20 &
 incr(\hhole{\zwsel{incr(\hnum{3})}}) &
@@
 \text{(\ref{r:moveparent-hole})}
-\\ 20 &
+\\ 21 &
 incr(\zwsel{\hhole{incr(\hnum{3})}})&
@@
 \text{(\ref{r:finishana})}
-\\ 21 &
+\\ 22 &
 incr(\zwsel{incr(\hnum{3})}) &
@@
 
-We begin on Line 13 by constructing the variable $incr$. Line 14 then
+We begin on Line 14 by constructing the variable $incr$. Line 15 then
 constructs the application form with $incr$ in function position, leaving
@@
 
-We now need to apply $incr$ again, so we perform the same action on Line 15
-as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic
+We now need to apply $incr$ again, so we perform the same action on Line 16
+as we did on Line 14, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic
 structure editor, performing such an action would result in the following
@@
 The programmer could alternatively have performed the $\aConstruct{\fap}$
-action on Line 15. This would result in the following edit state, which is
+action on Line 16. This would result in the following edit state, which is
 well-typed according to the static semantics that we will define in the
@@
 function before constructing the function application form. This stands in
-contrast to Lines 13-14.
+contrast to Lines 14-15.
 
 Hazelnut's action semantics addresses this problem: rather than disallowing
-the $\aConstruct{\fvar{incr}}$ action on Line 15, it leaves $incr$ inside a
+the $\aConstruct{\fvar{incr}}$ action on Line 16, it leaves $incr$ inside a
 hole:
@@
 The expression inside a non-empty hole must itself be well-typed, so the
-programmer can continue to edit it. Lines 16-17 proceed to apply the inner
+programmer can continue to edit it. Lines 17-18 proceed to apply the inner
 mention of $incr$ to a number literal, $\hnum{3}$. Finally, Lines 18-19
-move the cursor to the non-empty hole and Line 20 performs the $\aFinish$
+move the cursor to the non-empty hole and Line 21 performs the $\aFinish$
 action. The $\aFinish$ action removes the hole if the type of the
 expression inside the hole is consistent with the expected type, as it now
-is. This results in the final edit state on Line 20, as desired. In
+is. This results in the final edit state on Line 22, as desired. In
 practice, the editor might automatically perform the $\aFinish$ action as

### 2016-10-06 hazelnut-popl17.tex
if we're gonna be faux french, we might as well spell it right

@@
 \usepackage{rulelinks}
-
 \newtheorem{theorem}{Theorem}
@@
 bidirectionally typed lambda calculus extended with \emph{holes} and
-a \emph{cursor} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond
+a \emph{cursor} (\emph{\`{a} la} Huet's zipper.) Hazelnut goes one step beyond
 syntactic well-formedness: its {edit actions} operate over statically

### 2016-10-06 formally.tex
removing nextSib in favor of the smaller set of actions "move parent" and "move child n". this makes the math in the paper consistent with the agda formalization, which got this update earlier tonight.
i beleive, but should be checked, that i didn't mess up any of the line
number in the example. i think i also caught the text that talked about
nextsib, mostly by deleting the macro and seeing what broke.

someone else should review this commit and make sure that is correct.

we should also consider adding text how you can get next and prev sib as
derived actions, maybe in the section where we t

@@
   \aMove{\delta} ~\vert~
+  \aConstruct{\psi} ~\vert~
   \aDel ~\vert~
-  \aConstruct{\psi} ~\vert~
   \aFinish\\
 \mathsf{Dir} & \delta & ::= &
-  \dChild ~\vert~
-  \dParent ~\vert~
-  \dNext \\
+  \dChildnm{n} ~\vert~
+  \dParent\\
 \mathsf{Shape} & \psi & ::= &
@@
     }{
-      \aMove{\dChild}
+      \aMove{\dChildn{1}}
     }{
@@
 \end{equation}
-\begin{equation}
+\begin{equation}\label{rule:move-arr-c2}
+  \inferrule{ }{
+    \performTyp{
+      \zwsel{\tarr{\htau_1}{\htau_2}}
+    }{
+      \aMove{\dChildn{2}}
+    }{
+      \tarr{\htau_1}{\zwsel{\htau_2}}
+    }
+  }
+\end{equation}
+\begin{equation}\label{rule:move-parent-arr-left}
   \inferrule{ }{
@@
       \zwsel{\tarr{\htau_1}{\htau_2}}
-    }
-  }
-\end{equation}
-\begin{equation}\label{rule:move-nextSib-arr}
-  \inferrule{ }{
-    \performTyp{
-      \tarr{\zwsel{\htau_1}}{{\htau_2}}
-    }{
-      \aMove{\dNext}
-    }{
-      {\tarr{\htau_1}{\zwsel{\htau_2}}}
     }
@@
   \begin{equation}
-    \label{r:movefirstchild}
+    \label{r:movefirstchild-asc}
   \inferrule{ }{
@@
     }{
-      \aMove{\dChild}
+      \aMove{\dChildn{1}}
     }{
       \zwsel{\hexp} : \htau
+    }
+  }
+  \end{equation}
+  \begin{equation}
+    \label{r:movesecondchild-asc}
+    \inferrule{ }{
+    \performTyp{
+      \zwsel{\hexp : \htau}
+    }{
+      \aMove{\dChildn{2}}
+    }{
+      \hexp : \zwsel{\htau}
     }
@@
 \end{equation}
-\begin{equation}
-  \label{r:movenextsib}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\hexp} : \htau
-    }{
-      \aMove{\dNext}
-    }{
-      \hexp : \zwsel{\htau}
-    }
-  }
-\end{equation}
 
@@
   }{
-    \aMove{\dChild}
+    \aMove{\dChildn{1}}
   }{
@@
     }{
-      \aMove{\dChild}
+      \aMove{\dChildn{1}}
     }{
       \hap{\zwsel{\hexp_1}}{\hexp_2}
+    }
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \zwsel{\hap{\hexp_1}{\hexp_2}}
+    }{
+      \aMove{\dChildn{2}}
+    }{
+      \hap{\hexp_1}{\zwsel{\hexp_2}}
     }
@@
 \end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performMove{
-      \hap{\zwsel{\hexp_1}}{\hexp_2}
-    }{
-      \aMove{\dNext}
-    }{
-      \hap{\hexp_1}{\zwsel{\hexp_2}}
-    }
-  }
-\end{equation}
 
@@
     }{
-      \aMove{\dChild}
+      \aMove{\dChildn{1}}
     }{
       \hadd{\zwsel{\hexp_1}}{\hexp_2}
+    }
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{ }{
+    \performMove{
+      \zwsel{\hadd{\hexp_1}{\hexp_2}}
+    }{
+      \aMove{\dChildn{2}}
+    }{
+      \hadd{\hexp_1}{\zwsel{\hexp_2}}
     }
@@
 \end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performMove{
-      \hadd{\zwsel{\hexp_1}}{\hexp_2}
-    }{
-      \aMove{\dNext}
-    }{
-      \hadd{\hexp_1}{\zwsel{\hexp_2}}
-    }
-  }
-\end{equation}
 
@@
 cursor from the outermost position to any other position. This serves as a
-check on our $\aMove{\dChild}$ and $\aMove{\dNext}$ rules.
+check on our $\aMove{\dChildnm{n}}$ rules.
 \begin{lemma}[Reach Down]\label{lemma:reach-down} ~
@@
 
-\subsubsection{Deletion} The $\aDel$ action inserts an empty hole at the
-cursor, deleting what was there before.
-
-The type action rule for $\aDel$ is self-explanatory:
-\begin{equation}
-  \inferrule{ }{
-    \performTyp{
-      \zwsel{\htau}
-    }{
-      \aDel
-    }{
-      \zwsel{\tehole}
-    }
-  }
-\end{equation}
-
-Deletion within a Z-expression is similarly straightforward:
-\begin{subequations}
-\begin{equation}
-  \inferrule{ }{
-    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
-  }
-\end{equation}
-\begin{equation}
-  \inferrule{ }{
-    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
-  }
-\end{equation}
-\end{subequations}
-
 \subsubsection{Construction}\label{sec:construction} The construction
@@
 
+\subsubsection{Deletion} The $\aDel$ action inserts an empty hole at the
+cursor, deleting what was there before.
+
+The type action rule for $\aDel$ is self-explanatory:
+\begin{equation}
+  \inferrule{ }{
+    \performTyp{
+      \zwsel{\htau}
+    }{
+      \aDel
+    }{
+      \zwsel{\tehole}
+    }
+  }
+\end{equation}
+
+Deletion within a Z-expression is similarly straightforward:
+\begin{subequations}
+\begin{equation}
+  \inferrule{ }{
+    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
+  }
+\end{equation}
+\begin{equation}
+  \inferrule{ }{
+    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
+  }
+\end{equation}
+\end{subequations}
+Unlike the relative movement and construction actions, there is no
+``checksum'' theorem for deletion. The rules do not inspect the structure
+of the expression in the cursor, so they both match our intuition and will
+be correct in any extension of the language without modification.
+
 \subsubsection{Finishing}
@@
 \end{subequations}
+Like deletion, there is no need for a ``checksum'' theorem for the
+finishing.
 

### 2016-10-06 example.tex
adding text so people aren't confused by lack of zipper rules, close #2

@@
 discuss in this section using the implementation.
+
+For concision, the right column in both Fig. \ref{ex1} and Fig \ref{ex2}
+lists only the rules that describe the edit \emph{at the cursor}. This
+omits both the rules needed to work down the tree structure of the
+{Z-expression} to the cursor and the rules to propagate the change back
+up. The missing rules are described in detail in
+Sec. \ref{sec:zipper-cases}, and these examples are included in full in the
+Agda mechanization described in Sec. \ref{sec:mech}.
 
@@
 \begin{figure}[t!]
+  \label{ex1}
 \begin{center}
@@
 \begin{figure}[t!]
+  \label{ex2}
 \begin{center}

### 2016-10-06 formally.tex
updating movement erasure statements to actually fit the judgemental forms. these imply the lemmas that were stated before, as well as pin and a few other things. they also match the form of the other theorems given more closely, so i like them much better. #3
mentioning structural props close #12

@@
 includes the complete collection of rules, in definitional order. These
-rules, along with the proofs of two key metatheorems, have been mechanized
-in Agda \cite{norell:thesis}, also in the supplement. We will summarize
-this effort in Sec. \ref{sec:mech}.
+rules, along with the proofs of all of the metatheorems listed, have been
+mechanized in Agda \cite{norell:thesis}, also in the supplement. All the
+judgements that mention contexts described below enjoy the standard
+structural properties of weakening, contraction, and exchange, and this too
+is proven in the mechanization. We will summarize this effort in
+Sec. \ref{sec:mech}.
 
@@
 quite hazardous -- we might make a mistake absent-mindedly. One check
-against this is to prove a lemma that establishes that movement actions
-move the cursor but do not change the cursor erasure.
+against this is to prove that establishes that movement actions do not
+change the cursor erasure, as in Theorem \ref{lemma:movement-erasure}.
 
@@
 
-  \item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then
-    $\removeSel{\zexp}=\removeSel{\zexp'}$.
+  \item If
+    $\performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}$
+    then $\removeSel{\zexp}=\removeSel{\zexp'}$.
+
+  \item If $\performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}$
+    then $\removeSel{\zexp}=\removeSel{\zexp'}$.
+
 \end{enumerate}

### 2016-10-06 formally.tex
close #13

@@
 the multi-dimensional design space around the question of evaluating
-incomplete H-expressions.)
+incomplete H-expressions.) For completeness, a mapping from the
+bidirectional type system to a declarative one, and a of type safety for
+complete terms in that system, is included in the mechanization.
 

### 2016-10-06 agda.tex
working on #6
updating text to reflect that everything is proven

--- 2016-07-07 agda.tex
@@
 % !TEX root = hazelnut-popl17.tex
-So far, we have given an overview of the most important judgements and rules in
-the semantics of Hazelnut, and stated the critical metatheorem, Sensibility, and several auxiliary ``checks''. In a few
-cases, we have informally sketched out why these metatheorems will hold. In order to formally establish that our design meets our stated objectives, we
-have mechanized the semantics of Hazelnut using the Agda proof assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
+So far, we have given an overview of the most important judgements and
+rules in the semantics of Hazelnut, and stated the critical metatheorem,
+Sensibility, and several auxiliary ``checks''. In a few cases, we have
+informally sketched out why these metatheorems will hold. In order to
+formally establish that our design meets our stated objectives, we have
+mechanized the semantics of Hazelnut using the Agda proof
+assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
 at \url{http://wiki.portal.chalmers.se/agda/}.)
 
-This development is available in the supplemental material.
-At the time of submission, the full set of rules is available and the proofs of Theorem \ref{thrm:actsafe}, Theorem \ref{lemma:movement-erasure} and a number of technical lemmas that we elided or mentioned informally in the text are complete.
-We also have nearly all of the cases of Reachability and Constructability complete, and we see no major barrier to the remaining cases -- we prioritized developing a clear exposition of our contributions after proving representative cases and finding that they were straightforward.
-The remaining cases will be completed in the repository linked in the non-anonymous supplementary material.\todo{talk about determinism either here or there}
+This development is available in the supplemental material. It contains the
+full set of rules given in the paper, and proofs of all of the theorems
+described. It also contains the rules needed for the extension to the
+language described in Sec. \ref{sec:extending}, and extends the proofs to
+the constructs.
 
-The documentation 
-includes a more detailed discussion of the technical representation
-decisions that we made. The main idea is standard: we encode each judgement as a
-dependent type. The rules defining the judgements become the constructors of this
-type, and derivations are terms of these type. This is a rich
-setting that allows proofs to take advantage of pattern matching on the
-shape of derivations, closely matching standard on-paper proofs.
-No proof automation was used, because the proof structure itself is likely to be interesting to researchers who plan to build upon our work.
-% The formalization differs from the calculus defined here in a few small 
-% ways. Most interestingly, instead of giving separate movement rules for each 
-% form of Z-expression, we abstract
-% over different corresponding forms that happen to have the same arity, e.g. 
-% additions and applications. Formalizing this intuition
-% reduces the number of cases we need to consider 
-% somewhat, but more importantly allows us to write a slightly more
-% general calculus -- it will be easier to extend Hazelnut with more 
-% interesting language features with this generic infrastructure in place.
-We adopt Barendregt's convention for
-bound variables \cite{urban}. Hazelnut's semantics does not need substitution, so we do not need to adopt more sophisticated encodings (e.g. \cite{lh09unibind,Pouillard11}.)
+The documentation includes a more detailed discussion of the technical
+representation decisions that we made. The main idea is standard: we encode
+each judgement as a dependent type. The rules defining the judgements
+become the constructors of this type, and derivations are terms of these
+type. This is a rich setting that allows proofs to take advantage of
+pattern matching on the shape of derivations, closely matching standard
+on-paper proofs. No proof automation was used, because the proof structure
+itself is likely to be interesting to researchers who plan to build upon
+our work.
+
+We adopt Barendregt's convention for bound
+variables \cite{urban}. Hazelnut's semantics does not need substitution, so
+we do not need to adopt more sophisticated encodings
+(e.g. \cite{lh09unibind,Pouillard11}.)

### 2016-10-06 formally.tex
updating the section on determinism. i only describe the predicate intuitively in english, rather than giving a full description of it mathematically. i don't know if this is appropriate.

@@
 \end{subequations}
+
+\subsection{Determinism}
+A last useful property to consider is \emph{action determinism}, i.e. that
+performing an action produces a unique result. Formally, this would be
+established as follows:
+\begin{theorem}[Action Determinism]
+  \label{thrm:actdet} All of the following hold:
+  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+  \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
+    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
+  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
+    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
+    $\zexp' = \zexp''$ and $\htau' = \htau''$.
+  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
+    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
+    \zexp''$.
+  \end{enumerate}
+\end{theorem}
+This is not a theorem of the system exactly as described so far. The reason
+is quite subtle: the semantics of two construction actions,
+$\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the
+analytic case than they do in the synthetic case -- the difference has to
+do with what the generated ascription contains. Both cases can technically
+``fire'' due to action subsumption, but one corresponds to an over-eager
+use of subsumption: it's intended to be a rule of last resort.
+
+The mechanization defines a predicate on action derivations that bans
+exactly these over-eager uses of action subsumption, gives a mapping from
+any derivation into one that matches the source and action and satisfies
+the predicate, and proves determinism as stated above but only for
+derivations modulo that predicate. This fix is analogous to the difference
+between a chaining and focused logic in proof theory, which correspond
+closely with bidirectional type checking systems, and therefore is to be
+expected.\cite{Simmons11tr}

### 2016-10-07 formally.tex
updated version of movement erasure that resolves the question about pin, and is a much more uniform description. close #3.
typo

@@
 incomplete H-expressions.) For completeness, a mapping from the
-bidirectional type system to a declarative one, and a of type safety for
-complete terms in that system, is included in the mechanization.
+bidirectional type system to a declarative one, and a proof of type safety
+for complete terms in that system, is included in the mechanization.
 
@@
 
-  \item If
-    $\performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}$
-    then $\removeSel{\zexp}=\removeSel{\zexp'}$.
-
-  \item If $\performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}$
-    then $\removeSel{\zexp}=\removeSel{\zexp'}$.
+  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau'}$
+    then $\removeSel{\zexp}=\removeSel{\zexp'}$ and $\htau=\htau'$.
+
+  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}$ then
+    $\removeSel{\zexp}=\removeSel{\zexp'}$.
 

### 2016-10-07 agda.tex
enhanced determinism section

@@
 % !TEX root = hazelnut-popl17.tex
-So far, we have given an overview of the most important judgements and
-rules in the semantics of Hazelnut, and stated the critical metatheorem,
-Sensibility, and several auxiliary ``checks''. In a few cases, we have
-informally sketched out why these metatheorems will hold. In order to
+% So far, we have given an overview of the most important judgements and
+% rules in the semantics of Hazelnut, and stated the critical metatheorem,
+% Sensibility, and several auxiliary ``checks''. In a few cases, we have
+% informally sketched out why these metatheorems will hold.
+
+In order to
 formally establish that our design meets our stated objectives, we have
-mechanized the semantics of Hazelnut using the Agda proof
+mechanized the semantics and metatheory of Hazelnut as described above (and more concisely in the appendix) using the Agda proof
 assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
-at \url{http://wiki.portal.chalmers.se/agda/}.)
-
-This development is available in the supplemental material. It contains the
-full set of rules given in the paper, and proofs of all of the theorems
-described. It also contains the rules needed for the extension to the
-language described in Sec. \ref{sec:extending}, and extends the proofs to
-the constructs.
+at \url{http://wiki.portal.chalmers.se/agda/}.) This development is available in the supplemental material. The mechanization also includes the  extension to the
+language described in Sec. \ref{sec:extending}.
 

### 2016-10-07 formally.tex
enhanced determinism section

@@
 A last useful property to consider is \emph{action determinism}, i.e. that
-performing an action produces a unique result. Formally, this would be
-established as follows:
-\begin{theorem}[Action Determinism]
-  \label{thrm:actdet} All of the following hold:
+performing an action produces a unique result. The following theorem establishes determinism for type actions:
+\begin{theorem}[Type Action Determinism]
+\label{thrm:type-actdet} If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
+    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
+\end{theorem}
+
+The corresponding theorem for expression actions would be stated as follows:
   \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
-  \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
-    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
   \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
@@
   \end{enumerate}
+
+This is not a theorem of the system as described so far. The reason
+is somewhat subtle: two construction actions,
+$\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the
+analytic case than they do in the synthetic case. The problem is that both rules can  
+``fire'' when considering a Z-expression in analytic position due to action subsumption.
+This is a technically valid but ``morally'' invalid use of action subsumption: subsumption is included in the system to be used as a rule of last resort, i.e.
+it should only be applied when no other analytic action rule can fire.
+
+There are several possible ways to address this problem.
+One approach would be to modify the judgement forms to internalize this notion of ``rule of last resort''.
+This approach is related to focusing from proof theory \cite{Simmons11tr}.
+However, this approach would substantially complicate our presentation of the system.
+
+The approach that we take leaves the system unchanged.
+Instead, we define predicates over \emph{derivations} of the expression action judgements that excludes those derivations that apply subsumption prematurely, i.e.
+when another rule could have been applied.
+We call such derivations \emph{subsumption-minimal derivations}.
+We can establish determinism for subsumption-minimal derivations.
+\begin{theorem}[Expression Action Determinism] ~
+  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
+  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\mathcal{D}_1 : \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
+    $\mathcal{D}_2 : \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ and $\subminsyn{\mathcal{D}_1}$ and $\subminsyn{\mathcal{D}_2}$ then
+    $\zexp' = \zexp''$ and $\htau' = \htau''$.
+  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
+    $\mathcal{D}_1 : \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
+    $\mathcal{D}_2 : \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ and $\subminana{\mathcal{D}_1}$ and $\subminana{\mathcal{D}_2}$ then $\zexp' =
+    \zexp''$.
+  \end{enumerate}
 \end{theorem}
-This is not a theorem of the system exactly as described so far. The reason
-is quite subtle: the semantics of two construction actions,
-$\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the
-analytic case than they do in the synthetic case -- the difference has to
-do with what the generated ascription contains. Both cases can technically
-``fire'' due to action subsumption, but one corresponds to an over-eager
-use of subsumption: it's intended to be a rule of last resort.
-
-The mechanization defines a predicate on action derivations that bans
-exactly these over-eager uses of action subsumption, gives a mapping from
-any derivation into one that matches the source and action and satisfies
-the predicate, and proves determinism as stated above but only for
-derivations modulo that predicate. This fix is analogous to the difference
-between a chaining and focused logic in proof theory, which correspond
-closely with bidirectional type checking systems, and therefore is to be
-expected.\cite{Simmons11tr}
+
+The mechanization, discussed next, defines the predicates $\subminsyn{\mathcal{D}}$ and $\subminana{\mathcal{D}}$. In addition, it defines a mapping from
+any derivation into a corresponding subsumption-minimal derivation. Implementors of Hazelnut are expected only to implement this subsumption-minimal fragment.

### 2016-10-07 formally.tex
typo

@@
 The approach that we take leaves the system unchanged.
-Instead, we define predicates over \emph{derivations} of the expression action judgements that excludes those derivations that apply subsumption prematurely, i.e.
+Instead, we define predicates over \emph{derivations} of the expression action judgements that exclude those derivations that apply subsumption prematurely, i.e.
 when another rule could have been applied.

### 2016-10-07 implementation.tex
updated implementation section for AE

@@
 \subsection{HZ}
-We have developed a simple reference implementation, HZ, of Hazelnut.  In
+We have developed a simple reference implementation, HZ, of Hazelnut extended with sum types as described in Sec. \ref{sec:extending}.  In
 order to reach a wide audience, we decided to implement HZ in the web
@@
 
-At the time of the writing, HZ's view computation renders the model as a
-string embedded into HTML that approximates our typeset notation in this
-paper. The action palette is a collection of buttons and text boxes, which
-are disabled when an action is not well-defined or an invalid action
-argument is entered. We determine this by simply attempting to perform the
-action internally and catching the exception that is raised when an action
-is undefined. This design, of course, is not meant to score marks for
+The core semantics have
+been implemented in a functional style that follows the presentation in this paper closely.
+
+The view computation renders the model (i.e.
+a Z-expression paired with a type) as nested HTML \texttt{div} elements matching the tree structure of the corresponding Z-expression.
+This tree is stylized separately using CSS.
+The action palette is a collection of buttons and text boxes which
+are disabled when the corresponding action cannot be performed. We determine this by simply attempting to perform the
+action internally and handling the exception that is raised when an action
+is undefined. Each action form also has a corresponding keyboard shortcut. This design is not, of course, meant to score marks for
 usability or performance. Instead, as a simple reference implementation, we
 expect its source code to be of use to others who are interested in
-layering a more fluid user interface atop the core semantics (which have
-been implemented in a style that follows the presentation in this paper
-closely), or extensions thereof.
+layering a more fluid user interface atop the core semantics, or extensions thereof.

### 2016-10-07 implementation.tex
minor wordsmithing, and added sentence about how implementation is there to "allow the reader to interact with the concepts presented in this paper, to aid in their understanding. "

@@
 order to reach a wide audience, we decided to implement HZ in the web
-browser.  In order to take advantage of a mature implementation of the FRP
+browser.  To take advantage of a mature implementation of the FRP
 model, we chose to implement HZ using
@@
 is undefined. Each action form also has a corresponding keyboard shortcut. This design is not, of course, meant to score marks for
-usability or performance. Instead, as a simple reference implementation, we
-expect its source code to be of use to others who are interested in
+usability or performance.
+Instead, as a simple reference implementation, it allows the reader to interact with the concepts presented in this paper, to aid in their understanding.
+Additionally, we expect its source code to be of use to others who are interested in
 layering a more fluid user interface atop the core semantics, or extensions thereof.

### 2016-10-07 formally.tex
updated implementation section for AE
prepping paper for artifact evaluation

@@
 
-Our overview below omits certain ``uninteresting'' details. The supplement
+Our overview below omits certain ``uninteresting'' details. The appendix
 includes the complete collection of rules, in definitional order. These
-rules, along with the proofs of all of the metatheorems listed, have been
-mechanized in Agda \cite{norell:thesis}, also in the supplement. All the
-judgements that mention contexts described below enjoy the standard
-structural properties of weakening, contraction, and exchange, and this too
-is proven in the mechanization. We will summarize this effort in
-Sec. \ref{sec:mech}.
+rules, along with the proofs of all of the metatheorems discussed, have been
+mechanized in Agda \cite{norell:thesis} (see Sec. \ref{sec:mech}.)
 
@@
 defining a dynamic semantics for Hazelnut, but the dynamics for complete
-H-expressions would be entirely standard (in Sec. \ref{sec:rw} we discuss
+H-expressions would be entirely standard because complete expressions correspond  immediately to those of the typed lambda calculus with numbers (in Sec. \ref{sec:rw} we discuss
 the multi-dimensional design space around the question of evaluating
-incomplete H-expressions.) For completeness, a mapping from the
-bidirectional type system to a declarative one, and a proof of type safety
-for complete terms in that system, is included in the mechanization.
+incomplete H-expressions.)
 
@@
 \begin{mathpar}
-
 \inferrule{ }{
@@
 }
-~~~~~~~~
+~~~~~
 \inferrule{ }{
@@
 }
-~~~~~~~~
+~~~~~
 \inferrule{
@@
   \tcompat{(\tarr{\htau_1}{\htau_2})}{(\tarr{\htau_1'}{\htau_2'})}
-}~~~~~~~~\text{(\ref*{eqns:consistency}a-d)}
+}~~~~\text{\hspace{-2px}(\ref*{eqns:consistency}a-d)}
 \end{mathpar}
@@
 \end{minipage}
-\begin{minipage}{0.5\linewidth}
+\begin{minipage}{0.48\linewidth}
 \begin{equation}
@@
 \domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner. We
-identify the context up to exchange and adopt the standard identification
-convention for formal structures that differ only up to alpha-equivalence.
+identify the context up to exchange and contraction and adopt the standard identification
+convention for formal structures that differ only up to alpha-renaming of bound variables. All judgements that mention a context obey weakening.
 
@@
 i.e. Rule (\ref{rule:ana-subsume}). Algorithmically, subsumption should be
-the rule of last resort.
+the rule of last resort (see Sec. \ref{sec:determinism}.)
 
@@
 
-\subsection{Determinism}
+\subsection{Determinism}\label{sec:determinism}
 A last useful property to consider is \emph{action determinism}, i.e. that
@@
 The mechanization, discussed next, defines the predicates $\subminsyn{\mathcal{D}}$ and $\subminana{\mathcal{D}}$. In addition, it defines a mapping from
-any derivation into a corresponding subsumption-minimal derivation. Implementors of Hazelnut are expected only to implement this subsumption-minimal fragment.
+any derivation into a corresponding subsumption-minimal derivation. Implementations of Hazelnut need only implement this subsumption-minimal fragment.

### 2016-10-07 implementation.tex

@@
 usability or performance.
-Instead, as a simple reference implementation, it allows the reader to interact with the concepts presented in this paper, to aid in their understanding.
+Instead, as a simple reference implementation, it allows the reader to interact with the system presented in this paper, to aid in their understanding.
 Additionally, we expect its source code to be of use to others who are interested in

### 2016-10-12 acks.tex
Adding acks

Thanks to Ed Morehouse for counsel on mechanization and Barendrecht's
Convention, and to Vincent Zeng for \textit{pro bono} artistic services.

### 2016-11-01 hazelnut-popl17.tex
adding a little more meat to the action macros future work paragraph, closing #11
Adding acks
adding that sweet sweet AEC badge

@@
 \makeatletter
+
+\usepackage[firstpage]{draftwatermark}
+\SetWatermarkText{\hspace*{8in}\raisebox{6.5in}{\includegraphics[scale=0.1]{aec-badge-popl}}}
+\SetWatermarkAngle{0}
 
@@
 functional programs could themselves be lifted to the level of actions to
-compute non-trivial compound actions.
+compute non-trivial compound actions. Such compound actions would give a
+uniform description of transformations ranging from the simple -- like
+``move the cursor to the next hole to the right'' -- to quite complex whole
+program refactoring, while remaining subject to the core Hazelnut
+metatheory.
 
@@
 
+\acks
+\input{acks.tex}
+
 \clearpage

### 2016-11-07 hazelnut-popl17.tex
revised abstract and intro

@@
 \RequirePackage{etex}
-\documentclass[preprint,9pt]{sigplanconf}
+\documentclass[9pt]{sigplanconf}
+\toappear{}
 \usepackage[T1]{fontenc}
@@
 \usepackage{colortab}
-\usepackage{url}
+\usepackage[lowtilde]{url}
 \usepackage{todonotes}
@@
 \makeatletter
-
-\usepackage[firstpage]{draftwatermark}
-\SetWatermarkText{\hspace*{8in}\raisebox{6.5in}{\includegraphics[scale=0.1]{aec-badge-popl}}}
-\SetWatermarkAngle{0}
 
@@
 % known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}
+\usepackage[hidelinks,breaklinks,draft=false]{hyperref}
 
@@
 {
- $^1$~Carnegie Mellon University
+ $^1$~Carnegie Mellon University (USA)
  \and
- $^2$~Oregon State University
+ $^2$~Oregon State University (USA)
  \and
- $^3$~University of Colorado Boulder\vspace{-10px}
+ $^3$~University of Colorado Boulder (USA)\vspace{-10px}
 }
@@
 a program directly. This can have cognitive benefits, particularly for
-novice and end-user programmers (as evidenced by the popularity of
-structure editors like Scratch.) It also simplifies matters for tool
+novice and end-user programmers. It also simplifies matters for tool
 designers, because they do not need to contend with malformed program text.
 
-This paper defines Hazelnut, a {structure editor} based on a small
+This paper introduces Hazelnut, a {structure editor} based on a small
 bidirectionally typed lambda calculus extended with \emph{holes} and
-a \emph{cursor} (\emph{\`{a} la} Huet's zipper.) Hazelnut goes one step beyond
+a \emph{cursor}. Hazelnut goes one step beyond
 syntactic well-formedness: its {edit actions} operate over statically
-meaningful (i.e. well-typed) terms.  Na\"ively, this prohibition on
-ill-typed edit states would force the programmer to construct terms in a
+meaningful incomplete terms.  Na\"ively, this would force the programmer to construct terms in a 
 rigid ``outside-in'' manner. To avoid this problem, the {action semantics}
 automatically places terms assigned a type that is inconsistent with the
-expected type {inside} a {hole}. This safely defers the type consistency
+expected type \emph{inside} a {hole}. This meaningfully defers the type consistency
 check until the term inside the hole is \emph{finished}.
 
-Hazelnut is a foundational type-theoretic account of typed structure
-editing, rather than an end-user tool itself. To that end, we describe how
-Hazelnut's rich metatheory, which we have mechanized in Agda, guides the
-definition of an extension to the calculus. We also discuss various
-plausible evaluation strategies for terms with holes, and in so doing
+Hazelnut is not intended as an end-user tool itself. Instead, it serves as a foundational account of typed structure
+editing. To that end, we describe how
+Hazelnut's rich metatheory, which we have mechanized using the Agda proof assistant, serves as a guide when we extend the calculus to include binary sum types.
+We also discuss various interpretations of holes, and in so doing
 reveal connections with gradual typing and contextual modal type theory
 (the Curry-Howard interpretation of contextual modal logic.) Finally, we
-discuss how Hazelnut's semantics lends itself to implementation as a
-functional reactive program. Our reference implementation is written
-using \lstinline{js_of_ocaml}.
+discuss how Hazelnut's semantics lends itself to implementation as an event-based functional reactive program.
+Our simple reference implementation is written in OCaml and compiled using \lstinline{js_of_ocaml}.
 
@@
 
-%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
-%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
+\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}
+\category{D.2.3}{Software Engineering}{Coding Tools and Techniques}[Program Editors]
 %\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
-%\keywords
-%extensible languages; module systems; type abstraction; typed compilation; type-level computation
+\keywords
+structure editors, bidirectional type systems, gradual typing, mechanized metatheory
 
@@
 \section{Related Work and Discussion}\label{sec:rw}
+\todo{move into own file}
 \subsection{Structure Editors}
@@
 
-\acks
-\input{acks.tex}
-
 \clearpage

### 2016-11-07 intro.tex
pass through for the easy syntactic errors pointed out in the reviews. this closes #8.
revised abstract and intro

@@
 
-Programmers typically construct and manipulate well-typed expressions only
+Programmers typically interact with meaningful programs only
 indirectly, by editing text that is first parsed according to a textual
@@
 indirection has practical benefits, to be sure -- text editors and other
-text-based tools benefit from decades of development effort -- but it also
+text-based tools benefit from decades of development effort. However, this indirection through text also
 introduces some fundamental complexity into the programming process.
@@
 First, it requires that programmers learn the subtleties of the textual
-syntax (e.g. the precedence of the various forms.) This can be particularly
-challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}.
+syntax (e.g. operator precedence.) This is particularly 
+challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258,stefik2013empirical}, and even experienced programmers frequently make mistakes \cite{7548903,jones2006developer}.
 
-The fact that not every sequence of characters corresponds to a meaningful
-program also complicates matters for tool designers. In particular, program
-editors must contend with meaningless text on a regular basis. In a dataset
-analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of
-edit states were syntactically malformed \cite{6883030}. Some additional
-percentage of edit states were well-formed but ill-typed (the dataset did
-not contain enough information to determine the exact percentage.)  Some of
-these meaningless edit states arose because the programmer was in the midst
-of a sequence of edits that left the edit state transiently meaningless,
-while others arose because the programmer made a logical mistake. Because
-the language definitions is silent about these edit states, it is difficult
-to design useful editor services, e.g. syntax
+Second, the fact that many sequences of characters do not correspond to a meaningful
+program complicates the design of program editors and other interactive programming tools. 
+%In particular, program
+%editors must contend with meaningless text on a regular basis. 
+In a dataset  
+gathered by Yoon and Myers consisting of 1460 hours of Eclipse edit logs  \cite{6883030}, 44.2\% of
+edit states were syntactically malformed. Some additional
+percentage of edit states were well-formed but ill-typed (the dataset was
+not complete enough to determine the exact percentage.) Collectively, we refer to these edit states as \emph{meaningless edit states}, because they are not given static or dynamic meaning by the language definition.
+%Some of
+%these meaningless edit states arose because the programmer was in the midst
+%of a sequence of edits that left the edit state transiently meaningless,
+%while others arose because the programmer made a logical mistake. 
+As a consequence, it is difficult
+to design useful language-aware editor services, e.g. syntax
 highlighting~\cite{sarkar2015impact}, type-aware code
@@
 these editor services when they encounter meaningless edit states or
-develop \emph{ad hoc} heuristics, which can be misleading.
+deploy \emph{ad hoc} heuristics, e.g. by using whitespace to guess the programmer's intent \cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.
 
@@
 into \emph{structure editors}, i.e. program editors where every edit state
-is a program structure.
+corresponds to a program structure \cite{teitelbaum_cornell_1981}.\footnote{Structure editors are also variously known as \emph{structured editors}, \emph{structural editors}, \emph{syntax-directed editors} and \emph{projectional editors} (esp.
+when the user interface presents rich views of the program structure.)}
 
 Most structure editors are \emph{syntactic structure editors}, i.e. the
-edit state is a syntax tree with \emph{holes} that stand for branches of
+edit state corresponds to a syntax tree with \emph{holes} that stand for branches of
 the tree that have yet to be constructed, and the edit actions are
@@
 structure editor that has achieved success as a tool for teaching children
-how to program \cite{Resnick:2009:SP:1592761.1592779}.
+how to program \cite{Resnick:2009:SP:1592761.1592779}. %Every well-formed edit state is meaningful according to the semantics of Scratch (in particular, it dynamically continues past erroneous terms.)
 
-The Scratch language has a trivial static semantics, but researchers have
-also designed syntactic structure editors for languages with a non-trivial
-static semantics. For example, \texttt{mbeddr} is an editor for a C-like
-language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an
-object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an
+% Scratch does not impose any static requirements , i.e. every well-formed edit state is trivially statically meaningful. 
+Researchers have also 
+ designed  syntactic structure editors for more sophisticated languages with non-trivial  
+static type systems. Contemporary examples include \texttt{mbeddr}, a structure editor for a C-like
+language \cite{voelter_mbeddr:_2012}, TouchDevelop, a structure editor for an
+object-oriented language \cite{tillmann_touchdevelop:_2011}, and Lamdu, a structure 
 editor for a functional language similar to Haskell \cite{lamdu}. Each of
@@
 type and binding structure of the underlying language complicates its
-design. The reason is that a syntactic structure editor does not guarantee
-that every edit state is statically meaningful -- only that it is
-syntactically well-formed. As in textual program editors, these syntactic
-structure editors must either disable key editor services when they
-encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
+design. The reason is that syntactic structure editors do not assign static meaning to every edit state -- they guarantee only that every edit state corresponds to  a 
+syntactically well-formed tree. These editors must also either disable editor services that require an understanding of the semantics of the program or deploy \emph{ad hoc} heuristics when they
+encounter these meaningless edit states.
 
@@
 the invariant that every edit state is a statically meaningful
-(i.e. well-typed) H-expression with a single superimposed \emph{cursor}. We
+(i.e. well-H-typed) H-expression with a single superimposed \emph{cursor}. We
 call H-expressions and H-types with a cursor \textbf{Z-expressions}
@@
 
-Na\"ively, enforcing an injunction on ill-typed edit states would force
+Na\"ively, enforcing an injunction on ill-H-typed edit states would force
 programmers to construct programs in a rigid ``outside-in'' manner. For
@@
   \item We then give a detailed overview of Hazelnut's semantics and
-  metatheory, and our mechanization in the Agda proof assistant, in
+  metatheory, which has been mechanized using the Agda proof assistant, in
   Sec.  \ref{sec:hazel}.
@@
   language and editor designers are expected to extend with higher level
-  constructs. We extend Hazelnut with simple sum types in
+  constructs. We extend Hazelnut with binary sum types in
   Sec.  \ref{sec:extending} to demonstrate how Hazelnut's rich metatheory
-  guides such a development.
+  guides one such extension.
 
@@
   reactive program. Our reference implementation is written
-  using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.
+  using the OCaml \lstinline{React} library and \lstinline{js_of_ocaml}.
 
@@
   much of the technical machinery needed to handle type holes coincides
-  with machinery developed in the study of gradual type systems. Similarly,
+  with machinery developed for gradual type systems. Similarly,
   expression holes can be interpreted as the closures of contextual modal
   type theory, which, by its correspondence with contextual modal logic,
-  lays logical foundations beneath our work.
+  suggests logical foundations for the system.
 
@@
 \end{itemize}
-The supplemental material includes 1) a typeset listing of Hazelnut's rules
-in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our
-reference implementation of Hazelnut, both in source form and pre-compiled
-to JavaScript for use in a web browser.
+The supplemental material can be accessed from:
+\[\text{\url{http://cs.cmu.edu/~comar/hazelnut-popl17/}} \]
+% 1) a typeset listing of Hazelnut's rules
+%in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our
+%reference implementation of Hazelnut, both in source form and pre-compiled
+%to JavaScript for use in a web browser.\todo{how are we going to distribute supplemental material? maybe a GitHub URL like the one for AE?}

### 2016-11-07 hazelnut-popl17.tex
Fixed #19

@@
 We also discuss various interpretations of holes, and in so doing
-reveal connections with gradual typing and contextual modal type theory
-(the Curry-Howard interpretation of contextual modal logic.) Finally, we
+reveal connections with gradual typing and contextual modal type theory, the Curry-Howard interpretation of contextual modal logic. Finally, we
 discuss how Hazelnut's semantics lends itself to implementation as an event-based functional reactive program.

### 2016-11-08 acks.tex
added an acknowledgement for Danny's COPE grant
final popl17 submission

@@
-Thanks to Ed Morehouse for counsel on mechanization and Barendrecht's
-Convention, and to Vincent Zeng for \textit{pro bono} artistic services.
+% !TEX root = hazelnut-popl17.tex
+
+The authors would like to thank the anonymous referees at POPL and TFP 2016  for useful feedback on earlier drafts of this paper; Ed Morehouse and Carlo Angiuli for counsel on mechanization and Barendrecht's
+Convention; Vincent Zeng for \textit{pro bono} artistic services; and YoungSeok Yoon for reanalyzing the data from \cite{6883030}. 
+%This is Danny Dig's COPE grant. He asked me to add the grant.
+This work was partially funded through the NSF grant \#CCF-1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140.
+The views and conclusions contained herein are those of the authors and should not be interpreted as necessarily representing the official policies or endorsements, either expressed or implied, of the NSF, AFRL, DARPA, the NSA or the U.S.
+Government.

### 2016-11-08 example.tex
revised abstract and intro
final popl17 submission

--- 2016-09-27 example.tex
@@
 % !TEX root = hazelnut-popl17.tex
-The reader is encouraged to follow along with the examples that we will
-discuss in this section using the implementation.
-
 \subsection{Example 1: Constructing the Increment Function}
@@
 formal syntax and define the referenced rules in
-Sec. \ref{sec:hazel}. First, let us build some high-level intuitions.
+Sec. \ref{sec:hazel}.\footnote{For concision, the column labeled \textbf{Rule} in Figures \ref{fig:first-example} and \ref{fig:second-example} 
+indicates only the relevant \emph{non-zipper} rule (see Sec.
+\ref{sec:zipper-cases}.) The reader is encouraged to follow along with these examples using the reference implementation.
+The derivations for these examples are also available in the Agda mechanization.} First, let us build some high-level intuitions.
 
@@
 has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the
-{cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively. For
+{cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively.\footnote{The reference implementation omits the triangles, while the Agda mechanization necessarily omits the colors.} For
 example, on Line 1, the empty expression hole, $\hhole{}$, is under the
@@
 
-Actions act relative to the cursor. The first action that we perform is
-$\aConstruct{\flam{x}}$, which instantiates the hole with a lambda
-abstraction binding the variable $x$. This results in the Z-expression on
-Line 2, consisting of a lambda abstraction with an arrow type
-ascription. The cursor is placed on the argument type hole. Type holes are
-typeset like empty expression holes for visual consistency.
+Actions act relative to the cursor. The first action, on Line 1, is
+$\aConstruct{\flam{x}}$. This results in the Z-expression on
+Line 2, consisting of a lambda abstraction with argument $x$ under an arrow type
+ascription. The cursor is placed (arbitrarily) on the argument type hole.
 
@@
 $\aConstruct{\fnum}$ action constructs the $\tnum$ type at the cursor and
-composition of the $\aMove{\dParent}$ and $\aMove{\dChildn{2}}$ actions
-move the cursor to the next hole.
+the $\aMove{\dParent}$ and $\aMove{\dChildn{2}}$ action sequence 
+moves the cursor to the next hole. In practice, an editor would also define compound movement
+actions, e.g. an action that moves the cursor directly to the next
+ hole, in terms of these primitive movement actions.
 
 \begin{figure}[t!]
+  \label{ex1}
 \begin{center}
-$\arraycolsep=4px
-\begin{array}{|r||l|l||l|l|}
+$\arraycolsep=2px
+\begin{array}[p]{|c||l|l||l|l|}
 \hline
@@
 \\ 2 &
-\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} &
+{\hlam{x}{\hhole{}}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} &
 \aConstruct{\fnum{}} &
@@
 \\ 3 &
-\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
+{\hlam{x}{\hhole{}}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
 \aMove{\dParent} &
@@
 \\ 4 &
-\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\hhole{}}} &
+{\hlam{x}{\hhole{}}} : \zwsel{\tarr{\tnum{}}{\hhole{}}} &
 \aMove{\dChildn{2}} &
@@
 \\ 5 &
-\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
+{\hlam{x}{\hhole{}}} : \tarr{\tnum}{\zwsel{\hhole{}}}
 &
@@
 \\ 6 &
-\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
+{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} &
 \aMove{\dParent{}} &
@@
 \\ 7 &
-\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
+{\hlam{x}{\hhole{}}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
 &
@@
 \\ 8 &
-\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} &
+\zwsel{{\hlam{x}{\hhole{}}} : \tarr{\tnum}{\tnum}} &
 \aMove{\dChildn{1}} &
@@
 \\ 9 &
-\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
+{\zwsel{\hlam{x}{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
 \aMove{\dChildn{1}} &
@@
 \\ 10 &
-\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
+{\hlam{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
 \aConstruct{\fvar{x}} &
@@
 \\ 11 &
-\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
+{\hlam{x}{\zwsel{{x}}}} : \tarr{\tnum{}}{\tnum{}} &
 {\aConstruct{\fplus}}
@@
 \\ 12 &
-\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} &
+{\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}}} : \tarr{\tnum{}}{\tnum{}} &
 \aConstruct{\fnumlit{1}} &
@@
 \\ 13 &
-\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} &
+{\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}}} : \tarr{\tnum}{\tnum} &
 \textrm{---} &
@@
 $\end{center}\vspace{-6px}
-\caption{Constructing an increment function in Hazelnut.}
+\caption{Constructing the increment function in Hazelnut.}
 \label{fig:first-example}
@@
 
-The actions on Lines 6-9 move the cursor to the function body. For
-simplicity, Hazelnut defines only a small collection of primitive movement
-actions. In practice, an editor would also define compound movement
-actions, e.g. an action that moves the cursor directly to the next
-remaining hole, in terms of these primitive movement actions.
-
-Lines 10-12 complete the function body by first constructing the variable
-$x$, then constructing the plus form, and finally constructing the number
+After moving to the function body, Lines 10-12 complete it by first constructing the variable
+$x$, then constructing the plus form, and finally constructing the number literal 
 $\hnum{1}$. Notice that we did not need to construct the function body in
 an ``outside-in'' manner, i.e. we constructed $x$ before constructing the
-outer plus form. Luckily, the transient function bodies, $x$ and $x +
-\hhole{}$, could be checked against type $\tnum$ (as we will detail in
+outer plus form inside which $x$ ultimately appears. The transient function bodies, $x$ and $\hadd{x}{\hhole{}}$, can be checked against the given return type, $\tnum$ (as we will detail in
 Sec. \ref{sec:holes}.)
@@
 \begin{figure}[t!]
+  \label{ex2}
 \begin{center}
-\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
+\colorbox{light-gray}{\hspace{54px} now assume $incr : \tarrnp{\tnum}{\tnum}$ \hspace{53px}}
 $\arraycolsep=4px
@@
 $\end{center}\vspace{-6px}
-\caption{Appyling the increment function.}
+\caption{Applying the increment function.}
 \label{fig:second-example}
@@
 constructs the application form with $incr$ in function position, leaving
-the cursor on a hole in the argument position. Notice that we did not need
-to construct the outer application form before identifying the function
-being applied. Again, the transient edit state happens to be well-typed, so
-we needed to make no special allowances for this order of actions.
+the cursor on a hole in the argument position. Notice again that we did not construct the outer application form before identifying the function
+being applied. %Again, the transient H-expression happens to be well-typed.
 
@@
 $incr$ must be of type $\tnum$ but here it is of type
-$\tarr{\tnum}{\tnum}$. Hazelnut cannot allow such an edit state to arise.
+$\tarr{\tnum}{\tnum}$. Hazelnut does not allow such an edit state to arise.
 
-The programmer could alternatively have performed the $\aConstruct{\fap}$
+We could alternatively have performed the $\aConstruct{\fap}$
 action on Line 16. This would result in the following edit state, which is
@@
 The problem is that the programmer is not able to identify the intended
-function before constructing the function application form. This stands in
+function, $incr$, before constructing the function application form. This stands in
 contrast to Lines 14-15.

### 2016-11-08 intro.tex
final popl17 submission

@@
 syntax (e.g. operator precedence.) This is particularly 
-challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258,stefik2013empirical}, and even experienced programmers frequently make mistakes \cite{7548903,jones2006developer}.
+challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258,stefik2013empirical,Marceau:2011:VGT:2016911.2016921}, and even experienced programmers frequently make mistakes \cite{7548903,jones2006developer,Marceau:2011:VGT:2016911.2016921}.
 
-Second, the fact that many sequences of characters do not correspond to a meaningful
-program complicates the design of program editors and other interactive programming tools. 
+Second, many sequences of characters do not correspond to meaningful
+programs. This complicates the design of program editors and other interactive programming tools. 
 %In particular, program
@@
 completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324},
-and refactoring support \cite{mens2004survey}. Editors must either disable
+and refactoring services \cite{mens2004survey}. Editors must either disable
 these editor services when they encounter meaningless edit states or
-deploy \emph{ad hoc} heuristics, e.g. by using whitespace to guess the programmer's intent \cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.
+deploy \emph{ad hoc} heuristics, e.g. by using whitespace to guess the intent \cite{DBLP:conf/oopsla/KatsJNV09,DBLP:conf/sle/JongeNKV09}.
 
@@
 into \emph{structure editors}, i.e. program editors where every edit state
-corresponds to a program structure \cite{teitelbaum_cornell_1981}.\footnote{Structure editors are also variously known as \emph{structured editors}, \emph{structural editors}, \emph{syntax-directed editors} and \emph{projectional editors} (esp.
-when the user interface presents rich views of the program structure.)}
+corresponds to a program structure \cite{teitelbaum_cornell_1981}.\footnote{Structure editors are also variously known as \emph{structured editors}, \emph{structural editors}, \emph{syntax-directed editors} and \emph{projectional editors}.}
 
@@
 type and binding structure of the underlying language complicates its
-design. The reason is that syntactic structure editors do not assign static meaning to every edit state -- they guarantee only that every edit state corresponds to  a 
-syntactically well-formed tree. These editors must also either disable editor services that require an understanding of the semantics of the program or deploy \emph{ad hoc} heuristics when they
-encounter these meaningless edit states.
+design. The problem is that syntactic structure editors do not assign static meaning to every edit state -- they guarantee only that every edit state corresponds to  a 
+syntactically well-formed tree.
+These editors must also either selectively disable editor services that require an understanding of the semantics of the program being written, or deploy \emph{ad hoc} heuristics.
 

### 2016-11-11 acks.tex
added a second NSF grant. Danny asked me to add it. IF it ends up getting kicked back from the publisher we can add it, if not, no problem.  I did get the main grant in there so that is good enough for me.

@@
 %This is Danny Dig's COPE grant. He asked me to add the grant.
-This work was partially funded through the NSF grant \#CCF-1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140.
+This work was partially funded through the NSF grant \#CCF-1553741 and 1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140.
 The views and conclusions contained herein are those of the authors and should not be interpreted as necessarily representing the official policies or endorsements, either expressed or implied, of the NSF, AFRL, DARPA, the NSA or the U.S.

### 2016-11-15 acks.tex
added ack for Mozilla gift

--- 2016-11-08 acks.tex
@@
 %This is Danny Dig's COPE grant. He asked me to add the grant.
-This work was partially funded through the NSF grant \#CCF-1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140.
-The views and conclusions contained herein are those of the authors and should not be interpreted as necessarily representing the official policies or endorsements, either expressed or implied, of the NSF, AFRL, DARPA, the NSA or the U.S.
+This work was partially funded through a gift from Mozilla; the NSF grant \#CCF-1553741 and 1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140.
+The views and conclusions contained herein are those of the authors and should not be interpreted as necessarily representing the official policies or endorsements, either expressed or implied, of Mozilla, the NSF, AFRL, DARPA, the NSA or the U.S.
 Government.

### 2016-11-15 agda.tex

@@
 formally establish that our design meets our stated objectives, we have
-mechanized the semantics and metatheory of Hazelnut as described above (and more concisely in the appendix) using the Agda proof
+mechanized the semantics and metatheory of Hazelnut as described above using the Agda proof
 assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
-at \url{http://wiki.portal.chalmers.se/agda/}.) This development is available in the supplemental material. The mechanization also includes the  extension to the
-language described in Sec. \ref{sec:extending}.
+at \url{http://wiki.portal.chalmers.se/agda/}.) This development is available in the supplemental material.
+The mechanization also includes the  extension to Hazelnut described in Sec.
+\ref{sec:extending}.
 

### 2016-11-15 formally.tex
dropped construct arg per reviewer comment / lack of symmetry
revised abstract and intro

@@
 The previous section introduced Hazelnut by example. In this section, we
-systematically introduce the following structures:
+systematically define the following structures:
 \begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
@@
 
-Our overview below omits certain ``uninteresting'' details. The appendix
+Our overview below omits certain ``uninteresting'' details. The supplement 
 includes the complete collection of rules, in definitional order. These
-rules, along with the proofs of all of the metatheorems discussed, have been
-mechanized in Agda \cite{norell:thesis} (see Sec. \ref{sec:mech}.)
+rules, along with the proofs of all of the metatheorems discussed in this section (and several omitted auxiliary lemmas), have been
+mechanized using the Agda proof assistant \cite{norell:thesis} (discussed in Sec. \ref{sec:mech}.)
 
@@
 define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$. The form
-$\hexp : \htau$ is an explicit \emph{type ascription}.
-
-In addition to these standard forms, \emph{type holes} and \emph{empty
+$\hexp : \htau$ is an explicit \emph{type ascription}. In addition to these standard forms, \emph{type holes} and \emph{empty
   expression holes} are both drawn $\hehole$ and \emph{non-empty expression
-  holes} are drawn $\hhole{\hexp}$. We do not need non-empty type holes
-because every H-type is a valid classifier of H-expressions.
-
+  holes} are drawn $\hhole{\hexp}$. 
 Types and expressions that contain no holes are \emph{complete types} and
-\emph{complete expressions}, respectively. Formally, we can derive
-$\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when
-$\hexp$ is complete (see supplement.) We are not concerned here with
-defining a dynamic semantics for Hazelnut, but the dynamics for complete
-H-expressions would be entirely standard because complete expressions correspond  immediately to those of the typed lambda calculus with numbers (in Sec. \ref{sec:rw} we discuss
-the multi-dimensional design space around the question of evaluating
-incomplete H-expressions.)
+\emph{complete expressions}, respectively. 
+%We do not, therefore, need to separately define the dynamics of complete expressions here (see, e.g., Harper \cite{pfpl}.) We will not develop a dynamics for incomplete expressions (but see Sec.
+\ref{sec:rw} for a discussion of this intriguing idea.)
+%Formally, we can derive
+%$\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when
+%$\hexp$ is complete (see supplement.)
 
@@
 \mathsf{HExp} & \hexp & ::= &
-  \hexp : \htau ~\vert~
   x ~\vert~
-  \hlam{x}{\hexp} ~\vert~
+  {\hlam{x}{\hexp}} ~\vert~
   \hap{\hexp}{\hexp} ~\vert~
@@
   \hadd{\hexp}{\hexp} ~\vert~
+  \hexp : \htau ~\vert~
   \hehole ~\vert~
@@
 \end{figure}
+
 \begin{figure}
-\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\tau$ and $\tau'$ are consistent}
+\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
+\begin{subequations}
+\begin{equation}\label{rule:syn-var}
+\inferrule{ }{
+  \hsyn{\hGamma, x : \htau}{x}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-ap}
+\inferrule{
+  \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
+  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\
+  \hana{\hGamma}{\hexp_2}{\htau_2}
+}{
+  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-num}
+\inferrule{ }{
+  \hsyn{\hGamma}{\hnum{n}}{\tnum}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-plus}
+\inferrule{
+  \hana{\hGamma}{\hexp_1}{\tnum}\\
+  \hana{\hGamma}{\hexp_2}{\tnum}
+}{
+  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-asc}
+\inferrule{
+  \hana{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hexp : \htau}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-ehole}
+\inferrule{ }{
+  \hsyn{\hGamma}{\hehole}{\tehole}
+}
+\end{equation}
+\begin{equation}\label{rule:syn-hole}
+\inferrule{
+  \hsyn{\hGamma}{\hexp}{\htau}
+}{
+  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
+}
+\end{equation}
+\end{subequations}
+\fbox{$\hana{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ analyzes against $\htau$}
+\begin{subequations}
+\begin{equation}\label{rule:syn-lam}
+\inferrule{
+  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
+  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
+}{
+  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
+}
+\end{equation}
+\begin{equation}\label{rule:ana-subsume}
+\inferrule{
+  \hsyn{\hGamma}{\hexp}{\htau'}\\
+  \tcompat{\htau}{\htau'}
+}{
+  \hana{\hGamma}{\hexp}{\htau}
+}
+\end{equation}
+\end{subequations}
+\caption{H-type synthesis and analysis.}
+\label{fig:ana-syn}
+\end{figure}
+\begin{figure}
+\vspace{-1px}\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\htau$ and $\htau'$ are consistent}\vspace{-4px}
 \begin{subequations}\label{eqns:consistency}
@@
 }{
-  \tcompat{(\tarr{\htau_1}{\htau_2})}{(\tarr{\htau_1'}{\htau_2'})}
+  \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
 }~~~~\text{\hspace{-2px}(\ref*{eqns:consistency}a-d)}
@@
 \end{subequations}
-\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
+\noindent\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}\vspace{-4px}
+
 \begin{subequations}
-\begin{minipage}{0.5\linewidth}
+\begin{minipage}{0.43\linewidth}
+\begin{equation}
+\inferrule{ }{
+  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
+}
+\end{equation}
+\end{minipage}
+\begin{minipage}{0.55\linewidth}
 \begin{equation}
@@
   \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
-}
-\end{equation}
-\end{minipage}
-\begin{minipage}{0.48\linewidth}
-\begin{equation}
-\inferrule{ }{
-  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
 }
@@
 
-\begin{figure}
-\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
-\begin{subequations}
-\begin{equation}\label{rule:syn-asc}
-\inferrule{
-  \hana{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hexp : \htau}{\htau}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-var}
-\inferrule{ }{
-  \hsyn{\hGamma, x : \htau}{x}{\htau}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-ap}
-\inferrule{
-  \hsyn{\hGamma}{\hexp_1}{\htau}\\
-  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
-  \hana{\hGamma}{\hexp_2}{\htau_2}
-}{
-  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-num}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hnum{n}}{\tnum}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-plus}
-\inferrule{
-  \hana{\hGamma}{\hexp_1}{\tnum}\\
-  \hana{\hGamma}{\hexp_2}{\tnum}
-}{
-  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-ehole}
-\inferrule{ }{
-  \hsyn{\hGamma}{\hehole}{\tehole}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-hole}
-\inferrule{
-  \hsyn{\hGamma}{\hexp}{\htau}
-}{
-  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
-}
-\end{equation}
-\end{subequations}
-\fbox{$\hana{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ analyzes against $\htau$}
-\begin{subequations}
-\begin{equation}\label{rule:ana-subsume}
-\inferrule{
-  \hsyn{\hGamma}{\hexp}{\htau'}\\
-  \tcompat{\htau}{\htau'}
-}{
-  \hana{\hGamma}{\hexp}{\htau}
-}
-\end{equation}
-\begin{equation}\label{rule:syn-lam}
-\inferrule{
-  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
-  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
-}{
-  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
-}
-\end{equation}
-\end{subequations}
-\caption{Synthesis and analysis.}
-\label{fig:ana-syn}
-\end{figure}
-
 Hazelnut's static semantics is organized as a \emph{bidirectional type
@@
 around the two mutually defined judgements in Figure
-\ref{fig:ana-syn}. Typing contexts, $\hGamma$, map each variable $x \in
-\domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner. We
-identify the context up to exchange and contraction and adopt the standard identification
-convention for formal structures that differ only up to alpha-renaming of bound variables. All judgements that mention a context obey weakening.
-
-Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$,
+\ref{fig:ana-syn}. Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$,
 establish that $\hexp$ can appear where an expression of type $\htau$ is
 expected. Derivations of the type synthesis judgement,
-$\hsyn{\hGamma}{\hexp}{\htau}$, infer a type from $\hexp$, which is
+$\hsyn{\hGamma}{\hexp}{\htau}$, synthesize (a.k.a. \emph{locally infer} \cite{Pierce:2000:LTI:345099.345100}) a type from $\hexp$. Type synthesis is
 necessary in positions where an expected type is not available (e.g. at the
@@
 
-\begin{subequations}\label{rules:syn-ana}
-Type synthesis is stronger than type analysis in that if an expression is
+If an expression is 
 able to synthesize a type, it can also be analyzed against that type, or
 any other \emph{consistent} type, according to the \emph{subsumption rule},
-Rule (\ref{rule:ana-subsume}).
+Rule (\ref{rule:ana-subsume}).% Type analysis of every form other than the lambda form goes through subsumption.
 
@@
 
+Typing contexts, $\hGamma$, map each variable $x \in
+\domof{\hGamma}$ to an hypothesis $x : \htau$. We
+identify contexts up to exchange and contraction and adopt the standard identification
+convention for structures that differ only up to alpha-renaming of bound variables.
+All hypothetical judgements obey a standard weakening lemma.
+Rule (\ref{rule:syn-var}) establishes that variable expressions synthesize the hypothesized H-type, in the standard manner.
+
 Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions,
@@
   but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.}
-Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
+Rule (\ref{rule:syn-lam}) is not quite the standard rule, as reproduced below:
 \begin{equation*}
@@
 (\ref{rule:syn-lam}) and the matched arrow type judgement.  Encouragingly,
-this judgement also arises in the study of gradual type systems
+the matched arrow type judgement also arises in gradual type systems
 \cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.
-
-Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of
-ascription form, $\hexp : \htau$. This allows the user to explicitly state
-a type for the ascribed expression to be analyzed against.
-
-Rule (\ref{rule:syn-var}) is the standard rule for variables.
 
@@
 -- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$
-synthesizes the type hole. Indeed, Siek and Taha needed two application
-rules for the same fundamental reason \cite{Siek06a}. Later work on gradual
-typing introduced this notion of type matching to resolve this redundancy.
+synthesizes $\tehole$.
 
@@
 
+Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of
+ascription form, $\hexp : \htau$. This allows the programmer to explicitly state
+a type for the ascribed expression to be analyzed against.
+
 The rules described so far are sufficient to type complete
@@
 Rule (\ref{rule:syn-ehole}) states that the empty expression hole
-synthesizes the type hole.
-
-A non-empty hole contains an H-expression that is ``under construction'',
-as described in Sec. \ref{sec:example}. According to Rule
-(\ref{rule:syn-hole}), this inner expression must synthesize some type,
-but, like the empty hole, non-empty holes synthesize the type hole.
-
-Given these rules, it is instructive to derive the following (by subsumption):
+synthesizes the type hole. Non-empty holes, which contain an H-expression that is ``under construction'' 
+as described in Sec. \ref{sec:example}, also synthesize the hole type. According to Rule
+(\ref{rule:syn-hole}), the enveloped expression must synthesize some (arbitrary) type. (We do not need non-empty type holes because every H-type is a valid classifier of H-expressions.)
+
+Because the hole type is consistent with every type, expression holes can be analyzed against any type by subsumption. For example, it is instructive to derive the following:
 \[\hana{incr : \tarr{\tnum}{\tnum}}{\hhole{incr}}{\tnum}\]
 
-\end{subequations}
 \subsection{Z-types and Z-expressions}\label{sec:cursors}
@@
 \begin{figure}[t]
-\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
+$\arraycolsep=2pt\begin{array}{lllllll}
 \mathsf{ZTyp} & \ztau & ::= &
@@
   \zwsel{\hexp} \cvert
-  \zexp : \htau \cvert
-  \hexp : \ztau \cvert
   \hlam{x}{\zexp} \cvert
@@
   \hadd{\zexp}{\hexp} \cvert
-  \hadd{\hexp}{\zexp} \cvert
+  \hadd{\hexp}{\zexp} \\
+& & \cvert & 
+    \zexp : \htau \cvert
+  \hexp : \ztau \cvert
   \hhole{\zexp}
@@
 
-Hazelnut's action semantics operates over Z-types, $\ztau$, and
-Z-expressions, $\zexp$. Figure \ref{fig:zexp-syntax} defines the syntax of
-Z-types and Z-expressions. A Z-type (resp. Z-expression) represents an
+Figure \ref{fig:zexp-syntax} defines the syntax of
+Z-types, $\ztau$, and Z-expressions, $\zexp$. A Z-type (resp. Z-expression) represents an
 H-type (resp. H-expression) with a single superimposed \emph{cursor}.
@@
 \begin{align*}
-\removeSel{(\zwsel{\htau})} & = \htau\\
-\removeSel{(\tarr{\ztau}{\htau})} & = \tarr{\removeSel{\ztau}}{\htau}\\
-\removeSel{(\tarr{\htau}{\ztau})} & = \tarr{\htau}{\removeSel{\ztau}}
+\removeSel{\zwsel{\htau}} & = \htau\\
+\removeSel{\tarr{\ztau}{\htau}} & = \tarr{\removeSel{\ztau}}{\htau}\\
+\removeSel{\tarr{\htau}{\ztau}} & = \tarr{\htau}{\removeSel{\ztau}}
 \end{align*}
 
-Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by
-erasing the cursor from the Z-expression $\zexp$, i.e. the cursor erasure
-of $\zexp$. The definition of this metafunction is analogous, so we omit it
-for concision (see supplement.)
+Similarly, we write $\removeSel{\zexp}$ for cursor erasure of $\zexp$. The definition of this metafunction is analogous, so we omit it
+for concision.
+
+This zipper structure is not the only way to model a cursor, though we have found it to be the most elegant for our present purposes.
+Another plausible strategy would be to formalize the notion of a relative path into an H-expression.
+This would then require defining the notion of consistency between a relative path and an H-expression, so we avoid it.
 
@@
 In some Z-expressions, the cursor is in a type ascription, so we also need
-a \emph{type action judgement}, pronounced ``performing $\alpha$ on $\ztau$
-results in $\ztau'$'':
+a \emph{type action judgement}:
 \[
@@
 
-Before giving the rules for these judgements, let us state the key
-metatheorem: \emph{sensibility}. This metatheorem deeply informs the design
-of the rules, given starting in Sec. \ref{sec:action-subsumption}. Its
-proof, which is mechanized in Agda, is by straightforward induction, so the
-reader is encouraged to think about the relevant cases of these proofs as
-we present the rules.
+These judgements are governed by a critical 
+metatheorem, \emph{action sensibility} (or simply \emph{sensibility}): 
 \begin{theorem}[Action Sensibility]
-  \label{thrm:actsafe} Both of the following hold:
+  \label{thrm:actsafe} ~
   \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
@@
 \end{theorem}
-\noindent In other words, if an edit state (i.e. a Z-expression) is
+\noindent In other words, if a Z-expression is
 statically meaningful, i.e. its cursor erasure is well-typed, then
-performing an action on it leaves the resulting edit state statically
-meaningful. In particular, the first clause of Theorem \ref{thrm:actsafe}
-establishes that when an action is performed on an edit state whose cursor
-erasure synthesizes an H-type, the result is an edit state whose cursor
+performing an action on it leaves the resulting Z-expression statically
+meaningful. More specifically, the first clause of Theorem \ref{thrm:actsafe}
+establishes that when an action is performed on a Z-expression whose cursor
+erasure synthesizes an H-type, the result is a Z-expression whose cursor
 erasure also synthesizes some (possibly different) H-type. The second
@@
 
+This metatheorem deeply informs the design
+of the rules, given starting in Sec. \ref{sec:action-subsumption}. Its
+proof is by straightforward induction, so the
+reader is encouraged to think about the relevant proof case when considering each action rule below.
+
+No sensibility theorem is needed for the type action judgement because every syntactically well-formed type is meaningful in Hazelnut.
+(Adding type variables to the language would require defining both a type-level sensibility theorem and
+type-level non-empty holes.)
 \begin{figure}[t]
@@
   \fap ~\vert~
-  \farg ~\vert~
+  % \farg ~\vert~
   \fnumlit{n} ~\vert~
@@
 of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can
-define this notion either directly as the metatheoretic negation of type
-consistency, or as a separate inductively defined judgement. The supplement
-does the latter. The key rule establishes that arrow types are inconsistent
+define this notion either directly as the constructive negation of type
+consistency, or as a separate inductively defined judgement with the following
+key rule, which establishes that arrow types are inconsistent
 with $\tnum$:
@@
   \end{equation*}
-The mechanization proves that this judgemental definition of type
+The mechanization proves that the judgemental definition of type
 inconsistency is indeed the negation of type consistency.
@@
 \end{equation}
-In other words, if the cursor erasure of the edit state synthesizes a type,
-then we defer to the synthetic action judgement, as long as the type of the
-resulting cursor erasure is consistent with the type provided for analysis.
-The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem
+In other words, if the cursor erasure of the edit state synthesizes a type, $\htau'$, 
+then we defer to the synthetic action judgement.
+The cursor erasure of the Z-expression resulting from performing the action $\alpha$ synthetically could have a different type, $\htau''$, so we must check that it is consistent with the type provided for analysis, $\htau$.
+The case for Rule (\ref{rule:action-subsume}) in the proof of Theorem
 \ref{thrm:actsafe} goes through by induction and static subsumption,
 i.e. Rule (\ref{rule:ana-subsume}). Algorithmically, subsumption should be
-the rule of last resort (see Sec. \ref{sec:determinism}.)
+the rule of last resort (see Sec. \ref{sec:determinism} for further discussion.)
 
@@
 The rules above are numerous and fairly uninteresting. That makes them
-quite hazardous -- we might make a mistake absent-mindedly. One check
-against this is to prove that establishes that movement actions do not
+quite hazardous -- we might make a typo or forget a rule absent-mindedly. One check
+against this is to establish that movement actions do not
 change the cursor erasure, as in Theorem \ref{lemma:movement-erasure}.
@@
 Theorem \ref{thrm:reachability} follows by straightforward composition of
-these two lemmas.
+these two lemmas. (Of course, the witness produced is not usually the most efficient way
+to move between two locations.)
 
@@
 \end{equation}
+This choice is formally arbitrary -- it would have also been sensible to use the type under the cursor as the return type, for example.
+Indeed, we could consider defining both of these using different shapes.
+We avoid this for the sake of simplicity.
 
@@
 \end{equation}
-The rule above featured on Line 15 of Figure \ref{fig:second-example}.
+The rule above featured on Line 16 of Figure \ref{fig:second-example}.
 
@@
 ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the
-argument type hole:
+argument type hole (again, arbitrarily):
 \begin{equation}
   \label{r:conelamhole}
-  \inferrule{ }{
+  \inferrule{\alpha=\aConstruct{\flam{x}}}{
     \performSyn
@@
       {\tehole}
-      {\aConstruct{\flam{x}}}
+      {\alpha}
       {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
@@
 The type ascription is necessary because lambda expressions do not
-synthesize a type. If the empty hole is being analyzed against a type with
+synthesize a type. (Type-annotated function definitions often
+arise as a single syntactic construct in full-scale languages like ML.)
+
+If the empty hole is being analyzed against a type with
 matched arrow type, then no ascription is necessary:
@@
 
-The $\aConstruct{\farg}$ action instead places the expression under the
-cursor in the argument position of an application form. Because the
-function position is always an empty hole in this situation, we need only a
-single rule:
-\begin{equation}
-  \inferrule{ }{
-    \performSyn
-      {\hGamma}
-      {\zwsel{\hexp}}
-      {\htau}
-      {\aConstruct{\farg}}
-      {\hap{\zwsel{\hehole}}{\hexp}}
-      {\tehole}
-  }
-\end{equation}
+% The $\aConstruct{\farg}$ action instead places the expression under the
+% cursor in the argument position of an application form. Because the
+% function position is always an empty hole in this situation, we need only a
+% single rule:
+% \begin{equation}
+%   \inferrule{ }{
+%     \performSyn
+%       {\hGamma}
+%       {\zwsel{\hexp}}
+%       {\htau}
+%       {\aConstruct{\farg}}
+%       {\hap{\zwsel{\hehole}}{\hexp}}
+%       {\tehole}
+%   }
+% \end{equation}
 
@@
 The $\aConstruct{\fplus}$ action constructs a plus expression with the
-expression under the cursor as its first argument. If that expression
+expression under the cursor as its first argument (again, arbitrarily.) If that expression
 synthesizes a type consistent with $\tnum$, then the rule is
@@
 The $\fnehole$ shape is grayed out in Figure \ref{fig:action-syntax}
-because we do not expect the programmer to perform it explicitly -- other
+because we do not generally expect the programmer to perform it explicitly -- other
 actions automatically insert holes when a type inconsistency would
-arise. Its inclusion is mainly to make it easier to state another
-``checksum'' theorem: \emph{constructability}
+arise. 
+% Occasionally, it is useful to insert a non-empty hole explicitly 
+%to avoid multiple automatically inserted non-empty holes that would otherwise arise over a sequence of actions. 
+The inclusion of this rule 
+simplifies the statement of the constructability theorem, discussed next.
 
@@
 \end{theorem}
+Without the $\fnehole$ shape, this theorem as stated would not hold.
+For example, it is not possible to construct well-typed H-expressions where non-empty holes appear superfluously without the $\fnehole$ shape.
+Note also that although none of the shapes that we have defined can be dropped without losing this theorem, some construction rules could be dropped.
+In particular, rules that insert non-empty holes automatically could be dropped because the $\aConstruct{\fnehole}$ action can always be used instead.
+We included them because we are interested in the mechanics of automatic non-empty hole insertion.
 
@@
 Like deletion, there is no need for a ``checksum'' theorem for the
-finishing.
+finishing action.
 
@@
 
-When the cursor is in the expression position of an ascription, we use the
-analytic  judgement, mirroring Rule (\ref{rule:syn-asc}):
-\begin{equation}
-\inferrule{
-  \performAna
-    {\hGamma}
-    {\zexp}
-    {\htau}
-    {\alpha}
-    {\zexp'}
-}{
-  \performSyn
-    {\hGamma}
-    {\zexp : \htau}
-    {\htau}
-    {\alpha}
-    {\zexp' : \htau}
-    {\htau}
-}
-\end{equation}
-
-When the cursor is in the type position of an ascription, we must re-check
-the ascribed expression because the cursor erasure might have changed (in
-practice, one would optimize this check to only occur if the cursor erasure
-did change):
-\begin{equation}\label{rule:zipper-asc}
-\inferrule{
-  \performTyp{\ztau}{\alpha}{\ztau'}\\
-  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
-}{
-  \performSyn
-    {\hGamma}
-    {\hexp : \ztau}
-    {\removeSel{\ztau}}
-    {\alpha}
-    {\hexp : \ztau'}
-    {\removeSel{\ztau'}}
-}
-\end{equation}
-
 When the cursor is in the function position of an application, the rule
@@
       {\tnum}
+  }
+\end{equation}
+
+When the cursor is in the expression position of an ascription, we use the
+analytic  judgement, mirroring Rule (\ref{rule:syn-asc}):
+\begin{equation}
+\inferrule{
+  \performAna
+    {\hGamma}
+    {\zexp}
+    {\htau}
+    {\alpha}
+    {\zexp'}
+}{
+  \performSyn
+    {\hGamma}
+    {\zexp : \htau}
+    {\htau}
+    {\alpha}
+    {\zexp' : \htau}
+    {\htau}
+}
+\end{equation}
+
+When the cursor is in the type position of an ascription, we must re-check
+the ascribed expression because the cursor erasure might have changed (in
+practice, one would optimize this check to only occur if the cursor erasure
+did change):
+\begin{equation}\label{rule:zipper-asc}
+\inferrule{
+  \performTyp{\ztau}{\alpha}{\ztau'}\\
+  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
+}{
+  \performSyn
+    {\hGamma}
+    {\hexp : \ztau}
+    {\removeSel{\ztau}}
+    {\alpha}
+    {\hexp : \ztau'}
+    {\removeSel{\ztau'}}
 }

### 2016-11-15 hazelnut-popl17.tex

@@
 \usepackage{mathpartir}
+\usepackage{balance}
 \usepackage{stmaryrd}
@@
 {
- $^1$~Carnegie Mellon University (USA)
- \and
- $^2$~Oregon State University (USA)
- \and
- $^3$~University of Colorado Boulder (USA)\vspace{-10px}
+ \begin{tabular}{ccc}
+ 	\begin{tabular}{c}
+ $^1$Carnegie Mellon University (USA)\\
+ \textsf{\hphantom{$^1$}\{comar, iev, aldrich\}@cs.cmu.edu}
+ \end{tabular} & 
+ \begin{tabular}{c}
+ $^2$Oregon State University (USA)\\
+ \textsf{\hphantom{$^1$}hiltonm@eecs.oregonstate.edu}
+ \end{tabular} & 
+  \begin{tabular}{c}
+ $^3$University of Colorado Boulder (USA)\\
+ \textsf{\hphantom{$^1$}matthew.hammer@colorado.edu}
+ \end{tabular}
+ \end{tabular}
 }
 {
-%\email{comar@cs.cmu.edu}\\
-%\texttt{http://www.cs.cmu.edu/\homedir comar/}
-%\email{hiltonm@eecs.oregonstate.edu}\\
+ % \vspace{-12px}
 }
@@
 reveal connections with gradual typing and contextual modal type theory, the Curry-Howard interpretation of contextual modal logic. Finally, we
-discuss how Hazelnut's semantics lends itself to implementation as an event-based functional reactive program.
-Our simple reference implementation is written in OCaml and compiled using \lstinline{js_of_ocaml}.
+discuss how Hazelnut's semantics lends itself to implementation as an event-based functional reactive program. Our simple reference implementation is written using \lstinline{js_of_ocaml}.
 
@@
 \section{Related Work and Discussion}\label{sec:rw}
-\todo{move into own file}
 \subsection{Structure Editors}
@@
 
-The examples given so far either define a language with a trivial static
-semantics, or do not attempt to maintain well-typedness as an edit
-invariant. This can pose problems, for reasons discussed in the
-Introduction. One apparent exception is Unison~\cite{unison}, a structure
+The examples given so far either do not attempt to reason statically about
+types and binding, or do not attempt to maintain well-typedness as an edit
+invariant. This can pose problems, for reasons discussed in Sec. \ref{sec:introduction}. One apparent exception is Unison~\cite{unison}, a structure
 editor for a typed functional language similar to Haskell. Like Hazelnut,
@@
 consider it essential to start by building type theoretic foundations,
-which are independent of nearly all decisions about the user interface. In
+which are independent of nearly all decisions about the user interface (other than our choice to use an explicit cursor.) In
 contrast, these editors have developed innovative user interfaces (e.g. see
@@
 and have gone to great effort to develop a clear metatheory, like Standard
-ML \cite{mthm97-for-dart,Harper00atype-theoretic}.  In the future, we hope
+ML \cite{mthm97-for-dart,Harper00atype-theoretic,Lee:2007:TMM:1190216.1190245}.  In the future, we hope
 that these lines of research will merge to produce a human-usable typed
@@
 foundational calculus that could be extended to achieve this vision. 
+Our implementation resembles the minimal structure editor defined in 
+Haskell by Sufrin and De Moor \cite{sufrin1999modeless}.
 
-Some structure editor generators do operate on formal or semi-formal
+Some structure editor \emph{generators} do operate on formal or semi-formal
 definitions of an underlying language. For example, the Synthesizer
@@
 
+Other work has attempted to integrate structure editing features into 
+text editors. For example, recent work has used syntactic placeholders
+reminiscent of our expression holes to decrease the percentage of 
+edit states that are malformed \cite{Amorim:2016:PSC:2997364.2997374}. This work does not consider the semantics of placeholders.
+
+Prior work has also explored 
+formal definitions of text editor commands, e.g. using functional 
+combinators \cite{DBLP:journals/scp/Sufrin82}.
+
 \subsection{Gradual Type Systems}
@@
 introduced type consistency. Subsequent work developed the concept of type
-matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}. In
+matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15} and has further 
+studied the notion of type consistency \cite{Garcia:2016:AGT:2837614.2837670}. In
 retrospect, this relationship is perhaps unsurprising: gradual typing is,
 notionally, also motivated by the idea of iterated development of a program
-where every intermediate state is well-defined in some sense.
+where every intermediate state is well-defined in some sense, albeit at 
+different granularity.
 
@@
 concern with assigning a dynamics to programs where type information is not
-known, by inserting dynamic type casts. This would correspond to assigning
+known, by inserting dynamic type casts \cite{Siek06a} or deducing evidence for consistency during evaluation \cite{Garcia:2016:AGT:2837614.2837670}. This would correspond to assigning
 a dynamics to Hazelnut expressions with type holes such that a run-time
-error occurs when a type hole is found to be unfillable through evaluation.
+error occurs when a type hole is found to be unfillable through evaluation. This 
+may be useful as an exploratory programming tool.
 
-\subsection{Type Holes as Unification Variables}
-Another possible interpretation of type holes is as explicit unification
-variables. In other words, we might define the dynamics of Hazelnut such
-that a program cannot be run if some automated type inference procedure
-cannot statically fill in any type holes that arise. For a simple calculus,
-e.g. the STLC upon which Hazelnut is based, type inference is known to be
-decidable \cite{damas1982principal}. In more complex settings, e.g. in a
+\subsection{Bidirectional Type Systems}
+Hazelnut is bidirectionally typed \cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial,odersky2001colored}. Bidirectional type systems are notable in 
+that they are easy to define, easy to implement, produce simple error messages and support advanced language features \cite{dunfield2013complete}.
+For example, Scala \cite{odersky2001colored} and Agda \cite{norell:thesis} are both fundamentally bidirectionally typed languages.
+
+\subsection{Type Reconstruction}
+An alternative approach to type inference is to use a unification-based type reconstruction system, as in functional languages like ML and Haskell \cite{damas1982principal}.
+This is difficult to reconcile with the approach presented in this paper, because edit actions could introduce new  unification constraints that would require placing non-empty holes around terms far from the cursor.
+A whole-program hole insertion pass after each edit action could perhaps be used to recover invariants similar to those presented here, but we leave the details as future work.
+Our contention is that a bidirectional approach is a sweet spot in the design space of interactive systems like Hazelnut because it precludes
+``spooky errors at a distance''. Instead, the interaction is a sort of local
+dialog between the programmer and the system involving simple, familiar
+concepts -- types with holes -- rather than sets of constraints.
+
+An intermediate approach would be to 
+layer unification-based type generation features atop the bidirectional system. This would amount to  interpretting type holes as unification
+variables. For a simple calculus,
+e.g. the STLC upon which Hazelnut is based, type inference for complete expressions is known to be
+decidable, so type holes could be instantiated automatically once the expression that they appear within has been constructed.
+It would also be possible to flag expressions for which there does not exist any way to fill the type holes.
+In more complex settings, e.g.
+in a
 dependently typed language, a partial decision procedure may still be
 useful in this regard, both at edit-time and (just prior to)
-run-time. Indeed, text editor modes for proof assistants, e.g. for Agda,
+run-time. Indeed, text editor modes for dependently typed proof assistants, e.g. for Agda,
 attempt to do exactly this for indicated ``type holes'' (and do not always
@@
 \subsection{Exceptions}
-Expression holes, too, could dynamically be interpreted in several
-ways. One straightforward approach would be to dynamically raise an
-exception if evaluation encounters an expression hole. Indeed,
+Expression holes can be interpreted in several
+ways. One straightforward interpretation is to treat them 
+like expressions that raise exceptions. Indeed,
 placing \textt{raise Unimplemented} or similar in portions of an expression
 that are under construction is a common practice across programming
-languages.
+languages today. The GHC dialect of Haskell recently introduced an explicit notion of a 
+typed hole that behaves similarly \cite{GHC/holes}.
 
@@
 treated as sites where the system can be asked to automatically generate an
-expression of an appropriate type. These systems are not statically
-well-defined themselves (though see below.)
+expression of an appropriate type. %These systems are not statically
+% well-defined themselves (though see below.)
 
-The Graphite system used an informal model of typed holes in Java to allow
+The Graphite system borrowed Eclipse's heuristic model of typed holes for Java to allow
 library providers to associate interactive code generation interfaces with
@@
 
-\subsection{Contextual Modal Type Theory}
-Expression holes can also be understood by invoking the notion of
-a \emph{metavariable} as found in contextual modal type theory
+\subsection{Tactics}
+
+Interactive proof refinement systems, e.g. those in LCF \cite{Gordon:1978:MIP:512760.512773}, and more recent typed tactic systems, e.g. Mtac for Coq \cite{Ziliani:2013:MMT:2544174.2500579}, 
+support an explicit model of a ``current'' typed hole that serves as the 
+target of program synthesis. Hazelnut differs in that edits can occur anywhere within a term.
+
+A related approach is to interpret expression holes as the \emph{metavariables} of contextual modal type theory
 (CMTT) \cite{DBLP:journals/tocl/NanevskiPP08}. In particular, expression
 holes have types and are surrounded by contexts, just as metavariables in
-CMTT are associated with types and contexts. This helps to clarify the
+CMTT are associated with types and contexts. This begins to clarify the
 logical meaning of a typing derivation in Hazelnut -- it conveys
@@
 
-Making the modal context explicit in our semantics is not technically
-useful given our stated purpose -- interactive program editing is not
+We did not make the modal context explicit in our semantics because interactive program editing is not
 merely hole filling in Hazelnut (i.e. the cursor need not be on a
-hole). Moreover, the hole's type and context become apparent as our action
-semantics traverses the zipper structure on each step. Some interactive
-proof assistants, however, support a tactic model based directly on hole
-filling, so the connection to CMTT and similar systems is more useful. For
+hole.) Moreover, the hole's type and context become apparent as our action
+semantics traverses the zipper structure on each step. For interactive
+proof assistants that support a tactic model based directly on hole
+filling, as just discussed, the connection to CMTT and similar systems is more useful. For
 example, Beluga \cite{DBLP:conf/flops/Pientka10} is based on dependent CMTT
@@
 
-\subsection{Evaluation Strategies: A High-Dimensional Space}
-To summarize, we have discussed three different evaluation strategies in
-the presence of type holes:
-\begin{enumerate}[noitemsep]
-\item ...as preventing evaluation (the standard approach.)
-\item ...as unknown types, in the gradual sense.
-\item ...as unification variables.
-\end{enumerate}
-In addition, we have discussed four different evaluation strategies in the
-presence of expression holes:
-\begin{enumerate}[noitemsep]
-\item ...as preventing evaluation (the standard approach.)
-\item ...as causing exceptions.
-\item ...as sites for automatic program synthesis.
-\item ...as the closures of CMTT.
-\end{enumerate}
-
-Every combination of these choices could well be considered in the design
-of a full-scale programming system in the spirit of Hazelnut. Indeed, the
-user might be given several options from among these combinations,
-depending on their usage scenario. Many of these warrant further inquiry.
-
 \section{Conclusion}
@@
 language by extension. Our plan is to simultaneously maintain a
-mechanization and implementation (following, for example, Standard ML) as
+mechanization and implementation (following, for example, Standard ML \cite{Lee:2007:TMM:1190216.1190245}) as
 we proceed, ultimately producing the first large-scale, formally verified
@@
 
+\subsubsection{Evaluation Strategies: A High-Dimensional Space}
+The related work brought up in the previous section suggests three different evaluation strategies in
+the presence of type holes:
+\begin{enumerate}[noitemsep]
+\item ...as preventing evaluation (the standard approach.)
+\item ...as unknown types, in the gradual sense.
+\item ...as unification variables.
+\end{enumerate}
+In addition, we have discussed four different evaluation strategies in the
+presence of expression holes:
+\begin{enumerate}[noitemsep]
+\item ...as preventing evaluation (the standard approach.)
+\item ...as causing exceptions.
+\item ...as sites for automatic program synthesis.
+\item ...as the closures of CMTT.
+\end{enumerate}
+
+Every combination of these choices could well be considered in the design
+of a full-scale programming system in the spirit of Hazelnut. Indeed, the
+user might be given several options from among these combinations,
+depending on their usage scenario. Many of these warrant further inquiry.
+
 \subsubsection{Editor Services}
@@
 actually entered using, for example, key combinations or chords. In
-practice, we would want also to rank available actions in some reasonable
+practice, we would want also to suggest sensible compound actions, and 
+to rank these suggested actions in some reasonable
 manner (perhaps based on usage data gathered from other users or code
-repositories.) Designing a rigorous typed probability model over actions
-and H-expressions is one avenue of research that we have started to
+repositories.) Designing a action suggestion semantics and a rigorous typed probability model over actions is one avenue of research that we have started to
 explore, with intriguing initial results.
@@
 acts much like a simple imperative command language. This suggests future
-expansion to, for example, a true tactic language. Alternatively, it may be
-more useful to develop the notion of an \emph{action macro}, whereby
+expansion to, for example, a true \emph{action macro} language, whereby
 functional programs could themselves be lifted to the level of actions to
@@
 program refactoring, while remaining subject to the core Hazelnut
-metatheory.
+metatheory. Techniques like those in advanced tactic systems, e.g. Mtac, might be useful 
+in proving these action macros correct \cite{Ziliani:2013:MMT:2544174.2500579}.
 
@@
 
-It should also be possible to develop a semantics of semantic comments,
-i.e. comments that mention semantic structures. These would be subject to
+It should also be possible to develop the concept of semantic comments,
+i.e. comments that mention semantic structures or even contain values. These would be subject to
 the same operations, e.g. renaming, as other structures, helping to address
-the problem of comments becoming inconsistent with code.
+the problem of comments becoming inconsistent with code. This system, generalized sufficiently,
+could one day help unify document editing with program editing.
 
@@
 editors, are practical. We hope to conduct user studies once a richer
-semantics has been developed.
+semantics and a practical implementation thereof has been developed.
 
@@
 \clearpage
-
+\balance
 \bibliographystyle{abbrvnat}
@@
 
-\iftr
-\clearpage
+% \iftr
+% \clearpage
 %\onecolumn
-\appendix
-\input{appendix}
-\else
+% \appendix
+% \input{appendix}
+% \else
 % No Appendix Here
-\fi
+% \fi
 

### 2016-11-15 implementation.tex

@@
 row of Figure \ref{fig:impl-overview} diagrams this stream of edit
-states. By Theorem \ref{thrm:actsafe}, the editor does not need to
-typecheck each edit state (though in some of the rules, e.g. Rule
-(\ref{rule:zipper-asc}) which handles the situation where the type in a
-type ascription changes, portions of the program do need to be
-typechecked.) For example, the reader is encouraged to re-examine the
+states. For example, the reader is encouraged to re-examine the
 examples in Figure \ref{fig:first-example} and \ref{fig:second-example} --
@@
 
+Because Theorem \ref{thrm:actsafe} expresses an invariant, the editor does not need to
+typecheck the edit state anew on each action (though in some of the rules, e.g. Rule
+(\ref{rule:zipper-asc}) which handles the situation where the type in a
+type ascription changes, portions of the program would need to be
+typechecked again.) In other words, many of the problems related to incrementality 
+are simply not relevant. Storing the types of subtrees would allow for further
+optimizations (e.g. of the zipper rules.)
+
 The programmer examines a view generated from each edit state and produces
 actions in some implementation-defined manner (e.g. using a keyboard,
-mouse, touchscreen, voice interface, or neural implant.) Each new action
+mouse, touchscreen, voice interface, or neural interface), as diagrammed in Figure \ref{fig:impl-overview}. Each new action
 causes a new abstract edit state to arise according to an implementation of
@@
 action internally and handling the exception that is raised when an action
-is undefined. Each action form also has a corresponding keyboard shortcut. This design is not, of course, meant to score marks for
-usability or performance.
-Instead, as a simple reference implementation, it allows the reader to interact with the system presented in this paper, to aid in their understanding.
+is undefined.
+(An optimization that is not in HZ would be to implement a version of the action semantics that simply computes a boolean, rather than the resulting edit state, for the purposes of action validation.) Each action form also has a corresponding keyboard shortcut.
+For actions that take arguments, the keyboard shortcut moves the cursor into a text box.
+The action validation occurs on every change to the text box.
+
+This implementation is not, of course, meant to score marks for
+usability or performance (though both are surprisingly good for such a simple system.) Instead, as a simple reference implementation, it allows the reader to interact with the system presented in this paper, to aid in their understanding.
 Additionally, we expect its source code to be of use to others who are interested in

### 2016-11-15 acks.tex
added Hammer's NSF small grant number to acks

@@
 %This is Danny Dig's COPE grant. He asked me to add the grant.
-This work was partially funded through a gift from Mozilla; the NSF grant \#CCF-1553741 and 1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140.
+This work was partially funded through a gift from Mozilla; the NSF grant \#CCF-1619282, 1553741 and 1439957; by AFRL and DARPA under agreement \#FA8750-16-2-0042; and by NSA lablet contract \#H98230-14-C-0140.
 The views and conclusions contained herein are those of the authors and should not be interpreted as necessarily representing the official policies or endorsements, either expressed or implied, of Mozilla, the NSF, AFRL, DARPA, the NSA or the U.S.

### 2016-11-28 formally.tex
tweaks per publisher requests

@@
 
-A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$, defined in
-the supplement, establishes that $\bar\alpha$ consists only of actions of
+A simple auxiliary judgement, $\bar\alpha~\mathsf{movements}$ (not shown) establishes that $\bar\alpha$ consists only of actions of
 the form $\aMove{\delta}$.
@@
 \end{theorem}
-
-\begin{figure}
-$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\vspace{4px}\\
-\fbox{$\performTypI{\ztau}{\bar{\alpha}}{\ztau'}$}
-
-\vspace{-10px}\begin{subequations}
-\begin{minipage}{0.35\linewidth}
-\begin{equation}
-\inferrule{ }{
-    \performTypI{\ztau}{\cdot}{\ztau}
-}
-\end{equation}
-\end{minipage}
-\begin{minipage}{0.65\linewidth}
-\begin{equation}
-\inferrule{
-  \performTyp{\ztau}{\alpha}{\ztau'}\\
-  \performTypI{\ztau'}{\bar{\alpha}}{\ztau''}
-}{
-  \performTypI{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
-}
-\end{equation}
-\end{minipage}
-\end{subequations}
-
-\fbox{$\performSynI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
-\vspace{-10px}
-
-\begin{subequations}\label{rules:iterated-syn}
-\hspace{-12px}
-\begin{mathpar}
-\inferrule{ }{
-  \performSynI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
-}
-~~~~~~
-\inferrule{
-  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
-  \performSynI{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
-}{
-  \performSynI{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
-}
-~~~~~
-\text{(\ref*{rules:iterated-syn}a-b)}
-\end{mathpar}
-\end{subequations}
-
-\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
-\vspace{-12px}
-\begin{subequations}\label{rules:iterated-ana}
-\begin{mathpar}
-~~~~~~~\inferrule{ }{
-  \performAnaI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
-}
-~~~~~~~~~~~~~~
-\inferrule{
-  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
-  \performAnaI{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
-}{
-  \performAnaI{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
-}
-~~~~~~~~~~
-\text{(\ref*{rules:iterated-ana}a-b)}
-\end{mathpar}
-\end{subequations}
-\caption{Iterated Action Judgements}
-\label{fig:multistep}
-\end{figure}
 
@@
 
+\begin{figure}
+$\mathsf{ActionList}$~~$\bar{\alpha} ::= \cdot ~\vert~ \alpha; \bar{\alpha}$\vspace{4px}\\
+\fbox{$\performTypI{\ztau}{\bar{\alpha}}{\ztau'}$}
+
+\vspace{-10px}\begin{subequations}
+\begin{minipage}{0.35\linewidth}
+\begin{equation}
+\inferrule{ }{
+    \performTypI{\ztau}{\cdot}{\ztau}
+}
+\end{equation}
+\end{minipage}
+\begin{minipage}{0.65\linewidth}
+\begin{equation}
+\inferrule{
+  \performTyp{\ztau}{\alpha}{\ztau'}\\
+  \performTypI{\ztau'}{\bar{\alpha}}{\ztau''}
+}{
+  \performTypI{\ztau}{\alpha; \bar{\alpha}}{\ztau''}
+}
+\end{equation}
+\end{minipage}
+\end{subequations}
+
+\fbox{$\performSynI{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}{\htau'}$}
+\vspace{-10px}
+
+\begin{subequations}\label{rules:iterated-syn}
+\hspace{-12px}
+\begin{mathpar}
+\inferrule{ }{
+  \performSynI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}{\htau}
+}
+~~~~~~
+\inferrule{
+  \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}\\\\
+  \performSynI{\hGamma}{\zexp'}{\htau'}{\bar{\alpha}}{\zexp''}{\htau''}
+}{
+  \performSynI{\hGamma}{\zexp}{\htau}{\alpha; \bar{\alpha}}{\zexp''}{\htau''}
+}
+~~~~~
+\text{(\ref*{rules:iterated-syn}a-b)}
+\end{mathpar}
+\end{subequations}
+
+\fbox{$\performAna{\hGamma}{\zexp}{\htau}{\bar{\alpha}}{\zexp'}$}
+\vspace{-12px}
+\begin{subequations}\label{rules:iterated-ana}
+\begin{mathpar}
+~~~~~~~\inferrule{ }{
+  \performAnaI{\hGamma}{\zexp}{\htau}{\cdot}{\zexp}
+}
+~~~~~~~~~~~~~~
+\inferrule{
+  \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}\\\\
+  \performAnaI{\hGamma}{\zexp'}{\htau}{\bar\alpha}{\zexp''}
+}{
+  \performAnaI{\hGamma}{\zexp}{\htau}{\alpha; \bar\alpha}{\zexp''}
+}
+~~~~~~~~~~
+\text{(\ref*{rules:iterated-ana}a-b)}
+\end{mathpar}
+\end{subequations}
+\caption{Iterated Action Judgements}
+\label{fig:multistep}
+\end{figure}
 \subsubsection{Construction}\label{sec:construction} The construction
@@
 \end{equation}
+
 If the hole is being analyzed against a type that is consistent with the

### 2016-11-28 formally.tex
rewording comment about witnesses

@@
 Theorem \ref{thrm:reachability} follows by straightforward composition of
-these two lemmas. (Of course, the witness produced is not usually the most efficient way
-to move between two locations.)
+these two lemmas. The proofs we give of these three theorems in the
+mechanization do not produce the shortest sequence of actions to witness
+reachability, which would resemble something like a lowest common ancestor
+computation. In future versions of Hazelnut that use the produced witnesses
+for automatic tool support it may make sense to engineer these proofs
+differently; here we are only concerned with whether the theorems are true.
 

### 2016-11-28 formally.tex
rewording comment about composing with reach.

@@
 \paragraph{Constructability}
-To check that we have defined ``enough'' construct actions, we need to establish that we can start from an empty hole and arrive at any well-typed expression with, for simplicity, the cursor on the outside (Lemma \ref{lemma:reach-down} allows us to then move the cursor anywhere else.) As with reachability, we rely on the iterated action judgements defined in Figure \ref{fig:multistep}.
+To check that we have defined ``enough'' construct actions, we need to
+establish that we can start from an empty hole and arrive at any well-typed
+expression with the cursor on the outside. This simpler statement is
+sufficiently because Lemma \ref{lemma:reach-down} allows us to then move
+the cursor anywhere else inside the constructed term. As with reachability,
+we rely on the iterated action judgements defined in Figure
+\ref{fig:multistep}.
 \begin{theorem}[Constructability]\label{thrm:constructability} ~

### 2016-11-28 hazelnut-popl17.tex
tweaks per publisher requests
fixed some bib entry formatting things
fixing spacing and missing citation

@@
 \usepackage[T1]{fontenc}
+\usepackage{times}
 \usepackage{amsmath}
@@
 \AtBeginDocument{%
- \abovedisplayskip=2pt
+ \abovedisplayskip=2pt plus 4pt
  \abovedisplayshortskip=0pt
- \belowdisplayskip=2pt
+ \belowdisplayskip=2pt plus 4pt
  \belowdisplayshortskip=0pt
@@
 \usepackage[compact]{titlesec}
-\titlespacing*{\section}{0pt}{4pt}{2pt}
-\titlespacing*{\subsection}{0pt}{4pt}{2pt}
-\titlespacing*{\subsubsection}{0pt}{4pt}{2pt}
-\titlespacing*{\paragraph}{0pt}{4pt}{2pt}
+\titlespacing*{\section}{0pt plus 3pt}{4pt plus 3pt}{2pt plus 3pt}
+\titlespacing*{\subsection}{0pt plus 3pt}{4pt plus 3pt}{2pt plus 3pt}
+\titlespacing*{\subsubsection}{0pt plus 3pt}{4pt plus 3pt}{2pt plus 3pt}
+\titlespacing*{\paragraph}{0pt}{4pt}{3pt}
 \setlength{\skip\footins}{3pt plus 1px minus 5px}
@@
 
-Interactive proof refinement systems, e.g. those in LCF \cite{Gordon:1978:MIP:512760.512773}, and more recent typed tactic systems, e.g. Mtac for Coq \cite{Ziliani:2013:MMT:2544174.2500579}, 
+Interactive proof refinement systems, e.g. those in LCF \cite{Gordon:1978:MIP:512760.512773}, and more recent typed tactic systems, e.g. Mtac for Coq \cite{ziliani2015mtac}, 
 support an explicit model of a ``current'' typed hole that serves as the 
@@
 metatheory. Techniques like those in advanced tactic systems, e.g. Mtac, might be useful 
-in proving these action macros correct \cite{Ziliani:2013:MMT:2544174.2500579}.
+in proving these action macros correct \cite{ziliani2015mtac}.
 
@@
 
-\clearpage
 \balance

### 2016-11-28 implementation.tex
fixing spacing and missing citation

@@
 model, we chose to implement HZ using
-OCaml\footnote{\url{https://ocaml.org/}}, the \texttt{js\_of\_ocaml}
+OCaml\footnote{\url{https://ocaml.org/}}, the OCaml React
+library\footnote{\url{http://erratique.ch/software/react}}, and the \texttt{js\_of\_ocaml}
 compiler and associated libraries
-\cite{DBLP:conf/ml/Balat06}\footnote{\url{http://ocsigen.org/js\_of\_ocaml/}}
-and the OCaml React
-library\footnote{\url{http://erratique.ch/software/react}}.
+\cite{DBLP:conf/ml/Balat06}\footnote{\url{http://ocsigen.org/js\_of\_ocaml/}}.
 

### 2016-11-28 formally.tex

@@
 expression with the cursor on the outside. This simpler statement is
-sufficiently because Lemma \ref{lemma:reach-down} allows us to then move
-the cursor anywhere else inside the constructed term. As with reachability,
-we rely on the iterated action judgements defined in Figure
+sufficient because Lemma \ref{lemma:reach-down} allows us to then move the
+cursor anywhere else inside the constructed term. As with reachability, we
+rely on the iterated action judgements defined in Figure
 \ref{fig:multistep}.

### 2016-11-28 formally.tex
ran the tex files through aspell and found a few minor typos

@@
     then there exists some $\bar\alpha$ such that
-    $\bar\alpha~\mathsf{movements}$ nand
+    $\bar\alpha~\mathsf{movements}$ and
     $\performSynI{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.

### 2016-11-28 hazelnut-popl17.tex
fixed awkward gaps
ran the tex files through aspell and found a few minor typos

@@
  	\begin{tabular}{c}
- $^1$Carnegie Mellon University (USA)\\
+ $^1$Carnegie Mellon University, USA\\
  \textsf{\hphantom{$^1$}\{comar, iev, aldrich\}@cs.cmu.edu}
@@
  \begin{tabular}{c}
- $^2$Oregon State University (USA)\\
+ $^2$Oregon State University, USA\\
  \textsf{\hphantom{$^1$}hiltonm@eecs.oregonstate.edu}
@@
   \begin{tabular}{c}
- $^3$University of Colorado Boulder (USA)\\
+ $^3$University of Colorado Boulder, USA\\
  \textsf{\hphantom{$^1$}matthew.hammer@colorado.edu}
@@
 An intermediate approach would be to 
-layer unification-based type generation features atop the bidirectional system. This would amount to  interpretting type holes as unification
+layer unification-based type generation features atop the bidirectional system. This would amount to  interpreting type holes as unification
 variables. For a simple calculus,
@@
 with a notion of progress under a non-empty modal context), it would then
-be possible for the programmer to choose holes from amongst the visible
+be possible for the programmer to choose a hole from the visible
 holes (which may have been duplicated) to edit in their original
@@
 programming}, such as a packaging system, a differencing algorithm for use
-in a source control system, support for multiple simultaneous focii for
+in a source control system, support for multiple simultaneous cursors for
 different users, and so on. These are all interesting avenues for future
